### 2018-06-26 vignette.Rmd
Initial content commit

---
title: "Bayesian estimation of chronic disease epidemiology from incomplete data"
output: 
  html_document:
    number_sections: true
---

# Theoretical disease model

We represent a disease as a continuous-time, multi-state process with three states:

1. disease-free
2. disease
3. dead from the disease.

If we assume that mortality from other causes is independent of disease status, we can ignore deaths from other causes.   Assume also that recovery from the disease is not possible, but the framework is easily extensible to handle recurrence. The disease process is then fully defined by the

* disease incidence $i(t)$, and the

* case fatality rate $f(t)$,

which are both assumed to depend on age $t$.  Assume further that these rates are constant within integer years of age $a$, so they can be written $i_a$, $f_a$.

From these we can determine the _transition probability matrix_ $P_a$, whose $r,s$ entry $P_{ars}$ is the probability that a person is in state $s$ at age $a+1$ given they are in state $r$ at age $a$. The matrix $P_a$ is defined in terms of $i_a$ and $f_a$ as a differential equation, whose solution is written out explicitly in the DisMod II paper.

Further, let $sp_a$ be the proportion of individuals in a birth cohort who are in each state at age $a$.  This is a row vector with three elements $sp_{ar}$, one for each state $r$.  The transition probability matrix can then be used to obtain the corresponding proportions at age $a+1$, as

$$ sp_{a+1} = sp_a P_a $$

The prevalence of disease (among people who are alive) at each age $a$ is then obtained as $pr_a = sp_{a2} / (sp_{a1} + sp_{a2})$.

The disease-specific mortality rate at age $a$, or the probability that a person alive at age $a$ dies from the disease before age $a+1$, can also be expressed in terms of the transition probabilities and prevalence, as

$$ dm_a = P_{a23} pr_a + P_{a13} (1 - pr_a) $$

# Bayesian approach to estimating the model from data

Data are observed which give information about some, but not all, of the parameters in the theoretical disease model.   The form of the data available may be different in each application.  We then wish to estimate any remaining unknown parameters.   The Bayesian approach to estimating these unknowns can be described as four steps:

1. write down a theoretical model for underlying disease progression (as done above)

2. write down a statistical model for the observed data given the parameters of the underlying disease model.  An example of this is illustrated below.

3. write down prior distributions for the unknowns.  These may express prior ignorance, as in the example below.

4. compute the (unique) posterior distribution of the unknown parameters in the joint model, given the observed data.  Software discussed below.

This approach is used in the DisMod-MR software, as explained in the book by Flaxman et al, however the software itself is undocumented and not fully published.  The older (published) DisMod II used an optimisation approach to estimate parameters, which is a rough less statistically principled approximation.  Other advantages of the Bayesian method include

* straightforward to build in multiple sources of direct/indirect data.  This is enabled by the computational methods and available software for Bayesian modelling (illustrated below).   This allows the approach to generalise to settings with different forms of data available.  In contrast, DisMod II only allows limited forms of data.

* automatic quantification of uncertainty, given the information supplied (as illustrated below)

# Example: IHD in Greater London

The following data are given in the example spreadsheet

* incidence of IHD by age

* IHD-specific mortality by age

and we wish to estimate case fatality by age, given these inputs.

TODO PLOTS AND OR TABLES OF THE INPUT DATA

Note that these "data" are themselves estimates, based on underlying data from a finite population, that are not given.  In the Bayesian approach we can acknowledge that the incidence, mortality and/or case fatality are unknowns.   We illustrate this approach.  For simplicity, suppose that

* the incidence $i_a$ is known and fixed at the quantities supplied in the spreadsheet, but

* the disease-specific mortalities $dm_a$ are unknown, and have been estimated in terms of counts of individuals $n_a$ alive at age $a$ and the corresponding number $r_a$ who die of IHD before age $a+1$.

In this example, we can roughly recreate $n_a$ and $r_a$ as follows.  We are given population counts for five-year age groups, thus we obtain approximate one-year population counts $n_a$ by dividing these values by 5.  The number of IHD deaths $r_a$ is then approximately reconstructed by multiplying $n_a$ by the IHD-specific mortality rates supplied in the spreadsheet.

The four steps of the Bayesian modelling process are then be implemented as follows:

1. write down the theoretical disease model, as given above

2. write down the statistical model for the data.  This is $r_a \sim ~ Binomial(n_a, dm_a)$, that is, $r_a$ is assumed to have arisen from a Binomial distribution with denominator $n_a$ and probability $dm_a$, which is a deterministic function of the incidences and case fatalities $\{i_a,f_a\}$ up to age $a$, as described in the theoretical disease model.

3. define prior distributions for the unknown parameters.  For example, for each case fatality rate $f_a$, we assume a uniform distribution between 0 and 0.1.  Sensitivity to this should be assessed.  With enough data, the prior distribution will not matter.

4. compute the posterior distribution $p(\theta | \mathbf{y})$ for parameters $\theta = \{f_a\}$ given data $\mathbf{y} = \{i_a, r_a, n_a\}$

## Software for Bayesian computation

The Stan software ([mc-stan.org](http://mc-stan.org)) allows any Bayesian model to be written down in the Stan language.  The software then enables a sample from the posterior distribution to be drawn, using Hamiltonian Monte Carlo sampling.   It has an R interface, through the `rstan` package.

In this example, the theoretical disease model, statistical model and priors are written down in Stan language (link to Stan file on github)

The data $\mathbf{y} = \{i_a, r_a, n_a\}$ are supplied through R

(example of R code. data supplied in R data frame)

## Results

Posterior distribution for case fatality, overlay dismod outputs

Posterior for prevalence, disease-specific mortality

Load the Stan package, after installing it into R if necessary with `install.packages("rstan")`. 
```{r}
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

Load the IHD London dataset, put it in required format for Stan, and define initial values for 4 parallel MCMC simulations.   Finally, run the simulation with `stan()`, which should take less than a minute. 
```{r}
load(file="ihdlondon.rda")
datstan <- c(as.list(ihdlondon), nage=nrow(ihdlondon))
inits <- list(
    list(cf=rep(0.0101, datstan$nage)),
    list(cf=rep(0.0201, datstan$nage)),
    list(cf=rep(0.0056, datstan$nage)),
    list(cf=rep(0.0071, datstan$nage))
)
gbdcf <- stan("gbdcf-unsmoothed.stan", data=datstan, init=inits)
```

Check the MCMC simulations from the posteriors for six selected parameters has converged.  The four simulation chains should mix together and look like white noise.

```{r}
traceplot(gbdcf, pars=paste0("cf[", 60:65, "]"))
```

Extract all summary statistics (posterior median, 95% credible intervals, etc)
```{r}
summ <- summary(gbdcf)$summary
```

Extract and plot summary statistics for case fatality, and overlay the DisMod III estimates

```{r}
cfpars <- paste0("cf[",1:100,"]")
cf <- summ[cfpars,]
plot(1:100, cf[,"50%"], ylim=c(0, 0.1), pch=19, ylab="Estimated case fatality rate", xlab="Age")
title("Estimating case fatality given mortality and incidence data")
title("IHD, Greater London (?year?)", line=0.5)
segments(1:100, cf[,"2.5%"], 1:100, cf[,"97.5%"])
points(0:100, ihdlondon$cfout, col="purple")
legend("topright", lwd=c(1,NA), pch=c(19, 1), col=c("black","purple"), c("Bayesian posterior median and 95% interval", "DisMod II output"), bty="n")
```

### 2018-06-26 knitr.r

@@ -1,0 +1,3 @@
+library(rmarkdown)
+render("vignette.Rmd")
+browseURL("vignette.html")

### 2018-06-26 stan-unsmoothed.r

@@ -1,0 +1,74 @@
+
+library(rstan)
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+
+load(file="ihd_london.rda")
+
+dat <- cbind(outputs[,c("age","inc","rem")],
+             inputs[,c("pop","ndieddis","prevdenom","prevn")])
+dat$inc[1:4] <- 0.001/10*1:4 # don't really believe these are zero. 
+
+datstan <- c(as.list(ihd_london), nage=nrow(ihd_london))
+inits <- list(
+    list(cf=rep(0.0101, datstan$nage)),
+    list(cf=rep(0.0201, datstan$nage)),
+    list(cf=rep(0.0056, datstan$nage)),
+    list(cf=rep(0.0071, datstan$nage))
+)
+## inits <- list(list(cf=outputs$cfout))  # values estimated from dismod
+
+gbdcf <- stan("gbdcf-unsmoothed.stan", data=datstan, init=inits)
+summ <- summary(gbdcf)$summary
+
+traceplot(gbdcf, pars=paste0("cf[", 60:65, "]"))
+
+## Does it match Dismod estimates? 
+cfpars <- paste0("cf[",1:100,"]")
+cf <- summ[cfpars,]
+
+pdf("cf_bayes_ihd_london.pdf")
+plot(1:100, cf[,"50%"], ylim=c(0, 0.1), pch=19, ylab="Estimated case fatality rate", xlab="Age")
+title("Estimating case fatality given mortality and incidence data")
+title("IHD, Greater London (?year?)", line=0.5)
+segments(1:100, cf[,"2.5%"], 1:100, cf[,"97.5%"])
+points(0:100, outputs$cfout, col="purple")
+legend("topright", lwd=c(1,NA), pch=c(19, 1), col=c("black","purple"), c("Bayesian posterior median and 95% interval", "DisMod II output"), bty="n")
+## YES! we get proper estimates of uncertainty
+dev.off()
+
+## Does it recreate input prevalence estimates?
+prevpars <- paste0("prev[",1:100,"]")
+prev <- summ[prevpars,]
+plot(1:100, prev[,"mean"])
+segments(1:100, prev[,"2.5%"], 1:100, prev[,"97.5%"])
+points(0:100, outputs$prev, col="blue")
+points(0:100, outputs$prevout, col="purple")
+## Matches Dismod-adjusted prevalence estimates well, but not raw prevalence inputs after age 60
+
[…]

### 2018-06-26 vignette.Rmd
First vignette draft

@@
 title: "Bayesian estimation of chronic disease epidemiology from incomplete data"
+author: Chris Jackson <chris.jackson@mrc-bsu.cam.ac.uk>
 output: 
@@
     number_sections: true
+bibliography: disbayes.bib
 ---
+
+This document illustrates how case fatality can be estimated given incidence and mortality, using Bayesian modelling.
+
+This also serves to illustrate a general approach for estimating quantities describing disease epidemiology, based on indirect / incomplete data or multiple data sources.
+
+The method is implemented in freely-available software, and the code can be adapted easily to novel situations.
 
@@
 
-If we assume that mortality from other causes is independent of disease status, we can ignore deaths from other causes.
-Assume also that recovery from the disease is not possible, but the framework is easily extensible to handle recurrence.
+If we assume that mortality from other causes is independent of disease status, deaths from other causes are uninformative and can be ignored.
+Assume also for simplicity that remission from the disease is not possible (though the framework illustrated here is easily extensible to handle remission).
+
 The disease process is then fully defined by the
@@
 From these we can determine the _transition probability matrix_ $P_a$, whose $r,s$ entry $P_{ars}$ is the probability that a person is in state $s$ at age $a+1$ given they are in state $r$ at age $a$.
-The matrix $P_a$ is defined in terms of $i_a$ and $f_a$ as a differential equation, whose solution is written out explicitly in the DisMod II paper.
+The matrix $P_a$ is defined as a function of $i_a$ and $f_a$, which is the solution to a differential equation, and is written out explicitly in the DisMod II paper (@dismod2).
 
-Further, let $sp_a$ be the proportion of individuals in a birth cohort who are in each state at age $a$.
-This is a row vector with three elements $sp_{ar}$, one for each state $r$.
-The transition probability matrix can then be used to obtain the corresponding proportions at age $a+1$, as
+Further, let $S_a$ be the "state occupancy probabilities", or the proportion of individuals in a hypothetical birth cohort (of infinite size) who are in each state at age $a$.
+This is a row vector with three elements $S_{ar}$, one for each state $r$.
+Assume everyone is disease-free at age 0.
+The state occupancy probabilities at each subsequent age $a+1$ are then determined by mutiplying by the transition probability matrix:
 
-$$ sp_{a+1} = sp_a P_a $$
+$$ S_{a+1} = S_a P_a $$
 
-The prevalence of disease (among people who are alive) at each age $a$ is then obtained as $pr_a = sp_{a2} / (sp_{a1} + sp_{a2})$.
+The prevalence of disease (among people who are alive) at each age $a$ is then obtained as $pr_a = S_{a2} / (S_{a1} + S_{a2})$.
 
-The disease-specific mortality rate at age $a$, or the probability that a person alive at age $a$ dies from the disease before age $a+1$, can also be expressed in terms of the transition probabilities and prevalence, as
+The disease-specific mortality rate at age $a$, or the probability that a person alive at age $a$ dies from the disease before age $a+1$, can also be expressed in terms of the disease prevalence at age $a$ and the transition probabilities between ages $a$ and $a+1$, as
 
@@
 
-4. compute the (unique) posterior distribution of the unknown parameters in the joint model, given the observed data.  Software discussed below.
+4. compute the (unique) posterior distribution of the unknown parameters in the joint model, given the observed data.
 
-This approach is used in the DisMod-MR software, as explained in the book by Flaxman et al, however the software itself is undocumented and not fully published.
-The older (published) DisMod II used an optimisation approach to estimate parameters, which is a rough less statistically principled approximation.
+This approach is used in the DisMod-MR software, as explained by @flaxman2015, however the software itself is undocumented and not fully published.
+The older (published) DisMod II (@dismod2) used an optimisation approach to estimate parameters, which is a less statistically-principled rough approximation.
 Other advantages of the Bayesian method include
 
-* straightforward to build in multiple sources of direct/indirect data.
-This is enabled by the computational methods and available software for Bayesian modelling (illustrated below).
+* the ease of including multiple sources of direct/indirect data.
+This is enabled by the computational methods and available software for Bayesian modelling, illustrated below.
 This allows the approach to generalise to settings with different forms of data available.
@@
 
-* automatic quantification of uncertainty, given the information supplied (as illustrated below)
+* uncertainty about any quantity is quantified automatically through the posterior distribution, given the data, model assumptions and prior distributions supplied.
 
-# Example: IHD in Greater London
+The use of general-purpose free software for this computation is described below.
 
-The following data are given in the example spreadsheet
+# Example: IHD for men in Greater London
 
-* incidence of IHD by age
+The following data are given in the R data frame `ihdlondon` supplied here.
 
-* IHD-specific mortality by age
+* `inc` incidence of IHD by age
+
+* `mort` IHD-specific mortality by age
 
@@
 
-TODO PLOTS AND OR TABLES OF THE INPUT DATA
+The first few rows of the full data frame are shown here.  The estimates from DisMod II are given in the variable `cfout`.  The other variables are explained below.
 
-Note that these "data" are themselves estimates, based on underlying data from a finite population, that are not given.
-In the Bayesian approach we can acknowledge that the incidence, mortality and/or case fatality are unknowns.
+```{r}
+load(file="ihdlondon.rda")
+head(ihdlondon)
+```
+
+Note that the "data" `inc` and `mort` are themselves estimates, based on underlying data from a finite population that are not given.
+In the Bayesian approach, we can acknowledge that these quantities, along with the case fatality, are unknowns.
 We illustrate this approach.
@@
 
-* the incidence $i_a$ is known and fixed at the quantities supplied in the spreadsheet, but
+* the incidence $i_a$ is known and fixed at the quantities supplied in `inc`, but
 
-* the disease-specific mortalities $dm_a$ are unknown, and have been estimated in terms of counts of individuals $n_a$ alive at age $a$ and the corresponding number $r_a$ who die of IHD before age $a+1$.
+* the disease-specific mortalities $dm_a$ are not exactly equal to `mort`, but have been estimated in terms of counts of individuals $n_a$ alive at age $a$ and the corresponding number $r_a$ who die of IHD before age $a+1$.
 
-In this example, we can roughly recreate $n_a$ and $r_a$ as follows.
-We are given population counts for five-year age groups, thus we obtain approximate one-year population counts $n_a$ by dividing these values by 5.
-The number of IHD deaths $r_a$ is then approximately reconstructed by multiplying $n_a$ by the IHD-specific mortality rates supplied in the spreadsheet.
+In this example, $n_a$ and $r_a$ were not provided in the original data source, but can be roughly recreated, as follows.
+We were given population counts for five-year age groups, thus we obtain approximate one-year population counts $n_a$ by dividing these values by 5 (`pop` in the R data).
+The number of IHD deaths $r_a$ (`ndieddis` in the R data) is then approximately reconstructed by multiplying $n_a$ by the IHD-specific mortality rates supplied in the spreadsheet.
 
-The four steps of the Bayesian modelling process are then be implemented as follows:
+The four steps of the Bayesian modelling process are then implemented as follows:
 
@@
 write down the statistical model for the data.
-This is $r_a \sim ~ Binomial(n_a, dm_a)$, that is, $r_a$ is assumed to have arisen from a Binomial distribution with denominator $n_a$ and probability $dm_a$, which is a deterministic function of the incidences and case fatalities $\{i_a,f_a\}$ up to age $a$, as described in the theoretical disease model.
+This is $r_a \sim ~ Binomial(n_a, dm_a)$, that is, $r_a$ is assumed to have arisen from a Binomial distribution with denominator $n_a$ and probability $dm_a$.
+The disease-specific mortality $dm_a$ is a deterministic function of the incidences and case fatalities $\{i_j,f_j\}$ for ages $j$ up to $a$, as described in the theoretical disease model.
 
@@
 
-## Software for Bayesian computation
+Note also that estimates of prevalence are available in the source data, though these are also implicitly based on observing the number of people with IHD in a population.
+If the numerators $n^{(prev)}_a$ and denominator $r^{(prev)}_a$ corresponding to each age $a$ prevalence were known, they could be included in the model in the same way, as $r^{(prev)}_a \sim ~ Binomial(n^{(prev)}_a, pr_a)$.
+This acknowledges that the prevalence is an uncertain estimate, rather than a constant.
+The numerators and denominators are not provided in the source data, so we assume values for the denominators, given in the variable `prevdenom`.
+The corresponding numerators `prevn` are then obtained by multiplying `prevdenom` by the provided prevalence estimates, and rounding to the nearest integer.
+
+# Fitting Bayesian models with the Stan software
 
@@
 It has an R interface, through the `rstan` package.
+Its use is demonstrated briefly here.
 
-In this example, the theoretical disease model, statistical model and priors are written down in Stan language (link to Stan file on github)
-
-The data $\mathbf{y} = \{i_a, r_a, n_a\}$ are supplied through R
-
-(example of R code. data supplied in R data frame)
-
-## Results
-
-Posterior distribution for case fatality, overlay dismod outputs
-
-Posterior for prevalence, disease-specific mortality
-
-Load the Stan package, after installing it into R if necessary with `install.packages("rstan")`. 
-```{r}
+First load Stan and its R interface, after installing it into R if necessary with `install.packages("rstan")`. 
+```{r,message=FALSE}
 library(rstan)
@@
 ```
+In this example, the theoretical disease model, statistical model and priors are written down in Stan language in the file [gbdcf-unsmoothed.stan](gbdcf-unsmoothed.stan).
 
-Load the IHD London dataset, put it in required format for Stan, and define initial values for 4 parallel MCMC simulations.
-Finally, run the simulation with `stan()`, which should take less than a minute.
-```{r}
-load(file="ihdlondon.rda")
+The data $\mathbf{y} = \{i_a, r_a, n_a\}$ are supplied in the `ihdlondon` data frame.
+These are converted in to the required format for Stan, and initial values are defined for 4 parallel MCMC simulations.
+Finally, the simulation is run with `stan()`, which should take less than a minute.
+```{r, cache=TRUE, warning=FALSE, message=FALSE}
 datstan <- c(as.list(ihdlondon), nage=nrow(ihdlondon))
@@
 
-Check the MCMC simulations from the posteriors for six selected parameters has converged.  The four simulation chains should mix together and look like white noise.
-
+We check the MCMC simulations have converged, by examining the "trace plot" of the simulation progress for six selected parameters.
+The four simulated chains should mix together and look like white noise.
 ```{r}
@@
 
-Extract all summary statistics (posterior median, 95% credible intervals, etc)
+# IHD example: Results
+
+The summary statistics (posterior median, 95% credible intervals, etc) are extracted from the fitted model.
 ```{r}
@@
 
-Extract and plot summary statistics for case fatality, and overlay the DisMod III estimates
+The summary statistics for case fatality are extracted and plotted in R, and the corresponding estimates from DisMod II are overlaid.
 
@@
 ```
+
+The uncertainty about case fatality is biggest for the youngest and oldest ages, where either prevalence is lowest or the proportion of the cohort who are still alive is lowest.
+
+Agreement with the DisMod II estimates is good, except for the youngest ages, presumably because there was some additional source of information which was included in the DisMod estimates but omitted here.
+
+## Other things we could do
+
+For chronic diseases in general:
+
+* Account for variations in incidence and prevalence with time as well as age.
+DisMod II allows time trends to be modelled, but assumes that time trends are independent of age.
+Relaxing this assumption would be feasible, but possibly computationally expensive.
+	* For IHD in the UK, data on time trends may come from the Health Survey for England, years 1993, 1994, 1988, 2003, 2006, 2011.
+
+* Account for variations between populations in different areas.  Hierarchical modelling might be used here - DISMOD-MR does this, I believe.
+
+* Jointly model multiple diseases.  i.e. relaxing the assumption that mortality from causes other than the disease is the same for people with and without the disease.  Multimorbidity. 
+* In general, to transparently represent uncertainty, and explicitly build in other sources of information, in situations where we suspect the data may not be representative
+
+All of these would be theoretically feasible in the above framework. The limits will be computational, and difficult to tell before getting hands on.
+
+# References

### 2018-06-26 vignette.Rmd
vignette edits

@@
 This approach is used in the DisMod-MR software, as explained by @flaxman2015, however the software itself is undocumented and not fully published.
-The older (published) DisMod II (@dismod2) used an optimisation approach to estimate parameters, which is a less statistically-principled rough approximation.
-Other advantages of the Bayesian method include
+The older (published) DisMod II (@dismod2) used an ad-hoc optimisation approach to estimate parameters.
+
+Advantages of the Bayesian method include
 
@@
 and we wish to estimate case fatality by age, given these inputs.
-
-The first few rows of the full data frame are shown here.  The estimates from DisMod II are given in the variable `cfout`.  The other variables are explained below.
+The first few rows of the full data frame are shown here.
+The estimates of case fatality from DisMod II are given in the variable `cfout`.
+The other variables are explained below.
 

### 2018-06-26 vignette.Rmd
fix stan link

@@
 ```
-In this example, the theoretical disease model, statistical model and priors are written down in Stan language in the file [gbdcf-unsmoothed.stan](gbdcf-unsmoothed.stan).
+In this example, the theoretical disease model, statistical model and priors are written down in Stan language in the file [gbdcf-unsmoothed.stan](https://github.com/chjackson/disbayes/blob/master/gbdcf-unsmoothed.stan).
 

### 2019-11-11 R/ci2num.R
First upload as a proper R package

@@ -1,0 +1,33 @@
+##' Estimate binomial numerator and denominator given an estimate and uncertainty interval for a proportion
+##' 
+##' Based on least squares method provided by the SHELF package TODO REF
+##'
+##' @param est Point estimate
+##'
+##' @param lower Lower credible limit 
+##'
+##' @param upper Upper credible limit 
+##' 
+##' @examples 
+##' est <- 3.00 / 100 
+##' upper <- 3.52 / 100 
+##' lower <- 2.60 / 100 
+##' ci2num(est, lower, upper)
+##'
+##' @export
+
+ci2num <- function(est, lower, upper){
+    vals <- c(lower, est, upper) 
+    probs <- c(0.025, 0.5, 0.975)
+    bet <- SHELF::fitdist(vals=vals, probs=probs, lower=0, upper=1)$Beta
+    apost <- bet$shape1
+    bpost <- bet$shape2
+    aprior <- bprior <- 0.5
+    r <- round(apost - aprior)
+    n <- round(bpost - bprior + r)
+    list(num=r, denom=n)
+}
+
+est2num <- function(est, n){
+    list(num=est/n, denom=n)
+}

### 2019-11-11 R/disbayes-package.R

@@ -1,0 +1,25 @@
+#' The 'disbayes' package.
+#'
+#' @description Bayesian evidence synthesis for chronic disease epidemiology
+#'
+#' @docType package
+#' @name disbayes-package
+#' @useDynLib disbayes, .registration = TRUE
+#' @import methods
+#' @import Rcpp
+#' @importFrom rstan sampling
+#' @importFrom stats lm sd
+#'
+#' @references
+#' Stan Development Team (2019). RStan: the R interface to Stan. R package version 2.19.2. https://mc-stan.org
+#'
+NULL
+
+##' Ischaemic heart disease in London
+##'
+##' @format TODO 
+##' 
+##' @source Global Burden of Disease study 
+##' 
+##' @keywords datasets
+"ihdlondon"

### 2019-11-11 R/disbayes.R

@@ -1,0 +1,267 @@
+##' Bayesian estimation of chronic disease epidemiology
+##'
+##' Estimates a three-state disease model given data on incidence, prevalence and mortality.
+##'
+##' Currently it is designed to estimate case fatality given data on at least mortality and incidence.
+##'
+##' It could be modified to estimate incidence given prevalence and mortality (this would need a new Stan template model with incidence as a smooth function of age)
+##'
+##' Intended to extend to hierarchical models - will test in context of METAHIT project
+##'
+##' Other extensions to be motivated by applications 
+##'
+##' @rdname disbayes
+##' 
+##' @param data  Data frame containing some of the variables below.  The variables below are provided as character strings naming columns in this data frame.   For incidence and mortality, one of the three combinations of variables must be specified
+##'
+##' (1) numerator and denominator
+##' (2) estimate and denominator
+##' (3) estimate with lower and upper credible limits
+##'
+##' For estimates based on registry data that cover the whole population, then the denominator will be the population size. 
+##'
+##' @param inc_num Numerator for the incidence data, assumed to represent the observed number of new cases within a year among a population of size \code{inc_denom}. 
+##' 
+##' @param inc_denom Denominator for the incidence data.
+##'
+##' The function \code{\link{ci2num}} can be used to convert a published estimate and interval for a proportion to an implicit numerator and denominator.
+##'
+##' Note that to include extra uncertainty beyond that implied by a published interval, the numerator and denominator could be multiplied by a constant, for example, multiplying both the numerator and denominator by 0.5 would give the data source half its original weight. 
+##' 
+##' @param inc Estimate of incidence.
+##' @param inc_lower Lower credible limit for the incidence estimate
+##' @param inc_upper Upper credible limit for the incidence estimate
+##'
+##' @param prev_num Numerator for the estimate of prevalence 
+##' @param prev_denom Numerator for the estimate of prevalence
+##' @param prev Estimate of prevalence
[…]

### 2019-11-11 R/stanmodels.R

@@ -1,0 +1,26 @@
+# Generated by rstantools.  Do not edit by hand.
+
+# names of stan models
+stanmodels <- c("disbayes", "disbayes_unsmoothed")
+
+# load each stan module
+Rcpp::loadModule("stan_fit4disbayes_mod", what = TRUE)
+Rcpp::loadModule("stan_fit4disbayes_unsmoothed_mod", what = TRUE)
+
+# instantiate each stanmodel object
+stanmodels <- sapply(stanmodels, function(model_name) {
+  # create C++ code for stan model
+  stan_file <- if(dir.exists("stan")) "stan" else file.path("inst", "stan")
+  stan_file <- file.path(stan_file, paste0(model_name, ".stan"))
+  stanfit <- rstan::stanc_builder(stan_file,
+                                  allow_undefined = TRUE,
+                                  obfuscate_model_name = FALSE)
+  stanfit$model_cpp <- list(model_cppname = stanfit$model_name,
+                            model_cppcode = stanfit$cppcode)
+  # create stanmodel object
+  methods::new(Class = "stanmodel",
+               model_name = stanfit$model_name,
+               model_code = stanfit$model_code,
+               model_cpp = stanfit$model_cpp,
+               mk_cppmodule = function(x) get(paste0("model_", model_name)))
+})

### 2019-11-11 knitr.r

@@ -1,3 +1,0 @@
-library(rmarkdown)
-render("vignette.Rmd")
-browseURL("vignette.html")

### 2019-11-11 src/RcppExports.cpp

@@ -1,0 +1,22 @@
+// Generated by using Rcpp::compileAttributes() -> do not edit by hand
+// Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+#include <RcppEigen.h>
+#include <Rcpp.h>
+
+using namespace Rcpp;
+
+
+RcppExport SEXP _rcpp_module_boot_stan_fit4disbayes_mod();
+RcppExport SEXP _rcpp_module_boot_stan_fit4disbayes_unsmoothed_mod();
+
+static const R_CallMethodDef CallEntries[] = {
+    {"_rcpp_module_boot_stan_fit4disbayes_mod", (DL_FUNC) &_rcpp_module_boot_stan_fit4disbayes_mod, 0},
+    {"_rcpp_module_boot_stan_fit4disbayes_unsmoothed_mod", (DL_FUNC) &_rcpp_module_boot_stan_fit4disbayes_unsmoothed_mod, 0},
+    {NULL, NULL, 0}
+};
+
+RcppExport void R_init_disbayes(DllInfo *dll) {
+    R_registerRoutines(dll, NULL, CallEntries, NULL, NULL);
+    R_useDynamicSymbols(dll, FALSE);
+}

### 2019-11-11 src/stanExports_disbayes_unsmoothed.cc

@@ -1,0 +1,32 @@
+// Generated by rstantools.  Do not edit by hand.
+
+#include <Rcpp.h>
+using namespace Rcpp ;
+#include "stanExports_disbayes_unsmoothed.h"
+
+RCPP_MODULE(stan_fit4disbayes_unsmoothed_mod) {
+
+
+    class_<rstan::stan_fit<stan_model, boost::random::ecuyer1988> >("model_disbayes_unsmoothed")
+
+    .constructor<SEXP,SEXP,SEXP>()
+
+
+    .method("call_sampler", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::call_sampler)
+    .method("param_names", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::param_names)
+    .method("param_names_oi", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::param_names_oi)
+    .method("param_fnames_oi", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::param_fnames_oi)
+    .method("param_dims", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::param_dims)
+    .method("param_dims_oi", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::param_dims_oi)
+    .method("update_param_oi", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::update_param_oi)
+    .method("param_oi_tidx", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::param_oi_tidx)
+    .method("grad_log_prob", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::grad_log_prob)
+    .method("log_prob", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::log_prob)
+    .method("unconstrain_pars", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::unconstrain_pars)
+    .method("constrain_pars", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::constrain_pars)
+    .method("num_pars_unconstrained", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::num_pars_unconstrained)
+    .method("unconstrained_param_names", &rstan::stan_fit<stan_model, boost::random::ecuyer1988> ::unconstrained_param_names)
[…]

### 2019-11-11 src/stanExports_disbayes_unsmoothed.h

@@ -1,0 +1,1121 @@
+// Generated by rstantools.  Do not edit by hand.
+
+/*
+    disbayes is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    disbayes is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with disbayes.  If not, see <http://www.gnu.org/licenses/>.
+*/
+#ifndef MODELS_HPP
+#define MODELS_HPP
+#define STAN__SERVICES__COMMAND_HPP
+#include <rstan/rstaninc.hpp>
+// Code generated by Stan version 2.19.1
+#include <stan/model/model_header.hpp>
+namespace model_disbayes_unsmoothed_namespace {
+using std::istream;
+using std::string;
+using std::stringstream;
+using std::vector;
+using stan::io::dump;
+using stan::math::lgamma;
+using stan::model::prob_grad;
+using namespace stan::math;
+static int current_statement_begin__;
+stan::io::program_reader prog_reader__() {
+    stan::io::program_reader reader;
+    reader.add_event(0, 0, "start", "model_disbayes_unsmoothed");
+    reader.add_event(3, 3, "include", "/include/trans_probs.stan");
+    reader.add_event(3, 0, "start", "/include/trans_probs.stan");
+    reader.add_event(78, 75, "end", "/include/trans_probs.stan");
+    reader.add_event(78, 4, "restart", "model_disbayes_unsmoothed");
+    reader.add_event(130, 54, "end", "model_disbayes_unsmoothed");
+    return reader;
+}
+template <typename T0__, typename T1__, typename T2__>
+Eigen::Matrix<typename boost::math::tools::promote_args<T0__, T1__, T2__>::type, Eigen::Dynamic, Eigen::Dynamic>
+trans_probs_rem(const T0__& i,
+                    const T1__& f,
+                    const T2__& r, std::ostream* pstream__) {
+    typedef typename boost::math::tools::promote_args<T0__, T1__, T2__>::type local_scalar_t__;
+    typedef local_scalar_t__ fun_return_scalar_t__;
+    const static bool propto__ = true;
[…]

### 2019-11-14 doc/ihdbristol.Rmd
UPdated vignette, summary fixes

title: "Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package"
author: Chris Jackson <chris.jackson@mrc-bsu.cam.ac.uk>
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
    number_sections: true
vignette: >
 %\VignetteIndexEntry{Vignette Title}
 %\VignetteEngine{knitr::rmarkdown}
 \usepackage[utf8]{inputenc}
bibliography: disbayes.bib
---

This document illustrates how case fatality can be estimated given incidence and mortality, using Bayesian modelling.

This also serves to illustrate a general approach for estimating quantities describing disease epidemiology, based on indirect / incomplete data or multiple data sources.

The method is implemented in the R package `disbayes`, and could easily be adapted to novel situations.

# Theoretical disease model

We represent a disease as a continuous-time, multi-state process with three states:

1. disease-free
2. disease
3. dead from the disease.

If we assume that mortality from other causes is independent of disease status, deaths from other causes are uninformative and can be ignored.   Assume also for simplicity that remission from the disease is not possible (though the framework illustrated here is easily extensible to handle remission).

The disease process is then fully defined by the

* disease incidence $i(t)$, and the

* case fatality rate $f(t)$,

which are both assumed to depend on age $t$.  Assume further that these rates are constant within integer years of age $a$, so they can be written $i_a$, $f_a$.

From these we can determine the _transition probability matrix_ $P_a$, whose $r,s$ entry $P_{ars}$ is the probability that a person is in state $s$ at age $a+1$ given they are in state $r$ at age $a$. The matrix $P_a$ is defined as a function of $i_a$ and $f_a$, which is the solution to a differential equation, and is written out explicitly in the DisMod II paper (@dismod2).

Further, let $S_a$ be the "state occupancy probabilities", or the proportion of individuals in a hypothetical birth cohort (of infinite size) who are in each state at age $a$.  This is a row vector with three elements $S_{ar}$, one for each state $r$.  Assume everyone is disease-free at age 0.  The state occupancy probabilities at each subsequent age $a+1$ are then determined by mutiplying by the transition probability matrix:

$$ S_{a+1} = S_a P_a $$

The prevalence of disease (among people who are alive) at each age $a$ is then obtained as $pr_a = S_{a2} / (S_{a1} + S_{a2})$.

The disease-specific mortality rate at age $a$, or the probability that a person alive at age $a$ dies from the disease before age $a+1$, can also be expressed in terms of the disease prevalence at age $a$ and the transition probabilities between ages $a$ and $a+1$, as

$$ dm_a = P_{a23} pr_a + P_{a13} (1 - pr_a) $$

# Bayesian approach to estimating the model from data

Data are observed which give information about some, but not all, of the parameters in the theoretical disease model.   The form of the data available may be different in each application.  We then wish to estimate any remaining unknown parameters.   The Bayesian approach to estimating these unknowns can be described as four steps:

1. write down a theoretical model for underlying disease progression (as done above)

2. write down a statistical model for the observed data given the parameters of the underlying disease model.    In this model, observed data need to be expressed as counts of individuals and associated denominators, e.g.

  * (incidence): $n_a^{(inc)}, r_a^{(inc)}$: given a population of size $n_a$^{(mort)}, $r_a^{(inc)}$ of these are observed to get the disease within the next year

  * (mortality): $n_a^{(mort)}, r_a^{(mort)}$: given a population of size $n_a^{(mort)}$ (with or without the disease), $r_a^{(mort)}$ of these are observed to die from the disease within the next year

  * (prevalence): $n_a^{(prev)}, r_a^{(prev)}$: from a sample of $n_a^{(prev)}$ individuals, $r_a^{(prev)}$ are known to have the disease at age $a$ (and $n_a^{(prev)} - r_a^{(prev)} are known to not have the disease.

The next section discusses how these data can be derived, given the typical forms of data available from burden of disease studies.   A denominator is required because it enables uncertainty associated with each estimate to be quantified, as we will discuss later.

3. write down prior distributions for the unknowns.  These may express prior ignorance, as in the example below.

4. compute the (unique) posterior distribution of the unknown parameters in the joint model, given the observed data.

This approach is used in the DisMod-MR software, as explained by @flaxman2015, however the software itself is undocumented and not fully published.  The older (published) DisMod II (@dismod2) used an ad-hoc optimisation approach to estimate parameters.

Advantages of the Bayesian method, implemented in `disbayes`, include

* the ease of including multiple sources of direct/indirect data.  This is enabled by the computational methods and available software for Bayesian modelling, illustrated below.   This allows the approach to generalise to settings with different forms of data available.  In contrast, DisMod II only allows limited forms of data.

* uncertainty about any quantity is quantified automatically through the posterior distribution, given the data, model assumptions and prior distributions supplied.

The use of general-purpose free software for this computation, through the `disbayes` package, is described below.

# Converting data into the required form for the Bayesian model

## Given estimates and denominators

The following data are given in the R data frame `ihdbristol` supplied in the package, obtained from the Global Burden of Disease (REF) in a typical format:

* `inc` estimates of incidence of IHD by age

* `mort` estimates of IHD-specific mortality by age

* `pop` estimates of the population by age

and we wish to estimate case fatality by age, given these inputs.  A selection of rows from the full data frame are shown here.

```{r,show=FALSE}
library(disbayes)
ihdbristol[ihdbristol$age %in% 50:55, c("age","pop","inc","mort")]
```

Note that the published "data" `inc` and `mort` are themselves estimates, based on underlying data from a finite population. To use the Bayesian approach, we need to recreate the _numerator_ and _denominator_ underlying each of these estimates.   We assume that

* the incidence has been estimated as $r_a^{(inc)}/n_a$, where out of $n_a$ people who are alive at age $a$, $r_a^{(inc)}$ of these get IHD before age $a+1$.

* the mortality has been estimated as $r_a^{(mort)}/n_a$, where out of $n_a$ people who are alive at age $a$, $r_a^{(mort)}$ of these die of IHD before age $a+1$.  These ratios are estimates of the true, underlying disease-specific mortalities $dm_a$.

In this example, $n_a$ and $r_a$ were not provided in the original data source, but can be roughly recreated, as follows.  We were given population counts for five-year age groups, thus we obtain approximate one-year population counts $n_a$ by dividing these values by 5 (`pop` in the R data).

The number of incident IHD cases $r_a^{(inc)}$ can then be reconstructed by multiplying $n_a$ by the incidence rates `inc` supplied in the data.   Likewise, the number of IHD deaths $r_a$ (`ndieddis` in the R data) is then approximately reconstructed by multiplying $n_a$ by the IHD-specific mortality rates supplied in the spreadsheet.

The data also contain estimates of prevalence, implicitly based on observing the number of people with IHD $r^{(prev)}_a$ in a sample $n^{(prev)}_a$.    The assumed sample size $n^{(prev)}_a$ underlying these estimates is given in the variable `prevdenom`, which is multiplied by the prevalence estimates (given as `prev`) to obtain estimated prevalence counts $r^{(prev)}_a$ (given as `prevn`).   Typically, prevalence is estimated from survey data, though for diseases with comprehensive registers, such as cancer, the denominator underlying the estimate might be assumed to equal the population size.

```{r,eval=TRUE}
ihdbristol[ihdbristol$age %in% 50:55, c("age","prev","prevn","prevdenom")]
```

## Given point and interval estimates

Sometimes a point estimate $\hat{p}$ for a quantity such as incidence is published alongside an (e.g. 95%) interval estimate $(\hat{p}^{(lower)},\hat{p}^{(upper)})$.   The interval estimate can be assumed to express the uncertainty associated with the estimate.   Such information can be converted to an implicit numerator $r$ and denominator $n$  as follows.   We assume the point and interval estimate are summaries of a Beta posterior distribution which has been obtained by combining a vague prior with an observation of $r$ events occurring out of a sample of $n$ individuals.   If a Beta(0,0) prior is used, then the posterior is known to be Beta(r, n-r).  We can then search for the best-fitting Beta(r,n-r)$ distribution which has median $\hat{p}$ and (2.5,97.5) quantiles $(\hat{p}^{(lower)},\hat{p}^{(upper)})$, and set $r=a, n=a+b$.   A utility to perform this search is provided in the SHELF package for expert elicitation (REF).

## Expressing additional uncertainty

The uncertainty inherent in the information supplied about each of incidence, prevalence and mortality is measured by the denominator.  In the example above, this is the just size of the population used to produce the estimate.   However, if we also suspected that one of the data sources may be biased, but were unsure about the direction of bias, we could downweight that data source by multiplying both the numerator and denominator by the same amount, e.g. 0.5 if we wanted to give a data source half its original weight.  Note that if counts are scaled in this way, they should then be rounded to the nearest integer.

## Bayesian modelling process

The four steps of the Bayesian modelling process are then implemented as follows:

1. write down the theoretical disease model, as given above

2. write down the statistical model for the data.  All count data are assumed to arise from a Binomial distribution with the corresponding denominator, and a probability parameter which is a function of the parameters in the theoretical disease model.

	* (incidence) $r_a^{(inc)} \sim ~ Binomial(n_a, P_{a12})$, where $P_{a12}$ is the annual transition probability from no disease to disease.

	* (mortality)  $r_a^{(mort)} \sim ~ Binomial(n_a, dm_a)$, where the disease-specific mortality $dm_a$ is a deterministic function of the incidences and case fatalities $\{i_j,f_j\}$ for ages $j$ up to $a$, as described in the theoretical disease model.

	* (prevalence)  $r_a^{(prev)} \sim ~ Binomial(n_a^{(prev)}, pr_a)$, where $pr_a$ is the true prevalence, defined as a deterministic function of the incidences and case fatalities.

3. define prior distributions for the unknown parameters.  For incidence, currently the `disbayes` package assumes an unbounded uniform prior for each age independently.  For case fatality, the package allows two alternatives, explained below.

4. compute the posterior distribution $p(\theta | \mathbf{y})$ for parameters $\theta = \{i_a,f_a\}$ given data $\mathbf{y} = \{i_a, r_a, n_a\}$

## Alternative models / prior assumptions

In example such as this one, where quantities are estimated from indirect data, it is important to consider what substantive prior information is available about the unknown quantities, and what influence the assumed prior has on the resulting estimates.

In these examples, the case fatality is only informed indirectly.  The `disbayes` package implements two alternative models for how case fatality depends on age.

1. the case fatality rate $f_a$ for each year of age $a$ is assumed to have an independent vague prior distribution, taken to be exponential(1).

2. the case fatality rate is assumed to be a smooth function $g()$ of age, $f_a = g(a)$.   This smooth function is defined by a spline basis $log(f_a) = \sum_k \beta_k g_k(a)$, where $g_k()$ are basis functions.  A "thin plate" spline is used, following the default in the `mgcv` package (Wood), and the amount of smoothness is determined automatically through a hierarchical prior on the coefficients (REF Wood, `jagam` paper).

In addition, for all ages below a given cut-off age $a_{base}$, case fatalities are assumed to be equal to a constant $f_{base}$.   This cut-off age $a_{base}$ needs to be supplied when calling the `disbayes` function.  $f_{base}$ does not need to be supplied however, and is estimated from the data under the assumption that the dependence on age is a smooth function.

When running `disbayes`, both models are fitted by default.  Model (2) is more realistic.  However it may not give sensible estimates if the information provided by the data is too weak. The results of fitting Model (1) can help to diagnose where the indirect information on case fatality provided by the rest of the data is weaker or stronger.   If the information on $f_a$ for a particular age is too weak, then the posterior distribution will be identical to the prior.   In those cases, substantive prior information about case fatality at that age is necessary for the estimates to be meaningful.

This information might come from nearby ages that are better-informed, through extrapolation from the smoothing model.  However extrapolating in this way is only feasible for a limited time span (perhaps around 10-20 years) before stronger assumptions are necessary, e.g. that case fatality is constant below a certain age.

# Fitting Bayesian models with the Stan software via disbayes

The Stan software ([mc-stan.org](http://mc-stan.org)) allows any Bayesian model to be written down in the Stan language.  The software then enables a sample from the posterior distribution to be drawn, using Hamiltonian Monte Carlo sampling.

The `disbayes` package uses Stan to fit the Bayesian disease model given here to the data $\mathbf{y} = \{i_a, r_a, n_a\}$ supplied in the `ihdbristol` data frame.   The function `disbayes` does the work of converting the data to the required format for Stan, and initializing and a set of Markov Chain Monte Carlo simulations, and calling the Stan software to run the simulations.

The `disbayes` function requires the following arguments

 * `dat`: a data frame where all the variables are stored

 * additional arguments indicating which columns of the data contain which variables.

The required variables include at least information about incidence and mortality, optionally information about prevalence, and some indication of uncertainty around this information.   This information can be supplied in three alternative ways.  For example, for incidence,

1. estimate with denominator: through arguments `inc` and `inc_denom`.  The numerator is computed automatically, as described above.

2. numerator and denominator, through arguments `inc_num` and `inc_denom`

3. estimate with lower and upper 95% credible limits, through arguments `inc`, `inc_lower` and `inc_upper`. The numerator and denominator are comptue computed automatically, as described above.

The value given for the argument is a character string identifying the corresponding variable in the data.  For mortality and prevalence, arguments with names beginning `mort` and `prev` are supplied in the same manner.

Before running the `disbayes` function to fit the models, we set up the computer for parallel processing.  This allows multiple processor cores to be used, which enables the models to fit substantially faster. 
```{r}
options(mc.cores = parallel::detectCores())
```

The following call to the `disbayes` function is then made to fit the model to the `ihdbristol` data.   The argument `eqage=40` defines an assumption that case fatality is constant for all ages below 40.

```{r, eval=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
dbres <- disbayes(dat = ihdbristol,
                 inc = "inc", 
                 inc_denom = "pop", 
                 prev_num = "prevn", 
                 prev_denom = "prevdenom",
                 mort = "mort",
                 mort_denom = "pop",
                 eqage = 40
                 )
```

The computation may take a few minutes, or less depending on the number of cores and their speed.

The `disbayes` function returns a list with components `fit` and `fitu`, containing the smoothed and unsmoothed model fits respectively.  Each of these components is an object in the `stanfit` format defined by the `rstan` R interface to Stan

# IHD example: Results

The results could be summarised and manipulated using functions provided by rstan, e.g. the `summary` and `extract` method for `stanfit` objects.    For example, we can check the MCMC simulations have converged, by examining the "trace plot" of the simulation progress for selected parameters. The simulated chains should mix together and look like white noise.

```{r,eval=TRUE}
rstan::traceplot(dbres$fit, pars=paste0("cf[", 60:65, "]"))
```

The `disbayes` package provides some shortcuts for extracting summary statistics.   The posterior medians and 95% credible intervals can be extracted from the fitted smooth model using the `summary` method for `disbayes` objects.

The result is a data frame with the variables indicated in the row names.   These variables include `inc` (incidence), `cf` (case fatality), `mort` (mortality), `prev` (prevalence), each indexed by year of age.   The first few rows of the summary below contain estimates of incidence for the first few years of age.

```{r, eval=TRUE}
summ <- summary(dbres)
head(summ)
```

To extract results for a specific variable, the `vars` argument can be provided, e.g. for case fatality:

```{r,eval=TRUE}
summ <- summary(dbres, vars="cf")
head(summ)
```

To extract results from the unsmoothed model, the `summary_disbayes_fit` method is used.

```{r,eval=TRUE}
summ <- summary_disbayes_fit(dbres$fit, vars="cf")
summu <- summary_disbayes_fit(dbres$fitu, vars="cf")
head(summu)
```

The default `plot` method for objects returned by `disbayes` plots the posterior summaries for case fatality.  Both the smoothed and unsmoothed estimates from age 40 are shown.   This plot also overlays the corresponding estimates from DisMod II are overlaid in purple, included in the `ihdbristol` data in variable `dismod_cf`.

```{r, eval=TRUE}
library(ggplot2)
plot(dbres) +  ylab("Case fatality") + xlab("Age") + ylim(0,0.5) + xlim(40,100) + 
  geom_line(aes(y=dismod_cf), data=ihdbristol, col="purple")
```

Under the unsmoothed model, the uncertainty about case fatality is biggest for the youngest and oldest ages, where either prevalence is lowest or the proportion of the cohort who are still alive is lowest.

The smoothed model givese more precise estimates, aided by the assumption that case fatalities are similar between similar ages, and that case fatality is constant under the age of 40.  The smoothed model also removes the artefacts from the 5-year age grouping that are seen in the unsmoothed results.

The smoothed model agrees with the DisMod II estimates, and has the benefit of characterising uncertainty as a posterior distribution.  Uncertainty about case fatality in this example arises from the disease prevalence.   If the prevalence is lower, fewer people have the disease, giving less information from which to estimate the case fatality for people with the disease.

## Other things that could be done

For chronic diseases in general:

* More ways of supplying substantive prior information

* Account for variations between populations in different areas.  Hierarchical modelling might be used here - DISMOD-MR does this, I believe.

* Use data from larger areas as priors for quantities for smaller areas

* Borrow information between men/women?

* Jointly model multiple diseases.  i.e. relaxing the assumption that mortality from causes other than the disease is the same for people with and without the disease.  Multimorbidity.

* Account for variations in incidence and prevalence with time as well as age.  DisMod II allows time trends to be modelled, but assumes that time trends are independent of age.  Relaxing this assumption would be feasible, but possibly computationally expensive. 
	* For IHD in the UK, data on time trends may come from the Health Survey for England, years 1993, 1994, 1988, 2003, 2006, 2011.

* In general, to transparently represent uncertainty, and explicitly build in other sources of information, in situations where we suspect the data may not be representative

All of these would be theoretically feasible in the above framework. The limits will be computational, and difficult to tell before getting hands on.

### 2019-11-14 R/disbayes-package.R

@@ -8,18 +8,18 @@
 #' @import methods
 #' @import Rcpp
 #' @importFrom rstan sampling
-#' @importFrom stats lm sd
+#' @importFrom stats lm sd rlnorm rnorm
 #'
 #' @references
 #' Stan Development Team (2019). RStan: the R interface to Stan. R package version 2.19.2. https://mc-stan.org
 #'
 NULL
 
-##' Ischaemic heart disease in London
+##' Ischaemic heart disease in Bristol
 ##'
 ##' @format TODO 
 ##' 
 ##' @source Global Burden of Disease study 
 ##' 
 ##' @keywords datasets
-"ihdlondon"
+"ihdbristol"

### 2019-11-14 R/disbayes.R

@@ -123,7 +123,7 @@ disbayes <- function(data,
                  lambda = rlnorm(length(lam), mean=log(lam), sd=lam/10))
         }
         fits <- rstan::sampling(stanmodels$disbayes, data=datstans, 
-                     init=inits, iter=5000, ...)
+                     init=inits, iter=10000, ...)
         res <- list(fit=fits, fitu=fitu)
     } else res <- list(fit=fitu)
     
@@ -212,22 +212,26 @@ process_data <- function(data, prefix, num_str, denom_str, est_str, lower_str, u
 ##'
 ##' @param object Object returned by \code{\link{disbayes}}
 ##'
-##' @param variable Name of the variable of interest to return.  If not supplied, all estimated quantities are returned. 
+##' @param vars Names of the variables of interest to return.  If not supplied, all estimated quantities are returned. 
 ##'
 ##' @param ... Other arguments. Currently unused
 ##' 
 ##' @export
 summary.disbayes <- function(object, vars=NULL, ...){
     fit <- object$fit
-    summ <- summary.disbayes.fit(fit, vars=vars) 
+    summ <- summary_disbayes_fit(fit, vars=vars) 
     class(summ) <- c("summary.disbayes", class(summ))
     summ
 }
 
 ##' Summarise estimates from the model  TODO DOC 
 ##'
+##' @param fit A Stan fitted model object
+##'
+##' @param vars Names of variables indexed by age, e.g. \code{vars="cf"} if you want the case fatality estimates for all ages.   TODO describe other variables that can be extracted
+##' 
 ##' @export
-summary.disbayes.fit <- function(fit, vars=NULL){
+summary_disbayes_fit <- function(fit, vars=NULL){
     summ <- rstan::summary(fit)$summary
     summ <- as.data.frame(summ[, c("2.5%","50%","97.5%")])
     names(summ) <- c("lower95","med","upper95")
@@ -242,11 +246,13 @@ summary.disbayes.fit <- function(fit, vars=NULL){
 
 ##' Plot estimates from the model against age
 ##'
-##' Posterior medians and 95% credible intervals for a quantity of interest are plotted against year of age. 
+##' Posterior medians and 95% credible intervals for a quantity of interest are plotted against year of age.  
 ##'
 ##' @param x Object returned by \code{\link{disbayes}}
 ##'
[…]

### 2019-11-14 doc/ihdbristol.R

@@ -1,0 +1,42 @@
+## ----show=FALSE---------------------------------------------------------------
+library(disbayes)
+ihdbristol[ihdbristol$age %in% 50:55, c("age","pop","inc","mort")]
+
+## ----eval=TRUE----------------------------------------------------------------
+ihdbristol[ihdbristol$age %in% 50:55, c("age","prev","prevn","prevdenom")]
+
+## -----------------------------------------------------------------------------
+options(mc.cores = parallel::detectCores())
+
+## ---- eval=TRUE, cache=TRUE, warning=FALSE, message=FALSE---------------------
+dbres <- disbayes(dat = ihdbristol,
+                 inc = "inc", 
+                 inc_denom = "pop", 
+                 prev_num = "prevn", 
+                 prev_denom = "prevdenom",
+                 mort = "mort",
+                 mort_denom = "pop",
+                 eqage = 40
+                 )
+
+## ----eval=TRUE----------------------------------------------------------------
+rstan::traceplot(dbres$fit, pars=paste0("cf[", 60:65, "]"))
+
+## ---- eval=TRUE---------------------------------------------------------------
+summ <- summary(dbres)
+head(summ)
+
+## ----eval=TRUE----------------------------------------------------------------
+summ <- summary(dbres, vars="cf")
+head(summ)
+
+## ----eval=TRUE----------------------------------------------------------------
+summ <- summary_disbayes_fit(dbres$fit, vars="cf")
+summu <- summary_disbayes_fit(dbres$fitu, vars="cf")
+head(summu)
+
+## ---- eval=TRUE---------------------------------------------------------------
+library(ggplot2)
+plot(dbres) +  ylab("Case fatality") + xlab("Age") + ylim(0,0.5) + xlim(40,100) + 
+  geom_line(aes(y=dismod_cf), data=ihdbristol, col="purple")
+

### 2019-12-04 doc/hier.Rmd
Updated vignette and new hierarchical discussion

---
title: "Hierarchical models for estimating case fatality from multiple areas"
author: "Chris Jackson <chris.jackson@mrc-bsu.cam.ac.uk>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical models for estimating case fatality from multiple areas}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Motivation

* The main `disbayes` package vignette explains the main model behind the package.  This is a Bayesian multi-state model for estimating case fatality for a disease given information about the mortality, incidence, and optionally also the prevalence, in a _single group_.   In the example, this was defined by area and gender.

* This vignette explains how the method from the main vignette can be generalised to make the best use of data from multiple groups.  This is illustrated for multiple areas.

* Most simply, the basic method could be employed once for each area to estimate case fatality independently for each area.  However, for some diseases and areas, the information in the data alone is too weak to give meaningful estimates without excessively strong prior assumptions.   This happens for diseases where the numbers of cases of the disease observed in the area is too small to disentangle mortality (rate of death from the disease averaged over the whole population) from case fatality (rate of death from the disease for people with the disease).

* At the other extreme, we could assume that the case fatality is the same in every area, and estimate it by combining the data from multiple areas to form a larger area where there is enough information to provide a useful estimate.   However, there may be systematic variabilities between the areas.

* A _hierarchical model_ is a way of compromising between these two implausible extremes.

# Hierarchical modelling: principles

* A hierarchical model acknowledges that the data from _other areas_ give useful prior information about what we expect the estimate from one _area of interest_ to be.  If we estimate for each area independently, then this information is neglected.

* Essentially it produces an area-specific estimate as a _weighted average_ of the area-specific data and the data from the other areas.

* The consequence is that if the data from the area of interest is weaker, then a higher weight is given to the data from other areas.  This is desirable since an estimate just using the single area's data would be excessively noisy - thus we "borrow strength" from other areas.

* An appropriate weight is constructed automatically under the assumption that the true case fatalities for each area are generated from a between-area distribution of case fatalities, which we call a "random effects" distribution.

* The simplest random effects distributions are "exchangeable", that is, we assume the area case fatalities are conditionally independent given the random effects distribution.   This is what is done in the example below.  In other examples we might want to account for potential correlations arising from the structure of the data - for example:

	* spatial distributions for small-area estimates - which assume that closer areas are expected to be more similar than distant areas

	* random-walk type distributions for data grouped in ordered categories rather than areas, e.g. socioeconomic status / income.

# Example

* We want to estimate the case fatality for multiple diseases, for 9 city regions in England (Bristol, Liverpool, London, West Midlands, Sheffield, North East, Greater Manchester, Nottingham, Leeds).

* We have data on incidence, mortality and prevalence from these nine city regions, and additionally from the nine regions of England (East Midlands, North East England, North West England, South West England, West Midlands, Yorkshire and the Humber, East of England, South East England, Greater London)

	* Note that some of these regions contain the city regions of interest.  We can form incidence / mortality / prevalence count data for these regions, that _excludes_ the data from the city regions inside them, by converting all data into count format (using the method described in the main vignette) and subtracting the city region counts from the region counts.   Note that for Greater London the region is identical to the city region, so the region data provide no extra information.

	* This results in disease data from 18 mutually exclusive areas covering the whole of England.  We fit a hierarchical Bayesian multi-state model to estimate the case fatality from all the areas.

* As an example disease and area, take stomach cancer in the Bristol city region.   This plot shows case fatality estimates under the two alternative models described in the main vignette - one where case fatality is estimated independently for each year of age, and another where case fatality is assumed to be a smooth spline function of age, assuming a common value for ages under 50 (this age was arbitrarily chosen - in further work we should seek evidence / expert input on what different assumptions might be used for different diseases)

* It is clear there is insufficient information -- the results for ages under 80 are implausibly high, as they are dominated by a weak prior that allows case fatality to be high, and no further information is provided by the data.

```{r}
knitr::opts_chunk$set(fig.width=7, echo=FALSE, warning=FALSE)
```

```{r}
load(file="cfres.rda")
library(dplyr)
library(ggplot2)
source("constants.R")

p <- cfres %>%
    filter(disease == "stmc" & 
           area == "bristol" &
           model %in% c("unsmoothed", "smoothed")) %>% 
ggplot(aes(x=age, col=model)) +
    geom_pointrange(aes(y=med, ymin=lower95, ymax=upper95),
                    position = position_dodge(0.2)) + 
    facet_grid(cols=vars(gender)) +  
    coord_cartesian(ylim=c(0, 4)) +
    scale_x_continuous(breaks=seq(0,100,10)) + 
    ylab("Case fatality (median, 95% CrI)")  +
    xlab("Age (years)") +
    ggtitle("Bristol, stomach cancer")
p
```

* We demonstrate how improved estimates can be obtained with a hierarchical model.

# Technical details of the hierarchical model

* The same three-state stochastic model is used for disease progression, in terms of incidence and case fatality, as described in the main package vignette.

* In each area, the case fatality is assumed to be a different smooth spline function of age.   The hierarchical model is constructed as follows, following methods described by Wood (REF Generalized Additive Models with R, 2nd Edition).   The log case fatality for area $i$ is the following function of age $a$

$$ log(f_i(a)) = \alpha_i + \beta_i  a  + s_i(a)$$

where

* $\alpha_i$ is the random intercept, describing the average log case fatality over different ages in area $i$

* $\beta_i$ is a random slope, describing how log case fatality increases with age in area $i$

* $s_i(a) = \sum \gamma_{ik} h_k(a)$ is a smooth function describing how log case fatality deviates from a simple linear trend, as a function of age, in area $i$.

* The sum of the three terms above is defined in exactly the same way as in the area-specific model, to equal a linear function of "basis" terms.  The only difference here is in the presentation - here the parameters defining the intercept and the coefficient of age (the slope) are explicitly labelled, so we can define a model for how they vary between areas.   The function $s_i(a)$ is a linear function of higher-order or nonlinear terms in $a$ (quadratic, cubic and logarithmic functions), each with coefficient $\gamma_{ik}$. As for the single-area example, "thin plate" splines are used - the exact definitions of the basis terms are described by Wood (REF).

* $\alpha_i$ and $\beta_i$ are assumed to be generated from normal "random effects" distributions with mean $\mu_a$ and $\mu_b$ respectively, representing the average intercept and slope over all areas $i$.  Note there is no spatial / geographical information in this model - closer areas are not necessarily assumed to be more similar than distant areas. For an example with smaller/closer areas we might want to build geographical information into the random effects model.

* Weakly informative prior distributions (normal with mean 0, standard deviation 5, and normal with mean 10, standard deviation 10) are placed on $\mu_a$ and $\mu_b$ respectively.

* The $\gamma_{ik}$, for all areas $i$ and basis terms $k$ are assumed to be independent draws from a normal distribution with mean zero and unknown variance $\lambda$.  The parameter $\lambda$ governs the degree of smoothness in the function of age, so that lower values of $\lambda$ imply complete smoothing, thus log case fatality is a linear function of age, and higher values of $\lambda$ imply no smoothing, and no restriction on the wiggliness of the deviation from linearity.

* $\lambda$ is given a given a vague exponential(1) prior, and estimated along with the other parameters as part of the Bayesian model, so that the optimal level of smoothing is determined from the data.

* The basis terms are defined to be orthogonal, essentially to avoid collinearity, and make computation more efficient, as described by Wood (REF).

* The joint posterior distribution of the parameters describing case fatality $\alpha_i, \beta_i, \gamma_{ik}, \lambda, \mu_\alpha, \mu_\beta$, and incidence rates $i_{ai}$, for all areas $i$ and basis terms $k$, is estimated using Hamiltonian Monte Carlo sampling in the Stan software, via the `rstan` package.  No interface is provided yet in `disbayes`.

# Illustration of hierarchical model results

## Variations between ages, within an area

The estimated case fatality for stomach cancer in Bristol, by age, under two alternative models, is shown below.   The model labelled "smoothed" is the same smooth model as shown before which uses just the data from Bristol, and "hier50" is the corresponding hierarchical model that also borrows information from other areas.

* For females, the hierarchical model gives no improvement - the estimates for younger ages are still dominated by the vague prior that allows high values, with no data coming in to contradict that.

* For males, the hierarchical model produces precise estimates - the information in the data seems to have overcome the vague prior.

* _[ This suggests we might also borrow information about case fatality between genders as well as between areas - should discuss with epidemiologists - how do we expect case fatality for a disease to differ between men and women, given the disease, and given variations between ages and areas? ]_

```{r,echo=FALSE}

p <- cfres %>%
    filter(disease %in% "stmc" & 
           area == "bristol" &
           model %in% c("smoothed", "hier50")) %>% 
ggplot(aes(x=age, col=model)) +
    geom_pointrange(aes(y=med, ymin=lower95, ymax=upper95),
                    position = position_dodge(0.2)) + 
    facet_grid(rows=vars(disease), cols=vars(gender)) +  
    coord_cartesian(ylim=c(0, 4)) +
    scale_x_continuous(breaks=seq(0,100,10)) + 
    ylab("Case fatality (median, 95% CrI)")  +
    xlab("Age (years)") +
    ggtitle("Bristol, stomach cancer")
p

```

## Variations between areas, for a given age

The plot below shows how the estimates of case fatality, under the same two models, vary between areas. 
Bristol is shown at the bottom.

* Under the hierarchical model, all estimates become more precise, and the variability between areas in the results is smaller.

* While we can see that the estimate for Bristol becomes precise when hierarchical between-area information is included, it's not clear how that information gets used.   Typically the area-specific estimates get "shrunk" towards the rest of the data, but in this case, the Bristol estimate after shrinkage is among the lowest of the areas - we'd expect it to be more in the middle.

```{r, echo=FALSE, warning=FALSE}

hierdata <- readRDS("hierdata.rds")
hdat <- hierdata[["male"]][["stmc"]]
dat <- data.frame(
    area = c(cityregions, regions_exc, regions_inc),
    prevn = hdat$prev_num[60,],
    prevd = hdat$prev_denom[60,],
    mortn = hdat$mort_num[60,],
    mortd = hdat$mort_denom[60,],
    incn = hdat$inc_num[60,],
    incd = hdat$prev_denom[60,]
)
dat$mortl <- qbeta(0.025, dat$mortn + 0.5, dat$mortd - dat$mortn + 0.5)*1000
dat$mort <- qbeta(0.5, dat$mortn + 0.5, dat$mortd - dat$mortn + 0.5)*1000
dat$mortu <- qbeta(0.975, dat$mortn + 0.5, dat$mortd - dat$mortn + 0.5)*1000

p <- cfres %>%
    filter(age == 60,
           disease == "stmc", 
           model %in% c("smoothed", "hier30")) %>%
    left_join(dat, by="area") %>% 
ggplot(aes(x=med, y=area, col=model)) +
    coord_cartesian(xlim=c(0,0.15)) + 
    facet_grid(rows=vars(disease), cols=vars(gender)) + 
    geom_point(position=ggstance::position_dodgev(height=0.4),
               aes(size=mortd)) +
    geom_errorbarh(aes(xmin=lower95, xmax=upper95),
                   position=ggstance::position_dodgev(height=0.4)) +
    labs(size = "Age 60 population") + 
    xlab(sprintf("Case fatality for age 60") )
p

```

## Transparency - where do these estimates come from?

Taking Bristol, male, age 60 as an example, the case fatality estimates from the hierarchical model above come from the following data and assumptions:

* Mortality for age 60 in Bristol.  These data are very weak in this example.  Given the population size, the published mortality rate is estimated to be based on just 1 observed death out of around 6000 people - see the graph below.

* Incidence for all years up to 60 for Bristol.  Again, given the population sizes, the published incidence rates reflect about one incident case per year of age starting at about age 50.

* Prevalence for age 60: the published prevalence is again very low (but in this example, it is unclear what data this is based on. If it is cancer registry data, the denominator is the whole population, but the denominator should be smaller for survey data)

* Mortality/incidence/prevalence for neighbouring ages in Bristol, through smooth spline model for case fatality, and assuming constant case fatality below age 50

* Data for other areas of England, through the hierarchical model

Since the data from Bristol, age 60 alone are so sparse, the estimates will be largely based on borrowing strength from other ages and areas.   The information from other areas shrinks down the vague prior for Bristol, but it seems to shrink too far, further down than the estimates from the other areas.   So this result may be biased by some artefact of the model assumptions and how they combine.    For example, the prior assumption of constant case fatality for below age 50 is likely to be influential in all models where it is used in settings of weak data.

### "Raw" mortality data for stomach cancer, age 60 in different areas

Estimated number/denominator of deaths used to produce the published data - presented to the left.

```{r,echo=FALSE}

dat$numdenom <- sprintf("%s/%s",dat$mortn, dat$mortd)
ggplot(dat, aes(y=area, x=mort)) +
    geom_point(aes(size=mortn)) +
    xlim(-0.1, 0.8) +
    ylab("") + 
    geom_errorbar(aes(xmin=mortl, xmax=mortu)) +
    xlab("Annual mortality per 1000 for age 60" ) +
    labs(size = "Number of deaths") + 
    geom_text(aes(x=0, label=numdenom), size=3, hjust=1)

```

### 2019-12-04 R/disbayes-package.R

@@ -9,6 +9,7 @@
 #' @import Rcpp
 #' @importFrom rstan sampling
 #' @importFrom stats lm sd rlnorm rnorm
+#' @importFrom mgcv smoothCon jagam s
 #'
 #' @references
 #' Stan Development Team (2019). RStan: the R interface to Stan. R package version 2.19.2. https://mc-stan.org

### 2019-12-04 doc/hier.R

@@ -1,0 +1,90 @@
+## ------------------------------------------------------------------------
+knitr::opts_chunk$set(fig.width=7, echo=FALSE, warning=FALSE)
+
+## ------------------------------------------------------------------------
+load(file="cfres.rda")
+library(dplyr)
+library(ggplot2)
+source("constants.R")
+
+p <- cfres %>%
+    filter(disease == "stmc" & 
+           area == "bristol" &
+           model %in% c("unsmoothed", "smoothed")) %>% 
+ggplot(aes(x=age, col=model)) +
+    geom_pointrange(aes(y=med, ymin=lower95, ymax=upper95),
+                    position = position_dodge(0.2)) + 
+    facet_grid(cols=vars(gender)) +  
+    coord_cartesian(ylim=c(0, 4)) +
+    scale_x_continuous(breaks=seq(0,100,10)) + 
+    ylab("Case fatality (median, 95% CrI)")  +
+    xlab("Age (years)") +
+    ggtitle("Bristol, stomach cancer")
+p
+
+## ----echo=FALSE----------------------------------------------------------
+
+p <- cfres %>%
+    filter(disease %in% "stmc" & 
+           area == "bristol" &
+           model %in% c("smoothed", "hier50")) %>% 
+ggplot(aes(x=age, col=model)) +
+    geom_pointrange(aes(y=med, ymin=lower95, ymax=upper95),
+                    position = position_dodge(0.2)) + 
+    facet_grid(rows=vars(disease), cols=vars(gender)) +  
+    coord_cartesian(ylim=c(0, 4)) +
+    scale_x_continuous(breaks=seq(0,100,10)) + 
+    ylab("Case fatality (median, 95% CrI)")  +
+    xlab("Age (years)") +
+    ggtitle("Bristol, stomach cancer")
+p
+
+
+## ---- echo=FALSE, warning=FALSE------------------------------------------
+
+hierdata <- readRDS("hierdata.rds")
+hdat <- hierdata[["male"]][["stmc"]]
+dat <- data.frame(
+    area = c(cityregions, regions_exc, regions_inc),
+    prevn = hdat$prev_num[60,],
+    prevd = hdat$prev_denom[60,],
+    mortn = hdat$mort_num[60,],
+    mortd = hdat$mort_denom[60,],
+    incn = hdat$inc_num[60,],
[…]

### 2019-12-04 doc/ihdbristol.R

@@ -1,41 +1,30 @@
-## ----show=FALSE---------------------------------------------------------------
+## ----show=FALSE----------------------------------------------------------
 library(disbayes)
 ihdbristol[ihdbristol$age %in% 50:55, c("age","pop","inc","mort")]
 
-## ----eval=TRUE----------------------------------------------------------------
+## ----eval=TRUE-----------------------------------------------------------
 ihdbristol[ihdbristol$age %in% 50:55, c("age","prev","prevn","prevdenom")]
 
-## -----------------------------------------------------------------------------
+## ------------------------------------------------------------------------
 options(mc.cores = parallel::detectCores())
 
-## ---- eval=TRUE, cache=TRUE, warning=FALSE, message=FALSE---------------------
-dbres <- disbayes(dat = ihdbristol,
-                 inc = "inc", 
-                 inc_denom = "pop", 
-                 prev_num = "prevn", 
-                 prev_denom = "prevdenom",
-                 mort = "mort",
-                 mort_denom = "pop",
-                 eqage = 40
-                 )
-
-## ----eval=TRUE----------------------------------------------------------------
+## ----eval=TRUE-----------------------------------------------------------
 rstan::traceplot(dbres$fit, pars=paste0("cf[", 60:65, "]"))
 
-## ---- eval=TRUE---------------------------------------------------------------
+## ---- eval=TRUE----------------------------------------------------------
 summ <- summary(dbres)
 head(summ)
 
-## ----eval=TRUE----------------------------------------------------------------
+## ----eval=TRUE-----------------------------------------------------------
 summ <- summary(dbres, vars="cf")
 head(summ)
 
-## ----eval=TRUE----------------------------------------------------------------
+## ----eval=TRUE-----------------------------------------------------------
 summ <- summary_disbayes_fit(dbres$fit, vars="cf")
 summu <- summary_disbayes_fit(dbres$fitu, vars="cf")
 head(summu)
 
-## ---- eval=TRUE---------------------------------------------------------------
+## ---- eval=TRUE, warning=FALSE-------------------------------------------
 library(ggplot2)
 plot(dbres) +  ylab("Case fatality") + xlab("Age") + ylim(0,0.5) + xlim(40,100) + 
   geom_line(aes(y=dismod_cf), data=ihdbristol, col="purple")

### 2019-12-04 src/RcppExports.cpp

@@ -8,10 +8,12 @@ using namespace Rcpp;
 
 
 RcppExport SEXP _rcpp_module_boot_stan_fit4disbayes_mod();
+RcppExport SEXP _rcpp_module_boot_stan_fit4disbayes_linear_mod();
 RcppExport SEXP _rcpp_module_boot_stan_fit4disbayes_unsmoothed_mod();
 
 static const R_CallMethodDef CallEntries[] = {
     {"_rcpp_module_boot_stan_fit4disbayes_mod", (DL_FUNC) &_rcpp_module_boot_stan_fit4disbayes_mod, 0},
+    {"_rcpp_module_boot_stan_fit4disbayes_linear_mod", (DL_FUNC) &_rcpp_module_boot_stan_fit4disbayes_linear_mod, 0},
     {"_rcpp_module_boot_stan_fit4disbayes_unsmoothed_mod", (DL_FUNC) &_rcpp_module_boot_stan_fit4disbayes_unsmoothed_mod, 0},
     {NULL, NULL, 0}
 };

### 2019-12-04 src/stanExports_disbayes.h

@@ -37,7 +37,7 @@ stan::io::program_reader prog_reader__() {
     reader.add_event(0, 0, "start", "/include/trans_probs.stan");
     reader.add_event(75, 75, "end", "/include/trans_probs.stan");
     reader.add_event(75, 1, "restart", "model_disbayes");
-    reader.add_event(174, 98, "end", "model_disbayes");
+    reader.add_event(164, 88, "end", "model_disbayes");
     return reader;
 }
 template <typename T0__, typename T1__, typename T2__>
@@ -386,7 +386,6 @@ private:
         std::vector<int> inc_denom;
         vector_d rem;
         int K;
-        vector_d beta_fix;
         matrix_d X;
         double sprior;
 public:
@@ -527,24 +526,14 @@ public:
                 rem(j_1__) = vals_r__[pos__++];
             }
             check_greater_or_equal(function__, "rem", rem, 0);
-            current_statement_begin__ = 91;
+            current_statement_begin__ = 89;
             context__.validate_dims("data initialization", "K", "int", context__.to_vec());
             K = int(0);
             vals_i__ = context__.vals_i("K");
             pos__ = 0;
             K = vals_i__[pos__++];
             check_greater_or_equal(function__, "K", K, 0);
-            current_statement_begin__ = 92;
-            validate_non_negative_index("beta_fix", "K", K);
-            context__.validate_dims("data initialization", "beta_fix", "vector_d", context__.to_vec(K));
-            beta_fix = Eigen::Matrix<double, Eigen::Dynamic, 1>(K);
-            vals_r__ = context__.vals_r("beta_fix");
-            pos__ = 0;
-            size_t beta_fix_j_1_max__ = K;
-            for (size_t j_1__ = 0; j_1__ < beta_fix_j_1_max__; ++j_1__) {
-                beta_fix(j_1__) = vals_r__[pos__++];
-            }
-            current_statement_begin__ = 93;
+            current_statement_begin__ = 91;
             validate_non_negative_index("X", "nage", nage);
             validate_non_negative_index("X", "K", K);
             context__.validate_dims("data initialization", "X", "matrix_d", context__.to_vec(nage,K));
@@ -558,7 +547,7 @@ public:
                     X(j_1__, j_2__) = vals_r__[pos__++];
                 }
             }
-            current_statement_begin__ = 95;
+            current_statement_begin__ = 93;
             context__.validate_dims("data initialization", "sprior", "double", context__.to_vec());
[…]

### 2019-12-05 doc/ihdbristol.Rmd
vignette tweaks

@@
+---
 title: "Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package"
@@
 output: rmarkdown::html_vignette
-    number_sections: true
 vignette: >
- %\VignetteIndexEntry{Vignette Title}
+ %\VignetteIndexEntry{Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package}
  %\VignetteEngine{knitr::rmarkdown}
@@
 ---
+
+```{r}
+knitr::opts_chunk$set(fig.width=7)
+```
 
@@
 
-  * (incidence): $n_a^{(inc)}, r_a^{(inc)}$: given a population of size $n_a$^{(mort)}, $r_a^{(inc)}$ of these are observed to get the disease within the next year
+  * (incidence): given a population of size $n_a^{(mort)}$, $r_a^{(inc)}$ of these are observed to get the disease within the next year
 
-  * (mortality): $n_a^{(mort)}, r_a^{(mort)}$: given a population of size $n_a^{(mort)}$ (with or without the disease), $r_a^{(mort)}$ of these are observed to die from the disease within the next year
+  * (mortality): given a population of size $n_a^{(mort)}$ (with or without the disease), $r_a^{(mort)}$ of these are observed to die from the disease within the next year
 
-  * (prevalence): $n_a^{(prev)}, r_a^{(prev)}$: from a sample of $n_a^{(prev)}$ individuals, $r_a^{(prev)}$ are known to have the disease at age $a$ (and $n_a^{(prev)} - r_a^{(prev)} are known to not have the disease.
+  * (prevalence): from a sample of $n_a^{(prev)}$ individuals, $r_a^{(prev)}$ are known to have the disease at age $a$ (and $n_a^{(prev)} - r_a^{(prev)}$ are known to not have the disease.
 
@@
 
-## Given estimates and denominators
+## ...Given estimates and denominators
 
@@
 
-## Given point and interval estimates
+## ...Given point and interval estimates
 
@@
 
-## Expressing additional uncertainty
+## ...Expressing additional uncertainty
 
@@
 The `disbayes` package uses Stan to fit the Bayesian disease model given here to the data $\mathbf{y} = \{i_a, r_a, n_a\}$ supplied in the `ihdbristol` data frame.
-The function `disbayes` does the work of converting the data to the required format for Stan, and initializing and a set of Markov Chain Monte Carlo simulations, and calling the Stan software to run the simulations.
+The function `disbayes` does the work of converting the data to the required format for Stan, and calling the Stan software to run the simulations.
 
@@
 
-```{r, eval=TRUE}
+```{r, eval=TRUE, warning=FALSE}
 library(ggplot2)
@@
 
-For chronic diseases in general:
-
 * More ways of supplying substantive prior information
@@
 
-* Use data from larger areas as priors for quantities for smaller areas
+* Hierarchical modelling to account for variations between ordered socioeconomic categories / income groups
 
-* Borrow information between men/women?
+* Borrow information between men/women
 
@@
 Relaxing this assumption would be feasible, but possibly computationally expensive.
-	* For IHD in the UK, data on time trends may come from the Health Survey for England, years 1993, 1994, 1988, 2003, 2006, 2011.
+	* For IHD in the UK, data on time trends may come from the Health Survey for England, years 1993, 1994, 1988, 2003, 2006, 2011?
 
@@
 All of these would be theoretically feasible in the above framework. The limits will be computational, and difficult to tell before getting hands on.
+
+## References

### 2019-12-05 doc/ihdbristol.R

@@ -1,3 +1,6 @@
+## ------------------------------------------------------------------------
+knitr::opts_chunk$set(fig.width=7)
+
 ## ----show=FALSE----------------------------------------------------------
 library(disbayes)
 ihdbristol[ihdbristol$age %in% 50:55, c("age","pop","inc","mort")]
@@ -7,6 +10,17 @@ ihdbristol[ihdbristol$age %in% 50:55, c("age","prev","prevn","prevdenom")]
 
 ## ------------------------------------------------------------------------
 options(mc.cores = parallel::detectCores())
+
+## ---- eval=TRUE, cache=TRUE, warning=FALSE, message=FALSE----------------
+dbres <- disbayes(dat = ihdbristol,
+                 inc = "inc", 
+                 inc_denom = "pop", 
+                 prev_num = "prevn", 
+                 prev_denom = "prevdenom",
+                 mort = "mort",
+                 mort_denom = "pop",
+                 eqage = 40
+                 )
 
 ## ----eval=TRUE-----------------------------------------------------------
 rstan::traceplot(dbres$fit, pars=paste0("cf[", 60:65, "]"))

### 2020-03-04 R/ci2num.R
Added metahit results doc

@@ -1,6 +1,8 @@
-##' Estimate binomial numerator and denominator given an estimate and uncertainty interval for a proportion
+##' Estimate binomial numerator and denominator given an estimate and uncertainty interval for a proportion.  
 ##' 
 ##' Based on least squares method provided by the SHELF package TODO REF
+##'
+##' Requires that the estimate and upper and lower limits are all distinct. 
 ##'
 ##' @param est Point estimate
 ##'

### 2021-07-01 metahit/paper_analyses.Rmd
Commit new paper WIP

---
output: html_document
editor_options: 
  chunk_output_type: console
---
Keep this nice and tidy.

```{r}

setwd("~/work/chronic/disbayes/metahit")
source("paper_analyses_header.r")
library(directlabels) # this is nice but do i even use it ?

```

# Data summary

Raw point estimates of mort, inc and prev.   Just women for now.  Curves compared between areas, faceted by disease.   Shows diseases where prevalence and incidence are very low.  Mild differences between areas.

```{r,eval=FALSE}

gbdplot <- gbd %>% 
  filter(cause %in% diseases, 
         location %in% cityregions) %>%
  mutate(mort_prob = num_mort/denom_mort,
         mort_lower = qbeta(0.025, num_mort+0.5, denom_mort+0.5), 
         mort_upper = qbeta(0.975, num_mort+0.5, denom_mort+0.5), 
         prev_prob = num_prev/denom_prev,
         prev_lower = qbeta(0.025, num_prev+0.5, denom_prev+0.5), 
         prev_upper = qbeta(0.975, num_prev+0.5, denom_prev+0.5), 
         inc_prob = num_inc/denom_inc,
         inc_lower = qbeta(0.025, num_inc+0.5, denom_inc+0.5), 
         inc_upper = qbeta(0.975, num_inc+0.5, denom_inc+0.5), 
         location = fct_recode(location, 
                               "North East"="North East (city region)",
                               "West Midlands"="West Midlands (city region)"),
         cause = fct_recode(cause, "Dementia" = "Alzheimer's disease and other dementias",
                            "Lung cancer" = "Tracheal, bronchus, and lung cancer",
                            "COPD" = "Chronic obstructive pulmonary disease"),
         cause = factor(cause, levels = c("Ischemic heart disease", "Stroke", "Dementia",
                                          "COPD", "Breast cancer", "Lung cancer", 
                                          "Colon and rectum cancer","Stomach cancer",
                                          "Uterine cancer")))
gbdplotf <- gbdplot %>% filter(sex=="Female")
gbdplotm <- gbdplot %>% filter(sex=="Male", cause != "Breast cancer")

ht <- 3
pdf("../write/mort_data.pdf", height=ht)
ggplot(gbdplotf, aes(x=ageyr, y=mort_prob, col=location, group=location)) + 
  facet_wrap(~cause, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual mortality probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=mort_lower, ymax=mort_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

pdf("../write/inc_data.pdf", height=ht)
ggplot(gbdplotf, aes(x=ageyr, y=inc_prob, col=location, group=location)) + 
  facet_wrap(~cause, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual incidence probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=inc_lower, ymax=inc_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

pdf("../write/prev_data.pdf", height=ht)
ggplot(gbdplotf, aes(x=ageyr, y=prev_prob, col=location, group=location)) + 
  facet_wrap(~cause, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Prevalence probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=prev_lower, ymax=prev_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5))
dev.off()

pdf("../write/mort_datam.pdf", height=ht)
ggplot(gbdplotm, aes(x=ageyr, y=mort_prob, col=location, group=location)) + 
  facet_wrap(~cause, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual mortality probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=mort_lower, ymax=mort_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

pdf("../write/inc_datam.pdf", height=ht)
ggplot(gbdplotm, aes(x=ageyr, y=inc_prob, col=location, group=location)) + 
  facet_wrap(~cause, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual incidence probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=inc_lower, ymax=inc_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

pdf("../write/prev_datam.pdf", height=ht)
ggplot(gbdplotm, aes(x=ageyr, y=prev_prob, col=location, group=location)) + 
  facet_wrap(~cause, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Prevalence probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=prev_lower, ymax=prev_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5))
dev.off()

```

# Run the Bayesian modelling analyses

```{r,eval=FALSE}
# These are done on a cluster using SLURM so we don't source these files 
# source("paper_analyses_nonhier.r")
# source("paper_analyses_hier.r")
# source("paper_analyses_hier_gender.r")
```

# Read and join all model results

```{r}
reslist <- loolist <- vector(nrow(rundf), mode="list")
for (i in 1:nrow(rundf)){
  fname <- sprintf("results_nonhier/res%s.rds",i)
  if (file.exists(fname)) {
    resi <- readRDS(fname)
    reslist[[i]] <- resi$res
    loolist[[i]] <- resi$loo
  }
}

dput(as.numeric(which(sapply(reslist, is.null)))) # which ones didn't work due to tmpfile conflicts

resnh <- do.call("rbind", reslist) 
resns <- resnh %>% 
  filter(model=="unsmoothed")
resnh <- resnh %>%
  filter(model=="smoothed") %>% 
  mutate(model="Independent areas")
loonh <- do.call("rbind", loolist) %>%
  mutate(model="Independent areas")

resi <- readRDS("results_nonhier/res_eng_utrc_female.rds")
res_utrc <- resi$res; loo_utrc <- resi$loo
resi <- readRDS("results_nonhier/res_eng_stmc_male.rds")
res_stmc_m <- resi$res; loo_stmc_m <- resi$loo
resi <- readRDS("results_nonhier/res_eng_stmc_female.rds")
res_stmc_f <- resi$res; loo_stmc_f <- resi$loo
resnat <- rbind(res_utrc, res_stmc_m, res_stmc_f) %>%
  filter(model=="smoothed") %>% 
  mutate(model="National")
resnh <- resnh %>% 
  dplyr::filter(!(disease %in% c("Stomach cancer", "Uterine cancer"))) %>%
  full_join(resnat)

loonat <- rbind(loo_utrc, loo_stmc_m, loo_stmc_f) %>%
  mutate(model="National")
loonh <- loonh %>% 
  dplyr::filter(!(disease %in% c("Stomach cancer", "Uterine cancer"))) %>%
  full_join(loonat)

reslist <- loolist <- vector(nrow(hierrundf), mode="list")
for (i in setdiff(1:nrow(hierrundf), c())){
  resi <- readRDS(sprintf("results_hier/res%s.rds",i))
  reslist[[i]] <- resi[[1]]
  loolist[[i]] <- resi[[2]]
}
resh <- do.call("rbind", reslist)
resh$model <- "Hierarchical"
resh$area <- factor(as.character(unique(gbd$location))[as.numeric(resh$area)])  # TODO will not be needed if disbayes_hier returns original factor levels for areas instead of mucking with codes
looh <- do.call("rbind", loolist) %>%
  mutate(area = factor(area, labels=areas),
         model = "Hierarchical")

reslist <- loolist <- vector(nrow(hierrungdf), mode="list")
for (i in setdiff(1:nrow(hierrungdf), 7)){ # exclude stomach cancer
  resi <- readRDS(sprintf("results_hier_gender/res%s.rds",i))
  reslist[[i]] <- resi[[1]]
  loolist[[i]] <- resi[[2]]
}
resg <- do.call("rbind", reslist) %>%
  mutate(gender = fct_recode(gender, "Female"="1", "Male"="2"),
         model = "Hierarchical joint gender",
         area = factor(as.character(unique(gbd$location))[as.numeric(area)])
  )
loog <- do.call("rbind", loolist) %>%
  mutate(area = factor(area, labels=areas),
         model = "Hierarchical joint gender",
         gender = factor(gender, labels=genders))

reslist <- vector(nrow(trendrundf), mode="list")
for (i in setdiff(1:nrow(trendrundf), c(2))){
  load(sprintf("results_trend/res%s.rda",i))  
  reslist[[i]] <- res
}

restr <- do.call("rbind", reslist)
restr$model <- "Time trends"

disease_shorten <- c("Dementia" = "Alzheimer's disease and other dementias",
                     "Lung cancer" = "Tracheal, bronchus, and lung cancer",
                     "COPD" = "Chronic obstructive pulmonary disease")
disease_order <- c("Ischemic heart disease", "Stroke", "Dementia", "COPD", "Breast cancer", "Lung cancer", 
                   "Colon and rectum cancer","Stomach cancer","Uterine cancer")

resall <- full_join(resnh, resh) %>% 
  full_join(resg) %>%  # remove after new run
  full_join(restr) %>%
  mutate(disease = fct_recode(disease, !!!disease_shorten),
         disease = factor(disease, levels = disease_order),
         year = if_else(model == "Time trends", year + 1917, 2017))

looall <- full_join(loonh, looh) %>%
  full_join(loog) %>%
  mutate(disease = fct_recode(disease, !!!disease_shorten),
         disease = factor(disease, levels = disease_order))

```

# Independent areas

* Results from fitting disbayes model by disease, gender and area separately

```{r}

resnh %>% filter(disease == "Stomach cancer", var=="cf", area=="Bristol", gender=="Female")

## Convergence checks.  
## Hasn't converged for stomach and uterine cancer, in areas where there were no deaths or small numbers of deaths.
## We use national data for these instead. 
## What's happened now for stroke?

resnc <- resnh %>% filter(Rhat > 2) 
resnc <- resnh %>% filter(Rhat > 1.2) 
resnc <- resnh %>% filter(Rhat > 1.1) 
resnc <- resnh %>% filter(Rhat > 1.05) 
table(resnc$disease, resnc$gender)
table(resnc$disease, resnc$gender, resnc$area)["Stomach cancer",,]
gbd %>% filter(location=="Bristol", cause=="Uterine cancer") %>% select(ageyr, num_mort, num_inc, num_rem)  # No deaths. 
gbd %>% filter(cause=="Uterine cancer", location %in% cityregions) %>% pull(num_mort) %>% sum
gbd %>% filter(cause=="Uterine cancer", location %in% cityregions) %>% pull(num_inc) %>% sum

## Age curves for all diseases and areas

pdf("../write/all_cf_indep.pdf",width=6,height=6)

gg <- resall %>% 
  dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
  mutate(p50 = 1 - exp(-`50%`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  ggplot(aes(x=age, y=p50, group=area)) + 
  geom_ribbon(aes(ymin=plow, ymax=pupp), 
              fill="deepskyblue") +
  geom_line() + 
  facet_grid(cols=vars(gender), rows=vars(disease), scales = "free_y", switch = "y") + 
  xlab("Age (years)") + 
  ylab("Annual case fatality probability") + 
  xlim(60,100) + 
  theme(strip.text.y.left = element_text(angle = 0))

grob <- ggplotGrob(gg)
# gtable::gtable_show_layout(grob)
idx <- which(grob$layout$name %in% c("panel-5-2","panel-9-2"))
for (i in idx) grob$grobs[[i]] <- grid::nullGrob()
grid::grid.draw(grob)

dev.off()

#  TODO  more investigation of results for tbalc, carc 
## where theres between area var and unc 
## Lung cancer - lots of BAV. CARC less so. 
## What to we get from hier model in each case?

example_age <- 90
resall %>% 
  dplyr::filter(model %in% c("Independent areas"), var=="cf",
                disease == "Lung cancer", gender=="Male",age==example_age) %>%
ggplot(aes(x=`50%`, y=area, col=model)) +
    coord_cartesian() +
    geom_point(position=ggstance::position_dodgev(height=0.4)) +
    geom_errorbarh(aes(xmin=`2.5%`, xmax=`97.5%`),
                   position=ggstance::position_dodgev(height=0.4)) +
    xlab(sprintf("Case fatality for age %s", example_age) )

```

# Hierarchical model

```{r,cache=TRUE}

## convergence checks. look at Rhat
## i=16 only one with divergent- utrc - we're not using this model anyway
## stomach cancer is only one that hasn't converged

summary(resh$Rhat) 
resh %>% filter(gender=="Male", disease=="Stomach cancer", var=="cf", age>60, age<70) 
resnc <- resh %>% filter(Rhat > 2) 
resnc <- resh %>% filter(Rhat > 1.1) 
table(resnc$disease, resnc$gender)
table(resnc$disease, resnc$gender, resnc$area)
resh %>% filter(gender=="Male", disease=="Ischemic heart disease", age==60, var=="cf") %>% head
resh %>% filter(gender=="Male", disease=="Stroke", age==60, var=="cf") %>% head
resh %>% filter(gender=="Male", disease=="Stomach cancer", age==60, var=="cf") %>% head
resh %>% filter(disease=="Uterine cancer", age==60, var=="cf") %>% head
resh %>% filter(gender=="Male", area=="Bristol", age==60, var=="cf") %>% head

## Between area variability in point and interval ests, hier and nonhier models  
## Specific age and gender, compare between areas 
## Can use this to assess benefit from hier models. 
## Shrinkage is moderate.

example_age <- 90
dat <- resall %>% filter(age==example_age, var=="cf", 
                         !(disease %in% c("Stomach cancer","Uterine cancer")),
                         #      disease %in% c("Colon and rectum cancer"),
                         model %in% c("Independent areas","Hierarchical"))
ggplot(dat, aes(x=`50%`, y=area, col=model)) +
  #   coord_cartesian(xlim=c(0.01, 0.05)) +
  facet_grid(rows=vars(disease), cols=vars(gender)) + 
  geom_point(position=ggstance::position_dodgev(height=0.4)) +
  geom_errorbarh(aes(xmin=`2.5%`, xmax=`97.5%`),
                 position=ggstance::position_dodgev(height=0.4)) +
  xlab(sprintf("Case fatality for age %s", example_age) )

## Between area variability in age curves, point ests.  CARC.  Shrinkage for oldest

pdf("../write/hier_crc.pdf", width=6, height=4)

resall %>% 
  filter(var=="cf", 
         !disease %in% c("Stomach cancer","Uterine cancer"),
         disease %in% c(
           #   "Dementia"
           "Colon and rectum cancer"
           #       "Lung cancer"
         ),
         `97.5%` < 10,
         area %in% cityregions, 
         model %in% c("Independent areas","Hierarchical")) %>%
  mutate(model = relevel(factor(model), "Independent areas")) %>%
  ggplot(aes(x=age, col=area, group=area)) +
  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), fill="gray80", col="gray80") +
  geom_line(aes(y=`50%`)) + 
  coord_cartesian(ylim=c(0, 1), 
                  xlim=c(70,  100)) +
  scale_x_continuous(breaks=seq(0,100,10)) + 
  ylab("Case fatality rate")  +
  facet_grid(cols=vars(model), rows=vars(gender)) + 
  xlab("Age (years)") + 
  theme_bw() + labs(col="")

dev.off()

```

# Hierarchical model with gender effect independent of area

```{r}

resg <- resg %>% filter(disease != "Stomach cancer") # only national estimates
# convergence check
summary(resg$Rhat)  # 
resnc <- resg %>% filter(Rhat > 2) 
resnc <- resg %>% filter(Rhat > 1.1) 
resnc <- resg %>% filter(Rhat > 1.01) 
table(resnc$disease)
table(resnc$disease, resnc$area)

## point est curves compared between areas 
rescf <- resg %>% filter(gender=="Male", var=="cf")
rescf <- resg %>% filter(gender=="Female", var=="cf")
ggplot(rescf, aes(x=age, col=area)) +
  geom_line(aes(y=`50%`)) + 
  coord_cartesian(ylim=c(0, 1), 
                  xlim=c(50,  100)) +
  scale_x_continuous(breaks=seq(0,100,10)) + 
  ylab("Case fatality (median, 95% CrI)")  +
  facet_wrap(~disease, ncol=2) + 
  xlab("Age (years)")

## Show whether gender effect is common between areas 
## todo pick diseases to illustrate.  IHD, LC biggesc contrast 
## how to show for all ages.

dat <- resall %>%
  filter(var=="cf",
         model %in% c("Hierarchical","Hierarchical joint gender"),
         disease %in% c("Ischemic heart disease", "Stroke", "Lung cancer", "Colon and rectum cancer", "Dementia", "COPD")) %>%
  filter(disease %in% c("Ischemic heart disease", "Lung cancer")) %>%
  mutate(model = fct_recode(model, 
                            "Different between areas" = "Hierarchical",
                            "Common effect" = "Hierarchical joint gender"))
example_age <- 70

#DEBUGME SHOW RAW MORT DATA FOR LC FOR AREAS, SEE IF GENDER DIFFERENCE
gbd %>% filter(ageyr==70, 
               cause=="Tracheal, bronchus, and lung cancer", 
               location %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
  mutate(p_mort = num_mort/denom_mort, p_prev=num_prev/denom_prev, mpratio=p_mort/p_prev) %>%
  select(sex, location, num_mort, denom_mort, p_mort, p_prev, mpratio)
## YES: MORT/PREV AGREES WITH CF

i <- 3
resi <- readRDS(sprintf("results_hier_gender/res%s.rds",i))
resi <- readRDS(sprintf("results_hier/res%s.rds",3))
resi <- readRDS(sprintf("results_hier/res%s.rds",10))

resi[[1]] %>% 
  filter(var=="cf", between(age, 68, 72), area==1)
resi[[1]] %>% 
  filter(var=="cf", between(age, 68, 72), area==1)

gbd %>% filter(ageyr==70, 
               cause=="Tracheal, bronchus, and lung cancer", 
               location %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
  group_by(location) %>%
  mutate(num_mort=sum(num_mort), denom_mort=sum(denom_mort), 
         num_prev=sum(num_prev), denom_prev=sum(denom_prev)) %>%
  mutate(p_mort = num_mort/denom_mort, p_prev=num_prev/denom_prev, mpratio=p_mort/p_prev) %>%
  select(sex, location, num_mort, denom_mort, p_mort, p_prev, mpratio)

pdf("../write/gender.pdf", width=6, height=4) # NOTE one for IHD, one for lung cancer

dat %>% filter(age==example_age) %>%
ggplot(aes(x=`50%`, y=area, lty=model, col=gender)) +
  #    coord_cartesian(xlim=c(0.001, 0.2)) +
  facet_wrap(~disease, nrow=1, scales="free_x") + 
  geom_point(position=ggstance::position_dodgev(height=0.6), size=1.5) +
  geom_errorbarh(aes(xmin=`2.5%`, xmax=`97.5%`),
                 position=ggstance::position_dodgev(height=0.6),
                 lwd = 0.7) +
  xlab(sprintf("Case fatality for age %s", example_age) ) + ylab("") + 
  scale_linetype_manual(values = c("Common effect" = "solid",
                                   "Different between areas" = "dashed"
                                   )) +
  labs(col="", lty="Gender effect") +
  theme_bw() + 
  theme(legend.text = element_text(size=7)) +  
  guides(linetype = guide_legend(reverse = TRUE))

dev.off()

```

## Lookup table for all results
100 * 9 * 18 * 2 * 3 results for each of inc and CF

How to distribute this on github?  A data file in the disbayes package?  
A separate package with code and results?  yea should be standard format for these.  what does hadley do 
Just data and vignettes, no code.  yeah. 
where to put the analysis scripts and slurm script? inst/scripts or suchlike. 
yeah

```{r}
app_tab <- resall %>%
  filter(var %in% c("cf","inc","rem")) %>%
  select(disease, var, gender, age, area, model, `25%`, `50%`, `97.5%`) %>%
  arrange(disease, var, gender, age, area, model)
app_tab
```

# Cross validatory comparison

First compare nonhier with hier.

```{r}
looall <- loonh %>% mutate(model="Independent areas") %>%
  full_join(looh %>% mutate(model="Hierarchical")) %>%
  filter(!(model=="National"))

loosub <- looall %>% 
#  filter(disease=="Colon and rectum cancer", var=="mort", gender=="Male",
  filter(disease=="Ischemic heart disease", var=="inc", gender=="Male",
         area %in% cityregions) %>%
  select(age, elpd_loo, model, area) %>%
  pivot_wider(names_from=model, values_from=elpd_loo) %>%
  mutate(elpd_diff = `Hierarchical` - `Independent areas`) %>%
  ## sum ELPD by five year age group
  mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
  group_by(area, agegroup) %>%
  summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
  mutate(agemid = as.numeric(as.character(factor(agegroup, labels = seq(2.5, 97.5, by=5)))))

ggplot(loosub, aes(x=agemid, y=elpd_diff, colour=area)) +
  geom_hline(yintercept=0, col="gray", lwd=2) +
    geom_line() + 
  geom_point() + 
  scale_x_continuous(limits=c(30,100), breaks = seq(0,100,10)) + 
  geom_text(data=data.frame(x=c(50,50), y=c(-Inf,Inf)), 
            aes(x=x, y=y), 
            inherit.aes=FALSE,
            label=c("Independent areas preferred","Hierarchical preferred"), 
            vjust=c(0,1), hjust=c(0,0)) +
  geom_dl(aes(label=area), method= list(dl.combine("first.points", "last.points")))

## so why is hier better for mid ages?  worse for below 
## surely age 95 value is iffy due to weak data.

## why two groups of areas?  It's not the area numerator or denominator size.
## Labels are right way round. higher ELPD good

gbd %>% filter(cause=="Colon and rectum cancer") %>% 
  group_by(location) %>% filter(areatype=="cityregion") %>% 
  summarise(n=sum(num_mort)) %>% arrange(n)

looall %>% filter(age>90, age <92, gender=="Male", disease=="Colon and rectum cancer", var=="mort") %>%
  select(area, p_loo, model) %>%
  pivot_wider(names_from=model, values_from=p_loo) %>% as.data.frame()

## These results are hard to explain
## Suspect there is no point using cross validation to compare hier and nonhier
## The results are the same where there is data
## Hier only helps for extrapolation to oldest ages where there is no data
## That is a matter of judgement that can use exchangeability.

```

Then compare hier with hier additive gender

```{r}

# elpd : high is good. positive elpd diff favours additive gender model 
loohg <- looall %>% 
  filter(model %in% c("Hierarchical","Hierarchical joint gender")) %>%
  select(var, area, age, gender, disease, model, elpd_loo) %>%
  pivot_wider(names_from=model, values_from=elpd_loo) %>%
  mutate(elpd_diff = `Hierarchical joint gender` - `Hierarchical`) %>%
  mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
  group_by(var, area, gender, disease, agegroup) %>%
  summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
  mutate(agemid = as.numeric(as.character(factor(agegroup, labels = seq(2.5, 97.5, by=5)))))

loohg %>% 
  filter(disease %in% c("Ischemic heart disease","Lung cancer")) %>% 
  group_by(agemid, disease) %>%
  summarise(elpd = sum(elpd_diff)) %>%
ggplot(aes(x=agemid, y=elpd, col=disease, group=disease)) + 
  geom_hline(yintercept=0) + geom_line() + geom_point()

# ELPD all over the place for different ages 
# positive: common effects.  negative: different effects 
# Different effects for lung cancer and dem
# not transparent though. 
# IHD vs LC is still biggest contrast 
# not sure makes sense to split by age.  model defines risk as smooth function of age

loohg %>%
  group_by(var, disease) %>%
  summarise(elpd = sum(elpd_diff)) %>%
  arrange(disease) %>%
  pivot_wider(names_from=var, values_from=elpd) %>%
  filter(!is.na(inc)) %>%
  mutate(rem=ifelse(is.na(rem), 0, rem),
         overall = inc+mort+prev+rem)

```

# Time trends

Damn, Absolute incidence estimates are way off both the model results, 
but closest to the non trend one where they match up to age 70.

why.  We are adjusting them to be consistent with prevalence and mortality data. 
But we want to give more weight somehow to current incidence data.  Why tho, if we also believe the current 
prev and mort data? are we trying to make them all fit into a bad model?

Right the fitted values for year 2010 and earlier match up with current data.  There's a big jump around 2010

But then they will carry on changing through time in the future if we use them in a model

```{r}

pdf("../write/trend_res.pdf")

datt <- gbd %>% 
  filter(location=="Leeds", cause=="Ischemic heart disease") %>%
  mutate(inc_prob = num_inc / denom_inc, 
         inc_rate = -log(1-inc_prob), 
         var = "Incidence", 
         gender = sex)

resoy <- resall %>% filter((model=="Time trends" & year > 2000 ),
                          area=="Leeds", 
                          disease=="Ischemic heart disease",
                          var %in% c("inc")) %>% mutate(var="Incidence")

rest <- resall %>% filter((model == "Independent areas") | (model=="Time trends" & year==2017),
                          area=="Leeds", 
                          disease=="Ischemic heart disease",
                          var %in% c("inc","cf")
#                          gender=="Male"
)
rest %>% 
  filter(between(age, 50, 90)) %>% 
  mutate(var = fct_recode(var, "Incidence"="inc", "Case fatality"="cf"),
         model = fct_recode(model, "No time trends" = "Independent areas")) %>%
ggplot(aes(x=age, y=`50%`, col=model, fill=model, 
           group=interaction(gender, model))) +
  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), alpha=0.2) +
  facet_wrap(vars(gender, var), nrow=2, scales="free_y") +
  geom_line() + 
  xlab("Age") + ylab("") + 
  labs(col="", fill="") + 
  geom_line(data=datt, aes(x=ageyr, y=inc_rate), inherit.aes = FALSE) + 
  geom_line(data=resoy, aes(x=age, y=`50%`, col=factor(year), group=year), lty=2, inherit.aes = FALSE)
dev.off()

```

### 2021-07-01 metahit/cfest.r

@@ -1,0 +1,97 @@
+## simple cf estimator 
+
+## assuming 0.5 discrete time unit 
+## people getting disease in a year get it at halfway point 
+## incidence = p12
+## case fatality = p23 
+
+## * mort = (1-prev)*p12*p23 + prev*(p23 + (1-p23)*p23)
+## * mort = (1-prev)*p12*p23 + prev*(2*p23 - p23^2)
+## * mort = p23* ( (1-prev)*p12 + prev + (1 - p23))
+## quadratic equation for p23 with
+## a = -prev 
+## b = (1-prev)*p12 + 2*prev
+## c = -dm
+
+## But this needs prevalence to be known. 
+## For cancers that's fine I suppose as prev from register? 
+
+## TODO does even simpler version work ? 
+##  dm = (1-prev)*p12*p23 + prev*p23
+## p23 = dm / ((1-prev)*p12 + prev)
+## That does even better!! 
+
+source("constants.R")
+load("cfres.rda")
+
+gender <- "male"
+area <- "bristol"
+
+## Compare single-area spline model-based CF estimate to naive estimate 
+## for various diseases
+## Agrees well except in cases where there no information 
+## Naive estimate shows increasing trend for stmc - clearly shows bias in prior 
+## And for lvrc, tbalc naive estimate clearly noisy and unsafe.
+## Articulate what assumptions naive estimate is based on
+## discrete time, only move at halfway point
+## rates equal probs 
+
+## needs nondecreasing constraint.  constant would do but evidence from naive est of increasing rate.
+
+resall <- NULL
+for (disease in diseases){
+    dat <- readRDS(file.path(datadir,sprintf("%s_%s_%s.rds", area, gender, disease)))
+    a = dat$prev
+    b = (1 - dat$prev)*dat$inc + 2*dat$prev
+    c = -dat$mort
+    cf <- (-b + sqrt(b^2 - 4*a*c))/(2*a)
+    cf <- dat$mort / ((1 - dat$prev)*dat$inc + dat$prev)
+    cf
+    resmod <- cfres_nonhier[cfres_nonhier$area=="bristol" & cfres_nonhier$gender=="male" & cfres_nonhier$disease==disease,]
+    resmod$cf <- cf
+    resall <- rbind(resall, resmod)
+}
+
+ggplot(resall, aes(x=age, y=cf)) +
[…]

### 2021-07-01 metahit/hpc.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+#SBATCH -J metahit
+#SBATCH --time=12:00:00
+#SBATCH --cpus-per-task=2
+#SBATCH --array=[1]%8
+
+# module load r-4.0.2-gcc-5.4.0-xyx46xb
+module load r-3.6.1-gcc-5.4.0-zrytncq
+
+Rscript paper_analyses_gender.r

### 2021-07-01 metahit/paper_analyses_gender.r

@@ -1,0 +1,48 @@
+
+source("paper_analyses_header.r")
+options(mc.cores = 2)
+
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+runs_todo <- 1:nrow(hierrungdf) 
+i <- runs_todo[task_id]
+
+mhi <- gbd %>%
+    filter(cause==hierrungdf$disease[i]) %>%
+    droplevels
+
+db <- disbayes_hier_gender(data=mhi,
+                    group = "location",
+                    gender = "sex",
+                    group_init = "London",
+                    inc_num = "num_inc", inc_denom = "denom_inc",
+                    mort_num = "num_mort", mort_denom = "denom_mort",
+                    prev_num = "num_prev", prev_denom = "denom_prev",
+                    rem_num = "num_rem", rem_denom = "denom_rem",
+                    model = hierrungdf$model[i], 
+                    remission = hierrungdf$remission[i],
+                    age="ageyr", smooth_cf=TRUE,
+                    eqage=hierrundf$eqage[i], 
+                    nfold_int_guess = 5,
+                    nfold_int_upper =  50,
+                    nfold_slope_guess = 2,
+                    nfold_slope_upper =  20,
+                    cf_init = 0.1,
+                    chains=2, iter=1000)
+
+if (0){
+  library(bayesplot) 
+  mcmc_trace(db$fit,pars=c("sd_inter"), 
+             np = nuts_params(db$fit))
+  pairs(db$fit, pars=c("sd_inter","sd_slope[1]"))
+  color_scheme_set("darkgray")
+  mcmc_parcoord(db$fit, np = nuts_params(db$fit), pars=c("sd_inter","sd_slope[1]"))
+}
+
+res <- tidy(db) %>%
+  mutate(disease=hierrungdf$disease[i])
+
+loo <- looi_disbayes_hier_gender(db) %>%
+    mutate(disease=hierrungdf$disease[i])
+
+saveRDS(list(res=res,loo=loo), file= paste0("results_hier_gender/res", i, ".rds"))
+cat(sprintf("Completed case %s\n", i), file="metahit-hpc-log.txt", append=TRUE)

### 2021-07-01 metahit/paper_analyses_header.r

@@ -1,0 +1,76 @@
+library(tidyverse)
+library(broom)
+library(rstan)
+library("disbayes")
+
+## Processed version of GBD data, produced as the final output of dataprep.Rmd 
+gbd <- readRDS("gbddb.rds")
+
+areas <- unique(gbd$location)
+cityregions <- unique(gbd$location[gbd$areatype=="cityregion"])
+genders <- unique(gbd$sex)
+diseases_male <- c("Ischemic heart disease", "Stroke", "Tracheal, bronchus, and lung cancer", "Colon and rectum cancer","Alzheimer's disease and other dementias", "Chronic obstructive pulmonary disease","Stomach cancer") # no remission data for liver 
+diseases_female <- c(diseases_male, "Breast cancer","Uterine cancer")
+diseases <- unique(c(diseases_male, diseases_female))
+cancers <- c("Tracheal, bronchus, and lung cancer", "Colon and rectum cancer", "Stomach cancer", 
+             "Breast cancer", "Uterine cancer")
+
+dcodes <- data.frame(disease=diseases, 
+                     dcode = c("ihd","stroke","tbalc","carc","adaod","copd","stmc","brc","utrc"))
+
+rundf <- data.frame( 
+  gender = rep(rep(c("Male","Female"),c(length(diseases_male),length(diseases_female))),length(areas)),
+  disease = rep(c(diseases_male, diseases_female), length(areas)),
+  area = rep(areas, each=length(diseases_male)+length(diseases_female))
+) %>%
+  mutate(remission = disease %in% cancers, 
+         runid = 1:n(),
+         dcode = dcodes$dcode[match(disease, dcodes$disease)],
+         increasing = ifelse(disease %in% c("Alzheimer's disease and other dementias",
+                                            "Stomach cancer",
+                                            "Tracheal, bronchus, and lung cancer",
+                                            "Uterine cancer"),
+                             TRUE, FALSE),
+         eqage = case_when(disease == "Ischemic heart disease" ~ 30, 
+                           disease == "Stroke" ~ 30, 
+                           disease == "Alzheimer's disease and other dementias" ~ 70, 
+                           disease == "Stomach cancer" ~ 80, 
+                           TRUE ~ 50))
+rundf$eqage[c(101,117)] <- 75 # ADAOD, london, northeast cr, male 
+
+hierrundf <- data.frame(gender = rep(c("Male","Female"), c(length(diseases_male), length(diseases_female))),
+                        disease = c(diseases_male, diseases_female),
[…]

### 2021-07-01 metahit/paper_analyses_hier.r

@@ -1,0 +1,56 @@
+print(unlist(R.Version()))
+
+source("paper_analyses_header.r")
+
+options(mc.cores = 2)
+
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+
+runs_todo <- 1:nrow(hierrundf) 
+i <- runs_todo[task_id]
+
+mhi <- gbd %>%
+  filter(sex==hierrundf$gender[i], cause==hierrundf$disease[i]) %>%
+  droplevels 
+
+db <- disbayes_hier(data=mhi,
+                    group = "location",
+                    group_init = "London",
+                    inc_num = "num_inc", inc_denom = "denom_inc",
+                    mort_num = "num_mort", mort_denom = "denom_mort",
+                    prev_num = "num_prev", prev_denom = "denom_prev",
+                    rem_num = "num_rem", rem_denom = "denom_rem",
+                    model = hierrundf$model[i], 
+                    remission = hierrundf$remission[i],
+                    age="ageyr", smooth_cf=TRUE,
+                    eqage=hierrundf$eqage[i], chains=2, iter=1000,
+                    nfold_int_guess = 5,
+                    nfold_int_upper =  50,
+                    nfold_slope_guess = 2,
+                    nfold_slope_upper =  20,
+                    sprior = 1,
+                    refresh = 1,  # TURN IT DOWN BELOW 100
+                    verbose = TRUE
+                    #, 
+                    #control=list(max_treedepth=20) # don't think adapt_delta adds anything, just slows
+                    )
+
+# saveRDS(db, file="~/scratch/chronic/db_hier_test.rds")
+
+if (0){
+  library(bayesplot) 
+  mcmc_trace(db$fit,pars=c("lambda_smooth","sd_inter"), 
+             np = nuts_params(db$fit))
+  pairs(db$fit, pars=c("sd_inter","sd_slope[1]"))
+  color_scheme_set("darkgray")
+  mcmc_parcoord(db$fit, np = nuts_params(db$fit), pars=c("sd_inter","sd_slope[1]"))
+}
+
+res <- tidy(db) %>%
+  mutate(gender=hierrundf$gender[i], disease=hierrundf$disease[i])
+
+loo <- looi_disbayes_hier(db) %>%
+    mutate(gender=rundf$gender[i], disease=rundf$disease[i])
+
[…]

### 2021-07-01 metahit/paper_analyses_national.r

@@ -1,0 +1,47 @@
+source("paper_analyses_header.r")
+nchains <- 2
+options(mc.cores = nchains)
+
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+runs_todo <- 1:nrow(natrundf)  
+i <- runs_todo[task_id]
+
+gbdeng <- gbd %>% select(-areatype) %>% 
+  group_by(ageyr, sex, cause) %>% # should contain all English city regions and mutually exclusive regions
+  summarise(num_mort=sum(num_mort), num_inc=sum(num_inc), 
+            num_prev=sum(num_prev), num_rem=sum(num_rem),
+            denom_mort=sum(denom_mort), denom_inc=sum(denom_inc), 
+            denom_prev=sum(denom_prev), denom_rem=sum(denom_rem),
+            .groups = "drop") 
+
+db <- disbayes(data = gbdeng %>% filter(cause==natrundf$disease[i],
+                                        sex==natrundf$gender[i]),
+               inc_num = "num_inc", inc_denom = "denom_inc",
+               mort_num = "num_mort", mort_denom = "denom_mort",
+               prev_num = "num_prev", prev_denom = "denom_prev",
+               rem_num = "num_rem", rem_denom = "denom_rem",
+               remission = natrundf$remission[i],
+               increasing_cf = natrundf$increasing[i],
+               const_cf = FALSE,
+               age="ageyr", smooth_cf=FALSE,
+               eqage= natrundf$eqage[i], 
+               chains=nchains, iter=1000)
+
+res <- tidy(db) %>%
+  mutate(gender=natrundf$gender[i], disease=natrundf$disease[i], area=natrundf$area[i])
+## area should all be "England"
+
+loo <- looi_disbayes(db) %>%
+    mutate(gender=natrundf$gender[i], disease=natrundf$disease[i], area=natrundf$area[i])
+
+saveRDS(list(res=res,loo=loo), file=sprintf("results_nonhier/res_eng_%s.rds",natrundf$runlabel[i]))
+
+if (0){
+    p <- res %>% filter(var=="cf") %>%
+        ggplot(aes(x=age, y=`50%`)) + 
+        geom_ribbon(aes(ymin=`2.5%`,ymax=`97.5%`), fill="deepskyblue") +
+        geom_line() + 
+        xlab("Age (years)") + 
+        ylab("Case fatality rate") 
+    p
+}

### 2021-07-01 metahit/paper_analyses_nonhier.r

@@ -1,0 +1,48 @@
+#SBATCH --array=[48,51,77,94,95,117,118,120,121,124,125,129,130,133,134,136,137,140,141,145,146,149,150,152,153,156,157,161,162,165,166,168,169,172,173,177,178,181,182,184,185,188,189,193,194,197,198,200,201,204,205,209,210,213,214,216,217,220,221,222,223,225,226,229,230,232,233,236,237,241,242,245,246,248,249,252,253,257,258,261,262,264,265,268,269]%7
+
+source("paper_analyses_header.r")
+nchains <- 2
+options(mc.cores = nchains)
+
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+runs_todo <- 1:nrow(rundf)  # 272, run in groups of 8. 
+i <- runs_todo[task_id]
+
+#exists <- if (file.exists(tempdir())) "exists" else "doesn`t exist"
+#cat(sprintf("Case %s tempdir %s %s\n", i, tempdir(), exists),
+#    file="metahit-hpc-log.txt", append=TRUE)
+
+# i <- 16
+
+mhi <- gbd %>%
+  filter(sex==rundf$gender[i], cause==rundf$disease[i], location==rundf$area[i])
+db <- disbayes(data=mhi,
+               inc_num = "num_inc", inc_denom = "denom_inc",
+               mort_num = "num_mort", mort_denom = "denom_mort",
+               prev_num = "num_prev", prev_denom = "denom_prev",
+               rem_num = "num_rem", rem_denom = "denom_rem",
+               remission = rundf$remission[i],
+               increasing_cf = rundf$increasing[i],
+               const_cf = FALSE,
+               age="ageyr", smooth_cf=TRUE,
+                eqage=rundf$eqage[i], 
+               chains=nchains, iter=1000)
+
+res <- tidy(db) %>%
+  mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
+
+loo <- looi_disbayes(db) %>%
+    mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
+    
+if (0){
+  res %>% filter(var=="cf") %>% select(age, Rhat)
+  library(bayesplot)
+  mcmc_trace(db$fit, pars=paste0("cf[",1:4,"]"))
+}
+
[…]

### 2021-07-01 metahit/paper_analyses_trend.r

@@ -1,0 +1,46 @@
+source("paper_analyses_header.r")
+nchains <- 2
+options(mc.cores = nchains)
+
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+runs_todo <- 1:nrow(trendrundf) 
+i <- runs_todo[task_id]
+
+# i <- 1
+
+mhi <- gbd %>% filter(cause=="Ischemic heart disease",
+                      location==trendrundf$area[i], 
+                      sex==trendrundf$gender[i])
+
+trends <- ihdtrends %>% 
+  filter(outcome=="Incidence", gender==trendrundf$gender[i]) %>%
+  pivot_wider(names_from="year", values_from="p2017") %>%
+  select(-age, -gender, -outcome) %>% 
+  as.matrix()
+
+cftrends <- ihdtrends %>% 
+  filter(outcome=="Case fatality", gender==trendrundf$gender[i]) %>%
+  pivot_wider(names_from="year", values_from="p2017") %>%
+  select(-age, -gender, -outcome) %>% 
+  as.matrix()
+
+db <- disbayes(data = mhi,
+               inc_num = "num_inc", inc_denom = "denom_inc",
+               mort_num = "num_mort", mort_denom = "denom_mort",
+               prev_num = "num_prev", prev_denom = "denom_prev",
+               rem_num = "num_rem", rem_denom = "denom_rem",
+               remission = trendrundf$remission[i], 
+               increasing_cf = trendrundf$increasing[i], 
+               const_cf = FALSE,
+               age="ageyr", smooth_cf=TRUE,
+               inc_trend_mult = trends, 
+               cf_trend_mult = cftrends, 
+               eqage= trendrundf$eqage[i], 
+               chains=nchains, iter=1000)
+
+# saveRDS(db, file="~/scratch/chronic/fit_trend.rds")
+
+res <- tidy(db) %>%
+  mutate(gender=trendrundf$gender[i], disease="Ischemic heart disease", area=trendrundf$area[i])
+
+saveRDS(res, file= paste0("results_trend_sens/res", i, ".rds"))

### 2021-07-01 metahit/slurm_bsu.sh

@@ -1,0 +1,9 @@
+#!/bin/bash
+#SBATCH -J metahit
+#SBATCH --time=12:00:00
+#SBATCH --cpus-per-task=2
+#SBATCH --array=[1-34]%8
+module purge
+module load default-login
+
+Rscript paper_analyses_trend.r

### 2021-07-01 data-raw/gbd_process.Rmd
Commit new package work in progress

---
title:  "Preparing data from the Global Burden of Disease for METAHIT modelling"
author:
- "Jackson, Chris <chris.jackson@mrc-bsu.cam.ac.uk>"
date:   "`r Sys.time()`"
output:
    html_document:
        code_folding: show
params:
    date:   !r Sys.Date() - 1
editor_options: 
  chunk_output_type: console
---

# Read in and combine data

First load required packages and specify local file paths

```{r}
library(tidyverse)
library(progress)
library(tempdisagg)
relative_path_gbd <- "~/scratch/chronic/metahit_data"

```

Read look-up table mapping between local authorities and city regions, and add names for regions that are not city regions.  No need to bother with matching regions to countries.

```{r,eval=FALSE}
regions <- c("East Midlands", "East of England", "Greater London", "North East England", 
                  'North West England', "South East England", "South West England", "Yorkshire and the Humber", "West Midlands")

# LTLAs to city regions and regions 
local_government_areas <- read_csv(file.path(relative_path_gbd, "mh_regions_lad_lookup.csv")) %>% rename(region="gordet")
lad <- 
    read.csv("~/work/chronic/belen/Local_Authority_District_to_Region__December_2017__Lookup_in_England_.csv")
utla <- 
  read.csv("~/work/chronic/belen/Lower_Tier_Local_Authority_to_Upper_Tier_Local_Authority__December_2017__Lookup_in_England_and_Wales.csv")

## Make UTLA to region lookup
lad_wales <- c("Isle of Anglesey", "Gwynedd", "Conwy", "Denbighshire", "Flintshire", 
"Wrexham", "Ceredigion", "Pembrokeshire", "Carmarthenshire", 
"Swansea", "Neath Port Talbot", "Bridgend", "Vale of Glamorgan", 
"Cardiff", "Rhondda Cynon Taf", "Caerphilly", "Blaenau Gwent", 
"Torfaen", "Monmouthshire", "Newport", "Powys", "Merthyr Tydfil")
utlareg <- utla %>% 
  filter(!duplicated(UTLA17NM)) %>%
  select(LTLA17NM, UTLA17NM) %>%
  rename("LAD17NM"=LTLA17NM) %>%
  left_join(lad, by="LAD17NM") %>%
  filter(!(LAD17NM %in% lad_wales))

```

Read Global Burden of Disease 2017 data

```{r,eval=FALSE}
gbd1 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-1.csv"))

gbd2 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-2.csv"))

gbd3 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-3.csv"))

gbd4 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-4.csv"))

gbd5 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-5.csv"))

gbd <- rbind(gbd1, gbd2, gbd3, gbd4, gbd5)

## Attach geographical info to each GBD estimate
## Manual step - this should cover them all 
gbd$location_name[gbd$location_name=="St Helens"] <- "St. Helens"
gbd <- gbd %>% mutate(
  areatype = case_when(location_name %in% c("United Kingdom","Australia")  ~ "country",
                       location_name %in% c("England","Scotland","Wales","Northern Ireland")  ~ "nation",
                       location_name %in% regions  ~ "region",
                       location_name %in% unique(utla$LTLA17NM) ~ "ltla",
                       location_name %in% unique(utla$UTLA17NM) ~ "utla"
  )
)
stopifnot(all(!is.na(gbd$areatype)))

## Determine which estimates are in city regions 
gbd <- gbd %>%
  left_join(local_government_areas, by=c("location_name"="lad11nm")) %>%
  mutate(in_cityregion = !is.na(cityregion)) %>%
  filter(areatype %in% c("ltla","utla"))

## Next we want to decide whether to use LTLA/UTLA or region for areas not in city regions

```

Read disease coding table

```{r,eval=FALSE}
disease_names_execute <- read_csv(file.path(relative_path_gbd, "disease_outcomes_lookup.csv")) %>%
    select(GBD_name, acronym) %>%
    mutate(disease = tolower(GBD_name))

DISEASE_SHORT_NAMES <- data.frame(disease = tolower(as.character(unique(gbd$cause_name))), 
                                  sname = tolower(abbreviate(unique(gbd$cause_name, max = 2))),
                                  stringsAsFactors = F) %>%
    mutate(is_not_dis = ifelse((str_detect(disease, "injuries") |
                                str_detect(disease, "All causes") |
                                str_detect(disease, "Lower respiratory infections")), 
                               1, 0) ) %>%
    mutate(is_not_dis = case_when(sname == "allc"  ~  2,
                                  sname == "lwri"  ~  1,
                                  ## Code for major depressive disorder (no deaths) and hypertensive heart disease (no incidence)
                                  sname == "hyhd"  ~  3,
                                  sname == "mjdd"  ~  3,
                                  TRUE  ~  is_not_dis)) %>%
    left_join(disease_names_execute, by="disease") %>%
    mutate(acronym = ifelse(str_detect(disease, "injuries"), disease, acronym),
           acronym = word(acronym, 1),
           males = ifelse(disease %in% c("uterine cancer", "breast cancer"), 0, 1),
           females = ifelse(disease %in% "prostate cancer", 0, 1),
           sname = gsub("'", '', sname),
           acronym = ifelse(is.na(acronym), "no_pif", acronym))
```

Join geographical data to GBD data

```{r,eval=FALSE}
names(gbd) <- gsub(pattern = "_name", replacement = "", x = names(gbd))
gbd <- gbd %>%
    select(-contains("id")) %>%
    mutate(cause = tolower(cause))

saveRDS(gbd, file=file.path(relative_path_gbd, "GBD2017.rds"))

gbd <- readRDS(file=file.path(relative_path_gbd, "GBD2017.rds"))
```

# Determining "effective sample sizes" behind estimates

<!-- NOTE COUNTS FOR AREAS CONTAINING CITY REGIONS ARE NOT CONSTRAINED TO BE GREATER THAN TOTAL COUNTS FOR CITY REGIONS WITHIN THAT AREA. we will exclude non-city region data from this analysis.  see notes in noncityregions.Rmd -->

This part is the most computationally intensive.  For each estimated "rate" published by the GBD (actually a proportion), the associated credible interval is converted to an "effective sample size" that describes the amount of information that the estimate is based on.

For example, for prevalence estimates, these are assumed to be based on a survey of n people in the area, and observing r people with the disease.   The point estimate is assumed to equal r/n, and the credible interval is interpreted as the posterior interval we would get from a Bayesian analysis of these data with a vague (Beta(0,0)) prior.    The implied values of r and n can then be computed based on the point estimate and credible interval.

n is referred to as the "effective sample size".   Note that this is generally not equal to the actual population of the area.    They would only be equal if a survey of the full population had been conducted.  We do not have access to the underlying data or methods that GBD used to calculate the credible interval, so we do not know the true value of n.

Before determining the effective sample sizes, we filter the data to include only the "Rate" estimates for specific diseases (excluding "all causes" results).   Note the published "Rates" are actually the expected number of events (e.g. prevalent cases at the start of a year, incident events within a year or deaths within a year) among a population of 100000 that includes all people in the demographic subgroup of interest, not just those at risk of the event.  The "Rates" are divided by 100000 so they can be interpreted as proportions.

Lower limits published as 0 and upper limits published as 1 are modified to be close but not equal to these bounds, while remaining consistent with the point estimate, since values of exactly 0 and 1 are inconsistent with the Beta distribution (TODO put this in disbayes::ci2num).

```{r,eval=FALSE}
gbdp <- 
    gbd %>%
    filter(metric == "Rate") %>% 
    filter(cause != "all causes") %>% 
    mutate(num = NA, denom = NA) %>%
    mutate(val = val/100000, lower = lower / 100000, upper = upper / 100000) %>%
    mutate(lower = if_else(lower==0, pmin(val/2, 0.00001), lower)) %>% 
    mutate(upper = if_else(upper==1, pmax((1+val)/2, 0.99999), upper))
```

The function `ci2num` in the `disbayes` package is then used to calculate the effective sample sizes.  This takes about 40 minutes on a fairly fast laptop.   It could be made a lot faster by using parallel processing.

```{r,eval=FALSE,cache=TRUE}
nest <- nrow(gbdp) 
pb <- progress_bar$new(total = nest) # show progress of computation
for (i in 1:nest){
  if (gbdp$val[i] < gbdp$upper[i] &
      gbdp$val[i] > gbdp$lower[i]) { 
    pb$tick()
    counts <- disbayes::ci2num(gbdp$val[i], gbdp$lower[i], gbdp$upper[i])
    gbdp$num[i] <- counts$num
    gbdp$denom[i] <- counts$denom
  }
}
saveRDS(gbdp, file=file.path(relative_path_gbd, "gbdp.rds"))
```

```{r,echo=FALSE}
gbdp <- readRDS(file=file.path(relative_path_gbd, "gbdp.rds"))
```

The estimates are very close to the implicit numerator divided by the implicit denominator in all cases - so those implicit counts can be used in place of the point estimates.

```{r}
summary(gbdp$num/gbdp$denom - gbdp$val)
```

The remaining counts still to be filled in are those where the point estimate is exactly 0 or 1, which is incompatible with a beta distribution.

There are also many estimates for which the implicit denominator is implausibly large.  These correspond to disease events which are very rare in particular subgroups, thus the estimates of the effective sample size are unstable.

```{r}
gbdp %>% select(val, num, denom) %>% arrange(desc(denom)) %>% head
```

If the point estimate is 0 or 1, or if the denominator obtained from `ci2num` is larger than the actual population size, we will simply use the actual population size of the subgroup as the denominator, which will be closer to the amount of information contributing the estimate.

# Determining actual population sizes

Therefore we reconstruct the actual population sizes of each subgroup, assumed to be equal to the estimates published by GBD as the estimated "Number" of cases, divided by the estimated "rates" per person.

```{r,eval=FALSE}
gbdnum <- gbd %>%
  filter(metric=="Number") %>%
  select(measure, location, sex, age, cause, Number=val)
gbdp <- gbdp %>%
  left_join(gbdnum, by=c("measure","location","sex","age","cause")) %>%
  mutate(pop = Number / val)
saveRDS(gbdp, file=file.path(relative_path_gbd, "gbdp.rds"))
gbdp <- readRDS(file=file.path(relative_path_gbd, "gbdp.rds"))

```

We can then use these to fill in the effective sample sizes "d"  that were missing or implausible, and deduce the effective numerators "n" by multipling "d" by the point estimate of the proportion.

```{r}
gbdp <- gbdp %>%
  mutate(pop = if_else(is.na(pop), 5000, pop)) %>% 
  mutate(nodenom = is.na(denom) | (denom > pop),
         denom = if_else(is.na(denom), pop, denom),
         denom = if_else(denom > pop, pop, denom),
         num = ifelse(nodenom, round(val*denom), num))
```

Note that the data are still grouped as originally published - estimates by five-year age groups (not one year) and local authorities (not city regions), as well as by gender and disease measure.

Now we have reconstructed the implicit count "data" on which these estimates are based, these counts can now be easily aggregated or disaggregated to produce estimates for smaller or larger subgroups.  The counts will retain their meaning as the implicit number of events, or number of individuals, observed in the subgroup.

#  Disaggregating by age groups

Firstly we can disaggregate the five year age groups to single years of age.   If we assume that there was an equal amount of information from each single year contributing to the five-year result, we can simply divide the numerators r and denominators n for the five-year estimates by 5 (and round to the nearest integer).

However a smarter method involves a temporal disaggregation model, using the `tempdisagg` package (Sax and Steiner, 2020).  For each measure, location, sex, and cause, the five year counts are disaggregated to one-year counts in a way that preserves the five year totals, but the resulting one-year counts vary smoothly.

```{r} 
# Working with the data that has one row per five-year age group, 
# first construct 1-year counts as an extra column. 
gbdp <- gbdp %>%
  extract(age, c("from_age", "to_age"), "(.+) to (.+)", remove=FALSE, convert=TRUE) %>%
  mutate(from_age = case_when(age=="95 plus"  ~  95L,
                              age=="Under 5"  ~  0L,
                              TRUE  ~  from_age),
         to_age = case_when(age=="95 plus"  ~  99L,
                            age=="Under 5"  ~  4L,
                            TRUE  ~  to_age),
         agediff = to_age - from_age + 1,  # this will equal 5 
         num1yr = round(num/agediff),
         denom1yr = round(denom/agediff)) %>%
  rename(agegroup = age)

## Now stretch the data out using an index, to create a data frame with 1 row per year of age and create a variable for year of age. 
index <- rep(1:nrow(gbdp), gbdp$agediff)
gbdpyrd5 <- gbdp[index,] %>%
    mutate(ageyr = from_age + sequence(gbdp$agediff) - 1)
gbdpyrd5 <- gbdpyrd5 %>% 
  select(measure, location, ageyr, sex, agegroup, from_age, to_age, cause, year, areatype, cityregion, region, in_cityregion, num1yr, denom1yr) 
saveRDS(gbdpyrd5, file=file.path(relative_path_gbd, "gbdpyrd5.rds"))

## More advanced method for smooth disaggregation 
## using the tempdisagg package 
# tmp <- gbdp %>% group_by(measure, location, sex, cause) %>% filter(cur_group_id()==1)
gbdp_grp <- gbdp %>% 
  group_by(measure, location, sex, cause,
           areatype, cityregion, region, in_cityregion) %>%  # filter(cur_group_id() %in% 1:3) %>%
  arrange(measure, location, sex, cause, from_age)

disagg <- function(dat, key){ 
  res <- with(dat, { 
    data.frame( 
      ageyr = rep(from_age,agediff) + sequence(agediff) - 1,
      num = predict(td(num ~ 1, to=5, method="fast")),
      denom = predict(td(denom ~ 1, to=5, method="fast"))
    ) } )
  res 
}

# This takes about a minute 
gbdpyr <- group_modify(gbdp_grp, disagg) %>% 
  ungroup %>%
  left_join(gbdpyrd5, by=c("measure","location","ageyr","sex","cause",
                           "areatype", "cityregion", "region", "in_cityregion"))

# Sometimes the results are negative.  
# Revert to dividing by 5 for all agegroups that contain a negative result 
neg_num <- gbdpyr %>%
  filter(num<0) %>%
  select("measure","location","sex","cause","agegroup") %>%
  distinct() %>%
  mutate(zeronum = TRUE)
neg_denom <- gbdpyr %>%
  filter(denom<0) %>%
  select("measure","location","sex","cause","agegroup") %>%
  distinct() %>%
  mutate(zerodenom = TRUE)
gbdpyr <- gbdpyr %>% 
  left_join(neg_num, by=c("measure", "location","sex","cause","agegroup")) %>%
  left_join(neg_denom, by=c("measure", "location","sex","cause","agegroup")) %>%
  mutate(zeronum = replace_na(zeronum, FALSE)) %>%
  mutate(zerodenom = replace_na(zerodenom, FALSE)) %>%
  mutate(num = if_else(zeronum, num1yr, round(num)),
         denom = if_else(zerodenom, denom1yr, round(denom)))

saveRDS(gbdpyr, file=file.path(relative_path_gbd, "gbdpyr.rds"))

## Check for a few cases that the series match with the naive dividing by 5 method and the smooth methods
tloc <- sample(unique(gbd_disagg$location), 3)
test2 <- gbdpyrd5 %>%
  filter(cause=="ischemic heart disease", sex=="Male", measure=="Deaths",
         location %in% tloc)
gbdpyr %>% 
  filter(cause=="ischemic heart disease", sex=="Male", measure=="Deaths",
         location %in% tloc) %>%
  ggplot(aes(x=ageyr, y=denom, col=location, group=location)) + 
  geom_line() + geom_point() + 
  geom_line(data=test2, aes(y=denom1yr))

```

#### Example

Data by 5-year age groups for one location/cause/measure of interest

```{r}
gbdp %>%
  filter(measure=="Incidence", location=="Bristol, City of", cause=="ischemic heart disease",
         agegroup %in% c("40 to 44", "45 to 49"), sex=="Female") %>% 
  select(agegroup, sex, num, denom, num1yr, denom1yr)
```

Equivalent data by single year of age.  Note due to rounding errors, the one-year counts will not always add up exactly to the five-year counts.

```{r}
gbdpyr %>%
  filter(measure=="Incidence", location=="Bristol, City of", cause=="ischemic heart disease",
         ageyr %in% 40:49, sex=="Female") %>% 
  select(ageyr, sex, num, denom)
```

# Aggregating by area

Secondly we can aggregate the data from local authorities to produce data by city regions (defined as groups of local authorities), and regions excluding city regions.   This covers the whole of England with mutually exclusive areas.   No extra assumptions are required to do this.   It doesn't matter whether this is done before or after converting from 5-year to 1-year age groups.

```{r}
options(dplyr.summarise.inform = FALSE)

## City regions
gbdpyrcr <- gbdpyr %>% 
  filter(in_cityregion) %>%
  group_by(measure, sex, ageyr, cause, cityregion) %>%
  summarise(num = sum(num), denom = sum(denom)) %>%
  mutate(areatype = "cityregion") %>%
  rename(location = cityregion) %>%
  ungroup %>%
  select(measure, sex, ageyr, cause, location, areatype, num, denom)

## UTLAs with estimates in GBD 
gbdpyru <- gbdpyr  %>%
  filter(!in_cityregion, areatype=="utla") %>%
  select(measure, sex, ageyr, cause, location, areatype, num, denom)

## LTLAs from GBD aggregated to UTLAs
gbdpyrlagg  <- gbdpyr  %>%
  filter(!in_cityregion, areatype=="ltla") %>%
  left_join(utla, by=c("location"="LTLA17NM")) %>%
  group_by(measure, sex, ageyr, cause, UTLA17NM) %>%
  summarise(num=sum(num), denom=sum(denom)) %>%
  rename(location = "UTLA17NM") %>%
  mutate(areatype = "utla_agg") %>%
  ungroup %>%
  select(measure, sex, ageyr, cause, location, areatype, num, denom)

## don't use this aggregation for now, but save for later if needed
gbdmodu <- rbind(gbdpyrcr, gbdpyru, gbdpyrlagg)

## LTLAs and UTLAs from GBD aggregated to regions, excluding areas that are in city regions
gbdpyrreg  <- gbdpyr  %>%
  mutate(region = case_when(areatype=="ltla"  ~  lad$RGN17NM[match(location, lad$LAD17NM)],
                            areatype=="utla"  ~  utlareg$RGN17NM[match(location, utlareg$UTLA17NM)],
                            TRUE  ~  region
                              )) %>%
  filter(!in_cityregion) %>%
  group_by(measure, sex, ageyr, cause, region) %>%
  summarise(num=sum(num), denom=sum(denom)) %>%
  rename(location = "region") %>%
  mutate(areatype = "region") %>%
  ungroup %>%
  select(measure, sex, ageyr, cause, location, areatype, num, denom)

gbdmod <- rbind(gbdpyrcr, gbdpyrreg)

saveRDS(gbdmod, file=file.path(relative_path_gbd, "gbdmod.rds"))
gbdmod <- readRDS(file=file.path(relative_path_gbd, "gbdmod.rds"))

```

#### Example

Data by local authority for one cause/measure/demography of interest

```{r}
gbdpyr %>%
  filter(measure=="Incidence", cityregion=="bristol", ageyr==44, cause=="ischemic heart disease") %>% 
  arrange(sex) %>% 
  select(location, areatype, sex, num, denom)
```

Equivalent data by city region: the numbers and denominators for each local authority are just added up

```{r}
gbdpyrcr %>%
  filter(measure=="Incidence", location=="bristol", ageyr==44, cause=="ischemic heart disease") %>% 
  select(location, sex, num, denom)
```

### Spreading so one row by area x subgroup

Put numerators and denominators for incidence, prevalence and mortality in different columns

```{r}

gbdmod <- readRDS(file=file.path(relative_path_gbd, "gbdmod.rds"))

gbddb <- gbdmod %>%
  filter(measure %in% c("Deaths","Incidence", "Prevalence")) %>%
  mutate(measure = fct_recode(measure, mort="Deaths", inc="Incidence", prev="Prevalence")) %>% 
  pivot_wider(names_from=measure,
              values_from=c("num","denom"),
              values_fill = list(num=0, denom=5000)) %>% 
  mutate(location = fct_recode(str_to_title(location),
                                 `Greater Manchester`="Greatermanchester",
                                 `North East (city region)`="Northeast",
                                 `North East (region)`="North East",
                                 `West Midlands (city region)`="Westmidlands",
                                 `West Midlands (region)`="West Midlands",
                                 `East of England`="East Of England")) %>%
  mutate(cause = str_to_sentence(cause)) %>%
  arrange(location, cause, sex, ageyr) %>%
  filter(! cause %in% c("Hypertensive heart disease"))

saveRDS(gbddb, file=file.path(relative_path_gbd, "gbddb.rds"))

gbdmod %>% 
  filter(ageyr==60, sex=="Male", cause=="ischemic heart disease", measure=="Incidence")  %>% 
  filter(areatype=="cityregion") %>%
  select(location, num, denom)

# Some age groups are not available.  Assume count zero, denom 5000 for these 
# Note again that counts for the wider regions / countries are not constrained to be consistent with the smaller areas
# but doesn't affect our analysis

```

# Cancer remission rates

10-year survival to use to inform remission rates, from the [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/cancersurvivalinengland/nationalestimatesforpatientsfollowedupto201).  These are not published by age.  They are aggregated over ages 15-99, whereas 1, 5 year survival is published by age group.

Methods [here](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/methodologies/cancersurvivalstatisticalbulletinsqmi).   Assumes "conditional probabilities of surviving for patients diagnosed in current year are equal to those diagnosed over the full 10-year period of available data."  Caveats about trends through time.

Our approach allows uncertainty around this - could represent expected variations between age groups, given variations for 5 year survival.

Annual remission probability is computed from the 10 year survival probability.  Assumes that remission happened in one of the 10 years, with some unknown annual probability r such that the probability of remission in 10 years is equal to the 10 year survival probability s = 1 - (1-r)^10, so that r = 1 -(1-s)^(1/10).

Convert remission "rates" to numerators and denominators by assuming a denominator defined by the number of cancer cases over all ages and areas, but dividing all numerators and denominators by 100 to represent uncertainty about the true area-specific prevalence. For example, this gives a 95% credible interval for the annual breast cancer remission probability of 0.11 to 0.16.

```{r}
library(readxl)
rem <- read_excel("~/work/chronic/Figure_10__Predicted_10-year_net-survival_using_the_hybrid_approach_for_men_and_women_(aged_15_to_99_years).xls", range="A7:C26", na=":") %>%
  pivot_longer(cols=c("Men","Women"), names_to = "sex", values_to = "surv10") %>% 
  mutate(prem1 = 1 - (1 - surv10/100)^(1/10)) %>% 
  filter(Cancer %in% c("Breast","Uterus","Colorectal","Liver","Stomach","Lung"),
         !is.na(prem1)) %>%
  mutate(Cancer = fct_recode(Cancer, 
                             "Breast cancer"="Breast",
                             "Uterine cancer"="Uterus",
                             "Colon and rectum cancer"="Colorectal",
                             "Stomach cancer"="Stomach",
                             "Tracheal, bronchus, and lung cancer"="Lung"
                             #,
                             #"Liver cancer"="Liver"
                             ),
         sex = fct_recode(sex, "Male"="Men", "Female"="Women")) %>% 
  rename(cause = Cancer)

cancers <- c("Tracheal, bronchus, and lung cancer", "Colon and rectum cancer", "Stomach cancer", "Liver cancer", "Breast cancer", "Uterine cancer")

num_cancer <- gbddb %>% 
  filter(cause %in% cancers, 
         ageyr %in% 15:99) %>% 
  group_by(sex, cause) %>% 
  summarise(ncases = sum(num_prev)) %>% 
  left_join(rem, by=c("cause","sex")) %>%
  mutate(num_rem = round(ncases*prem1 / 100),
         denom_rem = round(ncases / 100)) %>% 
  select(sex, cause, num_rem, denom_rem)
num_cancer

gbddb <- gbddb %>% 
  left_join(num_cancer) %>% 
  mutate(num_rem = replace_na(num_rem, 0),
         denom_rem = replace_na(denom_rem, 0))

saveRDS(gbddb, file=file.path(relative_path_gbd, "gbddb.rds"))
saveRDS(gbddb, file=file.path("~/work/chronic", "gbddb.rds"))

qbeta(c(0.025, 0.975), 128, 940-128) # 0.11 to 0.16.more plausible

```

# Population of city regions

23872000 in English city regions , just over half of the population of England

https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/articles/populationdynamicsofukcityregionssincemid2011/2016-10-11

54786300  for England.

```{r}
#Greater London 8,674,000	 
#West Midlands	2,834,000	 
#Greater Manchester	2,756,000	 
#West Yorkshire	2,282,000	 
#North East	1,957,000	 
#Liverpool	1,525,000
#Sheffield	1,375,000	 
#Edinburgh	1,350,000	 
#Bristol	1,119,000	
```

### 2021-07-01 R/cfnaive.R

@@ -1,0 +1,47 @@
+##' Simplistic estimator for case fatality given incidence, prevalence and mortality
+##'
+##' Estimator for case fatality given incidence, prevalence and mortality, under a simple discrete-time approximation. 
+##'
+##' @param inc Vector of estimates of the incidence rate by age
+##'
+##' @param prev Vector of estimates of the prevalence probability by age
+##'
+##' @param mort Vector of estimates of the mortality rate by age
+##'
+##' @details This estimator is based on a half-year discrete-time approximation and assumes that transitions only happen at half years. 
+##' 
+##' Let
+##'
+##' p12 = half-year transition probability from no disease to disease
+##' 
+##' p23 = half-year transition probability from disease to death
+##'
+##' then the probability of death within a year (two half-year cycles) averaging over people with and without the disease, 
+##'
+##' For a person who doesn't have the disease at the start, the probability that they die within a year is p12*p23.  For someone who has the disease at the start, they either die at the first cycle (probability p23) or survive the first cycle and die at the second (probability (1-p23)*p23)).   
+##' 
+##' Therefore the annual probability of death is computed by adding these three possibilities, weighted by the prevalence of disease at the start of the cycle. 
+##'
+##' mort = (1-prev)*p12*p23 + prev*(p23 + (1-p23)*p23)
+##'
+##' Therefore a compatible value for p23 can be obtained from
+##' published prevalence, incidence and mortality as the solution for
+##' x of the quadratic equation
+##'
+##' a x^2 + b*x + c = 0
+##'
+##' with a = -prev, b = (1-prev)*inc + 2*prev, c=-mort 
+##' 
+##' assuming the published incidence rate (inc) equals the probability
+##' p12, and the case fatality rate (cf) equals the corresponding
+##' probability p23, and the published mortality rate equals the
+##' probability (mort) defined above. 
+##'
+##' @export
+cfnaive <- function(inc, prev, mort){
+    a <- prev
+    b <- (1 - prev)*inc + 2*prev
+    c <- - mort
+    cf <- (-b + sqrt(b^2 - 4*a*c))/(2*a)
+    cf
+}

### 2021-07-01 R/disbayes-package.R

@@ -7,20 +7,89 @@
 #' @useDynLib disbayes, .registration = TRUE
 #' @import methods
 #' @import Rcpp
+#' @importFrom generics tidy
 #' @importFrom rstan sampling
 #' @importFrom stats lm sd rlnorm rnorm
 #' @importFrom mgcv smoothCon jagam s
+#' @import dplyr
+#' @importFrom tidyr extract pivot_wider pivot_longer
+#' @importFrom magrittr "%>%"
+#' @import tibble
 #'
 #' @references
 #' Stan Development Team (2019). RStan: the R interface to Stan. R package version 2.19.2. https://mc-stan.org
 #'
 NULL
 
-##' Ischaemic heart disease in Bristol
+#' @importFrom generics tidy
+#' @export
+generics::tidy
+
+##' Ischemic heart disease in England 
 ##'
-##' @format TODO 
+##' @format A data frame with columns:
+##'
+##' \code{sex}: \code{"male"} or \code{"female"}. 
+##'
+##' \code{ageyr}. Year of age. 
+##'
+##' \code{location}. Name of the location, which is either a city region or region in England. 
+##'
+##' \code{num_mort}. Numerator behind the estimate of mortality 
 ##' 
-##' @source Global Burden of Disease study 
+##' \code{num_inc}. Numerator behind the estimate of incidence
+##' 
+##' \code{num_prev}. Numerator behind the estimate of prevalence
+##' 
+##' \code{denom_mort}. Denominator behind the estimate of mortality 
+##' 
+##' \code{denom_inc}.  Denominator behind the estimate of incidence
+##' 
+##' \code{denom_prev}. Denominator behind the estimate of prevalence
+##'
+##' @source Global Burden of Disease, 2017
+##'
+##' @details The data were processed to
+##'
+##' * change the geography to refer to England city regions and the remaining English regions,
+##'
+##' * change counts by 5-year age groups to estimated 1-year counts,
+##'
+##' * obtain estimated numerators and denominators from the published point estimates and uncertainty intervals.
+##' A point estimate of the risk is equivalent to the numerator divided by the denominator. The denominator is
+##' related to the extent of uncertainty around this estimate. 
+##'
+##' The script used to do this is available from REF GITHUB TODO
+##'
+##' @references TODO our paper when available
 ##' 
 ##' @keywords datasets
-"ihdbristol"
+"ihdengland"
+
+
+
[…]

### 2021-07-01 R/disbayes_hier.R

@@ -1,0 +1,371 @@
+##' Bayesian estimation of chronic disease epidemiology from incomplete data - hierarchical model for case fatalities 
+##' This is computationally intensive.  
+##' 
+##' @param group Variable in the data representing the area (or other grouping factor). 
+##' 
+##' @param gender Variable in the data representing gender, or a binary grouping factor, that is treated as a fixed additive effect, so the linear effect of gender on log case fatality is the same in each area.   The data should then have one row per year of age and gender.  If \code{NULL} (the default) then the data are one homogenous gender, and there should be one row per year of age. 
+##'
+##' @param model Three alternative models for case fatality are supported:
+##'
+##' \code{"default"} (the default). Random intercepts and slopes.
+##'
+##' \code{"interceptonly"}.  Random intercepts, but common slopes.
+##'
+##' \code{"increasing"}. Case fatality is assumed to be an increasing function of age (note it is constant below \code{"eqage"} in all models).
+##'
+##'
+##' @param nfold_int_guess  Prior guess at the ratio of case fatality between a high risk (97.5\% quantile) and low risk (2.5\% quantile) area.  
+##'
+##' @param nfold_int_upper  Prior upper 95\% credible limit for the ratio in average case fatality between a high risk (97.5\% quantile) and low risk (2.5\% quantile) area. 
+##'
+##' @param nfold_slope_guess,nfold_slope_upper This argument and the next argument define the prior distribution for the variance in the random linear effects of age on log case fatality.   They define a prior guess and upper 95\% credible limit for the ratio of case fatality slopes between a high trend (97.5\% quantile) and low risk (2.5\% quantile) area.  (Note that the model is not exactly linear, since departures from linearity are defined through a spline model.  See the Jackson et al. paper for details.).
+##'
+##' @inheritParams disbayes
+##' 
+##' @details For full details see Jackson et al. REF.
+##'
+##' Note that an independent incidence is estimated for each area and age group.  In principle this function
+##' would be straightforward to extend to model incidences as a smooth function of age and hierarchically by area.
[…]

### 2021-07-01 R/disbayes_utils.R

@@ -1,0 +1,53 @@
+## For the moment, don't smooth either inc or CF
+## it's supposed to be fast 
+## may need to smooth one before estimating other later on 
+
+init_rate <- function(rate, dat){
+    default <- 0.001
+    rcrude <- dat[[sprintf("%s_num",rate)]] / dat[[sprintf("%s_denom",rate)]]
+    rcrude[is.na(rcrude)] <- default
+    rcrude[is.nan(rcrude)] <- default
+    rcrude <- pmax(default, rcrude)
+    rcrude 
+}
+
+train_rate <- function(dat,
+                       inc_init=NULL, cf_init=0.01, rem_init=NULL, remission,
+                       iter_train = 1000, 
+                       return="stan", ...){
+    if (is.null(inc_init)) inc_init <- init_rate("inc", dat)
+    if (is.null(rem_init)) rem_init <- init_rate("rem", dat)
+    nage <- dat$nage
+    Xdummy <- matrix(0, nrow=nage, ncol=2)
+    datstanu <- c(dat, list(smooth_cf=0, smooth_inc = 0, const_cf=0, trend=0, nyr = 1,
+                            increasing_cf=0, K=2, X=Xdummy, sprior=dat$sprior,
+                            inc_trend_mult = array(1, dim=c(nage,1)),
+                            cf_trend_mult = array(1, dim=c(nage,1)) 
+                            ))
+    initu <- function(){
+        list(cf_par = rnorm(nage, mean=cf_init, sd=cf_init/10),
+             rem_par = rnorm(nage*remission, mean=rem_init, sd=rem_init/10),
+             inc_par = rnorm(nage, mean=inc_init, sd=inc_init/10))
+    }
+    fitu <- rstan::sampling(stanmodels$disbayes, data = datstanu, 
+                            init = initu, include = FALSE, pars=c("beta","lambda"),
+                            iter = iter_train,
+                            ...)
+    fitu 
+}
+
+
+## Obtains spline basis terms and initial values for their coefficients
+## given estimates of incidence or CF from unsmoothed model.
+
+init_smooth <- function(y, eqage, s_opts=NULL){
+    nage <- length(y)
+    age <- agecons <- 1:nage 
+    agecons[1:eqage] <- eqage
+    sm <- mgcv::smoothCon(s(x),
+                          data=data.frame(x=agecons),
+                          diagonal.penalty=TRUE)[[1]]
+    X <- sm$X
+    beta <- coef(lm(y ~ X - 1))
+    list(X=X, beta=beta)
+}

### 2021-07-01 R/state_probs.R

@@ -1,0 +1,19 @@
+state_probs <- function(i, f)  {
+    l <- i + f
+    q <- sqrt(i*i - 2*i*f  + f*f)
+    w <- exp(-(l + q) / 2)
+    v <- exp(-(l - q) / 2)
+    P <- array(dim=c(3, 3))
+    P[1,1] <- (2*(v-w)*f + v*(q-l) + w*(q+l)) / (2*q)
+    P[2,1] <- 0 
+    P[3,1] <- 0 
+    
+    P[1,2] <- i*(v - w)/q
+    P[2,2] <- -((2*f - l)*(v-w) - q*(v+w)) / (2*q)
+    P[3,2] <- 0
+    
+    P[1,3] <- (-l*(v-w) - q*(v+w))/(2*q) + 1
+    P[2,3] <- ((v-w)*(2*f - l) - q*(v+w))/(2*q) + 1
+    P[3,3] <- 1
+    P
+}

### 2021-07-01 data-raw/ihdengland.R

@@ -1,0 +1,13 @@
+library(devtools)
+
+relative_path_gbd <- "~/scratch/chronic/metahit_data"
+
+ihdengland <- readRDS(file.path(relative_path_gbd,"gbddb.rds")) %>% 
+  filter(cause == "Ischemic heart disease") %>%
+  rename(age=ageyr, inc_num=num_inc, inc_denom=denom_inc, 
+         prev_num=num_prev, prev_denom=denom_prev,
+         mort_num=num_mort, mort_denom=denom_mort,
+         rem_num=num_rem, rem_denom=denom_rem) %>%
+  select(-cause, -areatype, rem_num, rem_denom)
+
+use_data(ihdengland, overwrite = TRUE)

### 2021-07-01 data-raw/trends.r

@@ -1,0 +1,605 @@
+library(readxl)
+library(tidyverse) 
+
+# BHF trends data for 2005-2017, downloaded from 
+#https://www.bhf.org.uk/-/media/files/research/heart-statistics/cvd-statistics-2020-chapter-2a-morbidity-incidence-final.xlsx
+datxl <- read_excel("~/work/chronic/trends/cvd-statistics-2020-chapter-2a-morbidity-incidence-final.xlsx", sheet="2.9")
+
+# All incidence rates are calculated per 100,000 of the population; population-weighted data from each of the UK's four nations
+# MI. United Kingdom.  England data is for all ages. CHD is in sheet 2.7
+# Age-standardised rate is calculated for adults aged 15+
+# The Health Improvement Network (THIN) 2020
+
+bhf <- datxl %>%
+    filter(row_number() %in% c(12:15, 20:23)) %>%
+    mutate(gender = rep(c("Male","Female"), each=4)) %>%
+    rename(age = 1) %>%
+    pivot_longer(names_to="year0", values_to="inc", cols=2:14) %>%
+    mutate(year = rep(2005:2017, 8))
+
+bhf <- bhf %>% 
+    left_join(bhf %>% filter(year==2017) %>%
+              mutate(inc2017 = inc) %>% select(age, gender, inc2017),
+              by=c("age","gender")) %>%
+  mutate(plast = inc / inc2017) %>%
+  tidyr::extract(age, c("agefrom","ageto"), "([0-9]+)-([0-9]+) years", remove=FALSE, convert=TRUE) %>%
+  mutate(agefrom = ifelse(age == "75+ years", 75, agefrom),
+         ageto = ifelse(age == "75+ years", 100, ageto))
+bhf %>% mutate(agesex = paste(age,gender,sep=",")) %>% 
+ggplot(aes(x=year, y=inc, group=agesex, col=age, lty=gender)) + 
+  geom_line() + geom_point() + 
+  scale_x_continuous(breaks = 2005:2017) + 
+  xlab("") + ylab("Incidence (per 100,000)")
+
+
+
+## Annual percentage decreases between 2002 and 2010
+# Taken from Smolina et al.  https://www.bmj.com/content/344/bmj.d8059.full
+
[…]

### 2021-07-01 doc/hier.R

@@ -1,90 +1,0 @@
-## ------------------------------------------------------------------------
-knitr::opts_chunk$set(fig.width=7, echo=FALSE, warning=FALSE)
-
-## ------------------------------------------------------------------------
-load(file="cfres.rda")
-library(dplyr)
-library(ggplot2)
-source("constants.R")
-
-p <- cfres %>%
-    filter(disease == "stmc" & 
-           area == "bristol" &
-           model %in% c("unsmoothed", "smoothed")) %>% 
-ggplot(aes(x=age, col=model)) +
-    geom_pointrange(aes(y=med, ymin=lower95, ymax=upper95),
-                    position = position_dodge(0.2)) + 
-    facet_grid(cols=vars(gender)) +  
-    coord_cartesian(ylim=c(0, 4)) +
-    scale_x_continuous(breaks=seq(0,100,10)) + 
-    ylab("Case fatality (median, 95% CrI)")  +
-    xlab("Age (years)") +
-    ggtitle("Bristol, stomach cancer")
-p
-
-## ----echo=FALSE----------------------------------------------------------
-
-p <- cfres %>%
-    filter(disease %in% "stmc" & 
-           area == "bristol" &
-           model %in% c("smoothed", "hier50")) %>% 
-ggplot(aes(x=age, col=model)) +
-    geom_pointrange(aes(y=med, ymin=lower95, ymax=upper95),
-                    position = position_dodge(0.2)) + 
-    facet_grid(rows=vars(disease), cols=vars(gender)) +  
-    coord_cartesian(ylim=c(0, 4)) +
-    scale_x_continuous(breaks=seq(0,100,10)) + 
-    ylab("Case fatality (median, 95% CrI)")  +
-    xlab("Age (years)") +
-    ggtitle("Bristol, stomach cancer")
-p
-
-
-## ---- echo=FALSE, warning=FALSE------------------------------------------
-
-hierdata <- readRDS("hierdata.rds")
-hdat <- hierdata[["male"]][["stmc"]]
-dat <- data.frame(
-    area = c(cityregions, regions_exc, regions_inc),
-    prevn = hdat$prev_num[60,],
-    prevd = hdat$prev_denom[60,],
-    mortn = hdat$mort_num[60,],
-    mortd = hdat$mort_denom[60,],
-    incn = hdat$inc_num[60,],
[…]

### 2021-07-01 tests/testthat/test_hier.R

@@ -1,0 +1,59 @@
+library(dplyr)
+library(disbayes)
+
+ihdmale <- ihdengland %>% 
+  filter(sex=="Male")
+
+db <- disbayes_hier(ihdmale, age="age", group="location", 
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                           prev_num = "prev_num", prev_denom = "prev_denom",
+                           mort_num = "mort_num", mort_denom = "mort_denom",
+                    iter=10, iter_train=10, chains=1, loo=FALSE, algorithm="Fixed_param")
+expect_s3_class(db, "disbayes_hier")
+
+test_that("errors in specifying hierarchical model", {
+  expect_error(disbayes_hier(ihdengland, age="age", group="location", 
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                             prev_num = "prev_num", prev_denom = "prev_denom",
+                             mort_num = "mort_num", mort_denom = "mort_denom"),
+               "rows in data, expected")
+  
+  ## call nonhier model by mistake 
+  expect_error(
+    disbayes(ihdmale, age="age", 
+             inc_num = "inc_num", inc_denom = "inc_denom",
+             prev_num = "prev_num", prev_denom = "prev_denom",
+             mort_num = "mort_num", mort_denom = "mort_denom"),
+    "rows in data")
+  ihdmale$badage <- ihdmale$age
+  ihdmale$badage[20:30] <- 92
+  
+  expect_error(disbayes_hier(ihdmale, age="nonexistent", group="location", 
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                             prev_num = "prev_num", prev_denom = "prev_denom",
+                             mort_num = "mort_num", mort_denom = "mort_denom"),
+               "age variable")
+  
+  expect_error(disbayes_hier(ihdmale, age = "badage",
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                             prev_num = "prev_num", prev_denom = "prev_denom",
+                             mort_num = "mort_num", mort_denom = "mort_denom"
+  ),
+  "`group` variable not supplied")
+  
+  expect_error(disbayes_hier(ihdmale, age = "badage", group="location",
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                             prev_num = "prev_num", prev_denom = "prev_denom",
+                             mort_num = "mort_num", mort_denom = "mort_denom"
+  ),
[…]

### 2021-07-01 tests/testthat/test_hier_gender.R

@@ -1,0 +1,62 @@
+library(dplyr)
+library(disbayes)
+
+db <- disbayes_hier(ihdengland, age="age", group="location", gender="sex",
+                           inc_num = "inc_num", inc_denom = "inc_denom",
+                           prev_num = "prev_num", prev_denom = "prev_denom",
+                           mort_num = "mort_num", mort_denom = "mort_denom",
+                           iter=10, iter_train=10, chains=1, loo=FALSE, algorithm="Fixed_param")
+expect_s3_class(db, "disbayes_hier_gender")
+
+test_that("errors in specifying gender model", {
+  expect_error(disbayes_hier(ihdengland, age="age", group="location",  
+                                    inc_num = "inc_num", inc_denom = "inc_denom",
+                                    prev_num = "prev_num", prev_denom = "prev_denom",
+                                    mort_num = "mort_num", mort_denom = "mort_denom"),
+               "`gender` variable not supplied")
+  
+  expect_error(disbayes_hier(ihdengland, age="age", group="location", gender="sex",
+                                    inc_num = "inc_num", inc_denom = "inc_denom",
+                                    prev_num = "prev_num", prev_denom = "prev_denom",
+                                    mort_num = "mort_num", mort_denom = "mort_denom"),
+               "rows in data, expected")
+  
+  ## call nonhier model by mistake 
+  expect_error(
+    disbayes(ihdengland, age="age", 
+             inc_num = "inc_num", inc_denom = "inc_denom",
+             prev_num = "prev_num", prev_denom = "prev_denom",
+             mort_num = "mort_num", mort_denom = "mort_denom"),
+    "rows in data")
+  ihdengland$badage <- ihdengland$age
+  ihdengland$badage[20:30] <- 92
+  
+  expect_error(disbayes_hier(ihdengland, age="nonexistent", group="location", gender="sex",
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                             prev_num = "prev_num", prev_denom = "prev_denom",
+                             mort_num = "mort_num", mort_denom = "mort_denom"),
+               "age variable")
+  
+  expect_error(disbayes_hier(ihdengland, age = "badage", gender="sex",
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                             prev_num = "prev_num", prev_denom = "prev_denom",
+                             mort_num = "mort_num", mort_denom = "mort_denom"
+  ),
+  "`group` variable not supplied")
+  
[…]

### 2021-07-01 metahit/paper_analyses.Rmd
A range of doc and code fixes

@@
 
+res %>% filter(disease=="Uterine cancer", var=="cf", between(age, 58, 70))
+res_utrc %>% filter(disease=="Uterine cancer", var=="cf", between(age, 90, 100))
+## Smoothed hasn't worked
+
 reslist <- loolist <- vector(nrow(hierrundf), mode="list")
@@
 
-pdf("../write/all_cf_indep.pdf",width=6,height=6)
+pdf("../../write/all_cf_indep.pdf",width=6,height=6)
 
@@
   dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
+  dplyr::filter(! (disease=="Uterine cancer" & age > 90)) %>%
   mutate(p50 = 1 - exp(-`50%`),

### 2021-07-01 R/disbayes.R

@@ -115,7 +115,7 @@
 ##'   average age.
 ##'
 ##' @param eqage Case fatality and incidence are assumed to be equal for all ages
-##' below this age.
+##' below this age when using the smoothed model
 ##'
 ##' @param loo Compute leave-one-out cross validation statistics. 
 ##'
@@ -135,7 +135,30 @@
 ##'
 ##' @param ... Further arguments passed to \pkg{rstan}{sampling} to control
 ##'   running of Stan, that are applied to both the "training" (unsmoothed) model
-##'   fit and the smoothed model fit. 
+##'   fit and the smoothed model fit.
+##'
+##' @return A list with the following components
+##'
+##' \code{fits}: An object containing posterior samples from the fitted model,
+##' in the \code{stanfit} format returned by the \code{\link[rstan]{stan}}
+##' function in the \pkg{rstan} package.
+##'
+##' \code{fitu}: Another \code{stanfit} object containing the equivalent "unsmoothed" model results
+##' where case fatality and incidence are given independent priors per year of age.
+##'
+##' \code{loos}: A list of objects containing leave-one-out cross-validation statistics.  There is
+##' one list component for each of the observed outcomes informing the model: incidence, prevalence,
+##' mortality and remission.   The component for each outcome is an object
+##' in the form returned by the \code{\link[loo]{loo}} function in the \pkg{loo} package. This can
+##' be used to assess how well the model predicts the data for that outcome, compared to other models. 
+##'
+##' The \code{\link{looi_disbayes}} function can be used to extract from this list
+##' a single tidy data frame with one row per observation. 
+##'
+##' \code{loou}: Equivalent cross-validation statistics for the unsmoothed model. 
+##'
+##' Use the \code{\link{tidy.disbayes}} method to return summary statistics from the
+##' fitted models. 
 ##'   
 ##' @export
 disbayes <- function(data,
@@ -207,10 +230,6 @@ disbayes <- function(data,
         inccrude <- summary_disbayes_fit(fitu, vars="inc")[, "med"]
         si <- init_smooth(log(inccrude), eqage, s_opts)
         betainc_init <- si$beta
-
-        for (i in 1:(eqage-1)){
-            X[i,] <- X[eqage,]
-        }
         
         datstans <- c(dat, list(smooth_cf=as.numeric(smooth_cf),
                                 increasing_cf=as.numeric(increasing_cf),
[…]

### 2021-07-01 metahit/paper_analyses_national.r

@@ -23,7 +23,7 @@ db <- disbayes(data = gbdeng %>% filter(cause==natrundf$disease[i],
                remission = natrundf$remission[i],
                increasing_cf = natrundf$increasing[i],
                const_cf = FALSE,
-               age="ageyr", smooth_cf=FALSE,
+               age="ageyr", smooth_cf=TRUE,
                eqage= natrundf$eqage[i], 
                chains=nchains, iter=1000)
 

### 2021-07-01 tests/slow/test.R

@@ -1,14 +1,47 @@
+ihdbristol <- ihdengland %>% filter(location=="Bristol", sex=="Male")
 
 ## Smooth incidence.  Noticeably slower than model with just smooth CF
 
-dbres <- disbayes(dat = ihdbristol, inc = "inc", 
-                  inc_denom = "pop", prev_num = "prevn", prev_denom = "prevdenom",
-                  mort = "mort", mort_denom = "pop", eqage = 40,
+dbres <- disbayes(dat = ihdbristol,
+                    inc_num = "inc_num", inc_denom = "inc_denom",
+                    prev_num = "prev_num", prev_denom = "prev_denom",
+                    mort_num = "mort_num", mort_denom = "mort_denom",
+                  eqage = 40,
                   smooth_inc = TRUE, 
                   chains = 1, iter_train = 400, iter=1000, loo=FALSE)
 
+
+## Prevalence not supplied
+
+dbres <- disbayes(dat = ihdbristol, 
+                  inc_num = "inc_num", inc_denom = "inc_denom",
+                  mort_num = "mort_num", mort_denom = "mort_denom",
+                  eqage = 40, smooth_inc = FALSE, 
+                  chains = 1, iter_train = 400, iter=1000, loo=FALSE)
+
+## Prevalence supplied, but not incidence. smooth incidence
+dbres2 <- disbayes(dat = ihdbristol, 
+                  prev_num = "prev_num", prev_denom = "prev_denom",
+                  mort_num = "mort_num", mort_denom = "mort_denom",
+                  eqage = 40, smooth_inc = TRUE, 
+                  chains = 1, iter_train = 400, iter=1000, loo=FALSE)
+
+# Unsmoothed model doesn't actually do too bad  
+dbres <- disbayes(dat = ihdbristol, 
+                  prev_num = "prev_num", prev_denom = "prev_denom",
+                  mort_num = "mort_num", mort_denom = "mort_denom",
+                  eqage = 40, smooth_inc = FALSE, 
+                  chains = 1, iter_train = 400, iter=1000, loo=FALSE)
+
 summ <- tidy(dbres)
+summ <- tidy(dbres2)
 summ %>% filter(var=="inc") %>%
+  ggplot(aes(x=age,y=`50%`,col=model,group=model)) + 
+  geom_line() + geom_point() +
[…]

### 2021-07-07 doc/disbayes.Rmd
Restore doc dir with built vignettes

--- 2019-12-05 doc/ihdbristol.Rmd
@@
 vignette: >
+ \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package}
  %\VignetteEngine{knitr::rmarkdown}
- \usepackage[utf8]{inputenc}
 bibliography: disbayes.bib
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
+A brief overview is given in this vignette, but a fuller explanation will be given in the paper _Bayesian multi-state modelling of incomplete chronic disease epidemiology data for health impact simulation models_ (Jackson et al., in progress).
+
 # Theoretical disease model
@@
 If we assume that mortality from other causes is independent of disease status, deaths from other causes are uninformative and can be ignored.
-Assume also for simplicity that remission from the disease is not possible (though the framework illustrated here is easily extensible to handle remission).
 
@@
 
-* disease incidence $i(t)$, and the
+* disease incidence $i(a)$, and the
 
-* case fatality rate $f(t)$,
+* case fatality rate $f(a)$,
 
-which are both assumed to depend on age $t$.  Assume further that these rates are constant within integer years of age $a$, so they can be written $i_a$, $f_a$.
+which are both assumed to depend on age $a$.
+Remission is also possible for some diseases, with rate $r(a)$, though for simplicity of illustration is ignored here.
+Assume further that the rates are constant within integer years of age $a$, so they can be written $i_a$, $f_a$.
 
-From these we can determine the _transition probability matrix_ $P_a$, whose $r,s$ entry $P_{ars}$ is the probability that a person is in state $s$ at age $a+1$ given they are in state $r$ at age $a$.
+From these, we can determine the _transition probability matrix_ $P_a$, whose $r,s$ entry $P_{ars}$ is the probability that a person is in state $s$ at age $a+1$ given they are in state $r$ at age $a$.
 The matrix $P_a$ is defined as a function of $i_a$ and $f_a$, which is the solution to a differential equation, and is written out explicitly in the DisMod II paper (@dismod2).
@@
 The next section discusses how these data can be derived, given the typical forms of data available from burden of disease studies.
-A denominator is required because it enables uncertainty associated with each estimate to be quantified, as we will discuss later.
+A denominator is required, because it enables uncertainty associated with each estimate to be quantified, as we will discuss later.
 
@@
 
-This approach is used in the DisMod-MR software, as explained by @flaxman2015, however the software itself is undocumented and not fully published.
+This approach is used in the DisMod-MR package, as explained by @flaxman2015, however the software itself is not fully documented.
 The older (published) DisMod II (@dismod2) used an ad-hoc optimisation approach to estimate parameters.
@@
 
+* uncertainty about any quantity is quantified automatically through the posterior distribution, given the data, model assumptions and prior distributions supplied.
+
 * the ease of including multiple sources of direct/indirect data.
-This is enabled by the computational methods and available software for Bayesian modelling, illustrated below.
+This is enabled by the generic computational methods and available software for Bayesian modelling, specifically the Stan package, illustrated below.
 This allows the approach to generalise to settings with different forms of data available.
+Currently implemented models include hierarchical models, models with additive area and gender effects, and models with age-specific time trends.
 In contrast, DisMod II only allows limited forms of data.
 
-* uncertainty about any quantity is quantified automatically through the posterior distribution, given the data, model assumptions and prior distributions supplied.
+Further discussion of the differences between `disbayes` and DisMod-MR are given in the paper [Jackson et al, to appear].
 
-The use of general-purpose free software for this computation, through the `disbayes` package, is described below.
-
-# Converting data into the required form for the Bayesian model
+# Data required by the Bayesian model
 
@@
 
-The following data are given in the R data frame `ihdbristol` supplied in the package, obtained from the Global Burden of Disease (REF) in a typical format:
+The following data are given in the R data frame `ihdengland` supplied in the package.
 
-* `inc` estimates of incidence of IHD by age
+* `inc_num`,`inc_denom`: incidence of IHD by age
 
-* `mort` estimates of IHD-specific mortality by age
+* `mort_num`,`mort_denom`: estimates of IHD-specific mortality by age
 
-* `pop` estimates of the population by age
+* `prev_num`,`prev_denom`: estimates of prevalence by age
 
-and we wish to estimate case fatality by age, given these inputs.  A selection of rows from the full data frame are shown here.
+and we wish to estimate case fatality by age, given these inputs.
+A selection of rows from the full data frame are shown here.
+We select only the data from one area and gender (Bristol, male) for this illustration.
 
@@
 library(disbayes)
-ihdbristol[ihdbristol$age %in% 50:55, c("age","pop","inc","mort")]
+library(dplyr)
+ihdbristol <- ihdengland %>% filter(location=="Bristol", sex=="Male")
+ihdbristol %>% filter(between(age, 50, 55))
 ```
 
-Note that the published "data" `inc` and `mort` are themselves estimates, based on underlying data from a finite population.
-To use the Bayesian approach, we need to recreate the _numerator_ and _denominator_ underlying each of these estimates.
-We assume that
+These data were obtained from the Global Burden of Disease, 2017, and transformed in three ways to obtain the data seen here.
+Firstly, the published point estimates and confidence intervals for each annual risk probability was converted to an implicit numerator and denominator.
+This is done by assuming that
 
-* the incidence has been estimated as $r_a^{(inc)}/n_a$, where out of $n_a$ people who are alive at age $a$, $r_a^{(inc)}$ of these get IHD before age $a+1$.
+* the incidence has been estimated as $r_a^{(inc)}/n_a^{(inc)}$, where out of $n_a^{(inc)}$ people who are alive at age $a$, $r_a^{(inc)}$ of these get IHD before age $a+1$.
 
-* the mortality has been estimated as $r_a^{(mort)}/n_a$, where out of $n_a$ people who are alive at age $a$, $r_a^{(mort)}$ of these die of IHD before age $a+1$.
+* the mortality has been estimated as $r_a^{(mort)}/n_a^{(mort)}$, where out of $n_a^{(mort)}$ people who are alive at age $a$, $r_a^{(mort)}$ of these die of IHD before age $a+1$.
 These ratios are estimates of the true, underlying disease-specific mortalities $dm_a$.
 
-In this example, $n_a$ and $r_a$ were not provided in the original data source, but can be roughly recreated, as follows.
-We were given population counts for five-year age groups, thus we obtain approximate one-year population counts $n_a$ by dividing these values by 5 (`pop` in the R data).
+* the prevalence has been estimated as $r_a^{(prev)}/n_a^{(prev)}$, where out of $n_a^{(prev)}$ people who are alive at age $a$, $r_a^{(prev)}$ of them have IHD.
 
-The number of incident IHD cases $r_a^{(inc)}$ can then be reconstructed by multiplying $n_a$ by the incidence rates `inc` supplied in the data.
-Likewise, the number of IHD deaths $r_a$ (`ndieddis` in the R data) is then approximately reconstructed by multiplying $n_a$ by the IHD-specific mortality rates supplied in the spreadsheet.
+The published estimates and uncertainty interval are then used to derive the implicit denominator and numerator, using the method described in the next subsection.
 
-The data also contain estimates of prevalence, implicitly based on observing the number of people with IHD $r^{(prev)}_a$ in a sample $n^{(prev)}_a$.
-The assumed sample size $n^{(prev)}_a$ underlying these estimates is given in the variable `prevdenom`, which is multiplied by the prevalence estimates (given as `prev`) to obtain estimated prevalence counts $r^{(prev)}_a$ (given as `prevn`).
-Typically, prevalence is estimated from survey data, though for diseases with comprehensive registers, such as cancer, the denominator underlying the estimate might be assumed to equal the population size.
+Then secondly, estimates for five-year age groups were converted to one-year groups.
+Most simply this could be done by assuming the counts are equal for each year of age, but a more sophisticated method was used based on "temporal disaggregation" using one of the methods implemented by Sax and Steiner in the `tempdisagg` R package (@RJ-2013-028).
+This ensures that the disaggregated counts vary smoothly with age, while preserving the aggregate totals.
+Thirdly, counts from smaller geographical areas were simply added up to obtain the corresponding counts for the larger areas (English city regions) required in this analysis.
 
 ```{r,eval=TRUE}
-ihdbristol[ihdbristol$age %in% 50:55, c("age","prev","prevn","prevdenom")]
+ihdbristol[ihdbristol$age %in% 50:55, ]
 ```
 
-## ...Given point and interval estimates
+## Determining numerators and denominators from point and/or interval estimates
 
@@
 We can then search for the best-fitting Beta(r,n-r)$ distribution which has median $\hat{p}$ and (2.5,97.5) quantiles $(\hat{p}^{(lower)},\hat{p}^{(upper)})$, and set $r=a, n=a+b$.
-A utility to perform this search is provided in the SHELF package for expert elicitation (REF).
+A utility to perform this search is provided in the SHELF package for expert elicitation (@ohagan:elic).
 
-## ...Expressing additional uncertainty
+In other applications, the denominators might be directly available, e.g.
+we might know that the prevalence was estimated from a survey of $r_a^{(prev)}$ people, or we might assume that the incidence and mortality were obtained from a comprehensive registry of the population, in which case the denominator would equal the population size (within the specific stratum, e.g.
+as defined by age and gender, that the estimate refers to).
 
 The uncertainty inherent in the information supplied about each of incidence, prevalence and mortality is measured by the denominator.
-In the example above, this is the just size of the population used to produce the estimate.
+In the example above, this is equated to the size of the population used to produce the estimate.
 However, if we also suspected that one of the data sources may be biased, but were unsure about the direction of bias, we could downweight that data source by multiplying both the numerator and denominator by the same amount, e.g.
@@
 
-	* (incidence) $r_a^{(inc)} \sim ~ Binomial(n_a, P_{a12})$, where $P_{a12}$ is the annual transition probability from no disease to disease.
+	* (incidence) $r_a^{(inc)} \sim ~ Binomial(n_a, 1 - P_{a11})$, where $P_{a11}$ is the annual probability of remaining free of the disease.
 
@@
 
-3.
-define prior distributions for the unknown parameters.
-For incidence, currently the `disbayes` package assumes an unbounded uniform prior for each age independently.
-For case fatality, the package allows two alternatives, explained below.
+3. define prior distributions for the unknown parameters, as explained in the next section.
 
@@
 
-In these examples, the case fatality is only informed indirectly.  The `disbayes` package implements two alternative models for how case fatality depends on age.
+In these examples, the case fatality is only informed indirectly.
+The `disbayes` package implements two alternative models for how case fatality depends on age.
+The same methods can be used for how incidence depends on age.
 
@@
 This smooth function is defined by a spline basis $log(f_a) = \sum_k \beta_k g_k(a)$, where $g_k()$ are basis functions.
-A "thin plate" spline is used, following the default in the `mgcv` package (Wood), and the amount of smoothness is determined automatically through a hierarchical prior on the coefficients (REF Wood, `jagam` paper).
+A "thin plate" spline is used, following the default in the `mgcv` R package (Wood), and the amount of smoothness is determined automatically through a hierarchical prior on the coefficients.
 
@@
 
-When running `disbayes`, both models are fitted by default.
 Model (2) is more realistic.
@@
 
-1. estimate with denominator: through arguments `inc` and `inc_denom`.  The numerator is computed automatically, as described above.
+1. risk probability estimate with denominator: through arguments `inc` and `inc_denom`.  The numerator is computed automatically, by multiplying the denominator by the risk estimate.
 
-2. numerator and denominator, through arguments `inc_num` and `inc_denom`
+2. numerator and denominator, through arguments `inc_num` and `inc_denom`.  This is what is done in the example here.
 
@@
 dbres <- disbayes(dat = ihdbristol,
-                 inc = "inc", 
-                 inc_denom = "pop", 
-                 prev_num = "prevn", 
-                 prev_denom = "prevdenom",
-                 mort = "mort",
-                 mort_denom = "pop",
-                 eqage = 40
+                  age = "age",
+                 inc_num = "inc_num", 
+                 inc_denom = "inc_denom", 
+                 prev_num = "prev_num", 
+                 prev_denom = "prev_denom",
+                 mort_num = "mort_num",
+                 mort_denom = "mort_denom",
+                 eqage = 40,
+                 chains = 2,
+                 iter = 1000
                  )
@@
 The computation may take a few minutes, or less depending on the number of cores and their speed.
+The number of iterations `iter` should be high enough that the results are not affected by Monte Carlo error.
 
 The `disbayes` function returns a list with components `fit` and `fitu`, containing the smoothed and unsmoothed model fits respectively.
-Each of these components is an object in the `stanfit` format defined by the `rstan` R interface to Stan
+Each of these components is an object in the `stanfit` format defined by the `rstan` R interface to Stan.
 
@@
 The simulated chains should mix together and look like white noise.
+Also if the `disbayes` call resulted in many reports of "divergent transitions", then the chains are unlikely to have converged to a valid posterior.
+If convergence hasn't occurred, then the model probably needs to be simplified (e.g.
+by increasing `eqage`), or the data are insufficient.
 
@@
 The `disbayes` package provides some shortcuts for extracting summary statistics.
-The posterior medians and 95% credible intervals can be extracted from the fitted smooth model using the `summary` method for `disbayes` objects.
+The posterior medians, 95% credible intervals and other statistics for all parameters of interest can be extracted from the fitted models using the `tidy` method for `disbayes` objects.
 
-The result is a data frame with the variables indicated in the row names.
-These variables include `inc` (incidence), `cf` (case fatality), `mort` (mortality), `prev` (prevalence), each indexed by year of age.
+The result is a data frame with the variables indicated in the variable called `var`.
+Variables include `inc` (incidence), `cf` (case fatality), `mort` (mortality), `prev` (prevalence), each indexed by year of age given in the `age` variable.
+`model` indicates the smoothed or unsmoothed model.
 The first few rows of the summary below contain estimates of incidence for the first few years of age.
@@
 ```{r, eval=TRUE}
-summ <- summary(dbres)
+summ <- tidy(dbres)
 head(summ)
@@
 
-To extract results for a specific variable, the `vars` argument can be provided, e.g. for case fatality:
+To extract results for a specific variable and model, it is convenient to use standard tidyverse functions, e.g.
+to get the median and interquartile range of case fatality under the smoothed model for people aged between 60 and 65:
 
 ```{r,eval=TRUE}
-summ <- summary(dbres, vars="cf")
-head(summ)
-```
-
-To extract results from the unsmoothed model, the `summary_disbayes_fit` method is used.
-
-```{r,eval=TRUE}
-summ <- summary_disbayes_fit(dbres$fit, vars="cf")
-summu <- summary_disbayes_fit(dbres$fitu, vars="cf")
-head(summu)
+library(dplyr)
+summ %>% 
+  filter(var=="cf", between(age,60,65), model=="smoothed") %>%
+  select(age, `25%`, `50%`, `75%`)
 ```
@@
 The default `plot` method for objects returned by `disbayes` plots the posterior summaries for case fatality.
-Both the smoothed and unsmoothed estimates from age 40 are shown.
-This plot also overlays the corresponding estimates from DisMod II are overlaid in purple, included in the `ihdbristol` data in variable `dismod_cf`.
+Both the smoothed and unsmoothed estimates from age 40 and above are shown.
+If other styles of plot are desired, the tidy data frame output of the `tidy` method above is designed so that it can be passed to `ggplot2` without too much effort.
 
@@
 library(ggplot2)
-plot(dbres) +  ylab("Case fatality") + xlab("Age") + ylim(0,0.5) + xlim(40,100) + 
-  geom_line(aes(y=dismod_cf), data=ihdbristol, col="purple")
+plot(dbres) +  ylab("Case fatality") + xlab("Age") + 
+  coord_cartesian(xlim=c(40,100), ylim=c(0,0.5)) 
 ```
@@
 The smoothed model also removes the artefacts from the 5-year age grouping that are seen in the unsmoothed results.
-
-The smoothed model agrees with the DisMod II estimates, and has the benefit of characterising uncertainty as a posterior distribution.
+While the point estimates agree with those from DisMod II (not shown here), the Bayesian method has the benefit of characterising uncertainty as a posterior distribution.
 Uncertainty about case fatality in this example arises from the disease prevalence.
@@
 
-## Other things that could be done
+# Advanced models
 
-* More ways of supplying substantive prior information
+`disbayes` provides a selection of more advanced models for more complex data structures.
+These are substantially more computationally intensive than the basic model shown in this vignette, typically taking over an hour for applications similar to the one shown here.
+See the paper for further information about how they are defined mathematically, and their R help pages for further information on how to invoke the functions.
 
-* Account for variations between populations in different areas.  Hierarchical modelling might be used here - DISMOD-MR does this, I believe.
+## Hierarchical models
 
-* Hierarchical modelling to account for variations between ordered socioeconomic categories / income groups
+`disbayes_hier` is used to fit a joint model for data from multiple areas.
 
-* Borrow information between men/women
+## Hierarchical models with additive area and gender effects
 
-* Jointly model multiple diseases.  i.e. relaxing the assumption that mortality from causes other than the disease is the same for people with and without the disease.  Multimorbidity.
+`disbayes_hier` can also be used to fit a joint model for data from multiple areas and different genders.
+It is assumed that the relative case fatality between women and men is the same for each area.
+While this is labelled `gender` it can be used for any single binary predictor, though usually we will want to consider gender.
 
-* Account for variations in incidence and prevalence with time as well as age.
-DisMod II allows time trends to be modelled, but assumes that time trends are independent of age.
-Relaxing this assumption would be feasible, but possibly computationally expensive.
-	* For IHD in the UK, data on time trends may come from the Health Survey for England, years 1993, 1994, 1988, 2003, 2006, 2011?
+## Trends through time (non-hierarchical models only)
 
-* In general, to transparently represent uncertainty, and explicitly build in other sources of information, in situations where we suspect the data may not be representative
-
-All of these would be theoretically feasible in the above framework. The limits will be computational, and difficult to tell before getting hands on.
+`disbayes` can include information about trends through calendar time, which can be age dependent.
+The ratio of incidence between previous years and the current year (the one represented by the data) can be supplied as a matrix, with one row per year of age and one column per calendar year.
+See the help page, arguments `inc_trend_cf` and `cf_trend_mat`.
 

### 2021-07-07 doc/disbayes.R

@@ -1,0 +1,47 @@
+## -----------------------------------------------------------------------------
+knitr::opts_chunk$set(fig.width=7)
+
+## ----show=FALSE---------------------------------------------------------------
+library(disbayes)
+library(dplyr)
+ihdbristol <- ihdengland %>% filter(location=="Bristol", sex=="Male")
+ihdbristol %>% filter(between(age, 50, 55))
+
+## ----eval=TRUE----------------------------------------------------------------
+ihdbristol[ihdbristol$age %in% 50:55, ]
+
+## -----------------------------------------------------------------------------
+options(mc.cores = parallel::detectCores())
+
+## ---- eval=TRUE, cache=TRUE, warning=FALSE, message=FALSE---------------------
+dbres <- disbayes(dat = ihdbristol,
+                  age = "age",
+                 inc_num = "inc_num", 
+                 inc_denom = "inc_denom", 
+                 prev_num = "prev_num", 
+                 prev_denom = "prev_denom",
+                 mort_num = "mort_num",
+                 mort_denom = "mort_denom",
+                 eqage = 40,
+                 chains = 2,
+                 iter = 1000
+                 )
+
+## ----eval=TRUE----------------------------------------------------------------
+rstan::traceplot(dbres$fit, pars=paste0("cf[", 60:65, "]"))
+
+## ---- eval=TRUE---------------------------------------------------------------
+summ <- tidy(dbres)
+head(summ)
+
+## ----eval=TRUE----------------------------------------------------------------
+library(dplyr)
+summ %>% 
+  filter(var=="cf", between(age,60,65), model=="smoothed") %>%
+  select(age, `25%`, `50%`, `75%`)
+
+## ---- eval=TRUE, warning=FALSE------------------------------------------------
+library(ggplot2)
+plot(dbres) +  ylab("Case fatality") + xlab("Age") + 
+  coord_cartesian(xlim=c(40,100), ylim=c(0,0.5)) 
+

### 2021-10-01 data-raw/gbd_process.Rmd
Major update, featuring fast estimation method

@@
 
+Rename variables for consistency with disbayes
+
 ```{r}
@@
               values_fill = list(num=0, denom=5000)) %>% 
-  mutate(location = fct_recode(str_to_title(location),
+  mutate(area = fct_recode(str_to_title(location),
                            `Greater Manchester`="Greatermanchester",
@@
                            `East of England`="East Of England")) %>%
-  mutate(cause = str_to_sentence(cause)) %>%
-  arrange(location, cause, sex, ageyr) %>%
-  filter(! cause %in% c("Hypertensive heart disease"))
+  mutate(disease = str_to_sentence(cause)) %>%
+  rename(age = ageyr,
+         gender = sex,
+         inc_num = num_inc, inc_denom = denom_inc,
+         prev_num = num_prev, prev_denom = denom_prev,
+         mort_num = num_mort, mort_denom = denom_mort
+  ) %>% 
+  select(age, gender, disease, area, areatype, inc_num, inc_denom, prev_num, prev_denom, mort_num, mort_denom) %>%
+  arrange(area, disease, gender, age) %>%
+  filter(! disease %in% c("Hypertensive heart disease"))
 
@@
 library(readxl)
-rem <- read_excel("~/work/chronic/Figure_10__Predicted_10-year_net-survival_using_the_hybrid_approach_for_men_and_women_(aged_15_to_99_years).xls", range="A7:C26", na=":") %>%
+rem <- read_excel("~/work/chronic/cancerdata/Figure_10__Predicted_10-year_net-survival_using_the_hybrid_approach_for_men_and_women_(aged_15_to_99_years).xls", range="A7:C26", na=":") %>%
   pivot_longer(cols=c("Men","Women"), names_to = "sex", values_to = "surv10") %>% 
@@
                              ),
-         sex = fct_recode(sex, "Male"="Men", "Female"="Women")) %>% 
-  rename(cause = Cancer)
+         gender = fct_recode(sex, "Male"="Men", "Female"="Women")) %>% 
+  rename(disease = Cancer)
 
@@
 num_cancer <- gbddb %>% 
-  filter(cause %in% cancers, 
-         ageyr %in% 15:99) %>% 
-  group_by(sex, cause) %>% 
-  summarise(ncases = sum(num_prev)) %>% 
-  left_join(rem, by=c("cause","sex")) %>%
-  mutate(num_rem = round(ncases*prem1 / 100),
-         denom_rem = round(ncases / 100)) %>% 
-  select(sex, cause, num_rem, denom_rem)
+  filter(disease %in% cancers, 
+         age %in% 15:99) %>% 
+  group_by(gender, disease) %>% 
+  summarise(ncases = sum(prev_num)) %>% 
+  left_join(rem, by=c("disease","gender")) %>%
+  mutate(rem_num = round(ncases*prem1 / 100),
+         rem_denom = round(ncases / 100)) %>% 
+  select(gender, disease, rem_num, rem_denom)
 num_cancer
@@
   left_join(num_cancer) %>% 
-  mutate(num_rem = replace_na(num_rem, 0),
-         denom_rem = replace_na(denom_rem, 0))
+  mutate(rem_num = replace_na(rem_num, 0),
+         rem_denom = replace_na(rem_denom, 0))
 
 saveRDS(gbddb, file=file.path(relative_path_gbd, "gbddb.rds"))
-saveRDS(gbddb, file=file.path("~/work/chronic", "gbddb.rds"))
+saveRDS(gbddb, file=file.path("~/work/chronic/disbayes/metahit", "gbddb.rds"))
 

### 2021-10-01 metahit/paper_analyses.Rmd
Major update, featuring fast estimation method

@@
 source("paper_analyses_header.r")
-library(directlabels) # this is nice but do i even use it ?
+source("read_model_results.r")
 
@@
 
-gbdplot <- gbd %>% 
-  filter(cause %in% diseases, 
-         location %in% cityregions) %>%
-  mutate(mort_prob = num_mort/denom_mort,
-         mort_lower = qbeta(0.025, num_mort+0.5, denom_mort+0.5), 
-         mort_upper = qbeta(0.975, num_mort+0.5, denom_mort+0.5), 
-         prev_prob = num_prev/denom_prev,
-         prev_lower = qbeta(0.025, num_prev+0.5, denom_prev+0.5), 
-         prev_upper = qbeta(0.975, num_prev+0.5, denom_prev+0.5), 
-         inc_prob = num_inc/denom_inc,
-         inc_lower = qbeta(0.025, num_inc+0.5, denom_inc+0.5), 
-         inc_upper = qbeta(0.975, num_inc+0.5, denom_inc+0.5), 
-         location = fct_recode(location, 
-                               "North East"="North East (city region)",
-                               "West Midlands"="West Midlands (city region)"),
-         cause = fct_recode(cause, "Dementia" = "Alzheimer's disease and other dementias",
-                            "Lung cancer" = "Tracheal, bronchus, and lung cancer",
-                            "COPD" = "Chronic obstructive pulmonary disease"),
-         cause = factor(cause, levels = c("Ischemic heart disease", "Stroke", "Dementia",
-                                          "COPD", "Breast cancer", "Lung cancer", 
-                                          "Colon and rectum cancer","Stomach cancer",
-                                          "Uterine cancer")))
-gbdplotf <- gbdplot %>% filter(sex=="Female")
-gbdplotm <- gbdplot %>% filter(sex=="Male", cause != "Breast cancer")
+foo <- function(x){
+  cat(deparse(substitute(x)))
+}
 
 ht <- 3
-pdf("../write/mort_data.pdf", height=ht)
-ggplot(gbdplotf, aes(x=ageyr, y=mort_prob, col=location, group=location)) + 
-  facet_wrap(~cause, ncol=3, scales="free_y") + 
+pdf("../../write/mort_data.pdf", height=ht)
+gbdplot$areagender <- interaction(gbdplot$area, gbdplot$gender)
+ggplot(gbdplot, aes(x=age, y=mort_prob, col=area, lty=gender, group=areagender)) + 
+  facet_wrap(~disease, ncol=3, scales="free_y") + 
   coord_cartesian(xlim = c(50,100)) +
@@
 
-pdf("../write/inc_data.pdf", height=ht)
-ggplot(gbdplotf, aes(x=ageyr, y=inc_prob, col=location, group=location)) + 
-  facet_wrap(~cause, ncol=3, scales="free_y") + 
+pdf("../../write/inc_data.pdf", height=ht)
+ggplot(gbdplot, aes(x=age, y=inc_prob, col=area, lty=gender, group=areagender)) + 
+  facet_wrap(~disease, ncol=3, scales="free_y") + 
   coord_cartesian(xlim = c(50,100)) +
@@
 
-pdf("../write/prev_data.pdf", height=ht)
-ggplot(gbdplotf, aes(x=ageyr, y=prev_prob, col=location, group=location)) + 
-  facet_wrap(~cause, ncol=3, scales="free_y") + 
+pdf("../../write/prev_data.pdf", height=ht)
+ggplot(gbdplot, aes(x=age, y=prev_prob, col=area, lty=gender, group=areagender)) + 
+  facet_wrap(~disease, ncol=3, scales="free_y") + 
   coord_cartesian(xlim = c(50,100)) +
@@
   geom_line() + 
-  labs(col = "City region") +
-  theme_bw() + 
-  theme(strip.text.x = element_text(size = 7.5))
-dev.off()
-
-pdf("../write/mort_datam.pdf", height=ht)
-ggplot(gbdplotm, aes(x=ageyr, y=mort_prob, col=location, group=location)) + 
-  facet_wrap(~cause, ncol=3, scales="free_y") + 
-  coord_cartesian(xlim = c(50,100)) +
-  ylab("Annual mortality probability") + 
-  xlab("Year of age") +
-  geom_ribbon(aes(ymin=mort_lower, ymax=mort_upper), fill="gray80", col="gray80") +
-  geom_line() + 
-  labs(col = "City region") +
+  labs(col = "City region", lty="") +
   theme_bw() + 
   theme(strip.text.x = element_text(size = 7.5),
-        legend.position = "none")
+        legend.spacing.y = unit(0.1, "cm"))
 dev.off()
-
-pdf("../write/inc_datam.pdf", height=ht)
-ggplot(gbdplotm, aes(x=ageyr, y=inc_prob, col=location, group=location)) + 
-  facet_wrap(~cause, ncol=3, scales="free_y") + 
-  coord_cartesian(xlim = c(50,100)) +
-  ylab("Annual incidence probability") + 
-  xlab("Year of age") +
-  geom_ribbon(aes(ymin=inc_lower, ymax=inc_upper), fill="gray80", col="gray80") +
-  geom_line() + 
-  labs(col = "City region") +
-  theme_bw() + 
-  theme(strip.text.x = element_text(size = 7.5),
-        legend.position = "none")
-dev.off()
-
-pdf("../write/prev_datam.pdf", height=ht)
-ggplot(gbdplotm, aes(x=ageyr, y=prev_prob, col=location, group=location)) + 
-  facet_wrap(~cause, ncol=3, scales="free_y") + 
-  coord_cartesian(xlim = c(50,100)) +
-  ylab("Prevalence probability") + 
-  xlab("Year of age") +
-  geom_ribbon(aes(ymin=prev_lower, ymax=prev_upper), fill="gray80", col="gray80") +
-  geom_line() + 
-  labs(col = "City region") +
-  theme_bw() + 
-  theme(strip.text.x = element_text(size = 7.5))
-dev.off()
-
-```
-
-# Run the Bayesian modelling analyses
-
-```{r,eval=FALSE}
-# These are done on a cluster using SLURM so we don't source these files 
-# source("paper_analyses_nonhier.r")
-# source("paper_analyses_hier.r")
-# source("paper_analyses_hier_gender.r")
-```
-
-# Read and join all model results
-
-```{r}
-reslist <- loolist <- vector(nrow(rundf), mode="list")
-for (i in 1:nrow(rundf)){
-  fname <- sprintf("results_nonhier/res%s.rds",i)
-  if (file.exists(fname)) {
-    resi <- readRDS(fname)
-    reslist[[i]] <- resi$res
-    loolist[[i]] <- resi$loo
-  }
-}
-
-dput(as.numeric(which(sapply(reslist, is.null)))) # which ones didn't work due to tmpfile conflicts
-
-resnh <- do.call("rbind", reslist) 
-resns <- resnh %>% 
-  filter(model=="unsmoothed")
-resnh <- resnh %>%
-  filter(model=="smoothed") %>% 
-  mutate(model="Independent areas")
-loonh <- do.call("rbind", loolist) %>%
-  mutate(model="Independent areas")
-
-resi <- readRDS("results_nonhier/res_eng_utrc_female.rds")
-res_utrc <- resi$res; loo_utrc <- resi$loo
-resi <- readRDS("results_nonhier/res_eng_stmc_male.rds")
-res_stmc_m <- resi$res; loo_stmc_m <- resi$loo
-resi <- readRDS("results_nonhier/res_eng_stmc_female.rds")
-res_stmc_f <- resi$res; loo_stmc_f <- resi$loo
-resnat <- rbind(res_utrc, res_stmc_m, res_stmc_f) %>%
-  filter(model=="smoothed") %>% 
-  mutate(model="National")
-resnh <- resnh %>% 
-  dplyr::filter(!(disease %in% c("Stomach cancer", "Uterine cancer"))) %>%
-  full_join(resnat)
-
-loonat <- rbind(loo_utrc, loo_stmc_m, loo_stmc_f) %>%
-  mutate(model="National")
-loonh <- loonh %>% 
-  dplyr::filter(!(disease %in% c("Stomach cancer", "Uterine cancer"))) %>%
-  full_join(loonat)
-
-res %>% filter(disease=="Uterine cancer", var=="cf", between(age, 58, 70))
-res_utrc %>% filter(disease=="Uterine cancer", var=="cf", between(age, 90, 100))
-## Smoothed hasn't worked
-
-reslist <- loolist <- vector(nrow(hierrundf), mode="list")
-for (i in setdiff(1:nrow(hierrundf), c())){
-  resi <- readRDS(sprintf("results_hier/res%s.rds",i))
-  reslist[[i]] <- resi[[1]]
-  loolist[[i]] <- resi[[2]]
-}
-resh <- do.call("rbind", reslist)
-resh$model <- "Hierarchical"
-resh$area <- factor(as.character(unique(gbd$location))[as.numeric(resh$area)])  # TODO will not be needed if disbayes_hier returns original factor levels for areas instead of mucking with codes
-looh <- do.call("rbind", loolist) %>%
-  mutate(area = factor(area, labels=areas),
-         model = "Hierarchical")
-
-reslist <- loolist <- vector(nrow(hierrungdf), mode="list")
-for (i in setdiff(1:nrow(hierrungdf), 7)){ # exclude stomach cancer
-  resi <- readRDS(sprintf("results_hier_gender/res%s.rds",i))
-  reslist[[i]] <- resi[[1]]
-  loolist[[i]] <- resi[[2]]
-}
-resg <- do.call("rbind", reslist) %>%
-  mutate(gender = fct_recode(gender, "Female"="1", "Male"="2"),
-         model = "Hierarchical joint gender",
-         area = factor(as.character(unique(gbd$location))[as.numeric(area)])
-  )
-loog <- do.call("rbind", loolist) %>%
-  mutate(area = factor(area, labels=areas),
-         model = "Hierarchical joint gender",
-         gender = factor(gender, labels=genders))
-
-reslist <- vector(nrow(trendrundf), mode="list")
-for (i in setdiff(1:nrow(trendrundf), c(2))){
-  load(sprintf("results_trend/res%s.rda",i))  
-  reslist[[i]] <- res
-}
-
-restr <- do.call("rbind", reslist)
-restr$model <- "Time trends"
-
-disease_shorten <- c("Dementia" = "Alzheimer's disease and other dementias",
-                     "Lung cancer" = "Tracheal, bronchus, and lung cancer",
-                     "COPD" = "Chronic obstructive pulmonary disease")
-disease_order <- c("Ischemic heart disease", "Stroke", "Dementia", "COPD", "Breast cancer", "Lung cancer", 
-                   "Colon and rectum cancer","Stomach cancer","Uterine cancer")
-
-resall <- full_join(resnh, resh) %>% 
-  full_join(resg) %>%  # remove after new run
-  full_join(restr) %>%
-  mutate(disease = fct_recode(disease, !!!disease_shorten),
-         disease = factor(disease, levels = disease_order),
-         year = if_else(model == "Time trends", year + 1917, 2017))
-
-looall <- full_join(loonh, looh) %>%
-  full_join(loog) %>%
-  mutate(disease = fct_recode(disease, !!!disease_shorten),
-         disease = factor(disease, levels = disease_order))
 
@@
 
-```{r}
+* Stomach cancer and uterine cancer are done using national data.
 
-resnh %>% filter(disease == "Stomach cancer", var=="cf", area=="Bristol", gender=="Female")
+```{r,eval=FALSE}
 
-## Convergence checks.  
-## Hasn't converged for stomach and uterine cancer, in areas where there were no deaths or small numbers of deaths.
-## We use national data for these instead. 
-## What's happened now for stroke?
+## Convergence checks.   All OK with 5000 samples and the occasional model simplification
+resnc <- resnh %>% filter(Rhat > 1.05)
+table(resnc$disease, resnc$area, resnc$gender)
 
-resnc <- resnh %>% filter(Rhat > 2) 
-resnc <- resnh %>% filter(Rhat > 1.2) 
-resnc <- resnh %>% filter(Rhat > 1.1) 
-resnc <- resnh %>% filter(Rhat > 1.05) 
-table(resnc$disease, resnc$gender)
-table(resnc$disease, resnc$gender, resnc$area)["Stomach cancer",,]
-gbd %>% filter(location=="Bristol", cause=="Uterine cancer") %>% select(ageyr, num_mort, num_inc, num_rem)  # No deaths. 
-gbd %>% filter(cause=="Uterine cancer", location %in% cityregions) %>% pull(num_mort) %>% sum
-gbd %>% filter(cause=="Uterine cancer", location %in% cityregions) %>% pull(num_inc) %>% sum
+## Compare modal approx with posterior 
+## Page of plots for a single area and gender.  Disease variation more important than area/gender variation  
+## stmc, utrc use national point estimates 
+## What are the current settings used for these?
+
+resall %>% 
+  filter(var=="cf", disease=="Dementia", gender=="Female", 
+         area=="Bristol", between(age, 70, 90), model=="Independent areas")
+
+resoptunc <- readRDS("resoptunc_nonhier.rds") %>% 
+  filter(!disease %in% c("Stomach cancer","Uterine cancer"))
+resoptnat <- readRDS("resopt_nat.rds")
+resoptunccf <- resoptunc %>%
+  full_join(resoptnat) %>% 
+  filter(var=="cf") %>%
+  mutate(p50 = 1 - exp(-`par`),
+         plow = 1 - exp(-`2.5%`),
+         pupp = 1 - exp(-`97.5%`)) %>%
+  filter(area=="Bristol" | area=="England", gender=="Female") %>%
+  mutate(disease = fct_recode(disease, !!!disease_shorten),
+           disease = factor(disease, levels = disease_order)) %>%
+  filter(age %in% c(70, 75, 80, 85, 90)) %>%
+  mutate(Algorithm = "Mode approximation")
+
+pdf("../../write/opt_compare.pdf")
+
+resp <- resall %>% 
+  dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
+  dplyr::filter(area=="Bristol" | (area=="England" & model=="National"), gender=="Female") %>%
+  mutate(Algorithm="MCMC") %>%
+  full_join(resoptunccf)
+resp %>%
+  filter(age %in% c(70, 75, 80, 85, 90)) %>%
+  mutate(par = 1 - exp(-`50%`),
+         plow = 1 - exp(-`2.5%`),
+         pupp = 1 - exp(-`97.5%`)) %>%
+  ggplot(aes(x=age, y=par, col=algorithm, lty=algorithm)) + 
+  geom_errorbar(aes(ymin=plow, ymax=pupp), width=1.5,
+              position=position_dodge(width=2), lwd=1.3) +
+  geom_point(position=position_dodge(width=2), size=1.3) + 
+  facet_wrap(vars(disease), ncol=2, scales="free_y") + 
+  xlab("Age (years)") + ylab("Case fatality probability")
+
+dev.off()
 
@@
 
-#  TODO  more investigation of results for tbalc, carc 
-## where theres between area var and unc 
-## Lung cancer - lots of BAV. CARC less so. 
-## What to we get from hier model in each case?
-
-example_age <- 90
-resall %>% 
-  dplyr::filter(model %in% c("Independent areas"), var=="cf",
-                disease == "Lung cancer", gender=="Male",age==example_age) %>%
-ggplot(aes(x=`50%`, y=area, col=model)) +
-    coord_cartesian() +
-    geom_point(position=ggstance::position_dodgev(height=0.4)) +
-    geom_errorbarh(aes(xmin=`2.5%`, xmax=`97.5%`),
-                   position=ggstance::position_dodgev(height=0.4)) +
-    xlab(sprintf("Case fatality for age %s", example_age) )
-
 ```
@@
 
-## convergence checks. look at Rhat
-## i=16 only one with divergent- utrc - we're not using this model anyway
-## stomach cancer is only one that hasn't converged
+## UTRC doesn't converge, so we use the empirical Bayes results
 
-summary(resh$Rhat) 
-resh %>% filter(gender=="Male", disease=="Stomach cancer", var=="cf", age>60, age<70) 
+resh %>%
+  filter(disease=="Stomach cancer", gender=="Male", var=="cf") %>%
+  ggplot(aes(x=age, y=`50%`, col=area, group=area)) + 
+  geom_line()
+
+resh %>% filter(gender=="Female", disease=="Stomach cancer", 
+                var=="cf", age>40, age<60) %>% select(age,`50%`,Rhat) 
 resnc <- resh %>% filter(Rhat > 2) 
-resnc <- resh %>% filter(Rhat > 1.1) 
+resnc <- resh %>% filter(Rhat > 1.05) 
 table(resnc$disease, resnc$gender)
-table(resnc$disease, resnc$gender, resnc$area)
+
+## Look at precision of CRC estimates as exemplar. 
+## Effective sample size of 1000 after a few hours
 resh %>% filter(gender=="Male", disease=="Ischemic heart disease", age==60, var=="cf") %>% head
-resh %>% filter(gender=="Male", disease=="Stroke", age==60, var=="cf") %>% head
-resh %>% filter(gender=="Male", disease=="Stomach cancer", age==60, var=="cf") %>% head
-resh %>% filter(disease=="Uterine cancer", age==60, var=="cf") %>% head
-resh %>% filter(gender=="Male", area=="Bristol", age==60, var=="cf") %>% head
+resh %>% filter(gender=="Male", disease=="Colon and rectum cancer", age==60, var=="cf") %>% head
 
@@
 ## Specific age and gender, compare between areas 
-## Can use this to assess benefit from hier models. 
-## Shrinkage is moderate.
+## Not used in paper
 
@@
 ## Between area variability in age curves, point ests.  CARC.  Shrinkage for oldest
+## Use this in paper
 
@@
 
+Fuck me the Rhats are fucked 
+Theyve basically all broken . 
+Must be some unused parameter or wev that has been left in to break it 
+Use the optim models to debug it 
+looks like the incidence related parameters
+beta_inc not in tidy but can't tell
+
 ```{r}
@@
 resnc <- resg %>% filter(Rhat > 1.01)
+
+resnc <- resg %>% filter(Rhat > 1.1, disease=="Ischemic heart disease") 
+resg %>% filter(grepl("beta_inc",var))
+
 table(resnc$disease)
@@
 gbd %>% filter(ageyr==70, 
-               cause=="Tracheal, bronchus, and lung cancer", 
-               location %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
-  mutate(p_mort = num_mort/denom_mort, p_prev=num_prev/denom_prev, mpratio=p_mort/p_prev) %>%
-  select(sex, location, num_mort, denom_mort, p_mort, p_prev, mpratio)
+               disease=="Tracheal, bronchus, and lung cancer", 
+               area %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
+  mutate(p_mort = mort_num/mort_denom, p_prev=prev_num/prev_denom, mpratio=p_mort/p_prev) %>%
+  select(sex, area, mort_num, mort_denom, p_mort, p_prev, mpratio)
 ## YES: MORT/PREV AGREES WITH CF
@@
 gbd %>% filter(ageyr==70, 
-               cause=="Tracheal, bronchus, and lung cancer", 
-               location %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
-  group_by(location) %>%
-  mutate(num_mort=sum(num_mort), denom_mort=sum(denom_mort), 
-         num_prev=sum(num_prev), denom_prev=sum(denom_prev)) %>%
-  mutate(p_mort = num_mort/denom_mort, p_prev=num_prev/denom_prev, mpratio=p_mort/p_prev) %>%
-  select(sex, location, num_mort, denom_mort, p_mort, p_prev, mpratio)
+               disease=="Tracheal, bronchus, and lung cancer", 
+               area %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
+  group_by(area) %>%
+  mutate(mort_num=sum(mort_num), mort_denom=sum(mort_denom), 
+         prev_num=sum(prev_num), prev_denom=sum(prev_denom)) %>%
+  mutate(p_mort = mort_num/mort_denom, p_prev=prev_num/prev_denom, mpratio=p_mort/p_prev) %>%
+  select(sex, area, mort_num, mort_denom, p_mort, p_prev, mpratio)
 
@@
 ## Lookup table for all results
+
 100 * 9 * 18 * 2 * 3 results for each of inc and CF
 
-How to distribute this on github?  A data file in the disbayes package?  
-A separate package with code and results?  yea should be standard format for these.  what does hadley do 
-Just data and vignettes, no code.  yeah. 
-where to put the analysis scripts and slurm script? inst/scripts or suchlike. 
-yeah
+TODO just save this data frame in the metahit directory. there's no standard format
+
+TODO append the results from optimisation.  OK to have multiple results for one case 
+or just save in different format, since one has mode
 
@@
   arrange(disease, var, gender, age, area, model)
-app_tab
+app_tab %>% head
 ```
@@
 
-First compare nonhier with hier.
+Compare nonhierarchical with hierarchical model, for all diseases and outcomes.
+
+Colon and rectum cancer is highlighted in the paper as a case where the hierarchical model makes a difference to the results, for ages 80+.   The hierarchical model gives better predictions here.
 
@@
   full_join(looh %>% mutate(model="Hierarchical")) %>%
-  filter(!(model=="National"))
+  filter(!(model=="National")) %>%
+  select(age, gender, elpd_loo, model, area, disease, var) %>%
+  group_by(model, gender, disease, var)
 
+looall %>% 
+  summarise(eisum = sum(elpd_loo)) %>%
+  pivot_wider(names_from=model, values_from=eisum) %>%
+  mutate(elpd_diff = `Hierarchical` - `Independent areas`)
+
+looall %>% 
+  filter(age > 80) %>%
+  summarise(eisum = sum(elpd_loo)) %>%
+  pivot_wider(names_from=model, values_from=eisum) %>%
+  mutate(elpd_diff = `Hierarchical` - `Independent areas`)%>%
+  filter(disease=="Colon and rectum cancer")
+```
+
+Graph of how predictive ability depends on age and area, for selected disease and outcome.
+Hierarchical model is better for older ages.
+Age 50 data seem to be strong enough that shrinkage is not needed.
+Note that age 50 parameters are also assumed to apply to younger ages.
+
+```{r}
 loosub <- looall %>% 
-#  filter(disease=="Colon and rectum cancer", var=="mort", gender=="Male",
-  filter(disease=="Ischemic heart disease", var=="inc", gender=="Male",
-         area %in% cityregions) %>%
-  select(age, elpd_loo, model, area) %>%
+  filter(disease=="Colon and rectum cancer", gender=="Male", var=="mort") %>%
+  ## sum ELPD by five year age group
   pivot_wider(names_from=model, values_from=elpd_loo) %>%
   mutate(elpd_diff = `Hierarchical` - `Independent areas`) %>%
-  ## sum ELPD by five year age group
   mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
-  group_by(area, agegroup) %>%
+  group_by(area, agegroup, gender, var) %>%
   summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
@@
             vjust=c(0,1), hjust=c(0,0)) +
-  geom_dl(aes(label=area), method= list(dl.combine("first.points", "last.points")))
-
-## so why is hier better for mid ages?  worse for below 
-## surely age 95 value is iffy due to weak data.
-
-## why two groups of areas?  It's not the area numerator or denominator size.
-## Labels are right way round. higher ELPD good
-
-gbd %>% filter(cause=="Colon and rectum cancer") %>% 
-  group_by(location) %>% filter(areatype=="cityregion") %>% 
-  summarise(n=sum(num_mort)) %>% arrange(n)
-
-looall %>% filter(age>90, age <92, gender=="Male", disease=="Colon and rectum cancer", var=="mort") %>%
-  select(area, p_loo, model) %>%
-  pivot_wider(names_from=model, values_from=p_loo) %>% as.data.frame()
-
-## These results are hard to explain
-## Suspect there is no point using cross validation to compare hier and nonhier
-## The results are the same where there is data
-## Hier only helps for extrapolation to oldest ages where there is no data
-## That is a matter of judgement that can use exchangeability.
+  directlabels::geom_dl(aes(label=area), 
+                        method=list(directlabels::dl.combine("first.points", "last.points")))
 
@@
 datt <- gbd %>% 
-  filter(location=="Leeds", cause=="Ischemic heart disease") %>%
-  mutate(inc_prob = num_inc / denom_inc, 
+  filter(area=="Leeds", disease=="Ischemic heart disease") %>%
+  mutate(inc_prob = inc_num / inc_denom, 
          inc_rate = -log(1-inc_prob), 
@@
 
-resoy <- resall %>% filter((model=="Time trends" & year > 2000 ),
+resoy <- resall %>% filter(model=="Time trends", 
+                            year == 2017,
                             area=="Leeds", 
@@
 #                          gender=="Male"
-)
+) %>% full_join(resoy)
+
+resopt <- readRDS(file="resopt_trend.rds") %>% 
+  filter(disease=="Ischemic heart disease",area=="Leeds",
+         var %in% c("inc","cf"), year==99) %>%
+  mutate(var = fct_recode(var, "Incidence"="inc", "Case fatality"="cf"))
+
 rest %>% 
@@
            group=interaction(gender, model))) +
+  coord_cartesian(ylim=c(0.0,0.04)) +
   geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), alpha=0.2) +
@@
   geom_line(data=datt, aes(x=ageyr, y=inc_rate), inherit.aes = FALSE) +
-  geom_line(data=resoy, aes(x=age, y=`50%`, col=factor(year), group=year), lty=2, inherit.aes = FALSE)
+  geom_line(data=resopt, aes(x=age, y=par), inherit.aes=FALSE, col="blue")
+
+## Point estimates are accurate, so we can use them to diagnose the bias
+
+#  geom_line(data=resoy, aes(x=age, y=`50%`, col=factor(year), group=year), lty=2, inherit.aes = FALSE)
 dev.off()
 
+## Experimental sensitivity analyses to different trends
+## what was it? 2003-2010 continues. saved in restrs 
+## where's the conclusion from this?  estimates vary a lot
+
+rest_saved <- rest %>% 
+  filter(between(age, 50, 90), model=="Time trends", 
+         gender=="Male", var %in% c("inc","cf")) 
+res_saved <- resnh %>% filter(between(age, 50, 90), model=="Independent areas", disease=="Ischemic heart disease", area=="Leeds", gender=="Male", var %in% c("inc","cf"))
+
+restrs <- restrs %>% 
+  mutate(trendin = ifelse(inctrend & cftrend, "both", 
+  ifelse(inctrend, "inc", ifelse(cftrend, "cf", "neither")))) %>%
+  mutate(trendin = ifelse(trsens=="zero", "neither", trendin))
+
+dat <- restrs %>% 
+  filter(var %in% c("inc","cf"), model=="smoothed", trsens %in% c("hi", "zero"))
+dat$age <- rep(1:100, length.out=nrow(dat))
+dat %>%
+  filter(age < 80) %>%
+ggplot(aes(x=age, y=`50%`, col=trendin)) + 
+  geom_line(lwd=2) + 
+  facet_grid(cols=vars(var)) + 
+  geom_line(data=datt %>% filter(gender=="Male") %>% mutate(var=fct_recode(var,"inc"="Incidence")), 
+            aes(x=ageyr, y=inc_rate), inherit.aes = FALSE)
+
++ 
+  geom_line(data=rest_saved, col="black") + 
+  geom_line(data=res_saved, col="blue")
+
+restrs %>% 
+  filter(var %in% c("inc","cf")) %>%
+  mutate(age = rep(1:100, length.out=n())) %>%
+  filter(model=="smoothed", age==60, var=="cf", cftrend==TRUE, inctrend==TRUE)
+
 ```
+
+Right this is saying that it doesn't make much difference to CF, but big diff to inc as in Scar
+were the original plots wrong.   OK fine just axis limit distortion.
+
+So the only problem left is disagreement with current empirical incidence 
+dismod2 types would say the model is "forcing consistency" 
+evidence of conflict between current inc and current CF  
+Hm is this evidence that the gbd data are consistent with themselves under a model that ignores trends.
+
+Hey wot it does match.   Wot about original independent-areas results
+
+cf - higher with inc or both, lower with cf or neither 
+inc - higher with cf only, then neither, then both, then lowest with inc.
+
+So dose response is OK 
+but they interact in unpredictable ways 
+and need to fix the 5 year tihng 
+and reconcile the modest changes here with the big changes from the real data
+
+Compare real with artificial trend data.  Where does the difference come from? 
+Looks reasonably representative of CF to me.  Age 70 value is generous 
+Incidence: trend data is more noisy.  Pick an even higher value that's on the high end of the data.
+
+```{r}
+ism <- inctrends_hi %>% as.data.frame() %>% 
+  mutate(age = 1:100) %>%
+  pivot_longer(names_to = "year", values_to = "p2017", cols = 1:100) %>%
+  mutate(year = as.numeric(year)) %>% 
+  filter(age %in% c(40,60,80,100), year > 1970) %>% 
+  mutate(age = factor(age), model = "sens")
+
+ihdtrends %>% 
+  filter(age %in% c(40,60,80,100), year > 1970, gender=="Male", outcome=="Incidence") %>% 
+  mutate(age = factor(age), model = "data") %>%
+  full_join(ism) %>%
+  ggplot(aes(x=year, y=p2017, col=age, lty=model)) + geom_line()
+
+```
+
+Well that's odd.
+
+With inc trends on. 
+There's much less of an effect than there was with the real data. 
+Consistent dose response effect 
+CF trend makes no difference to current CF ests , but affects current inc ests. Wot?
+
+With cf trends on
+Only an effect on cf if there is inc trends.
+Consistent dose response when there's an effect
+Inc est: both inc and cf trends make a difference.  moderate differences
+
+Is this a bug in the sens analysis setup??   cf trends are not affecting cf
+And why are there still those stupid 5 year artefacts for incidence?  didn't i transfer the new data?
+
+OK think og it a new way 
+TREND IN: 
+inc but not cf : 
+cf but not inc : reduced inc 
+both inc and cf:     reduced inc
+
+inc highest: 
+inctrend false, cftrend true
+inctrend true, cftrend true 
+cftrend false.   
+inc lowest
+
+cf highest with 
+inctrend true, cftrend false or true 
+inctrend false

### 2021-10-01 R/disbayes-package.R

@@ -7,10 +7,11 @@
 #' @useDynLib disbayes, .registration = TRUE
 #' @import methods
 #' @import Rcpp
+#' @importFrom stats coef lm qbeta qexp qnorm quantile rlnorm rnorm
 #' @importFrom generics tidy
 #' @importFrom rstan sampling
-#' @importFrom stats lm sd rlnorm rnorm
 #' @importFrom mgcv smoothCon jagam s
+#' @importFrom matrixStats colQuantiles
 #' @import dplyr
 #' @importFrom tidyr extract pivot_wider pivot_longer
 #' @importFrom magrittr "%>%"

### 2021-10-01 R/disbayes_utils.R

@@ -1,53 +1,0 @@
-## For the moment, don't smooth either inc or CF
-## it's supposed to be fast 
-## may need to smooth one before estimating other later on 
-
-init_rate <- function(rate, dat){
-    default <- 0.001
-    rcrude <- dat[[sprintf("%s_num",rate)]] / dat[[sprintf("%s_denom",rate)]]
-    rcrude[is.na(rcrude)] <- default
-    rcrude[is.nan(rcrude)] <- default
-    rcrude <- pmax(default, rcrude)
-    rcrude 
-}
-
-train_rate <- function(dat,
-                       inc_init=NULL, cf_init=0.01, rem_init=NULL, remission,
-                       iter_train = 1000, 
-                       return="stan", ...){
-    if (is.null(inc_init)) inc_init <- init_rate("inc", dat)
-    if (is.null(rem_init)) rem_init <- init_rate("rem", dat)
-    nage <- dat$nage
-    Xdummy <- matrix(0, nrow=nage, ncol=2)
-    datstanu <- c(dat, list(smooth_cf=0, smooth_inc = 0, const_cf=0, trend=0, nyr = 1,
-                            increasing_cf=0, K=2, X=Xdummy, sprior=dat$sprior,
-                            inc_trend = array(1, dim=c(nage,1)),
-                            cf_trend = array(1, dim=c(nage,1)) 
-                            ))
-    initu <- function(){
-        list(cf_par = rnorm(nage, mean=cf_init, sd=cf_init/10),
-             rem_par = rnorm(nage*remission, mean=rem_init, sd=rem_init/10),
-             inc_par = rnorm(nage, mean=inc_init, sd=inc_init/10))
-    }
-    fitu <- rstan::sampling(stanmodels$disbayes, data = datstanu, 
-                            init = initu, include = FALSE, pars=c("beta","lambda"),
-                            iter = iter_train,
-                            ...)
-    fitu 
-}
-
-
-## Obtains spline basis terms and initial values for their coefficients
-## given estimates of incidence or CF from unsmoothed model.
-
-init_smooth <- function(y, eqage, s_opts=NULL){
-    nage <- length(y)
-    age <- agecons <- 1:nage 
-    agecons[1:eqage] <- eqage
-    sm <- mgcv::smoothCon(s(x),
-                          data=data.frame(x=agecons),
-                          diagonal.penalty=TRUE)[[1]]
-    X <- sm$X
-    beta <- coef(lm(y ~ X - 1))
-    list(X=X, beta=beta)
-}

### 2021-10-01 R/inits.R

@@ -1,0 +1,175 @@
+## mdat <- remission, eqage, const_rem 
+## min <- cf_init
+
+init_rates <- function(dat, mdata, idata,...){
+    inc_init <- init_rate("inc", dat)
+    rem_init <- init_rate("rem", dat, agg=mdata$const_rem)
+    optu <- fit_unsmoothed_opt(dat, inc_init=inc_init, rem_init=rem_init,
+                               mdata=mdata, idata=idata)
+    optdf <- tidy_disbayes_opt(optu, .disbayes_vars)
+    cfcrude <- optdf$mode[optdf$var=="cf"] 
+    inccrude <- optdf$mode[optdf$var=="inc"]
+    eqage <- mdata$eqage
+    for (i in 1:eqage){
+        cfcrude[i] <- cfcrude[eqage+1]
+        inccrude[i] <- inccrude[eqage+1]
+    }
+    init_eqage_hi <- 90 
+    for (i in init_eqage_hi:dat$nage){
+        cfcrude[i] <- cfcrude[init_eqage_hi-1]
+        inccrude[i] <- inccrude[init_eqage_hi-1]
+    }
+    list(cf=cfcrude, inc=inccrude, rem=rem_init, optu=optu)
+}
+
+init_rate <- function(rate, dat, agg=FALSE){
+    default <- 0.001
+    nums <- dat[[sprintf("%s_num",rate)]]
+    denoms <- dat[[sprintf("%s_denom",rate)]]
+    if (agg) { nums <- sum(nums); denoms <- sum(denoms) }
+    rcrude <- nums / denoms 
+    rcrude[is.na(rcrude)] <- default
+    rcrude[is.nan(rcrude)] <- default
+    rcrude <- pmax(default, rcrude)
+    rcrude 
+}
+
+fit_unsmoothed_opt <- function(dat, inc_init=NULL, rem_init=NULL,
+                               mdata, idata){
+    if (is.null(inc_init)) inc_init <- init_rate("inc", dat)
+    if (is.null(rem_init)) rem_init <- init_rate("rem", dat, mdata$const_rem)
+    nage <- dat$nage
+    Xdummy <- matrix(0, nrow=nage, ncol=2)
+    datstanu <- c(dat, mdata)
+    ## quantities in the data that are different for the purpose of this training model
+    data_fixed <- list(smooth_cf =  0, smooth_inc = 0, const_cf = 0, trend = 0, nyr=1, nbias=1,
[…]

### 2021-10-01 R/tidy.R

@@ -1,0 +1,235 @@
+##' Form a tidy data frame from the estimates from a disbayes fit
+##'
+##' Simply call this after fitting disbayes, as, e.g.
+##' ```
+##' res <- disbayes(...)
+##' tidy(res)
+##' ```
+##' 
+##' @importFrom generics tidy
+##'
+##' @param x Object returned by \code{\link{disbayes}}
+##'
+##' @param startyear Only used for models with time trends.  Numeric year represented by year 1 in the data. For example, set this to 1918 to convert years 1-100 to years 1918-2017.
+##'
+##' @param ... Other arguments (currently unused)
+##'
+##' @return A data frame with one row per model parameter, giving summary statistics
+##' for the posterior distribution for that parameter.   For array parameters, e.g. those
+##' that depend on age or area, then the age and area are returned in separate columns,
+##' to make it easier to summarise and plot the results, e.g. using \pkg{ggplot2}.
+##'
+##' Model parameters might include, depending on the model specification, 
+##'
+##' * `cf` Case fatality rates
+##'
+##' * `inc` Incidence rates
+##'
+##' * `rem` Remission rates
+##'
+##' * `inc_par`, `cf_par`, `rem_par`.  Same as the above - should
+##'
+##' * `inc_prob`, `rem_prob`.  Annual incidence and remission probabilities. 
+##'
+##' * `mort` Annual mortality probability.
+##'
+##' * `prev` Prevalence probability. 
+##'
+##' * `state_probs` State occupancy probabilities.
+##'
+##' * `beta`, `beta_inc` Coefficients of the spline basis for case fatality and incidence respectively.
+##'
+##' * `lambda_cf`, `lambda_inc` Smoothness parameters of the spline functions.
+##'
+##' * `prevzero` Prevalence at age zero
+##'
+##' * `cfbase` Case fatality at the baseline age (only in models where case fatality is increasing).
+##'
+##' * `dcf` Annual increments in case fatality (only in models where case fatality is increasing).
+##'
+##' * `bias_loghr` Log hazard ratio describing bias in case fatality between datasets (only in models where `bias_model` has been set).
+##'
+##' For models with time trends:
+##'
[…]

### 2021-10-01 data-raw/ihdlondon.r

@@ -1,36 +1,0 @@
-library(devtools)
-
-inc <- read.csv("../../belen/second/GreaterLondon/Collection/IHDIncidence.csv", skip=4, nrow=20, header=FALSE)[,c(1,2,6)]
-mort <- read.csv("../../belen/second/GreaterLondon/Collection/IHDMortality.csv", skip=4, nrow=20, header=FALSE)[,c(2,6)]
-prev <- read.csv("../../belen/second/GreaterLondon/Collection/IHDprevalence.csv", skip=4, nrow=20, header=FALSE)[,c(2,6)]
-pop <- read.csv("../../belen/second/GreaterLondon/Collection/Population.csv", skip=4, nrow=20, header=FALSE)[,c(2,3)]
-names(inc) <- c("agegrp","incmale","incfemale")
-names(mort) <- c("mortmale","mortfemale")
-names(prev) <- c("prevmale","prevfemale")
-names(pop) <- c("popmale","popfemale")
-inputs <- cbind(inc, prev, mort, pop)
-inputs$remmale <- 0
-inputs$remfemale <- 0
-inputs$prevdenom <- c(100,100,500,500,500,500,500,500,500,500,500,500,500,500,500,500,200,200,100,100) / 10 # total sample size 3910, generous for London
-
-## Stretch data from 5-year age groups to by year of age 
-outage <- 0:100  # assume inc/prev/mort same in each year within a five-year age group
-ind <- findInterval(outage, inputs$agegrp)
-inputs <- inputs[ind,]
-inputs$age <- outage
-inputs <- within(inputs, {
-    ningrp <- rep(table(agegrp), table(agegrp))
-    popmale <- round(popmale/ningrp) ## assume population uniform between years within age group.
-    popfemale <- round(popfemale/ningrp) ## assume population uniform between years within age group.
-    ndieddismale <- round(popmale * (1 - exp(-mortmale)))
-    ndieddisfemale <- round(popfemale * (1 - exp(-mortfemale)))
-    prevnmale <- round(prevdenom * prevmale)
-    prevnfemale <- round(prevdenom * prevfemale)
-})
-ihdlondon <- inputs
-
-outputs <- read.csv("../../belen/second/outputs.csv", skip=103, nrow=101)[,-1]
-names(outputs) <- c("age","inc","prev","rem","mort","incout","prevout","remout","cfout","durout","mortout","rrout","ageonset")
-ihdlondon$dismod_cf <- outputs$cfout
-
-use_data(ihdlondon, overwrite=TRUE)

### 2021-10-01 data-raw/trends.r

@@ -196,7 +196,7 @@ trendplot <- trends %>%
 ## Perhaps because Oxfordshire is lower risk than ave
 ## but can't find any contemporary data on variations in incidence
 
-pdf("../write/trend_data.pdf", height=4, width=6)
+pdf("~/work/chronic/write/trend_data.pdf", height=4, width=6)
 
 trendplot %>% 
   mutate(age_gender = sprintf("%s,%s", agegroup, gender)) %>%
@@ -243,6 +243,7 @@ tryr <- rbind(trbackm, trbackf, tryr)
 ## learning opportunity for purrr 
 
 tryr_sm <- tryr %>% 
+  arrange(age, gender, year) %>% 
   group_by(gender, year) %>%
   nest() %>%
   mutate(mod = map(data, ~loess(p2017 ~ age, 
@@ -351,8 +352,6 @@ ggplot(oxmicfint, aes(x=year, y=cf, group=age, col=age)) +
 ## No data for after 2010. Could assume same as last year in Smolina
 # Could also extrapolate
 
-## eh don't bother mimicing BHF format, just use end format 
-
 smyjoin <- smcfy %>% 
   filter(year %in% 1999:2010) %>%
   mutate("ox_agefrom" = case_when(agefrom==30 ~ 45, 
@@ -405,7 +404,7 @@ cftrends <- rbind(smyjoin %>% filter(year %in% 2002:2010),
   arrange(agefrom, gender, year)
 
 ## Plot of absolute CF, excluding interpolated bit 
-pdf("../write/trendcf_data.pdf", height=4, width=6)
+pdf("~/work/chronic/write/trendcf_data.pdf", height=4, width=6)
 
 cftrendplot <- cftrends %>%
   filter(!(year %in% 1999:2001)) %>%
@@ -566,12 +565,13 @@ saveRDS(cftrends_female, file="cftrends_female_noextrap.rds")
 
 ## Data used in package and in analysis 
 cftrends <- cftryr_sm_sens %>% 
-    select(gender, age, year, p2017=p2017_sm) %>%
+    arrange(age, gender, year) %>% 
+    select(age, gender, year, p2017=p2017_sm) %>%
     mutate(outcome = "Case fatality") 
 
 ihdtrends <- rbind(inctrends, cftrends) 
 
-use_data(ihdtrends) 
+usethis::use_data(ihdtrends) 
 
 
 

### 2021-10-01 doc/disbayes.R

@@ -1,47 +1,0 @@
-## -----------------------------------------------------------------------------
-knitr::opts_chunk$set(fig.width=7)
-
-## ----show=FALSE---------------------------------------------------------------
-library(disbayes)
-library(dplyr)
-ihdbristol <- ihdengland %>% filter(location=="Bristol", sex=="Male")
-ihdbristol %>% filter(between(age, 50, 55))
-
-## ----eval=TRUE----------------------------------------------------------------
-ihdbristol[ihdbristol$age %in% 50:55, ]
-
-## -----------------------------------------------------------------------------
-options(mc.cores = parallel::detectCores())
-
-## ---- eval=TRUE, cache=TRUE, warning=FALSE, message=FALSE---------------------
-dbres <- disbayes(dat = ihdbristol,
-                  age = "age",
-                 inc_num = "inc_num", 
-                 inc_denom = "inc_denom", 
-                 prev_num = "prev_num", 
-                 prev_denom = "prev_denom",
-                 mort_num = "mort_num",
-                 mort_denom = "mort_denom",
-                 eqage = 40,
-                 chains = 2,
-                 iter = 1000
-                 )
-
-## ----eval=TRUE----------------------------------------------------------------
-rstan::traceplot(dbres$fit, pars=paste0("cf[", 60:65, "]"))
-
-## ---- eval=TRUE---------------------------------------------------------------
-summ <- tidy(dbres)
-head(summ)
-
-## ----eval=TRUE----------------------------------------------------------------
-library(dplyr)
-summ %>% 
-  filter(var=="cf", between(age,60,65), model=="smoothed") %>%
-  select(age, `25%`, `50%`, `75%`)
-
-## ---- eval=TRUE, warning=FALSE------------------------------------------------
-library(ggplot2)
-plot(dbres) +  ylab("Case fatality") + xlab("Age") + 
-  coord_cartesian(xlim=c(40,100), ylim=c(0,0.5)) 
-

### 2021-10-01 metahit/cfest.r

@@ -1,97 +1,0 @@
-## simple cf estimator 
-
-## assuming 0.5 discrete time unit 
-## people getting disease in a year get it at halfway point 
-## incidence = p12
-## case fatality = p23 
-
-## * mort = (1-prev)*p12*p23 + prev*(p23 + (1-p23)*p23)
-## * mort = (1-prev)*p12*p23 + prev*(2*p23 - p23^2)
-## * mort = p23* ( (1-prev)*p12 + prev + (1 - p23))
-## quadratic equation for p23 with
-## a = -prev 
-## b = (1-prev)*p12 + 2*prev
-## c = -dm
-
-## But this needs prevalence to be known. 
-## For cancers that's fine I suppose as prev from register? 
-
-## TODO does even simpler version work ? 
-##  dm = (1-prev)*p12*p23 + prev*p23
-## p23 = dm / ((1-prev)*p12 + prev)
-## That does even better!! 
-
-source("constants.R")
-load("cfres.rda")
-
-gender <- "male"
-area <- "bristol"
-
-## Compare single-area spline model-based CF estimate to naive estimate 
-## for various diseases
-## Agrees well except in cases where there no information 
-## Naive estimate shows increasing trend for stmc - clearly shows bias in prior 
-## And for lvrc, tbalc naive estimate clearly noisy and unsafe.
-## Articulate what assumptions naive estimate is based on
-## discrete time, only move at halfway point
-## rates equal probs 
-
-## needs nondecreasing constraint.  constant would do but evidence from naive est of increasing rate.
-
-resall <- NULL
-for (disease in diseases){
-    dat <- readRDS(file.path(datadir,sprintf("%s_%s_%s.rds", area, gender, disease)))
-    a = dat$prev
-    b = (1 - dat$prev)*dat$inc + 2*dat$prev
-    c = -dat$mort
-    cf <- (-b + sqrt(b^2 - 4*a*c))/(2*a)
-    cf <- dat$mort / ((1 - dat$prev)*dat$inc + dat$prev)
-    cf
-    resmod <- cfres_nonhier[cfres_nonhier$area=="bristol" & cfres_nonhier$gender=="male" & cfres_nonhier$disease==disease,]
-    resmod$cf <- cf
-    resall <- rbind(resall, resmod)
-}
-
-ggplot(resall, aes(x=age, y=cf)) +
[…]

### 2021-10-01 metahit/hpc.sh

@@ -1,10 +1,0 @@
-#!/bin/bash
-#SBATCH -J metahit
-#SBATCH --time=12:00:00
-#SBATCH --cpus-per-task=2
-#SBATCH --array=[1]%8
-
-# module load r-4.0.2-gcc-5.4.0-xyx46xb
-module load r-3.6.1-gcc-5.4.0-zrytncq
-
-Rscript paper_analyses_gender.r

### 2021-10-01 metahit/paper_analyses_hier.r

@@ -1,39 +1,31 @@
-print(unlist(R.Version()))
-
 source("paper_analyses_header.r")
-
-options(mc.cores = 2)
+nchains <- 2
+options(mc.cores = nchains)
 
 task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
-
 runs_todo <- 1:nrow(hierrundf) 
 i <- runs_todo[task_id]
 
 mhi <- gbd %>%
-  filter(sex==hierrundf$gender[i], cause==hierrundf$disease[i]) %>%
+  filter(gender==hierrundf$gender[i], disease==hierrundf$disease[i]) %>%
   droplevels 
 
 db <- disbayes_hier(data=mhi,
-                    group = "location",
-                    group_init = "London",
-                    inc_num = "num_inc", inc_denom = "denom_inc",
-                    mort_num = "num_mort", mort_denom = "denom_mort",
-                    prev_num = "num_prev", prev_denom = "denom_prev",
-                    rem_num = "num_rem", rem_denom = "denom_rem",
-                    model = hierrundf$model[i], 
-                    remission = hierrundf$remission[i],
-                    age="ageyr", smooth_cf=TRUE,
-                    eqage=hierrundf$eqage[i], chains=2, iter=1000,
-                    nfold_int_guess = 5,
-                    nfold_int_upper =  50,
-                    nfold_slope_guess = 2,
-                    nfold_slope_upper =  20,
-                    sprior = 1,
-                    refresh = 1,  # TURN IT DOWN BELOW 100
-                    verbose = TRUE
-                    #, 
-                    #control=list(max_treedepth=20) # don't think adapt_delta adds anything, just slows
+                    group = "area",
+                    inc_num = "inc_num", inc_denom = "inc_denom",
+                    mort_num = "mort_num", mort_denom = "mort_denom",
+                    prev_num = "prev_num", prev_denom = "prev_denom",
+                    rem_num = if (hierrundf$remission[i]) "rem_num" else NULL, 
+                    rem_denom = if (hierrundf$remission[i]) "rem_denom" else NULL,
+                    cf_model = hierrundf$model[i],
+                    eqage = hierrundf$eqage[i],
+                    nfold_int_guess = 5, nfold_int_upper =  50,
+                    nfold_slope_guess = 2, nfold_slope_upper =  20,
[…]

### 2021-10-01 metahit/paper_analyses_nonhier.r

@@ -1,48 +1,36 @@
-#SBATCH --array=[48,51,77,94,95,117,118,120,121,124,125,129,130,133,134,136,137,140,141,145,146,149,150,152,153,156,157,161,162,165,166,168,169,172,173,177,178,181,182,184,185,188,189,193,194,197,198,200,201,204,205,209,210,213,214,216,217,220,221,222,223,225,226,229,230,232,233,236,237,241,242,245,246,248,249,252,253,257,258,261,262,264,265,268,269]%7
-
 source("paper_analyses_header.r")
 nchains <- 2
 options(mc.cores = nchains)
 
 task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
-runs_todo <- 1:nrow(rundf)  # 272, run in groups of 8. 
+runs_todo <- 1:nrow(rundf)  # 211, run in groups of 8. 
 i <- runs_todo[task_id]
 
-#exists <- if (file.exists(tempdir())) "exists" else "doesn`t exist"
-#cat(sprintf("Case %s tempdir %s %s\n", i, tempdir(), exists),
-#    file="metahit-hpc-log.txt", append=TRUE)
-
-# i <- 16
-
 mhi <- gbd %>%
-  filter(sex==rundf$gender[i], cause==rundf$disease[i], location==rundf$area[i])
+  filter(gender==rundf$gender[i], disease==rundf$disease[i], area==rundf$area[i])
+
 db <- disbayes(data=mhi,
-               inc_num = "num_inc", inc_denom = "denom_inc",
-               mort_num = "num_mort", mort_denom = "denom_mort",
-               prev_num = "num_prev", prev_denom = "denom_prev",
-               rem_num = "num_rem", rem_denom = "denom_rem",
-               remission = rundf$remission[i],
-               increasing_cf = rundf$increasing[i],
-               const_cf = FALSE,
-               age="ageyr", smooth_cf=TRUE,
-                eqage=rundf$eqage[i], 
-               chains=nchains, iter=1000)
+               inc_num = "inc_num", inc_denom = "inc_denom",
+               mort_num = "mort_num", mort_denom = "mort_denom",
+               prev_num = "prev_num", prev_denom = "prev_denom",
+               rem_num = if (rundf$remission[i]) "rem_num" else NULL, 
+               rem_denom = if (rundf$remission[i]) "rem_denom" else NULL,
[…]

### 2021-10-01 metahit/paper_analyses_pointests.r

@@ -1,0 +1,437 @@
+### Analyses for the paper which use optimisation to produce posterior modes
+### with credible intervals based on a normal approximation
+
+source("paper_analyses_header.r")
+
+### Non-hierarchical model
+### Empirical Bayes for smoothness pars for simplicity
+### Some cases don't estimate the mode well, as mode is low. 
+### Fix smoothness pars at 0.1 then
+brks <- c(7, 96, 154, 160)
+rundf$scf_eb <- TRUE; rundf$scf_eb[brks] <- FALSE
+rundf$sinc_eb <- TRUE; rundf$sinc_eb[c(96,97,83)] <- FALSE
+
+resopt_nonhier <- vector(nrow(rundf), mode="list")
+
+for (i in 1:nrow(rundf)){ 
+    mhi <- gbd %>%
+        filter(gender==rundf$gender[i], disease==rundf$disease[i], area==rundf$area[i])
+    db <- disbayes(data=mhi,
+                   #inc_num = "inc_num", inc_denom = "inc_denom",
+                   mort_num = "mort_num", mort_denom = "mort_denom",
+                   prev_num = "prev_num", prev_denom = "prev_denom",
+                   rem_num = if (rundf$remission[i]) "rem_num" else NULL, 
+                   rem_denom = if (rundf$remission[i]) "rem_denom" else NULL,
+                   cf_model = if (rundf$increasing[i]) "increasing" else "smooth",
+                   eqage=rundf$eqage[i], 
+                   scf_fixed =  if (rundf$scf_eb[i]) TRUE else 0.1,
+                   sinc_fixed = if (rundf$sinc_eb[i]) TRUE else 0.1,
+                   rem_prior = c(1.1, 1), sprior = c(1,10),
+                   hessian=TRUE, draws=1000,
+                   seed = i) 
+    resopt_nonhier[[i]] <- tidy(db) %>%
+        mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
+    cat(sprintf("Completed case %s out of %s\n",i,nrow(rundf)))
+}
+
+
+if (0){
+    any(eigen(-db$fit$hessian)$values < 0)
+    diag(-db$fit$hessian)
+    sqrt(diag(solve(-db$fit$hessian)))
+}
+
+resopt_nonhier <- do.call("rbind", resopt_nonhier)
+saveRDS(resopt_nonhier, file="resopt_nonhier.rds")
[…]

### 2021-10-01 metahit/read_model_results.r

@@ -1,0 +1,95 @@
+### Read, tidy and combine all results obtained from running MCMC simulations (parallelised for different cases) on a HPC cluster
+
+reslist <- loolist <- vector(nrow(rundf), mode="list")
+for (i in 1:nrow(rundf)){
+  fname <- sprintf("results_nonhier/res%s.rds",i)
+  if (file.exists(fname)) {
+    resi <- readRDS(fname)
+    reslist[[i]] <- resi$res
+    loolist[[i]] <- resi$loo
+  }
+}
+dput(as.numeric(which(sapply(reslist, is.null)))) # check if any runs failed 
+
+resnh <- do.call("rbind", reslist) %>%
+  mutate(model="Independent areas")
+loonh <- do.call("rbind", loolist) %>%
+  mutate(model="Independent areas")
+
+resi <- readRDS("results_nonhier/res_eng_utrc_female.rds")
+res_utrc <- resi$res; loo_utrc <- resi$loo
+resi <- readRDS("results_nonhier/res_eng_stmc_male.rds")
+res_stmc_m <- resi$res; loo_stmc_m <- resi$loo
+resi <- readRDS("results_nonhier/res_eng_stmc_female.rds")
+res_stmc_f <- resi$res; loo_stmc_f <- resi$loo
+resnat <- rbind(res_utrc, res_stmc_m, res_stmc_f) %>%
+  mutate(model="National")
+resnh <- resnh %>% 
+  dplyr::filter(!(disease %in% c("Stomach cancer", "Uterine cancer"))) %>%
+  full_join(resnat)
+
+loonat <- rbind(loo_utrc, loo_stmc_m, loo_stmc_f) %>%
+  mutate(model="National")
+loonh <- loonh %>% 
+  dplyr::filter(!(disease %in% c("Stomach cancer", "Uterine cancer"))) %>%
+  full_join(loonat)
+
+#res %>% filter(disease=="Uterine cancer", var=="cf", between(age, 58, 70))
+#res_utrc %>% filter(disease=="Uterine cancer", var=="cf", between(age, 90, 100))
+## Smoothed hasn't worked 
+
+reslist <- loolist <- vector(nrow(hierrundf), mode="list")
+for (i in setdiff(1:nrow(hierrundf), c())){
+  resi <- readRDS(sprintf("results_hier/res%s.rds",i))
+  reslist[[i]] <- resi[[1]]
+  loolist[[i]] <- resi[[2]]
+}
[…]

### 2021-10-01 tests/testthat/test.R

@@ -1,34 +1,33 @@
 library(dplyr)
 library(tidyr)
 
-ihdbristol <- ihdengland %>% filter(location=="Bristol", sex=="Male")
+ihdbristol <- ihdengland %>% 
+  dplyr::filter(area=="Bristol", gender=="Male") %>%
+  mutate(mort_prob = mort_num/mort_denom,
+       mort_lower = plogis(qlogis(mort_prob)-1),
+       mort_upper = plogis(qlogis(mort_prob)+1))
 
-test_that("standard disbayes model",{
+test_that("standard disbayes model, MCMC",{
   dbres <- disbayes(dat = ihdbristol,
                     inc_num = "inc_num", inc_denom = "inc_denom",
                     prev_num = "prev_num", prev_denom = "prev_denom",
                     mort_num = "mort_num", mort_denom = "mort_denom",
-                    eqage = 40, loo=FALSE, algorithm="Fixed_param", chains=1, iter=100, iter_train = 100)
+                    eqage = 40, loo=FALSE, method="mcmc", algorithm="Fixed_param", 
+                    chains=1, iter=100)
   expect_s3_class(dbres, "disbayes")
 })
 
 test_that("data supplied as estimate and denominator",{
-  ihdbristol <- ihdbristol %>%
-    mutate(mort_prob = mort_num/mort_denom)
   dbres <- disbayes(dat = ihdbristol,
                     inc_num = "inc_num", inc_denom = "inc_denom",
                     prev_num = "prev_num", prev_denom = "prev_denom",
                     mort = "mort_prob", mort_denom = "mort_denom",
-                    eqage = 40, loo=FALSE, algorithm="Fixed_param", chains=1, iter=100, iter_train = 100)
+                    eqage = 40, loo=FALSE, method="mcmc", algorithm="Fixed_param", 
+                    chains=1, iter=100)
   expect_s3_class(dbres, "disbayes")
 })
 
 test_that("data supplied as estimate and credible limits",{
-  ihdbristol <- ihdbristol %>%
-    mutate(mort_prob = mort_num/mort_denom,
-           mort_lower = plogis(qlogis(mort_prob)-1),
-           mort_upper = plogis(qlogis(mort_prob)+1))
-  
   x <- with(ihdbristol, ci2num(mort_prob, mort_lower, mort_upper, denom0=1400))
   expect_equal(x$num[1], 0)
   expect_equal(x$denom[1], 1400)
[…]

### 2021-10-01 tests/testthat/test_hier_gender.R

@@ -1,25 +1,27 @@
 library(dplyr)
 library(disbayes)
 
-db <- disbayes_hier(ihdengland, age="age", group="location", gender="sex",
-                           inc_num = "inc_num", inc_denom = "inc_denom",
-                           prev_num = "prev_num", prev_denom = "prev_denom",
-                           mort_num = "mort_num", mort_denom = "mort_denom",
-                           iter=10, iter_train=10, chains=1, loo=FALSE, algorithm="Fixed_param")
-expect_s3_class(db, "disbayes_hier_gender")
+db <- disbayes_hier(ihdengland, age="age", group="area", gender="gender",
+                    inc_num = "inc_num", inc_denom = "inc_denom",
+                    prev_num = "prev_num", prev_denom = "prev_denom",
+                    mort_num = "mort_num", mort_denom = "mort_denom",
+                    iter=10, chains=1, loo=FALSE, algorithm="Fixed_param")
+expect_s3_class(db, "disbayes_hier")
 
 test_that("errors in specifying gender model", {
-  expect_error(disbayes_hier(ihdengland, age="age", group="location",  
-                                    inc_num = "inc_num", inc_denom = "inc_denom",
-                                    prev_num = "prev_num", prev_denom = "prev_denom",
-                                    mort_num = "mort_num", mort_denom = "mort_denom"),
-               "`gender` variable not supplied")
+  expect_error(disbayes_hier(ihdengland, age="age", group="area",  
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                             prev_num = "prev_num", prev_denom = "prev_denom",
+                             mort_num = "mort_num", mort_denom = "mort_denom"),
+               "rows in data")
   
-  expect_error(disbayes_hier(ihdengland, age="age", group="location", gender="sex",
-                                    inc_num = "inc_num", inc_denom = "inc_denom",
-                                    prev_num = "prev_num", prev_denom = "prev_denom",
-                                    mort_num = "mort_num", mort_denom = "mort_denom"),
-               "rows in data, expected")
+  expect_error(disbayes_hier(ihdengland %>% 
+                               dplyr::filter(gender=="Female"), 
+                             age="age", group="area", gender="gender",
+                             inc_num = "inc_num", inc_denom = "inc_denom",
+                             prev_num = "prev_num", prev_denom = "prev_denom",
+                             mort_num = "mort_num", mort_denom = "mort_denom"),
[…]

### 2021-10-02 metahit/paper_analyses.Rmd
Clean up

@@
 
-Fuck me the Rhats are fucked 
-Theyve basically all broken . 
-Must be some unused parameter or wev that has been left in to break it 
-Use the optim models to debug it 
-looks like the incidence related parameters
-beta_inc not in tidy but can't tell
-
 ```{r}
@@
 resnc <- resg %>% filter(Rhat > 1.01) 
-
-resnc <- resg %>% filter(Rhat > 1.1, disease=="Ischemic heart disease") 
-resg %>% filter(grepl("beta_inc",var))
-
 table(resnc$disease)

### 2021-10-02 metahit/paper_analyses.Rmd
Fixed hyperparameters / empirical Bayes rework

@@
 
-## Show whether gender effect is common between areas 
-## todo pick diseases to illustrate.  IHD, LC biggesc contrast 
-## how to show for all ages.
+## Get curves of m/f CF risk ratios by age and area, from the hier model with 
+## different gender effect for each area 
+reshi <- resh %>%  
+  mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
+  filter(! disease %in% c("Breast cancer", "Stomach cancer", "Uterine cancer"), 
+         area %in% cityregions, 
+         var=="cf") %>%
+  mutate(est = log(`50%`), se=(log(`97.5%`)-log(`2.5%`))/4) 
+reshie <- reshi %>%
+  select(age, gender, area, disease, est) %>% 
+  pivot_wider(names_from=gender, values_from=est) %>%
+  mutate(lhrmale = Male - Female) %>%
+  select(age, area, disease, lhrmale)
+reshis <- reshi %>%
+  select(age, gender, area, disease, se) %>% 
+  pivot_wider(names_from=gender, values_from=se) %>%
+  mutate(sediff = sqrt(Male^2 + Female^2)) %>%
+  select(age, area, disease, sediff) %>%
+  left_join(reshie, by=c("age","area","disease")) %>%
+  mutate(est = exp(lhrmale), 
+         lower=exp(lhrmale - 2*sediff),
+         upper=exp(lhrmale + 2*sediff))
 
-dat <- resall %>%
-  filter(var=="cf",
-         model %in% c("Hierarchical","Hierarchical joint gender"),
-         disease %in% c("Ischemic heart disease", "Stroke", "Lung cancer", "Colon and rectum cancer", "Dementia", "COPD")) %>%
-  filter(disease %in% c("Ischemic heart disease", "Lung cancer")) %>%
-  mutate(model = fct_recode(model, 
-                            "Different between areas" = "Hierarchical",
-                            "Common effect" = "Hierarchical joint gender"))
-example_age <- 70
+## Superimpose the equivalent common curve from the common effect model
+resgi <- resg %>%  
+  filter(! disease %in% c("Breast cancer", "Stomach cancer", "Uterine cancer"), 
+         area %in% cityregions, 
+         var=="cf") %>%
+  mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
+  mutate(est = log(`50%`), se=(log(`97.5%`)-log(`2.5%`))/4) 
+resgie <- resgi %>%
+  select(age, gender, area, disease, est) %>% 
+  pivot_wider(names_from=gender, values_from=est) %>%
+  mutate(lhrmale = Male - Female) %>%
+  select(age, area, disease, lhrmale)
+resgis <- resgi %>%
+  select(age, gender, area, disease, se) %>% 
+  pivot_wider(names_from=gender, values_from=se) %>%
+  mutate(sediff = sqrt(Male^2 + Female^2)) %>%
+  select(age, area, disease, sediff) %>%
+  left_join(resgie, by=c("age","area","disease")) %>%
+  mutate(est = exp(lhrmale), 
+         lower=exp(lhrmale - 2*sediff),
+         upper=exp(lhrmale + 2*sediff))
 
-#DEBUGME SHOW RAW MORT DATA FOR LC FOR AREAS, SEE IF GENDER DIFFERENCE
-gbd %>% filter(ageyr==70, 
-               disease=="Tracheal, bronchus, and lung cancer", 
-               area %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
-  mutate(p_mort = mort_num/mort_denom, p_prev=prev_num/prev_denom, mpratio=p_mort/p_prev) %>%
-  select(sex, area, mort_num, mort_denom, p_mort, p_prev, mpratio)
-## YES: MORT/PREV AGREES WITH CF
+## Put this in the paper 
+## Hazard ratio for men compared to women for case fatality 
+## This is good evidence that the general pattern is common among areas 
+## TODO Confirm this or otherwise by LOO CV
 
-i <- 3
-resi <- readRDS(sprintf("results_hier_gender/res%s.rds",i))
-resi <- readRDS(sprintf("results_hier/res%s.rds",3))
-resi <- readRDS(sprintf("results_hier/res%s.rds",10))
-
-resi[[1]] %>% 
-  filter(var=="cf", between(age, 68, 72), area==1)
-resi[[1]] %>% 
-  filter(var=="cf", between(age, 68, 72), area==1)
-
-gbd %>% filter(ageyr==70, 
-               disease=="Tracheal, bronchus, and lung cancer", 
-               area %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
-  group_by(area) %>%
-  mutate(mort_num=sum(mort_num), mort_denom=sum(mort_denom), 
-         prev_num=sum(prev_num), prev_denom=sum(prev_denom)) %>%
-  mutate(p_mort = mort_num/mort_denom, p_prev=prev_num/prev_denom, mpratio=p_mort/p_prev) %>%
-  select(sex, area, mort_num, mort_denom, p_mort, p_prev, mpratio)
-
-pdf("../write/gender.pdf", width=6, height=4) # NOTE one for IHD, one for lung cancer
-
-dat %>% filter(age==example_age) %>%
-ggplot(aes(x=`50%`, y=area, lty=model, col=gender)) +
-  #    coord_cartesian(xlim=c(0.001, 0.2)) +
-  facet_wrap(~disease, nrow=1, scales="free_x") + 
-  geom_point(position=ggstance::position_dodgev(height=0.6), size=1.5) +
-  geom_errorbarh(aes(xmin=`2.5%`, xmax=`97.5%`),
-                 position=ggstance::position_dodgev(height=0.6),
-                 lwd = 0.7) +
-  xlab(sprintf("Case fatality for age %s", example_age) ) + ylab("") + 
-  scale_linetype_manual(values = c("Common effect" = "solid",
-                                   "Different between areas" = "dashed"
-                                   )) +
-  labs(col="", lty="Gender effect") +
-  theme_bw() + 
-  theme(legend.text = element_text(size=7)) +  
-  guides(linetype = guide_legend(reverse = TRUE))
-
+pdf("../../write/gender_hr.pdf")
+ggplot(reshis %>% filter(age > 60), 
+       aes(x=age, y=est, group=area, col=area)) + 
+  geom_ribbon(aes(ymin=lower, ymax=upper), fill="gray", col=NA, alpha=0.2) + 
+  geom_hline(yintercept = 1) +
+  geom_line(lwd=0.8) +
+  geom_line(data=resgis %>% filter(age>60), col="black", lty=2) +
+  facet_wrap(~disease) +
+  labs(col="Area") +
+  coord_cartesian(ylim=c(0,3)) + 
+  ylab("Case fatality rate ratio (men/women)") + xlab("Age (years)")
 dev.off()
@@
 
+## FIXME rerun hier with bugfix in saving loo output .  was triggered when we changed rundf
+## to exclude (stmc and?) utrc
+
+looall %>% filter(var=="mort", area=="Bristol", gender=="Male", disease=="Lung cancer", 
+                  model=="Hierarchical", age==1)
+
 # elpd : high is good. positive elpd diff favours additive gender model 
@@
   select(var, area, age, gender, disease, model, elpd_loo) %>%
-  pivot_wider(names_from=model, values_from=elpd_loo) %>%
+  pivot_wider(names_from=model, values_from=elpd_loo, values_fn=length)
+
+%>%
   mutate(elpd_diff = `Hierarchical joint gender` - `Hierarchical`) %>%
@@
 ```
-
-# Time trends
-
-Damn, Absolute incidence estimates are way off both the model results, 
-but closest to the non trend one where they match up to age 70.
-
-why.  We are adjusting them to be consistent with prevalence and mortality data. 
-But we want to give more weight somehow to current incidence data.  Why tho, if we also believe the current 
-prev and mort data? are we trying to make them all fit into a bad model?
-
-Right the fitted values for year 2010 and earlier match up with current data.  There's a big jump around 2010
-
-But then they will carry on changing through time in the future if we use them in a model
-
-```{r}
-
-pdf("../write/trend_res.pdf")
-
-datt <- gbd %>% 
-  filter(area=="Leeds", disease=="Ischemic heart disease") %>%
-  mutate(inc_prob = inc_num / inc_denom, 
-         inc_rate = -log(1-inc_prob), 
-         var = "Incidence", 
-         gender = sex)
-
-resoy <- resall %>% filter(model=="Time trends", 
-                            year == 2017,
-                            area=="Leeds", 
-                            disease=="Ischemic heart disease",
-                            var %in% c("inc")) %>% mutate(var="Incidence")
-
-rest <- resall %>% filter((model == "Independent areas") | (model=="Time trends" & year==2017),
-                          area=="Leeds", 
-                          disease=="Ischemic heart disease",
-                          var %in% c("inc","cf")
-#                          gender=="Male"
-) %>% full_join(resoy)
-
-resopt <- readRDS(file="resopt_trend.rds") %>% 
-  filter(disease=="Ischemic heart disease",area=="Leeds",
-         var %in% c("inc","cf"), year==99) %>%
-  mutate(var = fct_recode(var, "Incidence"="inc", "Case fatality"="cf"))
-
-rest %>% 
-  filter(between(age, 50, 90)) %>% 
-  mutate(var = fct_recode(var, "Incidence"="inc", "Case fatality"="cf"),
-         model = fct_recode(model, "No time trends" = "Independent areas")) %>%
-ggplot(aes(x=age, y=`50%`, col=model, fill=model, 
-           group=interaction(gender, model))) +
-  coord_cartesian(ylim=c(0.0,0.04)) +
-  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), alpha=0.2) +
-  facet_wrap(vars(gender, var), nrow=2, scales="free_y") +
-  geom_line() + 
-  xlab("Age") + ylab("") + 
-  labs(col="", fill="") + 
-  geom_line(data=datt, aes(x=ageyr, y=inc_rate), inherit.aes = FALSE) +
-  geom_line(data=resopt, aes(x=age, y=par), inherit.aes=FALSE, col="blue")
-
-## Point estimates are accurate, so we can use them to diagnose the bias
-
-#  geom_line(data=resoy, aes(x=age, y=`50%`, col=factor(year), group=year), lty=2, inherit.aes = FALSE)
-dev.off()
-
-## Experimental sensitivity analyses to different trends
-## what was it? 2003-2010 continues. saved in restrs 
-## where's the conclusion from this?  estimates vary a lot
-
-rest_saved <- rest %>% 
-  filter(between(age, 50, 90), model=="Time trends", 
-         gender=="Male", var %in% c("inc","cf")) 
-res_saved <- resnh %>% filter(between(age, 50, 90), model=="Independent areas", disease=="Ischemic heart disease", area=="Leeds", gender=="Male", var %in% c("inc","cf"))
-
-restrs <- restrs %>% 
-  mutate(trendin = ifelse(inctrend & cftrend, "both", 
-  ifelse(inctrend, "inc", ifelse(cftrend, "cf", "neither")))) %>%
-  mutate(trendin = ifelse(trsens=="zero", "neither", trendin))
-
-dat <- restrs %>% 
-  filter(var %in% c("inc","cf"), model=="smoothed", trsens %in% c("hi", "zero"))
-dat$age <- rep(1:100, length.out=nrow(dat))
-dat %>%
-  filter(age < 80) %>%
-ggplot(aes(x=age, y=`50%`, col=trendin)) + 
-  geom_line(lwd=2) + 
-  facet_grid(cols=vars(var)) + 
-  geom_line(data=datt %>% filter(gender=="Male") %>% mutate(var=fct_recode(var,"inc"="Incidence")), 
-            aes(x=ageyr, y=inc_rate), inherit.aes = FALSE)
-
-+ 
-  geom_line(data=rest_saved, col="black") + 
-  geom_line(data=res_saved, col="blue")
-
-restrs %>% 
-  filter(var %in% c("inc","cf")) %>%
-  mutate(age = rep(1:100, length.out=n())) %>%
-  filter(model=="smoothed", age==60, var=="cf", cftrend==TRUE, inctrend==TRUE)
-
-```
-
-Right this is saying that it doesn't make much difference to CF, but big diff to inc as in Scar
-were the original plots wrong.   OK fine just axis limit distortion.
-
-So the only problem left is disagreement with current empirical incidence 
-dismod2 types would say the model is "forcing consistency" 
-evidence of conflict between current inc and current CF  
-Hm is this evidence that the gbd data are consistent with themselves under a model that ignores trends.
-
-Hey wot it does match.   Wot about original independent-areas results
-
-cf - higher with inc or both, lower with cf or neither 
-inc - higher with cf only, then neither, then both, then lowest with inc.
-
-So dose response is OK 
-but they interact in unpredictable ways 
-and need to fix the 5 year tihng 
-and reconcile the modest changes here with the big changes from the real data
-
-Compare real with artificial trend data.  Where does the difference come from? 
-Looks reasonably representative of CF to me.  Age 70 value is generous 
-Incidence: trend data is more noisy.  Pick an even higher value that's on the high end of the data.
-
-```{r}
-ism <- inctrends_hi %>% as.data.frame() %>% 
-  mutate(age = 1:100) %>%
-  pivot_longer(names_to = "year", values_to = "p2017", cols = 1:100) %>%
-  mutate(year = as.numeric(year)) %>% 
-  filter(age %in% c(40,60,80,100), year > 1970) %>% 
-  mutate(age = factor(age), model = "sens")
-
-ihdtrends %>% 
-  filter(age %in% c(40,60,80,100), year > 1970, gender=="Male", outcome=="Incidence") %>% 
-  mutate(age = factor(age), model = "data") %>%
-  full_join(ism) %>%
-  ggplot(aes(x=year, y=p2017, col=age, lty=model)) + geom_line()
-
-```
-
-Well that's odd.
-
-With inc trends on. 
-There's much less of an effect than there was with the real data. 
-Consistent dose response effect 
-CF trend makes no difference to current CF ests , but affects current inc ests. Wot?
-
-With cf trends on
-Only an effect on cf if there is inc trends.
-Consistent dose response when there's an effect
-Inc est: both inc and cf trends make a difference.  moderate differences
-
-Is this a bug in the sens analysis setup??   cf trends are not affecting cf
-And why are there still those stupid 5 year artefacts for incidence?  didn't i transfer the new data?
-
-OK think og it a new way 
-TREND IN: 
-inc but not cf : 
-cf but not inc : reduced inc 
-both inc and cf:     reduced inc
-
-inc highest: 
-inctrend false, cftrend true
-inctrend true, cftrend true 
-cftrend false.   
-inc lowest
-
-cf highest with 
-inctrend true, cftrend false or true 
-inctrend false

### 2021-10-02 R/disbayes.R

@@ -1,14 +1,13 @@
 ##' Bayesian estimation of chronic disease epidemiology from incomplete data
 ##'
-##' Estimates a three-state disease model given data on incidence, prevalence
-##' and mortality.
-##'
+##' Estimates a three-state disease model from incomplete data. 
 ##' It is designed to estimate case fatality and incidence, given
 ##' data on mortality and at least one of incidence and prevalence.
-##' Remission may also be included.
+##' Remission may also be included in the data and modelled.
 ##'
 ##'
 ##' @rdname disbayes
+##' @md
 ##'
 ##' @param data  Data frame containing some of the variables below.  The
 ##'   variables below are provided as character strings naming columns in this
@@ -149,10 +148,27 @@
 ##'   appropriately to the data, but higher values give stronger smoothing, or
 ##'   lower values give weaker smoothing,  if required.
 ##'
-##' @param scf_fixed  If this is set to a number, then the parameter determining the smoothness of case fatality with age (i.e. the standard deviation of the spline basis coefficients) is fixed at this number.  If \code{scf_fixed=TRUE}, then an "empirical Bayes" method is used, where the posterior mode of this parameter is determined (using the prior defined by \code{sprior}), and the model is fitted again with the parameter fixed at this mode.  If \code{scf_fixed=NULL} (the default) then a "full Bayes" method is used where the full posterior of this parameter is estimated.   NOT IMPLEMENTED YET FOR HIERARCHICAL MODELS 
+##' @param hp_fixed A list with one named element for each hyperparameter
+##' to be fixed.  The value should be either 
+##' 
+##' * a number (to fix the hyperparameter at this number) 
+##' 
+##' * \code{TRUE} (to fix the hyperparameter at the posterior mode from a training run
+##' where it is not fixed)
+##' 
+##' If the element is either \code{NULL}, \code{FALSE}, or omitted from the list, 
+##' then the hyperparameter is given a prior and estimated as part of the Bayesian model.  
+##' 
+##' The hyperparameters that can be fixed are 
+##' 
+##' * \code{scf} Smoothness parameter for the spline relating case fatality to age.
+##' 
[…]

### 2021-10-02 R/disbayes_hier.R

@@ -1,57 +1,114 @@
-##' Bayesian estimation of chronic disease epidemiology from incomplete data - hierarchical model for case fatalities 
-##' This is computationally intensive.  
+##' Bayesian estimation of chronic disease epidemiology from incomplete data -
+##' hierarchical model for case fatalities.  
 ##' 
-##' @param group Variable in the data representing the area (or other grouping factor). 
-##' 
-##' @param gender If \code{NULL} (the default) then the data are one homogenous gender, and there should be one row per year of age.  Otherwise, set \code{gender} to a character string naming the variable in the data representing gender (or other binary grouping factor).  Gender will then treated as a fixed additive effect, so the linear effect of gender on log case fatality is the same in each area.  The data should have one row per year of age and gender.  
+##' This is much more computationally 
+##' intensive than the basic model in \code{\link{disbayes}}.
 ##'
-##' @param cf_model The following alternative models for case fatality are supported:
+##' @param group Variable in the data representing the area (or other grouping
+##'   factor).
 ##'
-##' \code{"default"} (the default). Random intercepts and slopes, and no further restriction. 
+##' @param gender If \code{NULL} (the default) then the data are one homogenous
+##'   gender, and there should be one row per year of age.  Otherwise, set
+##'   \code{gender} to a character string naming the variable in the data
+##'   representing gender (or other binary grouping factor).  Gender will then
+##'   treated as a fixed additive effect, so the linear effect of gender on log
+##'   case fatality is the same in each area.  The data should have one row per
+##'   year of age and gender.
 ##'
-##' \code{"interceptonly"}.  Random intercepts, but common slopes.
+##' @param cf_model The following alternative models for case fatality are
+##'   supported:
 ##'
-##' \code{"increasing"}. Case fatality is assumed to be an increasing function of age (note it is constant below \code{"eqage"} in all models) with a common slope for all groups. 
+##'   \code{"default"} (the default). Random intercepts and slopes, and no
+##'   further restriction.
 ##'
-##' \code{"const"} Case fatality is assumed to be constant with age, for all ages.
+##'   \code{"interceptonly"}.  Random intercepts, but common slopes.
 ##'
-##' In all models, case fatality is a smooth function of age, and incidence rates are is estimated independently for each age.
[…]

### 2021-10-02 R/eb.R

@@ -1,23 +1,61 @@
 ## Empirical Bayes method
 ## Call disbayes to estimate posterior modes of hyperparameters
 ## Return the modes so that they can subsequently be fixed to fit the desired model 
-## args: names of the disbayes arguments indicating hyperparameters
 
-## TODO can we abstract this, it's used for the RE SD in hier.  then can do it in hier for smoothness too
-## arguments: call list, db_argnames, stan_parnames, stan_fixedparnames
-## TODO problems if outer call uses mcmc, and we try to recurse disbayes(method="opt", chains=...)
-## do we need to pass opt args to mode-finding function?  don't think so. if so could just
-## remove all args that are not formals of disbayes. 
-
-## If this recursion causes problems, could just write out all the arguments for safety
-
-eb_find_modes <- function(dbcall, dbfn=disbayes, args){ 
+eb_find_modes <- function(dbcall, dbfn=disbayes, hp_fixed_arg){ 
     dbcall$method <- "opt"
     dbcall$hessian <- FALSE
     dbcall$draws <- 0
-    for (arg_ind in args)
-        dbcall[[arg_ind]] <- NULL 
+    dbcall$hp_fixed <- hp_fixed_arg
     dbcall[[1]] <- NULL
+    ## Don't allow extra args to be passed through (e.g. to customise the sampler)
+    fargs <- names(dbcall)[names(dbcall) %in% names(formals(dbfn))]
+    dbcall <- dbcall[fargs]
     ebres <- do.call(dbfn, dbcall)
     ebres$fit$par
 }
+
+expand_hpfixed <- function(hp, hp_fixed){
+  hpfnew <- vector(nrow(hp), mode="list")
+  names(hpfnew) <- hp$pars
+  for (i in seq_along(hp$pars)) {
+    hpfnew[[i]] <- if (is.null(hp_fixed[[hp$pars[i]]])) FALSE else hp_fixed[[hp$pars[i]]]
+  }
+  hpfnew
+}
+
+.disbayes_hier_hp <- data.frame(
+  pars      = c("scf","sinc","scfmale","sd_int","sd_slope"),
+  row.names = c("scf","sinc","scfmale","sd_int","sd_slope"),
[…]

### 2021-10-02 R/tidy.R

@@ -178,9 +178,9 @@ attr(.disbayes_trend_vars, "order") <- c("age", "year", "bias", "state", "term")
         indnames = c("area"),
         varnames = c("lcfbase")),
     term = list(indnames="term",
-                varnames = c("bmale")),
-    const = c("lambda_cf","lambda_cf_male","lambda_inc","lambda_inc_male",
-              "mean_inter","mean_slope","sd_inter","sd_slope")
+                varnames = c("bmale", "sd_inter", "mean_slope", "sd_slope",
+                             "lambda_cf","lambda_cf_male","lambda_inc")),
+    const = c("mean_inter")
 )
 attr(.disbayes_hier_vars, "numerics") <- c("age", "state", "term")
 attr(.disbayes_hier_vars, "factors") <- data.frame(vars = c("area","gender"),

### 2021-10-02 metahit/paper_analyses_gender.r

@@ -8,14 +8,24 @@ runs_todo <- 1:nrow(hierrungdf)
 i <- runs_todo[task_id]
 
 if (0){
+  setwd("metahit")
   source("paper_analyses_header.r")
   i <- 1
   nchains <- 1
+  foo <- tidy(db)
+  db$modes 
+  foo %>% filter(grepl("lam",var))
 }
 
 mhi <- gbd %>%
     filter(disease==hierrungdf$disease[i]) %>%
     droplevels
+
+## TODO find posterior mode of lambda_cf_male so we can fix it at this plausible value in MCMC
+## OK mode is 1.3 here.  So how can we use it in the real run. cd just set TRUE
+## did we try this before 
+## TODO use new EB code from disbayes_hier in disbayes. any sharing ? needs to handle subset of 5 pars cleanly  
+## TODO test all EB cases, put in testthat
 
 db <- disbayes_hier(data=mhi,
                     group = "area",
@@ -28,14 +38,13 @@ db <- disbayes_hier(data=mhi,
                     cf_model = hierrungdf$model[i],
                     inc_model = "indep",
                     inc_prior = c(1.1, 1), 
-                    scf_fixed = 2.65, 
-                    scfmale_fixed = 2.65,
+                    hp_fixed = list(scf = 2.65, sint = 2.65),
                     eqage = hierrundf$eqage[i], 
                     nfold_int_guess = 5, nfold_int_upper =  50,
                     nfold_slope_guess = 2, nfold_slope_upper =  20,
-                    #method="opt", hessian=TRUE, draws=1000, iter=10000, verbose=TRUE
-                    method="mcmc", refresh = 1, chains=nchains, iter=1000,
-                    stan_control=list(max_treedepth=15)
+                    method="opt", hessian=TRUE, draws=1000, iter=10000, verbose=TRUE
+                    #method="mcmc", refresh = 1, chains=nchains, iter=1000,
+                    #stan_control=list(max_treedepth=15)
 )
 
 

### 2021-10-02 metahit/paper_analyses_hier.r

@@ -7,6 +7,8 @@ runs_todo <- 1:nrow(hierrundf)
 i <- runs_todo[task_id]
 
 if (0) { 
+  setwd("metahit")
+  source("paper_analyses_header.r")
   i <- 1 
   nchains <-1
 }
@@ -25,7 +27,7 @@ db <- disbayes_hier(data=mhi,
                     cf_model = hierrundf$model[i],
                     inc_model = "indep",
                     inc_prior = c(1.1, 1), 
-                    scf_fixed = 2.65, 
+                    hp_fixed = list(scf=2.65), 
                     eqage = hierrundf$eqage[i],
                     nfold_int_guess = 5, nfold_int_upper =  50,
                     nfold_slope_guess = 2, nfold_slope_upper =  20,
@@ -49,7 +51,7 @@ res <- tidy(db) %>%
   mutate(gender=hierrundf$gender[i], disease=hierrundf$disease[i])
 
 loo <- looi_disbayes_hier(db) %>%
-    mutate(gender=rundf$gender[i], disease=rundf$disease[i])
+    mutate(gender=hierrundf$gender[i], disease=hierrundf$disease[i])
 
 saveRDS(list(res=res,loo=loo), file= paste0("results_hier/res", i, ".rds"))
 cat(sprintf("Fitted model for case %s\n", i), file="metahit-hpc-log.txt", append=TRUE)

### 2021-10-07 metahit/paper_analyses.Rmd
Updated analysis scripts

@@
 
-foo <- function(x){
-  cat(deparse(substitute(x)))
-}
+source("paper_analyses_header.r")
+ymax_mort <- enframe(c("Ischemic heart disease"=0.2,"Stroke"=0.1,"Dementia"=0.2,
+                       "COPD"=0.1, "Breast cancer"=0.03, "Lung cancer"=0.02,
+                       "Colon and rectum cancer"=0.02, "Stomach cancer"=0.01, 
+                       "Uterine cancer"=0.002), name="disease", value="ymax_mort")
 
+ymax_inc <- enframe(c("Ischemic heart disease"=0.1, "Stroke"=0.1,"Dementia"=0.1,
+                      "COPD"=0.1, "Breast cancer"=0.03, "Lung cancer"=0.03,
+                      "Colon and rectum cancer"=0.03, "Stomach cancer"=0.01,
+                      "Uterine cancer"=0.002), name="disease", value="ymax_inc")
+
+gbdplot <- gbdplot %>%
+    left_join(ymax_mort, by="disease") %>%
+    left_join(ymax_inc, by="disease") %>%
+    mutate(mort_upper = pmin(mort_upper, ymax_mort),
+           inc_upper = pmin(inc_upper, ymax_inc),
+           disease = factor(disease, levels=disease_order))
 ht <- 3
@@
 
-## Convergence checks.   All OK with 5000 samples and the occasional model simplification
+## MCMC convergence checks.   All OK with 5000 samples and the occasional model simplification
 resnc <- resnh %>% filter(Rhat > 1.05)
@@
 
-## Compare modal approx with posterior 
-## Page of plots for a single area and gender.  Disease variation more important than area/gender variation  
+## Compare modal approximation with posterior 
+## Page of plots for a single area and gender.
+## Disease variation more important than area/gender variation  
 ## stmc, utrc use national point estimates
-## What are the current settings used for these?
 
-resall %>% 
-  filter(var=="cf", disease=="Dementia", gender=="Female", 
-         area=="Bristol", between(age, 70, 90), model=="Independent areas")
-
-resoptunc <- readRDS("resoptunc_nonhier.rds") %>% 
+resopt <- readRDS("resopt_nonhier.rds") %>% 
   filter(!disease %in% c("Stomach cancer","Uterine cancer"))
 resoptnat <- readRDS("resopt_nat.rds")
-resoptunccf <- resoptunc %>%
+resoptcf <- resopt %>%
   full_join(resoptnat) %>% 
   filter(var=="cf") %>%
-  mutate(p50 = 1 - exp(-`par`),
+  mutate(p50 = 1 - exp(-`mode`),
          plow = 1 - exp(-`2.5%`),
@@
   mutate(Algorithm="MCMC") %>%
-  full_join(resoptunccf)
+  full_join(resoptcf)
 resp %>%
@@
          pupp = 1 - exp(-`97.5%`)) %>%
-  ggplot(aes(x=age, y=par, col=algorithm, lty=algorithm)) + 
+  ggplot(aes(x=age, y=par, col=Algorithm, lty=Algorithm)) + 
   geom_errorbar(aes(ymin=plow, ymax=pupp), width=1.5,
@@
 gg <- resall %>% 
-  dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
+    dplyr::filter(model %in% c("Independent areas","National"), 
+                  var=="cf",
+                area %in% c(cityregions,"England")) %>%
+    mutate(area = forcats::fct_relevel(area, "England", after=Inf)) %>%
     dplyr::filter(! (disease=="Uterine cancer" & age > 90)) %>%
@@
          pupp = 1 - exp(-`97.5%`)) %>%
-  ggplot(aes(x=age, y=p50, group=area)) + 
+  ggplot(aes(x=age, y=p50, group=area, col=area)) + 
   geom_ribbon(aes(ymin=plow, ymax=pupp), 
-              fill="deepskyblue") +
+              fill="gray80", col="gray80") +
   geom_line() + 
@@
   xlim(60,100) + 
-  theme(strip.text.y.left = element_text(angle = 0))
+  theme(strip.text.y.left = element_text(angle = 0),
+        strip.text.y = element_text(size=6),
+        axis.text = element_text(size=8),
+        legend.text = element_text(size=6)) + 
+    labs(col="")
 
@@
 
-## UTRC doesn't converge, so we use the empirical Bayes results
-
-resh %>%
-  filter(disease=="Stomach cancer", gender=="Male", var=="cf") %>%
-  ggplot(aes(x=age, y=`50%`, col=area, group=area)) + 
-  geom_line()
-
-resh %>% filter(gender=="Female", disease=="Stomach cancer", 
-                var=="cf", age>40, age<60) %>% select(age,`50%`,Rhat) 
+## UTRC doesn't converge, and STMC estimates noisy, so we use national estimate results 
 resnc <- resh %>% filter(Rhat > 2) 
+resnc <- resh %>% filter(Rhat > 1.1) 
 resnc <- resh %>% filter(Rhat > 1.05) 
@@
 
-## Between area variability in point and interval ests, hier and nonhier models  
-## Specific age and gender, compare between areas 
-## Not used in paper
+## Between area variability in age curves, point ests.  All diseases
 
-example_age <- 90
-dat <- resall %>% filter(age==example_age, var=="cf", 
-                         !(disease %in% c("Stomach cancer","Uterine cancer")),
-                         #      disease %in% c("Colon and rectum cancer"),
-                         model %in% c("Independent areas","Hierarchical"))
-ggplot(dat, aes(x=`50%`, y=area, col=model)) +
-  #   coord_cartesian(xlim=c(0.01, 0.05)) +
-  facet_grid(rows=vars(disease), cols=vars(gender)) + 
-  geom_point(position=ggstance::position_dodgev(height=0.4)) +
-  geom_errorbarh(aes(xmin=`2.5%`, xmax=`97.5%`),
-                 position=ggstance::position_dodgev(height=0.4)) +
-  xlab(sprintf("Case fatality for age %s", example_age) )
-
-## Between area variability in age curves, point ests.  CARC.  Shrinkage for oldest
-## Use this in paper
-
-pdf("../write/hier_crc.pdf", width=6, height=4)
+pdf("../../write/hier.pdf", width=6, height=4)
 
 resall %>% 
-  filter(var=="cf", 
-         !disease %in% c("Stomach cancer","Uterine cancer"),
-         disease %in% c(
-           #   "Dementia"
-           "Colon and rectum cancer"
-           #       "Lung cancer"
-         ),
-         `97.5%` < 10,
+    filter(var=="cf", !(disease %in% c("Uterine cancer" , "Stomach cancer")),
            area %in% cityregions, 
+           gender=="Female",
            model %in% c("Independent areas","Hierarchical")) %>%
+    mutate(disease = fct_recode(disease, 
+                                "Colorectal cancer"="Colon and rectum cancer",
+                                "IHD" = "Ischemic heart disease")) %>%
+    mutate(p50 = 1 - exp(-`50%`),
+           plow = 1 - exp(-`2.5%`),
+           pupp = 1 - exp(-`97.5%`)) %>%
     mutate(model = relevel(factor(model), "Independent areas")) %>%
     ggplot(aes(x=age, col=area, group=area)) +
-  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), fill="gray80", col="gray80") +
-  geom_line(aes(y=`50%`)) + 
-  coord_cartesian(ylim=c(0, 1), 
-                  xlim=c(70,  100)) +
+    geom_ribbon(aes(ymin=plow, ymax=pupp), fill="gray80", col="gray80") +
+    geom_line(aes(y=p50)) + 
+    coord_cartesian(xlim=c(70,  100), ylim=c(0,NA)) +
     scale_x_continuous(breaks=seq(0,100,10)) + 
-  ylab("Case fatality rate")  +
-  facet_grid(cols=vars(model), rows=vars(gender)) + 
+    ylab("Annual case fatality probability")  +
+    facet_grid(cols=vars(model), rows=vars(disease),
+               scales="free_y", switch = "y") + 
     xlab("Age (years)") + 
-  theme_bw() + labs(col="")
+    labs(col="") +
+    theme(strip.text.y.left = element_text(angle = 0),
+          strip.text.y = element_text(size = 6.5),
+          legend.text = element_text(size = 6.5))
 
@@
 
-## point est curves compared between areas 
+## Point est curves compared between areas 
+## Not shown in paper
 rescf <- resg %>% filter(gender=="Male", var=="cf")
@@
   mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
-  filter(! disease %in% c("Breast cancer", "Stomach cancer", "Uterine cancer"), 
+  filter(! disease %in% c("Breast cancer", 
+                          "Stomach cancer", 
+                          "Uterine cancer"), 
          area %in% cityregions, 
@@
 resgi <- resg %>%  
-  filter(! disease %in% c("Breast cancer", "Stomach cancer", "Uterine cancer"), 
+  filter(! disease %in% c("Breast cancer", 
+                          "Stomach cancer", 
+                          "Uterine cancer"), 
          area %in% cityregions, 
@@
 ## This is good evidence that the general pattern is common among areas
-## TODO Confirm this or otherwise by LOO CV
 
-pdf("../../write/gender_hr.pdf")
+pdf("../../write/gender_hr.pdf", width=6, height=4)
+
 ggplot(reshis %>% filter(age > 60), 
@@
   labs(col="Area") +
-  coord_cartesian(ylim=c(0,3)) + 
-  ylab("Case fatality rate ratio (men/women)") + xlab("Age (years)")
+  coord_cartesian(ylim=c(0, 3)) + 
+  ylab("Case fatality rate ratio (men/women)") +
+  xlab("Age (years)") + 
+    theme(legend.text = element_text(size = 6.5),
+          strip.text = element_text(size = 6.5))
+
 dev.off()
@@
 
-## Lookup table for all results
+# Cross validatory comparison
 
-100 * 9 * 18 * 2 * 3 results for each of inc and CF
+Full table of model comparison statistics for all 4 relevant models for the paper
 
-TODO just save this data frame in the metahit directory. there's no standard format
-
-TODO append the results from optimisation.  OK to have multiple results for one case 
-or just save in different format, since one has mode
+No-incidence model is best for all diseases
 
 ```{r}
-app_tab <- resall %>%
-  filter(var %in% c("cf","inc","rem")) %>%
-  select(disease, var, gender, age, area, model, `25%`, `50%`, `97.5%`) %>%
-  arrange(disease, var, gender, age, area, model)
-app_tab %>% head
-```
+looall2 <- looall %>% 
+  filter(model %in% c("Independent areas",
+                      "Independent areas, no incidence data",
+                      "Hierarchical",
+                      "Hierarchical joint gender"),
+         !(disease %in% c("Uterine cancer","Stomach cancer")),
+         outcome %in% c("inc","prev","mort"))
 
-# Cross validatory comparison
+looallsum2 <- looall2 %>%
+  filter(outcome %in% c("prev","mort")) %>% 
+  group_by(age, area, gender, disease, model) %>%
+  summarise(looic = sum(looic), .groups="drop") %>%
+  mutate(outcome = "prevmort")
+looallsum3 <- looall2 %>%
+  group_by(age, area, gender, disease, model) %>%
+  summarise(looic = sum(looic), .groups="drop") %>%
+  mutate(outcome = "total")
 
-Compare nonhierarchical with hierarchical model, for all diseases and outcomes.
+lootab <- looall2 %>%
+  full_join(looallsum2) %>%
+  full_join(looallsum3) %>%
+  select(outcome, area, age, gender, disease, model, looic) %>%
+  pivot_wider(names_from=model, values_from=looic) %>%
+  group_by(disease, outcome) %>%
+  summarise(enh = sum(`Independent areas`),
+            enhi = sum(`Independent areas, no incidence data`),
+            eh = sum(`Hierarchical`),
+            ehg = sum(`Hierarchical joint gender`), .groups=c("drop")) %>%
+  mutate(enhi = ifelse(outcome %in% c("inc","total"), NA, enhi),
+         disease = ifelse(outcome=="total", "", as.character(disease)),
+         outcome = fct_recode(outcome, "Incidence"="inc", "Prevalence"="prev", 
+                              "Mortality"="mort", "Prevalence and mortality" = "prevmort",
+                              "Incidence, prevalence and mortality" ="total")) %>% 
+  filter(outcome %in% c("Prevalence and mortality","Incidence, prevalence and mortality")) %>%
+  mutate(across(3:6, ~ replace_na(as.character(round(.x)),"")))
 
-Colon and rectum cancer is highlighted in the paper as a case where the hierarchical model makes a difference to the results, for ages 80+.   The hierarchical model gives better predictions here.
-
-```{r}
-looall <- loonh %>% mutate(model="Independent areas") %>%
-  full_join(looh %>% mutate(model="Hierarchical")) %>%
-  filter(!(model=="National")) %>%
-  select(age, gender, elpd_loo, model, area, disease, var) %>%
-  group_by(model, gender, disease, var)
-
-looall %>% 
-  summarise(eisum = sum(elpd_loo)) %>%
-  pivot_wider(names_from=model, values_from=eisum) %>%
-  mutate(elpd_diff = `Hierarchical` - `Independent areas`)
-
-looall %>% 
-  filter(age > 80) %>%
-  summarise(eisum = sum(elpd_loo)) %>%
-  pivot_wider(names_from=model, values_from=eisum) %>%
-  mutate(elpd_diff = `Hierarchical` - `Independent areas`)%>%
-  filter(disease=="Colon and rectum cancer")
-```
-
-Graph of how predictive ability depends on age and area, for selected disease and outcome.
-Hierarchical model is better for older ages.
-Age 50 data seem to be strong enough that shrinkage is not needed.
-Note that age 50 parameters are also assumed to apply to younger ages.
-
-```{r}
-loosub <- looall %>% 
-  filter(disease=="Colon and rectum cancer", gender=="Male", var=="mort") %>%
-  ## sum ELPD by five year age group
-  pivot_wider(names_from=model, values_from=elpd_loo) %>%
-  mutate(elpd_diff = `Hierarchical` - `Independent areas`) %>%
-  mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
-  group_by(area, agegroup, gender, var) %>%
-  summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
-  mutate(agemid = as.numeric(as.character(factor(agegroup, labels = seq(2.5, 97.5, by=5)))))
-
-ggplot(loosub, aes(x=agemid, y=elpd_diff, colour=area)) +
-  geom_hline(yintercept=0, col="gray", lwd=2) +
-  geom_line() + 
-  geom_point() + 
-  scale_x_continuous(limits=c(30,100), breaks = seq(0,100,10)) + 
-  geom_text(data=data.frame(x=c(50,50), y=c(-Inf,Inf)), 
-            aes(x=x, y=y), 
-            inherit.aes=FALSE,
-            label=c("Independent areas preferred","Hierarchical preferred"), 
-            vjust=c(0,1), hjust=c(0,0)) +
-  directlabels::geom_dl(aes(label=area), 
-                        method=list(directlabels::dl.combine("first.points", "last.points")))
+lootab
+write.table(lootab, row.names = FALSE, sep="  &  ", eol="\\\\\n", quote=FALSE)
 
 ```
-
-Then compare hier with hier additive gender
-
-```{r}
-
-## FIXME rerun hier with bugfix in saving loo output .  was triggered when we changed rundf
-## to exclude (stmc and?) utrc
-
-looall %>% filter(var=="mort", area=="Bristol", gender=="Male", disease=="Lung cancer", 
-                  model=="Hierarchical", age==1)
-
-# elpd : high is good. positive elpd diff favours additive gender model 
-loohg <- looall %>% 
-  filter(model %in% c("Hierarchical","Hierarchical joint gender")) %>%
-  select(var, area, age, gender, disease, model, elpd_loo) %>%
-  pivot_wider(names_from=model, values_from=elpd_loo, values_fn=length)
-
-%>%
-  mutate(elpd_diff = `Hierarchical joint gender` - `Hierarchical`) %>%
-  mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
-  group_by(var, area, gender, disease, agegroup) %>%
-  summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
-  mutate(agemid = as.numeric(as.character(factor(agegroup, labels = seq(2.5, 97.5, by=5)))))
-
-loohg %>% 
-  filter(disease %in% c("Ischemic heart disease","Lung cancer")) %>% 
-  group_by(agemid, disease) %>%
-  summarise(elpd = sum(elpd_diff)) %>%
-ggplot(aes(x=agemid, y=elpd, col=disease, group=disease)) + 
-  geom_hline(yintercept=0) + geom_line() + geom_point()
-
-# ELPD all over the place for different ages 
-# positive: common effects.  negative: different effects 
-# Different effects for lung cancer and dem
-# not transparent though. 
-# IHD vs LC is still biggest contrast 
-# not sure makes sense to split by age.  model defines risk as smooth function of age
-
-loohg %>%
-  group_by(var, disease) %>%
-  summarise(elpd = sum(elpd_diff)) %>%
-  arrange(disease) %>%
-  pivot_wider(names_from=var, values_from=elpd) %>%
-  filter(!is.na(inc)) %>%
-  mutate(rem=ifelse(is.na(rem), 0, rem),
-         overall = inc+mort+prev+rem)
-
-```

### 2021-10-07 metahit/paper_analyses_gender.r

@@ -20,12 +20,8 @@ if (0){
 mhi <- gbd %>%
     filter(disease==hierrungdf$disease[i]) %>%
     droplevels
-
-## TODO find posterior mode of lambda_cf_male so we can fix it at this plausible value in MCMC
-## OK mode is 1.3 here.  So how can we use it in the real run. cd just set TRUE
-## did we try this before 
-## TODO use new EB code from disbayes_hier in disbayes. any sharing ? needs to handle subset of 5 pars cleanly  
-## TODO test all EB cases, put in testthat
+hpfixed <- if (hierrungdf$model[i]=="const") NULL else list(scf=2.5, sinc=5, scfmale=1)
+## mode of scfmale is 1.3, so 1 ensures no less smooth
 
 db <- disbayes_hier(data=mhi,
                     group = "area",
@@ -36,14 +32,14 @@ db <- disbayes_hier(data=mhi,
                     rem_num = if (hierrungdf$remission[i]) "rem_num" else NULL,
                     rem_denom = if (hierrungdf$remission[i]) "rem_denom" else NULL,
                     cf_model = hierrungdf$model[i],
-                    inc_model = "indep",
-                    inc_prior = c(1.1, 1), 
-                    hp_fixed = list(scf = 2.65, sint = 2.65),
-                    eqage = hierrundf$eqage[i], 
+                    inc_model = "smooth",
+                    rem_prior = c(1.1, 1),
+                    hp_fixed = hpfixed, 
+                    eqage = hierrungdf$eqage[i], 
                     nfold_int_guess = 5, nfold_int_upper =  50,
                     nfold_slope_guess = 2, nfold_slope_upper =  20,
-                    method="opt", hessian=TRUE, draws=1000, iter=10000, verbose=TRUE
-                    #method="mcmc", refresh = 1, chains=nchains, iter=1000,
+                    #method="opt", hessian=TRUE, draws=1000, iter=10000, verbose=TRUE
+                    method="mcmc", refresh = 1, chains=nchains, iter=1000,
                     #stan_control=list(max_treedepth=15)
 )
 

### 2021-10-07 metahit/paper_analyses_header.r

@@ -7,9 +7,9 @@ library("disbayes")
 gbd <- readRDS("~/work/chronic/disbayes/metahit/gbddb.rds")
 
 nage <- 100
-areas <- unique(gbd$area)
-cityregions <- unique(gbd$area[gbd$areatype=="cityregion"])
-genders <- unique(gbd$gender)
+areas <- unique(as.character(gbd$area))
+cityregions <- unique(as.character(gbd$area)[gbd$areatype=="cityregion"])
+genders <- unique(as.character(gbd$gender))
 diseases_male <- c("Ischemic heart disease", "Stroke", "Tracheal, bronchus, and lung cancer", "Colon and rectum cancer","Alzheimer's disease and other dementias", "Chronic obstructive pulmonary disease","Stomach cancer") # no remission data for liver 
 diseases_female <- c(diseases_male, "Breast cancer","Uterine cancer")
 diseases <- unique(c(diseases_male, diseases_female))
@@ -58,7 +58,9 @@ hierrundf <- data.frame(gender = rep(c("Male","Female"), c(length(diseases_male)
                            disease %in% "Uterine cancer"  ~  "const",
                            TRUE  ~  "interceptonly"),
          remission = disease %in% cancers,
-         eqage = case_when(disease == "Alzheimer's disease and other dementias" ~ 70, 
+         eqage = case_when(disease == "Ischemic heart disease" ~ 30, 
+                           disease == "Stroke" ~ 30,
+                           disease == "Alzheimer's disease and other dementias" ~ 70, 
                            disease == "Uterine cancer" ~ 70, 
                            disease == "Stomach cancer" ~ 70, 
                            TRUE ~ 50)
@@ -71,7 +73,9 @@ hierrungdf <- data.frame(disease = diseases_male) %>%
                                           "Uterine cancer") ~  "increasing",
                            TRUE  ~  "interceptonly"),
          remission = disease %in% cancers,
-         eqage = case_when(disease == "Alzheimer's disease and other dementias" ~ 70, 
+         eqage = case_when(disease == "Ischemic heart disease" ~ 30, 
+                           disease == "Stroke" ~ 30,
+                           disease == "Alzheimer's disease and other dementias" ~ 70, 
                            disease == "Uterine cancer" ~ 70, 
                            disease == "Stomach cancer" ~ 70, 
                            TRUE ~ 50))
@@ -106,6 +110,10 @@ gbdeng <- gbd %>% select(-areatype) %>%
             .groups = "drop") 
 
 
[…]

### 2021-10-07 metahit/paper_analyses_hier.r

@@ -9,13 +9,17 @@ i <- runs_todo[task_id]
 if (0) { 
   setwd("metahit")
   source("paper_analyses_header.r")
-  i <- 1 
+  i <- 1
   nchains <-1
+  mhi <- gbd %>%
+    filter(gender==hierrundf$gender[i], disease==hierrundf$disease[i]) %>%
+    droplevels 
 }
 
 mhi <- gbd %>%
   filter(gender==hierrundf$gender[i], disease==hierrundf$disease[i]) %>%
   droplevels 
+hpfixed <- if (hierrundf$model[i]=="const") list(sd_int=TRUE,sinc=1) else list(scf=2.5,sinc=5)
 
 db <- disbayes_hier(data=mhi,
                     group = "area",
@@ -25,12 +29,11 @@ db <- disbayes_hier(data=mhi,
                     rem_num = if (hierrundf$remission[i]) "rem_num" else NULL, 
                     rem_denom = if (hierrundf$remission[i]) "rem_denom" else NULL,
                     cf_model = hierrundf$model[i],
-                    inc_model = "indep",
-                    inc_prior = c(1.1, 1), 
-                    hp_fixed = list(scf=2.65), 
+                    inc_model = "smooth",
+                    rem_prior = c(1.1, 1),
                     eqage = hierrundf$eqage[i],
+                    hp_fixed = hpfixed, 
                     nfold_int_guess = 5, nfold_int_upper =  50,
-                    nfold_slope_guess = 2, nfold_slope_upper =  20,
                     # method="opt", hessian=TRUE, draws=1000
                     method="mcmc", refresh = 1, chains=nchains, iter=1000,
                     stan_control=list(max_treedepth=15)
@@ -55,11 +58,3 @@ loo <- looi_disbayes_hier(db) %>%
 
 saveRDS(list(res=res,loo=loo), file= paste0("results_hier/res", i, ".rds"))
 cat(sprintf("Fitted model for case %s\n", i), file="metahit-hpc-log.txt", append=TRUE)
-
-
-if (0){
-  
-  plot.disbayes_hier(db)
-  
-  
-}

### 2021-10-07 metahit/paper_analyses_national.r

@@ -14,8 +14,8 @@ db <- disbayes(data = gbdeng %>% filter(disease==natrundf$disease[i],
                rem_num = if (natrundf$remission[i]) "rem_num" else NULL, 
                rem_denom = if (natrundf$remission[i]) "rem_denom" else NULL,
                eqage= natrundf$eqage[i], 
-               cf_model = if (rundf$increasing[i]) "increasing" else "smooth",
-               method="mcmc", chains=nchains, iter=5000
+               cf_model = if (natrundf$increasing[i]) "increasing" else "smooth",
+               #method="mcmc", chains=nchains, iter=3000
                )
 
 res <- tidy(db) %>%

### 2021-10-07 metahit/paper_analyses_nonhier.r

@@ -8,6 +8,9 @@ i <- runs_todo[task_id]
 
 mhi <- gbd %>%
   filter(gender==rundf$gender[i], disease==rundf$disease[i], area==rundf$area[i])
+hpfixed <- list(scf=NULL, sinc=NULL)
+if (!is.na(rundf$scf_fixed[i])) hpfixed$scf <- rundf$scf_fixed[i]
+if (!is.na(rundf$sinc_fixed[i])) hpfixed$sinc <- rundf$sinc_fixed[i]
 
 db <- disbayes(data=mhi,
                inc_num = "inc_num", inc_denom = "inc_denom",
@@ -17,16 +20,16 @@ db <- disbayes(data=mhi,
                rem_denom = if (rundf$remission[i]) "rem_denom" else NULL,
                eqage = rundf$eqage[i],
                cf_model = if (rundf$increasing[i]) "increasing" else "smooth",
-               scf_fixed = if(is.na(rundf$scf_fixed[i])) NULL else rundf$scf_fixed[i],
-               sinc_fixed = if(is.na(rundf$sinc_fixed[i])) NULL else rundf$sinc_fixed[i],
-               method = "mcmc", chains=nchains, iter=5000, refresh=10
+               hp_fixed = hpfixed, 
+               method = "mcmc", chains=nchains, iter=100, refresh=10,
+               stan_control=list(max_treedepth=15)
                )
 
 res <- tidy(db) %>%
   mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
 loo <- looi_disbayes(db) %>%
     mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
-saveRDS(list(res=res, loo=loo), file= paste0("results_nonhier/res", i, ".rds"))
+saveRDS(list(res=res, loo=loo), file= paste0("results_nonhier_noinc/res", i, ".rds"))
 
 
 if (0){

### 2021-10-07 metahit/paper_analyses_pointests.r

@@ -1,6 +1,7 @@
 ### Analyses for the paper which use optimisation to produce posterior modes
 ### with credible intervals based on a normal approximation
 
+setwd("metahit")
 source("paper_analyses_header.r")
 
 ### Non-hierarchical model
@@ -43,17 +44,15 @@ if (0){
 
 resopt_nonhier <- do.call("rbind", resopt_nonhier)
 saveRDS(resopt_nonhier, file="resopt_nonhier.rds")
-readRDS(file="resopt_nonhier.rds") %>% 
-    filter(disease=="Ischemic heart disease", gender=="Female", area=="London")
+resopt_nonhier <- readRDS(file="resopt_nonhier.rds") 
 
-resopt_nonhier %>%
-    filter(disease=="Ischemic heart disease", gender=="Female", area=="London")
 
+    
 
 ## Point estimates of smoothness parameters without uncertainty,
 ## produced for specific cases where MCMC doesn't converge
-## An easier way might be to return the modes along with the tidied results 
-## for the empirical Bayes fits done above. 
+## Used to provide fixed values in paper_analyses_header.r 
+## It'd be easier to just request an EB fit directly when using MCMC - do this if we update the data
 
 resbad <- vector(3, mode="list")
 badids <- c(47,85,86)
@@ -80,9 +79,7 @@ resbad <- do.call("rbind", resbad)
 resbad %>% filter(grepl("lambda",var))
 
 
-
-
-### National estimates 
+### National estimates (uterine and stomach cancer)
 
 resopt_nat <- vector(nrow(natrundf), mode="list")
 
@@ -112,40 +109,35 @@ resopt_nat <- readRDS(file="resopt_nat.rds")
 
 
 
+### Hierarchical models for stomach and uterine cancers, using optimisation 
+### Not shown in paper 
+### MCMC doesn't work for uterine cancer, and gives noisy results for stomach cancer
 
-### Hierarchical.  EB normal approx for all, sprior = 100 and tighter re var priors 
-### We only need results for uterine cancer here, since MCMC works for the rest 
-resopt_hier <- vector(nrow(hierrundf), mode="list")
+hoptdf <- hierrundf %>% 
+    filter(disease %in% c("Stomach cancer","Uterine cancer"))
+resopt_hier <- vector(nrow(hoptdf), mode="list")
 
-## Settings needed for normal approx sampler to work in specific cases
[…]

### 2021-10-07 metahit/read_model_results.r

@@ -1,5 +1,8 @@
 ### Read, tidy and combine all results obtained from running MCMC simulations (parallelised for different cases) on a HPC cluster
 
+
+## Non-hierarchical models with areas fitted independently 
+
 reslist <- loolist <- vector(nrow(rundf), mode="list")
 for (i in 1:nrow(rundf)){
   fname <- sprintf("results_nonhier/res%s.rds",i)
@@ -15,6 +18,9 @@ resnh <- do.call("rbind", reslist) %>%
   mutate(model="Independent areas")
 loonh <- do.call("rbind", loolist) %>%
   mutate(model="Independent areas")
+
+
+## National estimates for stomach/uterine cancers
 
 resi <- readRDS("results_nonhier/res_eng_utrc_female.rds")
 res_utrc <- resi$res; loo_utrc <- resi$loo
@@ -34,16 +40,21 @@ loonh <- loonh %>%
   dplyr::filter(!(disease %in% c("Stomach cancer", "Uterine cancer"))) %>%
   full_join(loonat)
 
-#res %>% filter(disease=="Uterine cancer", var=="cf", between(age, 58, 70))
-#res_utrc %>% filter(disease=="Uterine cancer", var=="cf", between(age, 90, 100))
-## Smoothed hasn't worked 
+
+## Hierarchical models with genders fitted independently
 
 reslist <- loolist <- vector(nrow(hierrundf), mode="list")
 for (i in setdiff(1:nrow(hierrundf), c())){
-  resi <- readRDS(sprintf("results_hier/res%s.rds",i))
+  try(resi <- readRDS(sprintf("results_hier/res%s.rds",i)))
   reslist[[i]] <- resi[[1]]
   loolist[[i]] <- resi[[2]]
 }
+loolist[[16]] <- loolist[[16]] %>%
+mutate(gender=hierrundf$gender[16], disease=hierrundf$disease[16])
+
+
+## Hierarchical models with additive gender effects
+
 resh <- do.call("rbind", reslist)
 resh$model <- "Hierarchical"
 looh <- do.call("rbind", loolist) %>%
@@ -51,7 +62,7 @@ looh <- do.call("rbind", loolist) %>%
          model = "Hierarchical")
 
 reslist <- loolist <- vector(nrow(hierrungdf), mode="list")
-for (i in setdiff(1:nrow(hierrungdf), 7)){ # exclude stomach cancer
+for (i in setdiff(1:nrow(hierrungdf), c())){ # previously excluded stomach cancer
[…]

### 2021-10-07 metahit/slurm_bsu.sh

@@ -1,9 +1,9 @@
 #!/bin/bash
 #SBATCH -J metahit
-#SBATCH --time=12:00:00
+#SBATCH --time=24:00:00
 #SBATCH --cpus-per-task=2
-#SBATCH --array=[1-16]%8
+#SBATCH --array=[1-3]%8
 module purge
 module load default-login
 
-Rscript paper_analyses_hier.r
+Rscript paper_analyses_national.r

### 2021-10-18 data-raw/gbd_process.Rmd
Remission rates can be a smooth function of age

@@
 library(tempdisagg)
-relative_path_gbd <- "~/scratch/chronic/metahit_data"
+relative_path_gbd <- "~/scratch/chronic/metahit_data/2017"
 
@@
 
-saveRDS(gbddb, file=file.path(relative_path_gbd, "gbddb.rds"))
-saveRDS(gbddb, file=file.path("~/work/chronic/disbayes/metahit", "gbddb.rds"))
-
-qbeta(c(0.025, 0.975), 128, 940-128) # 0.11 to 0.16.more plausible
-
 ```

### 2021-10-18 R/disbayes_hier.R

@@ -49,14 +49,14 @@
 ##' @param inc_model Model for how incidence varies with age.
 ##'
 ##'   \code{"smooth"} (the default). Incidence is modelled as a smooth spline
-##'   function of age, independently for eahc area (and gender).
+##'   function of age, independently for each area (and gender).
 ##'
 ##'   \code{"indep"} Incidence rates for each year of age, area (and gender) are
 ##'   estimated independently.
 ##'
 ##' @param rem_model Model for how remission varies with age.  Currently
 ##'   supported models are \code{"const"} for a constant remission rate over all
-##'   ages, or \code{"indep"} for a different remission rates estimated
+##'   ages, \code{"const"} for a smooth spline,  or \code{"indep"} for a different remission rates estimated
 ##'   independently for each age with no smoothing.
 ##'
 ##' @param hp_fixed A list with one named element for each hyperparameter
@@ -140,14 +140,14 @@ disbayes_hier <- function(data,
                           rem_model = "const",
                           prev_zero = FALSE, 
                           loo = TRUE, 
-                          sprior = c(1,1),
+                          sprior = c(1, 1, 1),
                           hp_fixed = NULL,
                           nfold_int_guess = 5,  nfold_int_upper = 100,
                           nfold_slope_guess = 5, nfold_slope_upper = 100,
                           mean_int_prior = c(0,10), mean_slope_prior = c(5,5),
                           gender_int_priorsd = 0.82, gender_slope_priorsd = 0.82, 
-                          inc_prior = c(2, 0.1), 
-                          rem_prior = c(2, 1), 
+                          inc_prior = c(1.1, 0.1), 
+                          rem_prior = c(1.1, 1), 
                           method = "mcmc",
                           draws = 1000,
                           iter = 10000,
@@ -159,7 +159,7 @@ disbayes_hier <- function(data,
     cf_model <- match.arg(cf_model, c("default","interceptonly","increasing",
                                       "common", "increasing_common", "const", "const_common"))
     inc_model <- match.arg(inc_model, c("smooth", "indep"))
[…]

### 2021-10-18 R/eb.R

@@ -25,16 +25,17 @@ expand_hpfixed <- function(hp, hp_fixed){
 }
 
 .disbayes_hier_hp <- data.frame(
-  pars      = c("scf","sinc","scfmale","sd_int","sd_slope"),
-  row.names = c("scf","sinc","scfmale","sd_int","sd_slope"),
-  stannames = c("lambda_cf[1]","lambda_inc[1]","lambda_cf_male[1]", "sd_inter[1]","sd_slope[1]"),
+  pars      = c("scf","sinc","srem","scfmale","sd_int","sd_slope"),
+  row.names = c("scf","sinc","srem","scfmale","sd_int","sd_slope"),
+  stannames = c("lambda_cf[1]", "lambda_inc[1]", "lambda_rem[1]", 
+                "lambda_cf_male[1]", "sd_inter[1]", "sd_slope[1]"),
   stringsAsFactors=FALSE
 )
 
 .disbayes_hp <- data.frame(
-  pars      = c("scf","sinc"),
-  row.names = c("scf","sinc"),
-  stannames = c("lambda_cf[1]","lambda_inc[1]"),
+  pars      = c("scf","sinc","srem"),
+  row.names = c("scf","sinc","srem"),
+  stannames = c("lambda_cf[1]","lambda_inc[1]","lambda_rem[1]"),
   stringsAsFactors=FALSE
 )
 

### 2021-10-18 R/fit.R

@@ -3,23 +3,33 @@ tidy_obsdat <- function(dat){
     prev <- data.frame(var="prev", num = dat$prev_num, denom = dat$prev_denom)
     mort <- data.frame(var="mort", num = dat$mort_num, denom = dat$mort_denom)
     tdat <- rbind(inc, prev, mort)
+    if (dat$remission) {
+        rem <- data.frame(var="rem", num = dat$rem_num, denom = dat$rem_denom)
+        tdat <- rbind(tdat, rem)
+    }
     tdat$age <- rep(1:dat$nage, length.out = nrow(tdat))
     tdat$est <- tdat$num/tdat$denom
     tdat$lower <- qbeta(0.025, tdat$num, tdat$denom-tdat$num)
     tdat$upper <- qbeta(0.975, tdat$num, tdat$denom-tdat$num)
     tdat
 }
-    
+
 plotfit_data_disbayes <- function(x, startyear=NULL){
-    var <- age <- est <- lower <- upper <- source <- mode <- `2.5%` <- `97.5%` <- NULL
+    var <- age <- est <- lower <- upper <- source <- mode <- `50%` <- `2.5%` <- `97.5%` <- NULL
     datobs <- tidy_obsdat(x$dat) %>% 
         select(var, age, est, lower, upper) %>% 
         mutate(source="Observed")
-    res <- tidy(x, startyear=startyear) %>% 
-        filter(var %in% c("inc","prev","mort")) %>%
-        rename(est=mode, lower=`2.5%`, upper=`97.5%`) %>%
-##        select(var, age, est, lower, upper) %>% 
-        mutate(source="Fitted") %>%
+    vars <- c("inc_prob","prev","mort")
+    if (x$dat$remission) vars <- c(vars, "rem_prob")
+    res <- tidy(x, startyear=startyear)
+    res$est <- if(is.null(res$mode)) res$`50%` else res$mode
+    res <- res %>% 
+        filter(var %in% vars) %>%
+        rename(lower=`2.5%`, upper=`97.5%`) %>%
+        select(var, age, est, lower, upper) %>% 
+        mutate(source="Fitted",
+               var = ifelse(var=="inc_prob", "inc", var),
+               var = ifelse(var=="rem_prob", "rem", var)) %>%
         full_join(datobs, by = c("var", "age", "est", "lower", "upper", "source"))
     res
 }

### 2021-10-18 src/stanExports_disbayes.h

@@ -37,7 +37,7 @@ stan::io::program_reader prog_reader__() {
     reader.add_event(0, 0, "start", "/include/trans_probs.stan");
     reader.add_event(90, 90, "end", "/include/trans_probs.stan");
     reader.add_event(90, 1, "restart", "model_disbayes");
-    reader.add_event(364, 273, "end", "model_disbayes");
+    reader.add_event(385, 294, "end", "model_disbayes");
     return reader;
 }
 template <typename T0__, typename T1__, typename T2__>
@@ -450,6 +450,7 @@ class model_disbayes
 private:
         int smooth_cf;
         int smooth_inc;
+        int smooth_rem;
         int remission;
         int trend;
         int prev_zero;
@@ -480,8 +481,10 @@ private:
         std::vector<double> rem_prior;
         int scf_isfixed;
         int sinc_isfixed;
+        int srem_isfixed;
         double lambda_cf_fixed;
         double lambda_inc_fixed;
+        double lambda_rem_fixed;
 public:
     model_disbayes(stan::io::var_context& context__,
         std::ostream* pstream__ = 0)
@@ -525,38 +528,44 @@ public:
             pos__ = 0;
             smooth_inc = vals_i__[pos__++];
             current_statement_begin__ = 95;
+            context__.validate_dims("data initialization", "smooth_rem", "int", context__.to_vec());
+            smooth_rem = int(0);
+            vals_i__ = context__.vals_i("smooth_rem");
+            pos__ = 0;
+            smooth_rem = vals_i__[pos__++];
+            current_statement_begin__ = 96;
             context__.validate_dims("data initialization", "remission", "int", context__.to_vec());
             remission = int(0);
             vals_i__ = context__.vals_i("remission");
             pos__ = 0;
             remission = vals_i__[pos__++];
-            current_statement_begin__ = 96;
+            current_statement_begin__ = 97;
             context__.validate_dims("data initialization", "trend", "int", context__.to_vec());
             trend = int(0);
             vals_i__ = context__.vals_i("trend");
             pos__ = 0;
             trend = vals_i__[pos__++];
-            current_statement_begin__ = 97;
+            current_statement_begin__ = 98;
             context__.validate_dims("data initialization", "prev_zero", "int", context__.to_vec());
             prev_zero = int(0);
             vals_i__ = context__.vals_i("prev_zero");
             pos__ = 0;
             prev_zero = vals_i__[pos__++];
-            current_statement_begin__ = 98;
[…]

### 2021-10-25 R/disbayes.R
Rename inc,mort,prev to inc_prob,mort_prob,prev_prob in input and output to emphasise they are probabilities not rates

@@ -42,7 +42,7 @@
 ##'   for example, multiplying both the numerator and denominator by 0.5 would
 ##'   give the data source half its original weight.
 ##'
-##' @param inc Estimate of the incidence probability
+##' @param inc_prob Estimate of the incidence probability
 ##' @param inc_lower Lower credible limit for the incidence estimate
 ##' @param inc_upper Upper credible limit for the incidence estimate
 ##'
@@ -51,7 +51,7 @@
 ##' 
 ##' @param prev_denom Denominator for the estimate of prevalence (e.g. the size
 ##'   of the survey used to obtain the prevalence estimate)
-##' @param prev Estimate of the prevalence probability
+##' @param prev_prob Estimate of the prevalence probability
 ##' @param prev_lower Lower credible limit for the prevalence estimate
 ##' @param prev_upper Upper credible limit for the prevalence estimate
 ##'
@@ -62,7 +62,7 @@
 ##'   the population size, if the estimates were obtained from a comprehensive
 ##'   register)
 ##' 
-##' @param mort  Estimate of the mortality probability
+##' @param mort_prob  Estimate of the mortality probability
 ##' @param mort_lower Lower credible limit for the mortality estimate
 ##' @param mort_upper Upper credible limit for the mortality estimate
 ##'
@@ -78,7 +78,7 @@
 ##'   weaker prior information. 
 ##'   
 ##' @param rem_denom Denominator for the estimate of the remission probability
-##' @param rem  Estimate of the remission probability
+##' @param rem_prob  Estimate of the remission probability
 ##' @param rem_lower Lower credible limit for the remission estimate
 ##' @param rem_upper Upper credible limit for the remission estimate
 ##'
@@ -298,10 +298,10 @@
 ##'
 ##' @export
 disbayes <- function(data,
-                     inc_num=NULL, inc_denom=NULL, inc=NULL, inc_lower=NULL, inc_upper=NULL,
-                     prev_num=NULL, prev_denom=NULL, prev=NULL, prev_lower=NULL, prev_upper=NULL,
-                     mort_num=NULL, mort_denom=NULL, mort=NULL, mort_lower=NULL, mort_upper=NULL,
-                     rem_num=NULL, rem_denom=NULL, rem=NULL, rem_lower=NULL, rem_upper=NULL,
[…]

### 2021-10-25 R/tidy.R

@@ -21,19 +21,11 @@
 ##'
 ##' Model parameters might include, depending on the model specification, 
 ##'
-##' * `cf` Case fatality rates
+##' * `cf`, `inc`, `rem`: Case fatality, incidence, remission rates
 ##'
-##' * `inc` Incidence rates
+##' * `inc_prob`, `rem_prob`, `mort_prob`:  Annual incidence, remission and mortality probabilities. 
 ##'
-##' * `rem` Remission rates
-##'
-##' * `inc_par`, `cf_par`, `rem_par`.  Same as the above - should
-##'
-##' * `inc_prob`, `rem_prob`.  Annual incidence and remission probabilities. 
-##'
-##' * `mort` Annual mortality probability.
-##'
-##' * `prev` Prevalence probability. 
+##' * `prev_prob` Prevalence probability. 
 ##'
 ##' * `state_probs` State occupancy probabilities.
 ##'
@@ -125,32 +117,34 @@ tidy_disbayes_full <- function(fit, varlist, method, levs=NULL, ...) {
 .disbayes_vars <- list(
     age = list(indnames = "age",
                varnames = c("cf","inc_par","cf_par","rem_par",
-                            "dcf","rem","rem_prob","mort")), 
+                            "dcf","rem","rem_prob","mort_prob")), 
     agebias =  list(indnames = c("age", "bias"),
-                    varnames = c("inc", "inc_prob", "prev")),
+                    varnames = c("inc", "inc_prob", "prev_prob")),
     agebiasstate = list(indnames = c("age", "bias", "state"),
                         varnames = "state_probs"),
     term = list(indnames = "term", 
                 varnames = c("beta", "beta_inc")),
-    const = c("lambda_cf","lambda_inc","prevzero","cfbase","bias_loghr")
+    const = c("lambda_cf","lambda_inc","prevzero","cfbase","bias_loghr"),
+    redundant = c("inc_par","cf_par","rem_par")
 )
 attr(.disbayes_vars, "numerics") <- c("age", "bias", "state", "term")
 attr(.disbayes_vars, "order") <- c("age", "bias", "state", "term")
 
 .disbayes_trend_vars <- list(
     age = list(indnames="age",
-               varnames = c("cf", "inc_par", "rem_par", "rem", "rem_prob", "mort")),
+               varnames = c("cf", "inc_par", "rem_par", "rem", "rem_prob", "mort_prob")),
     ageyear = list(indnames = c("age", "year"),
[…]

### 2021-10-25 tests/testthat/test.R

@@ -14,21 +14,17 @@ test_that("data supplied as estimate and denominator",{
   dbres <- disbayes(dat = ihdbristol,
                     inc_num = "inc_num", inc_denom = "inc_denom",
                     prev_num = "prev_num", prev_denom = "prev_denom",
-                    mort = "mort_prob", mort_denom = "mort_denom",
+                    mort_prob = "mort_prob", mort_denom = "mort_denom",
                     eqage = 40, loo=FALSE, method="mcmc", algorithm="Fixed_param", 
                     chains=1, iter=100)
   expect_s3_class(dbres, "disbayes")
 })
 
 test_that("data supplied as estimate and credible limits",{
-  x <- with(ihdbristol, ci2num(mort_prob, mort_lower, mort_upper, denom0=1400))
-  expect_equal(x$num[1], 0)
-  expect_equal(x$denom[1], 1400)
-  
   dbres <- disbayes(dat = ihdbristol,
                     inc_num = "inc_num", inc_denom = "inc_denom",
                     prev_num = "prev_num", prev_denom = "prev_denom",
-                    mort = "mort_prob", mort_lower = "mort_lower", mort_upper="mort_upper",
+                    mort_prob = "mort_prob", mort_lower = "mort_lower", mort_upper="mort_upper",
                     eqage = 40, loo=FALSE, method="mcmc", algorithm="Fixed_param", chains=1, iter=100)
   expect_s3_class(dbres, "disbayes")
 })
@@ -38,7 +34,7 @@ test_that("errors when insufficient data supplied",{
    disbayes(dat = ihdbristol,
                         inc_num = "inc_num", inc_denom = "inc_denom",
                         prev_num = "prev_num", prev_denom = "prev_denom",
-                        mort = "mort_prob", 
+                        mort_prob = "mort_prob", 
                         eqage = 40, loo=FALSE, algorithm="Fixed_param"),
    "Not enough information"
   )
@@ -167,7 +163,7 @@ test_that("errors when data are invalid",{
     disbayes(dat = baddat,
              inc_num = "inc_num", inc_denom = "inc_denom",
              prev_num = "prev_num", prev_denom = "prev_denom",
-             mort = "mort_prob", 
+             mort_prob = "mort_prob", 
              eqage = 40),
     "should be <="
   )
@@ -189,7 +185,7 @@ test_that("errors when data are invalid",{
     disbayes(dat = baddat,
[…]

### 2021-10-25 tests/testthat/test_eb.R

@@ -8,7 +8,7 @@ test_that("fixing smoothness parameters",{
                     hp_fixed = list(scf = 1.234), 
                     eqage = 40, loo=FALSE, method="mcmc", algorithm="Fixed_param",
                     chains=1, iter=100)
-  dbres$hp_fixed
+  expect_equal(dbres$hp_fixed[["scf"]], 1.234)
   
   dbres <- disbayes(dat = ihdbristol,
                     inc_num = "inc_num", inc_denom = "inc_denom",
@@ -17,5 +17,20 @@ test_that("fixing smoothness parameters",{
                     hp_fixed = list(scf = 1.234, sinc=TRUE), 
                     eqage = 40, loo=FALSE, method="mcmc", algorithm="Fixed_param",
                     chains=1, iter=100)
-  dbres$hp_fixed
+  expect_equal(dbres$hp_fixed[["scf"]], 1.234)
+  expect_equal(dbres$hp_fixed[["sinc"]], 5.822661, tol=1e-03)
+  expect_equal(dbres$hp_fixed[["sinc"]], dbres$stan_data$lambda_inc_fixed, tol=1e-03)
+
+  dbres <- disbayes(dat = ihdbristol,
+                    inc_num = "inc_num", inc_denom = "inc_denom",
+                    prev_num = "prev_num", prev_denom = "prev_denom",
+                    mort_num = "mort_num", mort_denom = "mort_denom",
+                    rem_num = "rem_num", rem_denom = "rem_denom",
+                    rem_model = "smooth",
+                    hp_fixed = list(scf = 1.234, srem=TRUE), 
+                    eqage = 40, loo=FALSE, method="mcmc", algorithm="Fixed_param",
+                    chains=1, iter=100)
+  expect_equal(dbres$hp_fixed[["scf"]], 1.234)
+  expect_equal(dbres$hp_fixed[["srem"]], 0.9352172, tol=1e-03)
+  expect_equal(dbres$hp_fixed[["srem"]], dbres$stan_data$lambda_rem_fixed, tol=1e-03)
 })

### 2021-11-08 metahit/gbd_process_2019.Rmd
Upload 2019 results for METAHIT

--- 2021-10-18 data-raw/gbd_process.Rmd
@@
 ---
-title:  "Preparing data from the Global Burden of Disease for METAHIT modelling"
+title:  "Preparing data from the Global Burden of Disease 2019 for METAHIT modelling"
 author:
@@
 library(progress)
-library(tempdisagg)
-relative_path_gbd <- "~/scratch/chronic/metahit_data/2017"
-
+library(disbayes) ## Install
+relative_path_execute <- "~/scratch/chronic/mh-execute"
+relative_path_gbd <- "~/scratch/chronic/mh-mslt/input/gbd/GBD2019/METAHIT/"
 ```
 
-Read look-up table mapping between local authorities and city regions, and add names for regions that are not city regions.  No need to bother with matching regions to countries.
+Read look-up table mapping between local authorities and city regions, and add names for regions that are not city regions
 
 ```{r,eval=FALSE}
+names_non_cr <- c("United Kingdom", "England", "East Midlands", "East of England", "Greater London", "North East England", 
+                  'North West England', "South East England", "South West England", "West Midlands", "Yorkshire and the Humber", 
+                  "Northern Ireland", "Scotland", "Wales")
+local_government_areas <- read_csv(file.path(relative_path_execute, "inputs/mh_regions_lad_lookup.csv"), show_col_types = FALSE) %>% 
+  rename(region="gordet")
+for (i in names_non_cr){
+  local_government_areas <- rbind(local_government_areas, rep(i, ncol(local_government_areas)))
+}
+local_government_areas  <- local_government_areas %>%
+#    dplyr::filter(!is.na(cityregion)) %>%
+    dplyr::rename(location = lad11nm) %>%
+    dplyr::mutate(location = gsub('St. Helens', 'St Helens', location))
+
 regions <- c("East Midlands", "East of England", "Greater London", "North East England", 
                   'North West England', "South East England", "South West England", "Yorkshire and the Humber", "West Midlands")
-
-# LTLAs to city regions and regions 
-local_government_areas <- read_csv(file.path(relative_path_gbd, "mh_regions_lad_lookup.csv")) %>% rename(region="gordet")
 lad <- 
-    read.csv("~/work/chronic/belen/Local_Authority_District_to_Region__December_2017__Lookup_in_England_.csv")
+    read.csv("~/work/chronic/disbayes/metahit/geodata/Local_Authority_District_to_Region_(December_2018)_Lookup_in_England.csv")
 utla <- 
-  read.csv("~/work/chronic/belen/Lower_Tier_Local_Authority_to_Upper_Tier_Local_Authority__December_2017__Lookup_in_England_and_Wales.csv")
+  read.csv("~/work/chronic/disbayes/metahit/geodata/Lower_Tier_Local_Authority_to_Upper_Tier_Local_Authority_(December_2018)_Lookup_in_England_and_Wales_v2.csv")
 
@@
 utlareg <- utla %>% 
-  filter(!duplicated(UTLA17NM)) %>%
-  select(LTLA17NM, UTLA17NM) %>%
-  rename("LAD17NM"=LTLA17NM) %>%
-  left_join(lad, by="LAD17NM") %>%
-  filter(!(LAD17NM %in% lad_wales))
+  filter(!duplicated(UTLA18NM)) %>%
+  select(LTLA18NM, UTLA18NM) %>%
+  rename("LAD18NM"=LTLA18NM) %>%
+  left_join(lad, by="LAD18NM") %>%
+  filter(!(LAD18NM %in% lad_wales))
 
@@
 
-Read Global Burden of Disease 2017 data
+Read Global Burden of Disease data
+
+I'm doing the geographical manipulations bit differently here, to obtain a mutually exclusive set of LTLAs and UTLAs that cover the whole of England.
+We can aggregate some of those to get the city regions, and aggregate the remaining areas by region.
+
+GBD look to be using 2018 area codings in their 2019 data, e.g. Bournemouth and Poole are separate in 2019, but April 2019 created new unitary authority "Bournemouth, Christchurch and Poole".
 
 ```{r,eval=FALSE}
-gbd1 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-1.csv"))
-
-gbd2 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-2.csv"))
-
-gbd3 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-3.csv"))
-
-gbd4 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-4.csv"))
-
-gbd5 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-5.csv"))
-
-gbd <- rbind(gbd1, gbd2, gbd3, gbd4, gbd5)
+list_of_files <- list.files(path = relative_path_gbd,
+                            pattern = "\\.csv$",
+                            full.names = TRUE)
+gbd <- do.call(rbind, lapply(list_of_files, read.csv))
 
@@
                        location_name %in% regions  ~ "region",
-                       location_name %in% unique(utla$LTLA17NM) ~ "ltla",
-                       location_name %in% unique(utla$UTLA17NM) ~ "utla"
+                       location_name %in% unique(utla$LTLA18NM) ~ "ltla",
+                       location_name %in% unique(utla$UTLA18NM) ~ "utla"
   )
@@
 gbd <- gbd %>%
-  left_join(local_government_areas, by=c("location_name"="lad11nm")) %>%
+  left_join(local_government_areas, by=c("location_name"="location")) %>%
   mutate(in_cityregion = !is.na(cityregion)) %>%
@@
 
-## Next we want to decide whether to use LTLA/UTLA or region for areas not in city regions
-
+## Check that deaths add up to 500,000, and illustrate number of deaths by cause
+egd <- c("All causes", "Ischemic heart disease", "Multiple myeloma")
+gbd %>% 
+  filter(measure_name=="Deaths", 
+         metric_name=="Number" 
+         ) %>% 
+  group_by(cause_name) %>%
+  select(cause_name, val) %>%
+  summarise(num = sum(val)) %>%
+  arrange(desc(num)) %>%
+  as.data.frame()
 ```
@@
 ```{r,eval=FALSE}
-disease_names_execute <- read_csv(file.path(relative_path_gbd, "disease_outcomes_lookup.csv")) %>%
+disease_names_execute <- read_csv(file.path(relative_path_execute, "inputs/dose_response/disease_outcomes_lookup_new.csv")) %>% 
   select(GBD_name, acronym) %>%
@@
 
-DISEASE_SHORT_NAMES <- data.frame(disease = tolower(as.character(unique(gbd$cause_name))), 
-                                  sname = tolower(abbreviate(unique(gbd$cause_name, max = 2))),
-                                  stringsAsFactors = F) %>%
-    mutate(is_not_dis = ifelse((str_detect(disease, "injuries") |
-                                str_detect(disease, "All causes") |
-                                str_detect(disease, "Lower respiratory infections")), 
-                               1, 0) ) %>%
+DISEASE_SHORT_NAMES <- disease_names_execute %>%
+  mutate(sname = abbreviate(disease)) %>%
+  mutate(is_not_dis = ifelse((grepl("injuries", disease) | # grepl determines if the disease string contains 'injuries'
+                                grepl("all causes", disease) |
+                                grepl("lower respiratory infections", disease)), 
+                             1, 0)) %>%
   mutate(is_not_dis = case_when(sname == "allc"  ~  2,
                                 sname == "lwri"  ~  1,
-                                  ## Code for major depressive disorder (no deaths) and hypertensive heart disease (no incidence)
-                                  sname == "hyhd"  ~  3,
-                                  sname == "mjdd"  ~  3,
+                                sname == "npls" ~ 2,
+                                sname == "crdd" ~ 2,
                                 TRUE  ~  is_not_dis)) %>%
-    left_join(disease_names_execute, by="disease") %>%
-    mutate(acronym = ifelse(str_detect(disease, "injuries"), disease, acronym),
-           acronym = word(acronym, 1),
+  mutate(
     males = ifelse(disease %in% c("uterine cancer", "breast cancer"), 0, 1),
@@
     sname = gsub("'", '', sname),
-           acronym = ifelse(is.na(acronym), "no_pif", acronym))
+    acronym = ifelse(is.na(acronym), sapply(strsplit(disease, " "), head, 1), acronym))
+saveRDS(DISEASE_SHORT_NAMES, paste0(relative_path_gbd, "/DISEASE_SHORT_NAMES.rds"))
 ```
@@
 names(gbd) <- gsub(pattern = "_name", replacement = "", x = names(gbd))
-gbd <- gbd %>%
-    select(-contains("id")) %>%
-    mutate(cause = tolower(cause))
+gbd_temp <- gbd %>%
+    dplyr::select(-contains("id")) %>%
+    dplyr::mutate(cause = tolower(cause)) %>%
+    pivot_wider(names_from="metric", values_from=c("val", "upper", "lower")) %>%
+    mutate(pop = val_Number*100000/val_Rate) %>%
+    pivot_longer(cols = val_Percent:lower_Number,
+                 names_to = "metric",
+                 values_to = "val") %>%
+    separate(metric, c("metric2", "metric")) %>%
+    pivot_wider(names_from = "metric2", values_from = "val")
+save(gbd_temp, local_government_areas, DISEASE_SHORT_NAMES, 
+     file=file.path(relative_path_gbd, "/GBD2019.rda"))
 
-saveRDS(gbd, file=file.path(relative_path_gbd, "GBD2017.rds"))
-
-gbd <- readRDS(file=file.path(relative_path_gbd, "GBD2017.rds"))
+load(file=file.path(relative_path_gbd, "/GBD2019.rda"))
 ```
@@
 # Determining "effective sample sizes" behind estimates
-
-<!-- NOTE COUNTS FOR AREAS CONTAINING CITY REGIONS ARE NOT CONSTRAINED TO BE GREATER THAN TOTAL COUNTS FOR CITY REGIONS WITHIN THAT AREA.
-we will exclude non-city region data from this analysis.
-see notes in noncityregions.Rmd -->
 
@@
 
-Lower limits published as 0 and upper limits published as 1 are modified to be close but not equal to these bounds, while remaining consistent with the point estimate, since values of exactly 0 and 1 are inconsistent with the Beta distribution (TODO put this in disbayes::ci2num).
+Lower limits published as 0 and upper limits published as 1 are modified to be close but not equal to these bounds, while remaining consistent with the point estimate, since values of exactly 0 and 1 are inconsistent with the Beta distribution
 
@@
 gbdp <- 
-    gbd %>%
-    filter(metric == "Rate") %>% 
-    filter(cause != "all causes") %>% 
-    mutate(num = NA, denom = NA) %>%
-    mutate(val = val/100000, lower = lower / 100000, upper = upper / 100000) %>%
-    mutate(lower = if_else(lower==0, pmin(val/2, 0.00001), lower)) %>% 
-    mutate(upper = if_else(upper==1, pmax((1+val)/2, 0.99999), upper))
+    gbd_temp %>%
+    dplyr::filter(metric == "Rate") %>% 
+    dplyr::filter(!cause %in% c("all causes", "road injuries", "pedestrian road injuries", "cyclist road     injuries", "motorcyclist road injuries", "motor vechicle road injuries", "other road injuries", "motor vehicle road injuries", "depressive disorders" , "hypertensive heart disease")) %>%
+    dplyr::mutate(val = val/100000, lower = lower/100000, upper = upper/100000) %>%
+    dplyr::mutate(lower = if_else(lower<=0, pmin(val/2, 0.00001), lower)) %>% 
+    dplyr::mutate(upper = if_else(upper>=1, pmax((1+val)/2, 0.99999), upper))
+# CJ Breaks at i = 27954 since upper credible limit is greater than 1. 
+# There are 29 rows with upper > 1 
+# Lower respiratory infections, age 95. 
+# Perhaps from high chance of more than one event in a year. 
+# Restrict credible limits on indiv probs in these cases to be less than 1
+# Or could have used actual population size 
 ```
 
-The function `ci2num` in the `disbayes` package is then used to calculate the effective sample sizes.
-This takes about 40 minutes on a fairly fast laptop.
-It could be made a lot faster by using parallel processing.
+The function `ci2num` in the `disbayes` package is then used to calculate the effective sample sizes.  This is computationally intensive.
 
 ```{r,eval=FALSE,cache=TRUE}
+
+library(foreach)
+library(doParallel)
+registerDoParallel(3)
+
 nest <- nrow(gbdp)
-pb <- progress_bar$new(total = nest) # show progress of computation
-for (i in 1:nest){
+numdenom <- foreach (i=1:nest, .combine=rbind) %dopar% {
   if (gbdp$val[i] < gbdp$upper[i] &
       gbdp$val[i] > gbdp$lower[i]) { 
-    pb$tick()
-    counts <- disbayes::ci2num(gbdp$val[i], gbdp$lower[i], gbdp$upper[i])
-    gbdp$num[i] <- counts$num
-    gbdp$denom[i] <- counts$denom
+    counts <- disbayes:::ci2num(gbdp$val[i], gbdp$lower[i], gbdp$upper[i], denom0=gbdp$pop[i])
+    c(rowid = i, num=counts$num, denom=counts$denom)
   }
 }
-saveRDS(gbdp, file=file.path(relative_path_gbd, "gbdp.rds"))
+numdenom <- as.data.frame(numdenom)
+rownames(numdenom) <- NULL
+saveRDS(numdenom, file=file.path(relative_path_gbd, "numdenom.rds"))
+
+gbdp <- gbdp %>% 
+  mutate(rowid = row_number()) %>%
+  left_join(numdenom, by="rowid") %>% 
+  mutate(rowid= NULL)
+
+saveRDS(gbdp, file=file.path(relative_path_gbd, "GBD2019.rds"))
 ```
@@
 ```{r,echo=FALSE}
-gbdp <- readRDS(file=file.path(relative_path_gbd, "gbdp.rds"))
+gbdp <- readRDS(file=file.path(relative_path_gbd, "GBD2019.rds"))
 ```
@@
 ```{r}
-summary(gbdp$num/gbdp$denom - gbdp$val)
+summary(gbdp$num/gbdp$denom - gbdp$val, na.rm=TRUE)
 ```
@@
 ```{r}
-gbdp %>% select(val, num, denom) %>% arrange(desc(denom)) %>% head
+gbdp %>% dplyr::select(val, num, denom) %>% dplyr::arrange(desc(denom)) %>% head
 ```
@@
 ```{r,eval=FALSE}
-gbdnum <- gbd %>%
-  filter(metric=="Number") %>%
-  select(measure, location, sex, age, cause, Number=val)
+gbdnum <- gbd_temp %>%
+  dplyr::filter(metric=="Number") %>%
+  dplyr::select(measure, location, sex, age, cause, Number=val)
 gbdp <- gbdp %>%
   left_join(gbdnum, by=c("measure","location","sex","age","cause")) %>%
-  mutate(pop = Number / val)
-saveRDS(gbdp, file=file.path(relative_path_gbd, "gbdp.rds"))
-gbdp <- readRDS(file=file.path(relative_path_gbd, "gbdp.rds"))
-
+  dplyr::mutate(pop = Number / val)
 ```
 
-We can then use these to fill in the effective sample sizes "d"  that were missing or implausible, and deduce the effective numerators "n" by multipling "d" by the point estimate of the proportion.
+We can then use these to fill in the effective sample sizes "d"  that were missing or implausible,
+and deduce the effective numerators "n" by multipling "d" by the point estimate of the proportion.
+
+Use an arbitrary plausible number (5000) where the population size was indeterminate
 
@@
   mutate(pop = if_else(is.na(pop), 5000, pop)) %>% 
-  mutate(nodenom = is.na(denom) | (denom > pop),
+  dplyr::mutate(nodenom = is.na(denom) | (denom > pop),
          denom = if_else(is.na(denom), pop, denom),
@@
          num = ifelse(nodenom, round(val*denom), num))
+summary(gbdp$num/gbdp$denom - gbdp$val, na.rm=TRUE)
 ```
+
+That looks better.
 
@@
 gbdp <- gbdp %>%
-  extract(age, c("from_age", "to_age"), "(.+) to (.+)", remove=FALSE, convert=TRUE) %>%
+  tidyr::extract(age, c("from_age", "to_age"), "(.+) to (.+)", remove=FALSE, convert=TRUE) %>%
   mutate(from_age = case_when(age=="95 plus"  ~  95L,
@@
 # This takes about a minute 
+library(tempdisagg)
 gbdpyr <- group_modify(gbdp_grp, disagg) %>% 
@@
 ## Check for a few cases that the series match with the naive dividing by 5 method and the smooth methods
-tloc <- sample(unique(gbd_disagg$location), 3)
+tloc <- sample(unique(gbdpyr$location), 3)
 test2 <- gbdpyrd5 %>%
@@
   filter(!in_cityregion, areatype=="ltla") %>%
-  left_join(utla, by=c("location"="LTLA17NM")) %>%
-  group_by(measure, sex, ageyr, cause, UTLA17NM) %>%
+  left_join(utla, by=c("location"="LTLA18NM")) %>%
+  group_by(measure, sex, ageyr, cause, UTLA18NM) %>%
   summarise(num=sum(num), denom=sum(denom)) %>%
-  rename(location = "UTLA17NM") %>%
+  rename(location = "UTLA18NM") %>%
   mutate(areatype = "utla_agg") %>%
@@
 gbdpyrreg  <- gbdpyr  %>%
-  mutate(region = case_when(areatype=="ltla"  ~  lad$RGN17NM[match(location, lad$LAD17NM)],
-                            areatype=="utla"  ~  utlareg$RGN17NM[match(location, utlareg$UTLA17NM)],
+  mutate(region = case_when(areatype=="ltla"  ~  lad$RGN18NM[match(location, lad$LAD18NM)],
+                            areatype=="utla"  ~  utlareg$RGN18NM[match(location, utlareg$UTLA18NM)],
                             TRUE  ~  region
@@
 
-10-year survival to use to inform remission rates, from the [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/cancersurvivalinengland/nationalestimatesforpatientsfollowedupto201).
-These are not published by age.
-They are aggregated over ages 15-99, whereas 1, 5 year survival is published by age group.
+10-year survival to use to inform remission rates, from the [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/cancersurvivalinengland/nationalestimatesforpatientsfollowedupto2017
 
-Methods [here](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/methodologies/cancersurvivalstatisticalbulletinsqmi).
-Assumes "conditional probabilities of surviving for patients diagnosed in current year are equal to those diagnosed over the full 10-year period of available data."  Caveats about trends through time.
+https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/datasets/cancersurvivalratescancersurvivalinenglandadultsdiagnosed
 
-Our approach allows uncertainty around this - could represent expected variations between age groups, given variations for 5 year survival.
-
-Annual remission probability is computed from the 10 year survival probability.
-Assumes that remission happened in one of the 10 years, with some unknown annual probability r such that the probability of remission in 10 years is equal to the 10 year survival probability s = 1 - (1-r)^10, so that r = 1 -(1-s)^(1/10).
-
-Convert remission "rates" to numerators and denominators by assuming a denominator defined by the number of cancer cases over all ages and areas, but dividing all numerators and denominators by 100 to represent uncertainty about the true area-specific prevalence.
-For example, this gives a 95% credible interval for the annual breast cancer remission probability of 0.11 to 0.16.
+1-p is prob of dying in 10 years 
+1-r is prob of not remitting ie prob of stiking w cancer or dying in 1 year 
+(1-r)^10 is prob of not remitting in 10 years 
+p10 = 1 - (1-r)^10 is prob of remitting in 10 years 
+1 - (1-p10)  ^ (1/10)
 
 ```{r}
-library(readxl)
-rem <- read_excel("~/work/chronic/cancerdata/Figure_10__Predicted_10-year_net-survival_using_the_hybrid_approach_for_men_and_women_(aged_15_to_99_years).xls", range="A7:C26", na=":") %>%
-  pivot_longer(cols=c("Men","Women"), names_to = "sex", values_to = "surv10") %>% 
-  mutate(prem1 = 1 - (1 - surv10/100)^(1/10)) %>% 
-  filter(Cancer %in% c("Breast","Uterus","Colorectal","Liver","Stomach","Lung"),
-         !is.na(prem1)) %>%
+cancers_keep <- c("Breast","Colorectal","Larynx","Liver","Lung","Myeloma","Stomach","Uterus")
+rem <- readxl::read_excel("~/work/chronic/cancerdata/adultcancersurvivaltables.xlsx", 
+                  sheet="4. Predicted Survival", 
+                  skip=5,
+                  na = ":") %>%
+  select("Cancer", "Sex", standardised="Age group",  agegroup = "...4", 
+         surv10="10-year \r\nsurvival (%)",
+         surv10lower="Lower 95% CI...13", surv10upper="Upper 95% CI...14") %>%
+  mutate(surv10 = 1 - (1 - surv10/100)^(1/10),
+         surv10lower = 1 - (1 - surv10lower/100)^(1/10),
+         surv10upper = 1 - (1 - surv10upper/100)^(1/10)
+         ) %>% 
+  fill(Cancer, Sex, standardised) %>%
+  filter(standardised== "Non-standardised", 
+         Sex != "Persons",
+         agegroup != "All ages",
+         Cancer %in% cancers_keep,
+         !is.na(surv10)) %>%
+  tidyr::extract(agegroup, c("agefrom", "ageto"), "([[:digit:]]+)-([[:digit:]]+)", convert=TRUE) %>%
+  mutate(agediff = ageto - agefrom)
+
+```
+
+Convert these to numerators and denominators, using ci2num
+
+```{r}
+numdenom <- matrix(nrow=nrow(rem), ncol=2)
+for (i in 1:nrow(rem)){
+  numdenom[i,] <-  unlist(disbayes::ci2num(rem$surv10[i], rem$surv10lower[i], rem$surv10upper[i]))
+}
+numdenom <- as.data.frame(numdenom)
+names(numdenom) <- c("rem_num","rem_denom")
+rem <- cbind(rem, numdenom)
+```
+
+Disaggregate into single years of age.
+Can't work out how to use tempdisagg to get smooth disaggregations for this, since the age intervals are uneven.
+Just put equal numbers into each year of age for the moment.
+
+```{r}
+disagg <- function(dat, key){ 
+  res <- with(dat, { 
+    data.frame( 
+      ageyr = rep(agefrom,agediff) + sequence(agediff) - 1,
+      num = predict(td(rem_num ~ 1, to=5, method="fast")),
+      denom = predict(td(rem_denom ~ 1, to=5, method="fast"))
+    ) } )
+  res 
+}
+#dat1 <- rem[rem$Cancer=="Colorectal" & rem$Sex=="Men",]
+#td(dat1$rem_num ~ 0 + dat1$agediff, to=10, method = "fast")
+
+inds <- rep(1:nrow(rem), rem$agediff+1)
+remyr <- rem[inds,] %>%
+  mutate(ageyr = agefrom + sequence(rem$agediff+1) - 1, 
+           rem_num = round(rem_num / (agediff + 1)),
+         rem_denom = round(rem_denom / (agediff + 1))) %>%
+  select(Cancer, Sex, ageyr, rem_num, rem_denom)
+
+## Childhood cancer survival in England only published for all cancers combined 
+## Repeat the 15-year values.  Should also inflate the uncertainty
+
+remyr15 <- remyr %>% filter(ageyr==15)
+inds <- rep(1:nrow(remyr15), each=15)
+remyr014 <- remyr15[inds,] %>% 
+  mutate(ageyr = rep(0:14, nrow(remyr15)),
+         rem_num = round(rem_num/2),
+         rem_denom = round(rem_denom/2))
+
+remyr <- remyr %>% 
+  full_join(remyr014) %>%
+  arrange(Cancer, Sex, ageyr) %>%
   mutate(Cancer = fct_recode(Cancer, 
@@
                              "Stomach cancer"="Stomach",
-                             "Tracheal, bronchus, and lung cancer"="Lung"
-                             #,
-                             #"Liver cancer"="Liver"
+                             "Tracheal, bronchus, and lung cancer"="Lung",
+                             "Liver cancer"="Liver",
+                             "Head and neck cancer"="Larynx", 
+                             "Multiple myeloma"="Myeloma"
                              ),
-         gender = fct_recode(sex, "Male"="Men", "Female"="Women")) %>% 
-  rename(disease = Cancer)
+         gender = fct_recode(Sex, "Male"="Men", "Female"="Women")) %>%
+  select(-Sex) %>%
+  rename(disease = Cancer, age=ageyr)
 
-cancers <- c("Tracheal, bronchus, and lung cancer", "Colon and rectum cancer", "Stomach cancer", "Liver cancer", "Breast cancer", "Uterine cancer")
+# Merge the head and neck cancers, assume larynx remission applies to all
 
-num_cancer <- gbddb %>% 
-  filter(disease %in% cancers, 
-         age %in% 15:99) %>% 
-  group_by(gender, disease) %>% 
-  summarise(ncases = sum(prev_num)) %>% 
-  left_join(rem, by=c("disease","gender")) %>%
-  mutate(rem_num = round(ncases*prem1 / 100),
-         rem_denom = round(ncases / 100)) %>% 
-  select(gender, disease, rem_num, rem_denom)
-num_cancer
+hncancers <- c("Larynx cancer", "Lip and oral cavity cancer", "Nasopharynx cancer", "Other pharynx cancer")
+gbd_hnc <- gbddb %>% 
+  filter(disease %in% hncancers) %>% 
+  group_by(age, gender, area) %>%
+  summarise(mort_num=sum(mort_num), inc_num=sum(inc_num), 
+            prev_num=sum(prev_num), 
+            mort_denom=sum(mort_denom), inc_denom=sum(inc_denom), 
+            prev_denom=sum(prev_denom),
+            .groups = "drop") %>%
+  mutate(disease = "Head and neck cancer")
+gbddb <- gbddb %>%
+  filter(!(disease %in% hncancers)) %>% 
+  full_join(gbd_hnc)
 
 gbddb <- gbddb %>% 
-  left_join(num_cancer) %>% 
+  left_join(remyr, by=c("age","gender","disease")) %>% 
   mutate(rem_num = replace_na(rem_num, 0),
          rem_denom = replace_na(rem_denom, 0))
+
+saveRDS(gbddb, file=file.path(relative_path_gbd, "gbddb.rds"))
+gbddb <- readRDS(file=file.path(relative_path_gbd, "gbddb.rds"))
+saveRDS(gbddb, file=file.path("~/work/chronic/disbayes/metahit", "gbddb.rds"))
+
+gbddb <- gbddb %>% 
+  filter(!(disease %in% c("Chronic myeloid leukemia", # no deaths 
+                         "Lower respiratory infections", # not chronic 
+                         "Cyclist road injuries"     # not chronic 
+                         )))
 

### 2021-11-08 metahit/paper_analyses.Rmd
Upload 2019 results for METAHIT

@@
 
-Raw point estimates of mort, inc and prev.
-Just women for now.
-Curves compared between areas, faceted by disease.
-Shows diseases where prevalence and incidence are very low.
-Mild differences between areas.
+Raw point estimates of mort, inc and prev.  Curves compared between areas, faceted by disease.   Shows diseases where prevalence and incidence are very low.  Mild differences between areas.
+
+15 diseases in the plot, nice round number. Not including rheumatic heart disease and head and neck cancer, the two with the lowest mortality.
 
 ```{r,eval=FALSE}
+gbd %>% group_by(disease) %>% 
+  summarise(mort_num=sum(mort_num), inc_num=sum(inc_num), 
+            prev_num=sum(prev_num), rem_num=sum(rem_num),
+            mort_denom=sum(mort_denom), inc_denom=sum(inc_denom), 
+            prev_denom=sum(prev_denom), rem_denom=sum(rem_denom),
+            .groups = "drop") %>% arrange(desc(mort_num))
+
+ymax_mort <- enframe(c("Ischemic heart disease"=0.2,"Stroke"=0.1,"Dementia"=0.1,
+                       "COPD"=0.1, "Breast cancer"=0.03, "Lung cancer"=0.02,
+                       "Colon and rectum cancer"=0.02, "Stomach cancer"=0.01, 
+                       "Uterine cancer"=0.002,
+                       "Liver cancer"=0.002, 
+                       "Type 2 diabetes"=0.01, 
+                       "Parkinson's disease"=0.01, 
+                       "Cardiomyopathy and myocarditis" = 0.005,
+                       "Non-rheumatic valvular heart disease" = 0.01,
+                       "Multiple myeloma" = 0.002
+                       ), name="disease", value="ymax_mort")
+
+ymax_inc <- enframe(c("Ischemic heart disease"=0.05, "Stroke"=0.05,"Dementia"=0.1,
+                      "COPD"=0.1, "Breast cancer"=0.03, "Lung cancer"=0.03,
+                       "Type 2 diabetes"=0.02, 
+                       "Parkinson's disease"=0.01, 
+                      "Colon and rectum cancer"=0.03, "Stomach cancer"=0.01,
+                      "Uterine cancer"=0.002,
+                       "Liver cancer"=0.002, 
+                       "Cardiomyopathy and myocarditis" = 0.003,
+                       "Non-rheumatic valvular heart disease" = 0.01,
+                       "Multiple myeloma" = 0.002
+                      ), name="disease", value="ymax_inc")
+
+ymax_prev <- enframe(c("Ischemic heart disease"=0.25, "Stroke"=0.1,"Dementia"=0.4,
+                      "COPD"=1, "Breast cancer"=0.09, "Lung cancer"=0.03,
+                       "Type 2 diabetes"=1, 
+                       "Parkinson's disease"=0.07, 
+                      "Colon and rectum cancer"=0.1, "Stomach cancer"=0.02,
+                      "Uterine cancer"=0.02,
+                       "Liver cancer"=0.003, 
+                       "Cardiomyopathy and myocarditis" = 0.02,
+                       "Non-rheumatic valvular heart disease" = 0.1,
+                       "Multiple myeloma" = 0.005
+                      ), name="disease", value="ymax_prev")
 
 source("paper_analyses_header.r")
-ymax_mort <- enframe(c("Ischemic heart disease"=0.2,"Stroke"=0.1,"Dementia"=0.2,
-                       "COPD"=0.1, "Breast cancer"=0.03, "Lung cancer"=0.02,
-                       "Colon and rectum cancer"=0.02, "Stomach cancer"=0.01, 
-                       "Uterine cancer"=0.002), name="disease", value="ymax_mort")
-
-ymax_inc <- enframe(c("Ischemic heart disease"=0.1, "Stroke"=0.1,"Dementia"=0.1,
-                      "COPD"=0.1, "Breast cancer"=0.03, "Lung cancer"=0.03,
-                      "Colon and rectum cancer"=0.03, "Stomach cancer"=0.01,
-                      "Uterine cancer"=0.002), name="disease", value="ymax_inc")
-
 gbdplot <- gbdplot %>%
@@
     left_join(ymax_inc, by="disease") %>%
+    left_join(ymax_prev, by="disease") %>%
     mutate(mort_upper = pmin(mort_upper, ymax_mort),
            inc_upper = pmin(inc_upper, ymax_inc),
-           disease = factor(disease, levels=disease_order))
+           prev_upper = pmin(prev_upper, ymax_prev),
+           disease = factor(disease, levels=disease_order),
+           areagender = interaction(area, gender)) %>%
+    filter(disease %in% diseases_paper)
 ht <- 3
+
 pdf("../../write/mort_data.pdf", height=ht)
-gbdplot$areagender <- interaction(gbdplot$area, gbdplot$gender)
+
 ggplot(gbdplot, aes(x=age, y=mort_prob, col=area, lty=gender, group=areagender)) + 
@@
 
-* Stomach cancer and uterine cancer are done using national data.
+* 2017: Stomach cancer and uterine cancer are done using national data.
+
+* 2019: Cardiomyopathy, liver cancer, multiple myeloma with national data due to non convergence.
+Uterine and stomach cancers have noisy area-specific estimates, suggesting info drawn mostly from prior.
 
@@
 
-## MCMC convergence checks.   All OK with 5000 samples and the occasional model simplification
+## MCMC convergence checks
 resnc <- resnh %>% filter(Rhat > 1.05)
-table(resnc$disease, resnc$area, resnc$gender)
+resnc <- resnh %>% filter(Rhat > 1.1)
+table(resnc$disease, resnc$gender)
+
+resnc <- resnh %>% filter(Rhat > 2)
+table(resnc$disease, resnc$gender)
+table(resnc$disease, resnc$gender, resnc$area)["Multiple myeloma",,]
+table(resnc$disease, resnc$gender, resnc$area)["Liver cancer",,]
+## CM, LC, MM not worked for particular areas
+
+resnc <- resnat %>% filter(Rhat > 1.05)
+resnc <- resnat %>% filter(Rhat > 1.1)
+resnc <- resnat %>% filter(Rhat > 2)
+table(resnc$disease, resnc$gender)
+## all converges with national data
+
+## National results. Nice.
+resnat %>% filter(var=="cf") %>% 
+    ggplot(aes(x=`age`,y=`50%`,col=gender)) + 
+    geom_line() + 
+    geom_ribbon(aes(ymin=`2.5%`,ymax=`97.5%`,fill=gender)) +
+    facet_wrap(~disease, nrow=2, scales="free_y")
+
+resnh %>% filter(grepl("Liver",disease), var=="cf", 
+                 area=="Liverpool", gender=="Female") %>% 
+    ggplot(aes(x=age, y=`50%`)) + 
+    geom_ribbon(aes(ymin=`2.5%`,ymax=`97.5%`), fill="gray80") + 
+    coord_cartesian(ylim=c(0,0.1), xlim=c(50,90)) +
+    geom_line() 
+gbd %>% group_by(disease) %>% summarise(mort_num=sum(mort_num), inc_num=sum(inc_num)) %>% as.data.frame()
 
@@
 resopt <- readRDS("resopt_nonhier.rds") %>% 
-  filter(!disease %in% c("Stomach cancer","Uterine cancer"))
+  filter(!disease %in% diseases_nat)
 resoptnat <- readRDS("resopt_nat.rds")
@@
   dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
-  dplyr::filter(area=="Bristol" | (area=="England" & model=="National"), gender=="Female") %>%
+  dplyr::filter((area=="Bristol" & (!(disease %in% diseases_nat))) |
+                    (area=="England" & model=="National"), 
+                gender=="Female") %>%
   mutate(Algorithm="MCMC") %>%
@@
 resp %>%
-  filter(age %in% c(70, 75, 80, 85, 90)) %>%
+  filter(disease %in% diseases_paper, 
+         age %in% c(70, 75, 80, 85, 90)) %>%
   mutate(par = 1 - exp(-`50%`),
@@
 
-gg <- resall %>% 
+resplotnh <- resnh %>% 
     dplyr::filter(model %in% c("Independent areas","National"), 
@@
     mutate(area = forcats::fct_relevel(area, "England", after=Inf)) %>%
-    dplyr::filter(! (disease=="Uterine cancer" & age > 90)) %>%
+    dplyr::filter(! (disease=="Uterine cancer" & age > 90),
+                  disease %in% diseases_paper) %>%
     mutate(p50 = 1 - exp(-`50%`),
          plow = 1 - exp(-`2.5%`),
-         pupp = 1 - exp(-`97.5%`)) %>%
+         pupp = 1 - exp(-`97.5%`))
+
+resplotnh %>%
   ggplot(aes(x=age, y=p50, group=area, col=area)) + 
@@
 # gtable::gtable_show_layout(grob)
-idx <- which(grob$layout$name %in% c("panel-5-2","panel-9-2"))
+idx <- which(grob$layout$name %in% c("panel-2-2","panel-8-2"))
 for (i in idx) grob$grobs[[i]] <- grid::nullGrob()
@@
 
+```
+
+# Runs omitting incidence
+
+```{r}
+# Stroke NE (region) needs more runs, but we don't need this for city regions analysis
+resnc <- resnhi %>% 
+    filter(var=="cf", Rhat > 1.2, 
+           disease %in% diseases_paper) %>% 
+    droplevels()
+table(resnc$disease, resnc$gender, resnc$area)
 ```
@@
 
-## Look at precision of CRC estimates as exemplar. 
+## Look at precision of CRC estimates as exemplar for computational cost. 
 ## Effective sample size of 1000 after a few hours
@@
 ## Between area variability in age curves, point ests.  All diseases 
-
-pdf("../../write/hier.pdf", width=6, height=4)
-
-resall %>% 
-    filter(var=="cf", !(disease %in% c("Uterine cancer" , "Stomach cancer")),
-           area %in% cityregions, 
-           gender=="Female",
-           model %in% c("Independent areas","Hierarchical")) %>%
+rescfplot <- resall %>% 
+    filter(var=="cf", 
+           disease %in% diseases_paper_short, 
+           area %in% c(cityregions, "England"),
+           (model %in% c("Independent areas","National")) | 
+               (model %in% "Hierarchical" & !(disease %in% diseases_nat))
+    ) %>%
         mutate(disease = fct_recode(disease, 
@@
                                     "IHD" = "Ischemic heart disease")) %>%
-    mutate(p50 = 1 - exp(-`50%`),
+    mutate(model = ifelse(model=="National","Independent areas",model),
+           p50 = 1 - exp(-`50%`),
            plow = 1 - exp(-`2.5%`),
-           pupp = 1 - exp(-`97.5%`)) %>%
-    mutate(model = relevel(factor(model), "Independent areas")) %>%
+           pupp = 1 - exp(-`97.5%`),
+           gender_model = paste(gender, model, sep="\n")) %>%
+    mutate(gender_model = factor(gender_model, 
+                                 levels=c("Female\nIndependent areas",
+                                          "Male\nIndependent areas",
+                                          "Female\nHierarchical",
+                                          "Male\nHierarchical"))) 
+gg <- rescfplot %>%
     ggplot(aes(x=age, col=area, group=area)) +
@@
     ylab("Annual case fatality probability")  +
-    facet_grid(cols=vars(model), rows=vars(disease),
+    facet_grid(cols=vars(gender_model), rows=vars(disease),
                scales="free_y", switch = "y") + 
@@
     theme(strip.text.y.left = element_text(angle = 0),
+          strip.text.x = element_text(size = 5),
+          axis.text.x = element_text(size = 5.5),
+          axis.text.y = element_text(size = 5),
+          axis.title.x = element_text(size = 7),
+          axis.title.y = element_text(size = 7),
           strip.text.y = element_text(size = 6.5),
           legend.text = element_text(size = 6.5))
+gg
+
+pdf("../../write/cf.pdf", width=6, height=4)
+
+grob <- ggplotGrob(gg)
+# gtable::gtable_show_layout(grob)
+idx <- which(grob$layout$name %in% c("panel-5-2","panel-9-2",
+                                     "panel-8-3","panel-9-3",
+                                     "panel-5-4","panel-8-4","panel-9-4"))
+for (i in idx) grob$grobs[[i]] <- grid::nullGrob()
+grid::grid.draw(grob)
 
@@
 
-resg <- resg %>% filter(disease != "Stomach cancer") # only national estimates
-# convergence check
 summary(resg$Rhat)  # 
 resnc <- resg %>% filter(Rhat > 2) 
-resnc <- resg %>% filter(Rhat > 1.1) 
-resnc <- resg %>% filter(Rhat > 1.01) 
+resnc <- resg %>% filter(Rhat > 1.1) # DM and NRVHD didn't converge for all areas   
 table(resnc$disease)
-table(resnc$disease, resnc$area)
+table(resnc$disease, resnc$area)[c(4,6),]
 
@@
 reshi <- resh %>%  
-  mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
-  filter(! disease %in% c("Breast cancer", 
-                          "Stomach cancer", 
-                          "Uterine cancer"), 
+  filter(! disease %in% c("Breast cancer", "Stomach cancer", "Uterine cancer"),
+         disease %in% diseases_paper,
          area %in% cityregions, 
          var=="cf") %>%
+  mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
   mutate(est = log(`50%`), se=(log(`97.5%`)-log(`2.5%`))/4) 
@@
 resgi <- resg %>%  
-  filter(! disease %in% c("Breast cancer", 
-                          "Stomach cancer", 
-                          "Uterine cancer"), 
+  filter(! disease %in% c("Breast cancer", "Stomach cancer", "Uterine cancer"),
+         disease %in% diseases_paper, 
          area %in% cityregions, 
@@
 
-No-incidence model is best for all diseases
-
 ```{r}
@@
                       "Hierarchical joint gender"),
+         disease %in% diseases_paper_short,
          !(disease %in% c("Uterine cancer","Stomach cancer")),
@@
     full_join(looallsum3) %>%
+    filter(age >=50, age < 90) %>%
     select(outcome, area, age, gender, disease, model, looic) %>%
@@
     group_by(disease, outcome) %>%
-  summarise(enh = sum(`Independent areas`),
+    summarise(enh = sum(`Independent areas`, na.rm = TRUE),
               enhi = sum(`Independent areas, no incidence data`),
               eh = sum(`Hierarchical`),
-            ehg = sum(`Hierarchical joint gender`), .groups=c("drop")) %>%
-  mutate(enhi = ifelse(outcome %in% c("inc","total"), NA, enhi),
+              ehg = sum(`Hierarchical joint gender`), 
+              .groups=c("drop")) %>%
+    mutate(across(3:6, ~.x - enh)) %>%
+    mutate(
+        enhi = ifelse(outcome %in% c("inc","total"), NA, enhi),
         disease = ifelse(outcome=="total", "", as.character(disease)),
@@
                              "Incidence, prevalence and mortality" ="total")) %>% 
-  filter(outcome %in% c("Prevalence and mortality","Incidence, prevalence and mortality")) %>%
-  mutate(across(3:6, ~ replace_na(as.character(round(.x)),"")))
-
+    filter(outcome %in% c("Mortality"
+        #"Prevalence and mortality","Incidence, prevalence and mortality"
+        )) %>%
+    #  mutate(across(3:4, ~ replace_na(as.character(round(.x)),"")))
+    mutate(across(3:6, ~ replace_na(as.character(round(.x)),""))) %>%
+    select(-enhi, -outcome)
 lootab
@@
 
+## DIAGNOSTICS
+loonc <- looall %>% 
+    filter(age >= 0, age <110, 
+           influence_pareto_k>0.7) %>% 
+    mutate(agegroup = cut(age, seq(0,100,10)))
+table(loonc$agegroup, loonc$model, loonc$outcome)
+
+looall %>% 
+    filter(influence_pareto_k>1) %>% 
+    select(outcome, age, gender, disease, area, model, influence_pareto_k) %>% 
+    head
+
+## Effective number of parameters 
+loop <- looall2 %>%
+    full_join(looallsum2) %>%
+    full_join(looallsum3) %>%
+    filter(age >=50, age < 90, disease=="Ischemic heart disease") %>%
+    select(outcome, area, age, gender, disease, model, p_loo) %>%
+    pivot_wider(names_from=model, values_from=p_loo) %>%
+    group_by(disease, outcome) %>%
+    summarise(enh = sum(`Independent areas`, na.rm = TRUE),
+              enhi = sum(`Independent areas, no incidence data`),
+              eh = sum(`Hierarchical`),
+              ehg = sum(`Hierarchical joint gender`), 
+              .groups=c("drop"))
+loop
+
 ```
+
+How much difference does model choice make to the results?
+
+Estimated CF probs differ by 0.01 in most cases.
+
+```{r}
+rescomp <- resall %>% 
+    filter(disease=="Ischemic heart disease",
+           model %in% c("Independent areas", "Hierarchical"),
+           var == "cf", between(age, 50, 90)) %>%
+    select(area, gender, age, med=`50%`, model) %>%
+    mutate(med = 1 - exp(-med)) %>%
+    pivot_wider(names_from = model, values_from=med) %>%
+    mutate(hir = (Hierarchical - `Independent areas`))
+table(abs(rescomp$hir) < 0.01)
+```
+
+Comparing mortality predictions for diseases other than IHD
+
+Compare how well mortality probabilities fit the data, and check if case fatality probabilities agree between models that include vs exclude incidence data.
+
+```{r}
+
+obsdat <- gbd %>% 
+    mutate(mort_prob = mort_num/mort_denom, 
+           model = "Observed")
+fitdat <- resall %>% filter(
+           var %in% c("mort","mort_prob"), 
+           model %in% c("Independent areas", 
+                        "Independent areas, no incidence data")) %>%
+    mutate(var = "mort_prob", 
+           mort_prob = `50%`) %>%
+    full_join(obsdat)
+
+dis <- "Colon and rectum cancer"
+are <- "Bristol"
+gend <- "Female"
+fitdat %>%
+    filter(disease==dis, area==are, gender==gend, between(age, 50, 90)) %>%
+ggplot(aes(x=age, y=mort_prob, col=model)) + 
+    geom_line()
+
+mortcompdat <- fitdat %>%
+    select(disease, area, gender, age, mort_prob, model) %>%
+    pivot_wider(names_from=model, values_from=mort_prob) %>%
+    filter(!is.na(Observed), 
+           !is.na(`Independent areas`)) %>%
+    mutate(mdiff0 = `Independent areas` - `Observed`,
+           mdiff1 = `Independent areas, no incidence data` - `Observed`) %>%
+    mutate(mdiff0 = abs(mdiff0), 
+           mdiff1 = abs(mdiff1))
+mean(mortcompdat$mdiff0)    
+mean(mortcompdat$mdiff1)
+
+cfcompdat <- resall %>% 
+    filter(var %in% c("cf"), 
+           model %in% c("Independent areas", 
+                        "Independent areas, no incidence data")) %>%
+    mutate(cf_prob = 1 - exp(-`50%`)) %>%
+    select(disease, area, gender, age, cf_prob, model) %>%
+    filter(between(age, 50, 90)) %>%
+    pivot_wider(names_from=model, values_from=cf_prob) %>%
+    mutate(mdiff = `Independent areas, no incidence data` - `Independent areas`) %>%
+    group_by(disease, area, gender)
+
+summary(compdat$mdiff)    
+prop.table(table(abs(compdat$mdiff<0.01))) # 99% of estimates are within 0.01 of each other
+
+```
+
+# Plots for the Armitage talk
+
+```{r}
+pdf("~/work/uncertainty/pres/armitage/mort_data.pdf", height=3)
+gbdplot %>% 
+    filter(disease %in% c("Ischemic heart disease","Dementia",
+                          "Lung cancer","Breast cancer")) %>%
+ggplot(aes(x=age, y=mort_prob, col=area, lty=gender, group=areagender)) + 
+    facet_wrap(~disease, ncol=2, scales="free_y") + 
+    coord_cartesian(xlim = c(50,100)) +
+    ylab("Annual mortality probability") + 
+    xlab("Year of age") +
+    geom_ribbon(aes(ymin=mort_lower, ymax=mort_upper), fill="gray80", col="gray80") +
+    geom_line() + 
+    labs(col = "", lty="") +
+    theme_bw() + 
+    guides(lty=guide_legend(override.aes=list(fill=NA))) +
+    theme(legend.spacing.y = unit(-0.1, "cm"))
+dev.off()
+
+```
+
+Array of diseases
+
+Basic one for IHD
+
+```{r}
+
+armitage_width <- 6
+armitage_height <- 2.5
+
+pdf("~/work/uncertainty/pres/armitage/cf_indep.pdf",width=armitage_width, height=armitage_height)
+
+diseases_armitage <- "Ischemic heart disease"
+gg <- resnh %>% 
+    dplyr::filter(disease %in% diseases_armitage, 
+                  model %in% c("Independent areas","National"), 
+                  var=="cf",
+                area %in% c(cityregions,"England")) %>%
+    mutate(p50 = 1 - exp(-`50%`),
+           plow = 1 - exp(-`2.5%`),
+           pupp = 1 - exp(-`97.5%`)) %>%
+    ggplot(aes(x=age, y=p50, group=area, col=area)) + 
+  geom_ribbon(aes(ymin=plow, ymax=pupp), 
+              fill="gray80", col="gray80") +
+  geom_line() + 
+  facet_grid(cols=vars(gender), rows=vars(disease), scales = "free_y", switch = "y") + 
+  xlab("Age (years)") + 
+  ylab("Annual case fatality probability") + 
+  xlim(60,100) + 
+  theme(strip.text.y.left = element_text(angle = 0),
+        strip.text.y = element_text(size=6),
+        axis.text = element_text(size=8),
+        legend.text = element_text(size=6)) + 
+    labs(col="")
+gg
+
+dev.off()
+```
+
+A hier model - also discuss national estimates
+
+```{r}
+resall <- full_join(resnh, resh)
+
+diseases_armitage_hier <- c("Parkinson's disease")
+
+pdf("~/work/uncertainty/pres/armitage/cf_hier.pdf",width=armitage_width, height=armitage_height)
+
+resall %>% 
+    filter(var=="cf", disease %in% diseases_armitage_hier,
+           area %in% cityregions, 
+           gender=="Female",
+           model %in% c("Independent areas","Hierarchical")) %>%
+    mutate(p50 = 1 - exp(-`50%`),
+           plow = 1 - exp(-`2.5%`),
+           pupp = 1 - exp(-`97.5%`)) %>%
+    mutate(model = relevel(factor(model), "Independent areas")) %>%
+    ggplot(aes(x=age, col=area, group=area)) +
+    geom_ribbon(aes(ymin=plow, ymax=pupp), fill="gray80", col="gray80") +
+    geom_line(aes(y=p50)) + 
+    coord_cartesian(xlim=c(70,  100), ylim=c(0,NA)) +
+    scale_x_continuous(breaks=seq(0,100,10)) + 
+    ylab("Annual case fatality probability")  +
+    facet_grid(cols=vars(model), rows=vars(disease),
+               scales="free_y", switch = "y") + 
+    xlab("Age (years)") + 
+    labs(col="") +
+    theme(strip.text.y.left = element_text(angle = 0),
+          strip.text.y = element_text(size = 6.5),
+          legend.text = element_text(size = 6.5))
+
+dev.off()
+```
+
+Gender effects
+
+```{r}
+pdf("~/work/uncertainty/pres/armitage/cf_gender.pdf", width=armitage_width, height=armitage_height)
+
+diseases_armitage_gender <- c("COPD", "Colon and rectum cancer")
+
+ggplot(reshis %>% 
+           filter(age > 60, 
+                  disease %in% diseases_armitage_gender), 
+       aes(x=age, y=est, group=area, col=area)) + 
+  geom_ribbon(aes(ymin=lower, ymax=upper), fill="gray", col=NA, alpha=0.2) + 
+  geom_hline(yintercept = 1) +
+  geom_line(lwd=0.8) +
+  geom_line(data=resgis %>% filter(age>60,
+                                   disease %in% diseases_armitage_gender), 
+            col="black", lty=2) +
+  facet_wrap(~disease) +
+  labs(col="Area") +
+  coord_cartesian(ylim=c(0, 3)) + 
+  ylab("Case fatality rate ratio (men/women)") +
+  xlab("Age (years)") + 
+    theme(legend.text = element_text(size = 6.5),
+          strip.text = element_text(size = 6.5))
+
+dev.off()
+
+```

### 2021-11-08 metahit/paper_analyses_gender.r

@@ -33,7 +33,7 @@ db <- disbayes_hier(data=mhi,
                     rem_denom = if (hierrungdf$remission[i]) "rem_denom" else NULL,
                     cf_model = hierrungdf$model[i],
                     inc_model = "smooth",
-                    rem_prior = c(1.1, 1),
+                    rem_model = if (hierrundf$remission[i]) "smooth" else NULL,
                     hp_fixed = hpfixed, 
                     eqage = hierrungdf$eqage[i], 
                     nfold_int_guess = 5, nfold_int_upper =  50,

### 2021-11-08 metahit/paper_analyses_header.r

@@ -10,24 +10,109 @@ nage <- 100
 areas <- unique(as.character(gbd$area))
 cityregions <- unique(as.character(gbd$area)[gbd$areatype=="cityregion"])
 genders <- unique(as.character(gbd$gender))
-diseases_male <- c("Ischemic heart disease", "Stroke", "Tracheal, bronchus, and lung cancer", "Colon and rectum cancer","Alzheimer's disease and other dementias", "Chronic obstructive pulmonary disease","Stomach cancer") # no remission data for liver 
+diseases_male <- c("Ischemic heart disease", "Stroke", "Tracheal, bronchus, and lung cancer", 
+                   "Colon and rectum cancer","Alzheimer's disease and other dementias", 
+                   "Chronic obstructive pulmonary disease",
+                   "Diabetes mellitus type 2",
+                   "Parkinson's disease",
+                   "Stomach cancer",
+                   "Liver cancer",
+                   "Cardiomyopathy and myocarditis",
+                   "Non-rheumatic valvular heart disease",
+                   "Multiple myeloma")
+
+## Restricting to diseases with more than 1000 deaths in England 
+## Excludes rheumatic heart disease and head and neck cancer 
+
 diseases_female <- c(diseases_male, "Breast cancer","Uterine cancer")
 diseases <- unique(c(diseases_male, diseases_female))
 cancers <- c("Tracheal, bronchus, and lung cancer", "Colon and rectum cancer", "Stomach cancer", 
-             "Breast cancer", "Uterine cancer")
+             "Breast cancer", "Uterine cancer","Multiple myeloma", "Liver cancer")
 
-dcodes <- data.frame(disease=diseases, 
-                     dcode = c("ihd","stroke","tbalc","carc","adaod","copd","stmc","brc","utrc"))
+#dcodes <- data.frame(disease=diseases, 
+#                     dcode = c("ihd","stroke","tbalc","carc","adaod","copd","stmc","brc","utrc"))
+# greatest cause-specific mortality
+# six with the greatest cause-specific mortality (ihd, str, lc ,copd, carc, bc) 
+# plus dementia and 
+# pd,, nrv, dm have greater mortality 
+
+gbdeng <- gbd %>% select(-areatype) %>% 
+  group_by(age, gender, disease) %>% # should contain all English city regions and mutually exclusive regions
+  summarise(mort_num=sum(mort_num), inc_num=sum(inc_num), 
[…]

### 2021-11-08 metahit/paper_analyses_hier.r

@@ -6,20 +6,12 @@ task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
 runs_todo <- 1:nrow(hierrundf) 
 i <- runs_todo[task_id]
 
-if (0) { 
-  setwd("metahit")
-  source("paper_analyses_header.r")
-  i <- 1
-  nchains <-1
-  mhi <- gbd %>%
-    filter(gender==hierrundf$gender[i], disease==hierrundf$disease[i]) %>%
-    droplevels 
-}
-
 mhi <- gbd %>%
   filter(gender==hierrundf$gender[i], disease==hierrundf$disease[i]) %>%
   droplevels 
 hpfixed <- if (hierrundf$model[i]=="const") list(sd_int=TRUE,sinc=1) else list(scf=2.5,sinc=5)
+if (hierrundf$disease[i]=="Diabetes mellitus type 2" && hierrundf$gender[i]=="Female") hpfixed$sd_int <- TRUE
+if (hierrundf$disease[i]=="Non-rheumatic valvular heart disease" && hierrundf$gender[i]=="Female") hpfixed$sd_int <- TRUE
 
 db <- disbayes_hier(data=mhi,
                     group = "area",
@@ -30,16 +22,14 @@ db <- disbayes_hier(data=mhi,
                     rem_denom = if (hierrundf$remission[i]) "rem_denom" else NULL,
                     cf_model = hierrundf$model[i],
                     inc_model = "smooth",
-                    rem_prior = c(1.1, 1),
+                    rem_model = if (hierrundf$remission[i]) "smooth" else NULL,
                     eqage = hierrundf$eqage[i],
                     hp_fixed = hpfixed, 
                     nfold_int_guess = 5, nfold_int_upper =  50,
-                    # method="opt", hessian=TRUE, draws=1000
-                    method="mcmc", refresh = 1, chains=nchains, iter=1000,
+#                     method="opt", hessian=TRUE, draws=1000, verbose = TRUE
+                   method="mcmc", refresh = 1, chains=nchains, iter=1000,
                     stan_control=list(max_treedepth=15)
                     )
-
-# saveRDS(db, file="~/scratch/chronic/db_hier_test.rds")
 
 if (0){
   library(bayesplot) 

### 2021-11-08 metahit/paper_analyses_national.r

@@ -13,10 +13,12 @@ db <- disbayes(data = gbdeng %>% filter(disease==natrundf$disease[i],
                prev_num = "prev_num", prev_denom = "prev_denom",
                rem_num = if (natrundf$remission[i]) "rem_num" else NULL, 
                rem_denom = if (natrundf$remission[i]) "rem_denom" else NULL,
-               eqage= natrundf$eqage[i], 
-               cf_model = if (natrundf$increasing[i]) "increasing" else "smooth",
-               #method="mcmc", chains=nchains, iter=3000
-               )
+               eqage = natrundf$eqage[i], 
+               cf_model = if (natrundf$increasing[i]) "increasing" else if (natrundf$const[i]) "const" else "smooth",
+               rem_model = if (natrundf$remission[i]) natrundf$rem_model[i] else NULL,
+               method="mcmc", chains=nchains, iter=2000,
+               stan_control=list(max_treedepth=15)
+)
 
 res <- tidy(db) %>%
   mutate(gender=natrundf$gender[i], disease=natrundf$disease[i], area=natrundf$area[i])

### 2021-11-08 metahit/paper_analyses_nonhier.r

@@ -1,9 +1,13 @@
+if (0) { 
+  setwd("metahit")
+  i <- 3
+}
 source("paper_analyses_header.r")
 nchains <- 2
 options(mc.cores = nchains)
 
 task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
-runs_todo <- 1:nrow(rundf)  # 211, run in groups of 8. 
+runs_todo <- 1:nrow(rundf)
 i <- runs_todo[task_id]
 
 mhi <- gbd %>%
@@ -20,17 +24,18 @@ db <- disbayes(data=mhi,
                rem_denom = if (rundf$remission[i]) "rem_denom" else NULL,
                eqage = rundf$eqage[i],
                cf_model = if (rundf$increasing[i]) "increasing" else "smooth",
-               hp_fixed = hpfixed, 
+               rem_model = if (rundf$remission[i]) rundf$rem_model[i] else NULL,
+               hp_fixed = hpfixed,
                method = "mcmc", chains=nchains, iter=100, refresh=10,
                stan_control=list(max_treedepth=15)
                )
+
 
 res <- tidy(db) %>%
   mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
 loo <- looi_disbayes(db) %>%
     mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
 saveRDS(list(res=res, loo=loo), file= paste0("results_nonhier_noinc/res", i, ".rds"))
-
 
 if (0){
   res %>% filter(var=="cf") %>% select(age, Rhat)

### 2021-11-08 metahit/paper_analyses_pointests.r

@@ -5,81 +5,54 @@ setwd("metahit")
 source("paper_analyses_header.r")
 
 ### Non-hierarchical model
+### Results not used.  Just for exploratory analysis and comparison of algorithms
 ### Empirical Bayes for smoothness pars for simplicity
-### Some cases don't estimate the mode well, as mode is low. 
-### Fix smoothness pars at 0.1 then
-brks <- c(7, 96, 154, 160)
-rundf$scf_eb <- TRUE; rundf$scf_eb[brks] <- FALSE
-rundf$sinc_eb <- TRUE; rundf$sinc_eb[c(96,97,83)] <- FALSE
 
 resopt_nonhier <- vector(nrow(rundf), mode="list")
+brkid <- c(191, 219, 222, 231, 235, 298) # fix the smooth pars for these
+brkid2 <- c(191, 219, 235) # ignore these
 
-for (i in 1:nrow(rundf)){ 
+# 232:nrow(rundf)
+for (i in brkid){ 
     mhi <- gbd %>%
         filter(gender==rundf$gender[i], disease==rundf$disease[i], area==rundf$area[i])
-    db <- disbayes(data=mhi,
+    if (i %in% brkid) {
+        hpfixed <- list(scf=0.1, sinc=0.1, srem=0.1) 
+    } else { 
+        hpfixed <- list(scf=TRUE, sinc=TRUE, srem=TRUE) 
+    }
+    try(db <- disbayes(data=mhi,
                    #inc_num = "inc_num", inc_denom = "inc_denom",
                    mort_num = "mort_num", mort_denom = "mort_denom",
                    prev_num = "prev_num", prev_denom = "prev_denom",
                    rem_num = if (rundf$remission[i]) "rem_num" else NULL, 
                    rem_denom = if (rundf$remission[i]) "rem_denom" else NULL,
                    cf_model = if (rundf$increasing[i]) "increasing" else "smooth",
+                   rem_model = if (rundf$remission[i]) rundf$rem_model[i] else NULL,
+                   hp_fixed = hpfixed, 
                    eqage=rundf$eqage[i], 
-                   scf_fixed =  if (rundf$scf_eb[i]) TRUE else 0.1,
-                   sinc_fixed = if (rundf$sinc_eb[i]) TRUE else 0.1,
-                   rem_prior = c(1.1, 1), sprior = c(1,10),
+                   sprior = c(1,10,10),
                    hessian=TRUE, draws=1000,
[…]

### 2021-11-08 metahit/slurm_bsu.sh

@@ -2,7 +2,8 @@
 #SBATCH -J metahit
 #SBATCH --time=24:00:00
 #SBATCH --cpus-per-task=2
-#SBATCH --array=[1-3]%8
+#SBATCH --array=[1-13]%8
+
 module purge
 module load default-login
 

### 2021-11-08 metahit/trends2019.R

@@ -1,0 +1,635 @@
+library(readxl)
+library(tidyverse) 
+
+# BHF trends data for 2005-2017, downloaded from 
+#https://www.bhf.org.uk/-/media/files/research/heart-statistics/cvd-statistics-2020-chapter-2a-morbidity-incidence-final.xlsx
+datxl <- read_excel("~/work/chronic/trends/cvd-statistics-2020-chapter-2a-morbidity-incidence-final.xlsx", sheet="2.9")
+
+# All incidence rates are calculated per 100,000 of the population; population-weighted data from each of the UK's four nations
+# MI. United Kingdom.  England data is for all ages. CHD is in sheet 2.7
+# Age-standardised rate is calculated for adults aged 15+
+# The Health Improvement Network (THIN) 2020
+
+## BHF 2021 doesn't have data on MI or CHD incidence after 2019, just hospital admissions.
+## CF?   BHF 2021 has trends in mortality ie number of deaths, age-stand death rates, but not CF
+## mortality data also represents incidence
+
+bhf <- datxl %>%
+    filter(row_number() %in% c(12:15, 20:23)) %>%
+    mutate(gender = rep(c("Male","Female"), each=4)) %>%
+    rename(age = 1) %>%
+    pivot_longer(names_to="year0", values_to="inc", cols=2:14) %>%
+    mutate(year = rep(2005:2017, 8),
+           year0 = NULL)
+## assume 2018,2019 same as 2017
+bhf2018 <- bhf %>% filter(year==2017) %>% mutate(year=2018)
+bhf2019 <- bhf %>% filter(year==2017) %>% mutate(year=2019)
+bhf <- bhf %>% full_join(bhf2018) %>% full_join(bhf2019) %>% arrange(age, gender, year)
+
+bhf <- bhf %>% 
+    left_join(bhf %>% filter(year==2019) %>%
+              mutate(inc2019 = inc) %>% 
+                select(age, gender, inc2019),
+              by=c("age","gender")) %>%
+  mutate(plast = inc / inc2019) %>%
+  tidyr::extract(age, c("agefrom","ageto"), "([0-9]+)-([0-9]+) years", 
+                  remove=FALSE, convert=TRUE) %>%
+  mutate(agefrom = ifelse(age == "75+ years", 75, agefrom),
[…]

### 2021-11-08 metahit/paper_analyses.Rmd
Regenerate graphs

@@
 
-resplotnh %>%
+gg <- resplotnh %>%
   ggplot(aes(x=age, y=p50, group=area, col=area)) + 
@@
 # gtable::gtable_show_layout(grob)
-idx <- which(grob$layout$name %in% c("panel-2-2","panel-8-2"))
+idx <- which(grob$layout$name %in% c("panel-2-2","panel-9-2"))
 for (i in idx) grob$grobs[[i]] <- grid::nullGrob()

### 2021-11-24 metahit/paper_analyses.Rmd
Conflict p-value function and other scraps

@@
   coord_cartesian(xlim = c(50,100)) +
-  ylab("Prevalence probability") + 
+  ylab("Prevalence") + 
   xlab("Year of age") +
@@
 
-## National results. Nice.
+## National results.
 resnat %>% filter(var=="cf") %>% 

### 2021-11-24 metahit/paper_analyses_nonhier.r

@@ -26,7 +26,7 @@ db <- disbayes(data=mhi,
                cf_model = if (rundf$increasing[i]) "increasing" else "smooth",
                rem_model = if (rundf$remission[i]) rundf$rem_model[i] else NULL,
                hp_fixed = hpfixed,
-               method = "mcmc", chains=nchains, iter=100, refresh=10,
+               method = "mcmc", chains=nchains, iter=1000, refresh=10,
                stan_control=list(max_treedepth=15)
                )
 
@@ -41,4 +41,8 @@ if (0){
   res %>% filter(var=="cf") %>% select(age, Rhat)
   library(bayesplot)
   mcmc_trace(db$fit, pars=paste0("cf[",70:74,"]"))
+  
+  conflict_disbayes(db, var="inc")
+  conflict_disbayes(db, var="prev")
+  conflict_disbayes(db, var="mort")
 }

### 2021-11-24 tests/testthat/test_fit.R

@@ -1,0 +1,23 @@
+source("data.r")
+
+test_that("fit and conflict checks",{
+  set.seed(1)
+  db <- disbayes(dat = ihdbristol,
+                    inc_num = "inc_num", inc_denom = "inc_denom",
+                    prev_num = "prev_num", prev_denom = "prev_denom",
+                    mort_num = "mort_num", mort_denom = "mort_denom",
+                    eqage = 40, method="mcmc", algorithm="Fixed_param", 
+                    chains=1, iter=100)
+  expect_s3_class(db, "disbayes")
+  
+  od <- tidy_obsdat(db)
+  expect_equal(od$denom[1], ihdbristol$inc_denom[1])
+  if (interactive()){
+    plotfit_data_disbayes(db)
+    plotfit_disbayes(db)
+  }
+  
+  cd <- conflict_disbayes(db, varname="mort")
+  expect_lte(cd$p1[1], 1)
+  expect_gte(cd$p1[1], 0)
+})

### 2021-11-30 R/disbayes_hier.R
Add references to Arxiv paper

@@ -119,7 +119,9 @@
 ##'
 ##' @inheritParams disbayes
 ##'
-##' @details For full details see Jackson et al. (in progress...)
+##' @references Jackson C, Zapata-Diomedi B, Woodcock J. 
+##' "Bayesian multistate modelling of incomplete chronic disease burden data" 
+##' \url{http://arxiv.org/abs/2111.14100}
 ##'
 ##' @md
 ##' 

### 2022-07-12 metahit/paper_analyses.Rmd
Prepare for CRAN release

@@
   coord_cartesian(xlim = c(50,100)) +
-  ylab("Annual mortality probability") + 
+  ylab("Annual mortality risk") + 
   xlab("Year of age") +
@@
   coord_cartesian(xlim = c(50,100)) +
-  ylab("Annual incidence probability") + 
+  ylab("Annual incidence risk") + 
   xlab("Year of age") +
@@
   filter(var=="cf") %>%
-  mutate(p50 = 1 - exp(-`mode`),
-         plow = 1 - exp(-`2.5%`),
-         pupp = 1 - exp(-`97.5%`)) %>%
+  mutate(p50 = `mode`,
+         plow = `2.5%`,
+         pupp = `97.5%`) %>%
   filter(area=="Bristol" | area=="England", gender=="Female") %>%
@@
          age %in% c(70, 75, 80, 85, 90)) %>%
-  mutate(par = 1 - exp(-`50%`),
-         plow = 1 - exp(-`2.5%`),
-         pupp = 1 - exp(-`97.5%`)) %>%
+  mutate(par = `50%`,
+         plow = `2.5%`,
+         pupp = `97.5%`) %>%
   ggplot(aes(x=age, y=par, col=Algorithm, lty=Algorithm)) + 
@@
   facet_wrap(vars(disease), ncol=2, scales="free_y") + 
-  xlab("Age (years)") + ylab("Case fatality probability")
+  xlab("Age (years)") + ylab("Case fatality rate")
 
@@
     dplyr::filter(model %in% c("Independent areas","National"), 
-                  var=="cf",
+                  var=="cf_prob",
                 area %in% c(cityregions,"England")) %>%
@@
     dplyr::filter(! (disease=="Uterine cancer" & age > 90),
-                  disease %in% diseases_paper) %>%
-    mutate(p50 = 1 - exp(-`50%`),
-         plow = 1 - exp(-`2.5%`),
-         pupp = 1 - exp(-`97.5%`))
-
+                  disease %in% diseases_paper) 
 gg <- resplotnh %>%
-  ggplot(aes(x=age, y=p50, group=area, col=area)) + 
-  geom_ribbon(aes(ymin=plow, ymax=pupp), 
+  ggplot(aes(x=age, y=`50%`, group=area, col=area)) + 
+  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), 
               fill="gray80", col="gray80") +
@@
 ## UTRC doesn't converge, and STMC estimates noisy, so we use national estimate results 
+## Dementia is only other one where sampling performance is poor.
+## Trace plots for dementia look like convergence to plausible region, but poor mixing
 resnc <- resh %>% filter(Rhat > 2) 
-resnc <- resh %>% filter(Rhat > 1.1) 
+resnc <- resh %>% filter(Rhat > 1.1, age>80, var=="cf") 
 resnc <- resh %>% filter(Rhat > 1.05) 
@@
 rescfplot <- resall %>% 
-    filter(var=="cf", 
+    filter(var=="cf_prob", 
            disease %in% diseases_paper_short, 
@@
     mutate(model = ifelse(model=="National","Independent areas",model),
-           p50 = 1 - exp(-`50%`),
-           plow = 1 - exp(-`2.5%`),
-           pupp = 1 - exp(-`97.5%`),
            gender_model = paste(gender, model, sep="\n")) %>%
@@
     ggplot(aes(x=age, col=area, group=area)) +
-    geom_ribbon(aes(ymin=plow, ymax=pupp), fill="gray80", col="gray80") +
-    geom_line(aes(y=p50)) + 
+    geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), fill="gray80", col="gray80") +
+    geom_line(aes(y=`50%`)) + 
     coord_cartesian(xlim=c(70,  100), ylim=c(0,NA)) +
     scale_x_continuous(breaks=seq(0,100,10)) + 
-    ylab("Annual case fatality probability")  +
+    ylab("Annual case fatality risk")  +
     facet_grid(cols=vars(gender_model), rows=vars(disease),
@@
 ```
+
+# Discrete-time approximate estimate of case fatality.
+
+Mortality divided by prevalence.
+
+Leeds, female, all diseases.  Use standard disbayes model.
+
+```{r}
+setwd("metahit")
+source("paper_analyses_header.r")
+
+mhi <- gbd %>%
+  mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
+  filter(disease %in% diseases_paper_short) %>%
+  filter(area=="Leeds", gender=="Female") %>%
+  droplevels() %>%
+  mutate(mort_prob = mort_num / mort_denom, 
+         prev_prob = prev_num / prev_denom,
+         mort_rate = -log(1 - mort_prob),  
+         cf_crude = mort_rate / prev_prob)
+
+db <- resall %>% 
+  filter(model=="Independent areas") %>%
+  filter(area=="Leeds", gender=="Female") %>%
+  filter(disease %in% diseases_paper_short) %>%
+  filter(var=="cf") %>%
+  left_join(mhi %>% select(age, gender, disease, cf_crude))
+
+p1 <- ggplot(db %>% filter(age>60, age <80), 
+             aes(x=age, y=`50%`)) + 
+  facet_grid(rows=vars(disease), scales = "free_y", labeller = label_wrap_gen(15)) +
+  geom_line(lwd=1.5) + 
+  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), alpha=0.3) + 
+  geom_line(aes(y=cf_crude), col="red", lwd=1.5, alpha=0.6) + 
+  xlab("Age (years)") + 
+  ylab("Case fatality rate") +
+  theme(strip.text = element_text(size=7))
+p1
+
+p2 <- ggplot(db %>% filter(age>=80, age <100), 
+             aes(x=age, y=`50%`)) + 
+  facet_grid(rows=vars(disease), scales = "free_y", labeller = label_wrap_gen(15)) +
+  geom_line(lwd=1.5) + 
+  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), alpha=0.3) + 
+  geom_line(aes(y=cf_crude), col="red", lwd=1.5, alpha=0.6) + 
+  xlab("Age (years)") + 
+  ylab("Case fatality rate") +
+  theme(strip.text = element_text(size=7))
+p2
+
+pdf("~/work/chronic/write/cf_discrete.pdf")
+gridExtra::grid.arrange(p1, p2, nrow=1)
+dev.off()
+
+```
+
+Estimate is non-smooth, unreliable for older ages, doesn't account for uncertainty.
+
+# Comparison of MCMC and optimisation
+
+Timings.
+
+```{r}
+system.time(db <- disbayes(data=mhi,
+               inc_num = "inc_num", inc_denom = "inc_denom",
+               mort_num = "mort_num", mort_denom = "mort_denom",
+               prev_num = "prev_num", prev_denom = "prev_denom",
+               eqage = 40))
+options(mc.cores = 4)
+system.time(db <- disbayes(data=mhi,
+               inc_num = "inc_num", inc_denom = "inc_denom",
+               mort_num = "mort_num", mort_denom = "mort_denom",
+               prev_num = "prev_num", prev_denom = "prev_denom",
+               eqage = 40, method="mcmc",iter=1000))
+```
+
+Compare point estimates and uncertainty, and how these relate to the amount of data, and age.
+
+```{r}
+source("read_model_results.r")
+rescomp <- resall %>%
+  filter(model=="Independent areas") %>%
+  mutate(method="MCMC",
+         pointest = `50%`) %>%
+  full_join(readRDS("resopt_nonhier.rds") %>% 
+              mutate(method="Optimisation",
+                     pointest = `50%`)) %>%
+  filter(
+         disease %in% diseases_paper,
+         !(disease %in% c("Stomach cancer", "Uterine cancer"))
+         ) %>%
+  mutate(iqr = `75%` - `25%`,
+         ciw = `97.5%` - `2.5%`) %>%
+  filter(var == "cf",
+         age > 50) %>%
+  select(age, gender, disease, area, method, pointest, iqr, ciw) %>%
+  left_join(gbd %>% select(age, gender, disease, area, mort_num), 
+            by=c("age","gender","disease","area"))
+
+resc_bias <- rescomp %>% 
+  select(-iqr, -ciw) %>%
+  pivot_wider(names_from=method, 
+              values_from=pointest, 
+              id_cols = c("age","gender","disease","area","mort_num")) %>%
+  mutate(bias = (Optimisation - MCMC),
+         dga = interaction(disease, gender, area))
+
+resc_uncbias <- rescomp %>% 
+  select(-pointest, -iqr) %>%
+  pivot_wider(names_from=method, 
+              values_from=ciw, 
+              id_cols = c("age","gender","disease","area","mort_num")) %>%
+  mutate(bias = 100*(Optimisation - MCMC)/MCMC, 
+         dga = interaction(disease, gender, area))
+
+library(gridExtra)
+library(ggpubr)
+
+p1 <- ggplot(data=resc_bias, aes(x=age, y=bias, group=dga, col=disease)) + 
+  geom_line(alpha=0.5) +
+  geom_point() +
+  ylab("Median(optimisation) - median(MCMC)") + 
+  xlab("Age (years)")
+
+p2 <- ggplot(data=resc_uncbias, aes(x=age, y=bias, group=dga, col=disease)) + 
+  geom_line(alpha=0.3) +
+  geom_point(alpha=0.5) +
+  ylab("%Bias in CI width") +
+  xlab("Age (years)")
+
+p3 <- ggplot(data=resc_bias, aes(x=mort_num, y=bias, group=dga, col=disease)) + 
+  geom_line(alpha=0.3) +
+  geom_point(alpha=0.5) +
+  ylab("Median(optimisation) - median(MCMC)") +
+  xlab("Number of deaths in the subgroup") +
+  theme(legend.title = element_blank(), 
+        legend.text = element_text(size=8),
+        legend.position = "bottom") + 
+  guides(col=guide_legend(nrow=3,byrow=TRUE))
+
+p4 <- ggplot(data=resc_uncbias, aes(x=mort_num, y=bias, group=dga, col=disease)) + 
+  geom_line(alpha=0.3) +
+  geom_point(alpha=0.5) +
+  ylab("%Bias in CI width") + 
+  xlab("Number of deaths in the subgroup")
+
+p <- ggarrange(p3, p1, p4, p2, common.legend=TRUE, legend="bottom") # assign to object to work around blank page bug
+pdf("~/work/chronic/write/opt_compare_full.pdf")
+p
+dev.off()
+
+```

### 2022-07-12 R/disbayes_hier.R

@@ -121,7 +121,7 @@
 ##'
 ##' @references Jackson C, Zapata-Diomedi B, Woodcock J. 
 ##' "Bayesian multistate modelling of incomplete chronic disease burden data" 
-##' \url{http://arxiv.org/abs/2111.14100}
+##' \url{https://arxiv.org/abs/2111.14100}
 ##'
 ##' @md
 ##' 

### 2022-07-12 R/fit.R

@@ -31,13 +31,13 @@ tidy_obsdat_hier <- function(x){
   vars <- c("inc","prev","mort")
   if (dat$remission) vars <- c(vars, "rem")
   res <- vector(length(vars), mode="list")
-  for (vi in vars){
+  for (i in seq_along(vars)){
     tdat <- array_indvecs(dims)
     names(tdat) <- c("age", "area", "gender")
     tdat$age <- tdat$age - 1
-    tdat$num <- as.vector(dat[[sprintf("%s_num",vi)]])
-    tdat$denom <- as.vector(dat[[sprintf("%s_denom",vi)]])
-    tdat$var <- vi
+    tdat$num <- as.vector(dat[[sprintf("%s_num",vars[i])]])
+    tdat$denom <- as.vector(dat[[sprintf("%s_denom",vars[i])]])
+    tdat$var <- vars[i]
     tdat$prob <- tdat$num/tdat$denom
     tdat$lower <- qbeta(0.025, tdat$num+0.5, tdat$denom-tdat$num+0.5)
     tdat$upper <- qbeta(0.975, tdat$num+0.5, tdat$denom-tdat$num+0.5)
@@ -103,6 +103,7 @@ plotfit_disbayes <- function(x, agemin=50){
 ## Only implemented currently for parameters with three indices
 
 opt_extract_nonhier <- function(xfit, par){
+  varorig <- NULL
   sam <- xfit$theta_tilde
   parnames <- colnames(sam)
   sam <- sam[,grep(sprintf("^%s\\[.+", par), parnames)]
@@ -126,6 +127,7 @@ opt_extract_nonhier <- function(xfit, par){
 ## Not used currently
 
 opt_extract_hier <- function(xfit, par){
+  varorig <- NULL
   sam <- xfit$theta_tilde
   parnames <- colnames(sam)
   sam <- sam[,grep(sprintf("^%s\\[.+", par), parnames)]

### 2022-07-12 R/state_probs.R

@@ -1,4 +1,4 @@
-state_probs <- function(i, f)  {
+state_probs_norem <- function(i, f)  {
     l <- i + f
     q <- sqrt(i*i - 2*i*f  + f*f)
     w <- exp(-(l + q) / 2)
@@ -10,6 +10,27 @@ state_probs <- function(i, f)  {
     
     P[1,2] <- i*(v - w)/q
     P[2,2] <- -((2*f - l)*(v-w) - q*(v+w)) / (2*q)
+    P[3,2] <- 0
+    
+    P[1,3] <- (-l*(v-w) - q*(v+w))/(2*q) + 1
+    P[2,3] <- ((v-w)*(2*f - l) - q*(v+w))/(2*q) + 1
+    P[3,3] <- 1
+    P
+}
+
+state_probs_rem <- function(i, f, r)  {
+    l <- i + r + f
+    q <- sqrt(i*i + 2*i*r -  2*i*f  + r*r + 2*f*r + f*f)
+    w <- exp(-(l + q) / 2)
+    v <- exp(-(l - q) / 2)
+    P <- array(dim=c(3, 3))
+
+    P[1,1] <- (2*(v-w)*(f+r) + v*(q-l) + w*(q+l)) / (2*q)
+    P[2,1] <- (v-w)*r/q
+    P[3,1] <- 0 
+    
+    P[1,2] <- i*(v - w)/q
+    P[2,2] <- -((2*(f+r) - l)*(v-w) - q*(v+w)) / (2*q)
     P[3,2] <- 0
     
     P[1,3] <- (-l*(v-w) - q*(v+w))/(2*q) + 1

### 2022-07-12 R/tidy.R

@@ -23,9 +23,9 @@
 ##'
 ##' * `cf`, `inc`, `rem`: Case fatality, incidence, remission rates
 ##'
-##' * `inc_prob`, `rem_prob`, `mort_prob`:  Annual incidence, remission and mortality probabilities. 
+##' * `inc_prob`, `rem_prob`, `mort_prob`, `cf_prob`:  Annual incidence, remission, mortality and case fatality risks (probabilities). 
 ##'
-##' * `prev_prob` Prevalence probability. 
+##' * `prev_prob` Prevalence (probability). 
 ##'
 ##' * `state_probs` State occupancy probabilities.
 ##'
@@ -117,7 +117,7 @@ tidy_disbayes_full <- function(fit, varlist, method, levs=NULL, ...) {
 .disbayes_vars <- list(
     age = list(indnames = "age",
                varnames = c("cf","inc_par","cf_par","rem_par",
-                            "dcf","rem","rem_prob","mort_prob")), 
+                            "dcf","rem","rem_prob","mort_prob","cf_prob")), 
     agebias =  list(indnames = c("age", "bias"),
                     varnames = c("inc", "inc_prob", "prev_prob")),
     agebiasstate = list(indnames = c("age", "bias", "state"),
@@ -132,7 +132,8 @@ attr(.disbayes_vars, "order") <- c("age", "bias", "state", "term")
 
 .disbayes_trend_vars <- list(
     age = list(indnames="age",
-               varnames = c("cf", "inc_par", "rem_par", "rem", "rem_prob", "mort_prob")),
+               varnames = c("cf", "inc_par", "rem_par", "rem",
+                            "rem_prob", "mort_prob","cf_prob")),
     ageyear = list(indnames = c("age", "year"),
                    varnames = c("cf_yr")),
     agebias = list(indnames = c("age", "bias"),
@@ -152,7 +153,8 @@ attr(.disbayes_trend_vars, "order") <- c("age", "year", "bias", "state", "term")
 .disbayes_hier_vars <- list(
     ageareagender = list(
         indnames = c("age","area","gender"), 
-        varnames =  c("inc","cf","dcf","inc_prob","prev_prob","mort_prob","rem","rem_prob")
+        varnames =  c("inc","cf","dcf","inc_prob","prev_prob",
+                      "mort_prob","cf_prob","rem","rem_prob")
     ),
     ageareagenderstate = list(
         indnames = c("age","area","gender","state"), 

### 2022-07-12 metahit/paper_analyses_gender.r

@@ -62,7 +62,7 @@ if (0){
 res <- tidy(db) %>%
   mutate(disease=hierrungdf$disease[i])
 
-loo <- looi_disbayes_hier(db) %>%
+loo <- looi_disbayes(db) %>%
     mutate(disease=hierrungdf$disease[i])
 
 saveRDS(list(res=res,loo=loo), file= paste0("results_hier_gender/res", i, ".rds"))

### 2022-07-12 metahit/paper_analyses_nonhier.r

@@ -35,7 +35,7 @@ res <- tidy(db) %>%
   mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
 loo <- looi_disbayes(db) %>%
     mutate(gender=rundf$gender[i], disease=rundf$disease[i], area=rundf$area[i])
-saveRDS(list(res=res, loo=loo), file= paste0("results_nonhier_noinc/res", i, ".rds"))
+saveRDS(list(res=res, loo=loo), file= paste0("results_nonhier/res", i, ".rds"))
 
 if (0){
   res %>% filter(var=="cf") %>% select(age, Rhat)

### 2022-07-12 metahit/paper_analyses_pointests.r

@@ -170,7 +170,7 @@ rnh_noinc <- tidy(db) %>%
 
 rest_all <- rbind(rest, rest_noinc) %>%
     full_join(rnh) %>% full_join(rnh_noinc) %>% 
-    filter(var %in% c("inc_prob","cf","prev_prob","mort_prob"))
+    filter(var %in% c("inc_prob","cf_prob","prev_prob","mort_prob"))
 
 obsdat <- tidy_obsdat(dbt) %>% 
     mutate(model="Observed data") %>%
@@ -182,8 +182,7 @@ res_checkfit <- rest_all %>%
     rename(est=mode, lower=`2.5%`, upper=`97.5%`) %>%
     full_join(obsdat %>% 
                   mutate(lower=NULL, upper=NULL)) %>%
-    mutate(est = ifelse(var=="cf", 1 - exp(-est), est)) %>%
-    mutate(var = fct_recode(var, "Case fatality"="cf", "Incidence"="inc_prob",
+    mutate(var = fct_recode(var, "Case fatality"="cf_prob", "Incidence"="inc_prob",
                             "Mortality"="mort_prob","Prevalence"="prev_prob")) %>%
     mutate(model = relevel(factor(model), "Observed data")) %>%
     mutate(trend = ifelse(grepl("Time trends", model), "Time trends", "No trends"),
@@ -199,7 +198,7 @@ pdf("~/work/chronic/write/checkfit.pdf", width=6, height=3)
 hcols <- scales::hue_pal()(2)
 rcplot <-  res_checkfit %>% 
     filter(age>50) %>%
-    filter(!(var=="cf" & age > 90)) 
+    filter(!(var=="cf_prob" & age > 90)) 
 ggplot(rcplot %>% filter(model!="Observed data", 
                                !(var=="Case fatality" & est>0.1)
                                ), 
@@ -238,7 +237,7 @@ pobs <- ggplot(res_fit_arm,
                         breaks = c("Observed data", "No trends", "Time trends")) + 
     facet_wrap(~var, nrow=2, ncol=3, scales="free_y") + 
     scale_y_continuous(limits=c(0,NA)) +
-    ylab("Probability") + xlab("Age (years)") + 
+    ylab("Annual case fatality risk") + xlab("Age (years)") + 
     labs(col="", lty="Data sources") + 
     guides(fill="none", 
            lty = guide_legend(order = 2, override.aes = list(col=hcols[1])),
[…]

### 2023-03-07 R/disbayes.R
Updated reference to JRSSA paper

@@ -298,9 +298,10 @@
 ##'   Use the \code{\link{tidy.disbayes}} method to return summary statistics
 ##'   from the fitted models, simply by calling \code{tidy()} on the fitted model. 
 ##'   
-##' @references Jackson C, Zapata-Diomedi B, Woodcock J. 
-##' "Bayesian multistate modelling of incomplete chronic disease burden data" 
-##' \url{https://arxiv.org/abs/2111.14100}
+##' @references Jackson C, Zapata-Diomedi B, Woodcock J. (2023)
+##' "Bayesian multistate modelling of incomplete chronic disease burden data"
+##' Journal of the Royal Statistical Society, Series A, 186(1), 1-19
+##' \url{https://doi.org/10.1093/jrsssa/qnac015}
 ##'
 ##' @export
 disbayes <- function(data,
